%% ============================================================
%%  Chapter 11 --- Hybrid Retrieval Pipeline
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%%  Dependencies: rank-bm25 >= 0.2.2, flashrank >= 0.2.10,
%%                langchain-qdrant >= 0.2.0, litellm >= 1.40.0,
%%                instructor >= 1.3.0
%% ============================================================

\chapterquote{Recall without precision wastes context; precision without recall wastes potential.}{anon}

The production failure this chapter prevents is a silent one. An agent that
retrieves wrong or irrelevant memories does not raise an exception --- it
reasons from bad evidence and produces plausible-sounding but incorrect
output. A team running a customer-support assistant discovered this when
recall for technical queries dropped below 50\% after the corpus grew past
100\,000 memories. The dense vector index had not changed; the failure was
vocabulary mismatch. Model names, version strings, and product identifiers
appeared verbatim in user queries but were out-of-distribution for the
embedding encoder, which silently ranked semantically unrelated passages
higher. BM25 would have found those passages immediately. Neither retriever
alone was sufficient; the failure was architectural.

The solution this chapter builds is a two-stage hybrid pipeline. The first
stage is a \emph{recall layer}: BM25 lexical retrieval and dense vector
retrieval each produce a ranked candidate pool, and Reciprocal Rank Fusion
combines them without requiring score normalisation. The second stage is a
\emph{precision layer}: a \code{flashrank} cross-encoder reranker reorders
the fused pool using full query-passage interaction. The base pipeline
(Sections~\ref{sec:pipeline-architecture}--\ref{sec:reranker}) is a complete,
runnable system before any enhancements are introduced. HyDE
(Section~\ref{sec:hyde}) and CRAG (Section~\ref{sec:crag}) are opt-in
extensions applied only after the base pipeline passes the quality tests in
Section~\ref{sec:metrics}.

One architectural constraint from Chapter~10 governs the entire chapter:
the Qdrant collection established in \S10.1 uses a single unnamed dense
vector field (\code{VectorParams} directly, not a named\code{\{"dense": VectorParams(...)\}} dict). This means \code{RetrievalMode.HYBRID} from
\texttt{langchain-qdrant} is architecturally excluded here: that mode requires
a collection configured with both \code{vectors\_config} and
\code{sparse\_vectors\_config}, which is a different schema. Using it would
require deleting and recreating the Chapter~10 collection --- a breaking change.
BM25 is therefore implemented independently with \texttt{rank-bm25}, not
through Qdrant's sparse vector index.

This chapter introduces several pipeline-specific terms defined in
Table~\ref{tab:ch11-terms}.

\begin{table}[h]
\centering
\caption{Domain terms introduced in Chapter~11}
\label{tab:ch11-terms}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
\code{RetrievedPassage} & Shared TypedDict carrying a passage's \code{id}, \code{text}, \code{score}, \code{rank}, \code{source}, and \code{meta} through all pipeline stages. \\
BM25 first stage & Lexical retrieval using \code{BM25Okapi} over a tokenised corpus; excels at exact term match for product names, error codes, and version strings. \\
Dense stage & Approximate nearest-neighbour search via \code{QdrantVectorStore}; excels at semantic paraphrase retrieval. \\
RRF & Reciprocal Rank Fusion: rank-position-based score fusion that requires no score normalisation across heterogeneous retrievers. \\
Cross-encoder & A reranker model that scores (query, passage) pairs jointly; more precise than bi-encoder retrieval but too slow to run over a full corpus. \\
HyDE & Hypothetical Document Embedding: replaces a short or ambiguous query with a generated hypothetical answer whose embedding lies closer to relevant passages in vector space. \\
CRAG & Corrective Retrieval Augmented Generation: a LangGraph conditional loop that triggers re-retrieval when a grader node judges the first-pass candidates as insufficient. \\
\bottomrule
\end{tabularx}
\end{table}

The source layout for this chapter is:

\begin{verbatim}
src/
  retrieval/
    types.py           # §11.1
    bm25_stage.py      # §11.2
    dense_stage.py     # §11.3
    fusion.py          # §11.4
    reranker.py        # §11.5
    hyde.py            # §11.6
    crag_nodes.py      # §11.7
    crag_graph.py      # §11.7
    metrics.py         # §11.8
prompts/
  hyde_v1.jinja2           # §11.6
  retrieval_grader_v1.jinja2 # §11.7
tests/
  retrieval/
    test_retrieval_quality.py  # §11.8
  fixtures/
    retrieval_eval.json        # §11.8
\end{verbatim}


%% ------------------------------------------------------------
\section{Two-Stage Architecture: Recall Then Precision}
\label{sec:pipeline-architecture}
%% ------------------------------------------------------------

The design problem is that recall and precision are in direct tension when
only one retrieval mechanism is used. A retriever optimised for recall returns
many candidates, most of which are irrelevant; one optimised for precision
returns few candidates, risking the omission of genuinely relevant passages
that were phrased differently from the query. The governing principle is that
recall and precision are separate concerns that require separate stages, and
that each stage must receive only the output type the next stage is designed
to consume.

The data flow has five steps. First, BM25 retrieval over the tokenised memory
corpus produces a ranked list of up to 50 candidates
(\code{source="bm25"}). Second, dense vector retrieval via
\code{QdrantVectorStore} produces a second ranked list of up to 50 candidates
(\code{source="dense"}). Third, \code{reciprocal\_rank\_fusion()} merges the
two lists by rank position into a single pool of up to 100 candidates
(\code{source="fused"}). Fourth, \code{rerank\_passages()} applies a
cross-encoder to the fused pool and returns the top-5 reordered candidates
(\code{source="reranked"}). Fifth, the generation node receives exactly those
five passages as its evidence context.

The shared data carrier across all stages is \code{RetrievedPassage}. Every
stage reads from and writes to this type so that stage boundaries are clean
interfaces rather than implicit conventions.

\begin{lstlisting}[language=Python,
  caption={Shared passage type used across all pipeline stages},
  label={lst:retrieved-passage}]
# src/retrieval/types.py
from typing import Any, Literal
from typing_extensions import TypedDict

class RetrievedPassage(TypedDict):
    id: str
    text: str
    score: float
    rank: int
    source: Literal["bm25", "dense", "fused", "reranked"]
    meta: dict[str, Any]
\end{lstlisting}

Two decisions in Listing~\ref{lst:retrieved-passage} deserve explanation.
The \code{source} field carries a literal tag through the entire pipeline,
making it possible to trace which retriever originally contributed each
passage to the final context --- a detail that Chapter~23 uses to build
per-retriever recall metrics in Langfuse. The \code{rank} field is
reassigned at each stage, reflecting the passage's position within the
current stage's ordered output rather than its original position in the
corpus, which would have no pipeline-level meaning.

\begin{warningbox}
\code{RetrievalMode.HYBRID} from \texttt{langchain-qdrant} is not used in
this chapter. That mode requires the Qdrant collection to have both dense
and sparse vector fields (\code{sparse\_vectors\_config} at collection
creation). Chapter~10's \code{create\_agent\_memory\_collection()} creates
a single unnamed dense vector field; adding a sparse field would require
dropping and recreating the collection. BM25 is implemented with
\texttt{rank-bm25} as an independent lexical stage.
\end{warningbox}


%% ------------------------------------------------------------
\section{BM25 First Stage: rank\_bm25 with Tokenisation Pipeline}
\label{sec:bm25-stage}
%% ------------------------------------------------------------

The design problem is tokenisation asymmetry. BM25 scores are computed as
inner products between term-frequency vectors built from the corpus and from
the query. If the two are tokenised differently --- for example, punctuation is
stripped from corpus tokens but not from query tokens --- the overlapping term
sets diverge and scores collapse. The governing principle is that the same
\code{tokenize()} function is applied to every corpus passage at index time
and to every query at search time, with no exceptions.

The tokenisation pipeline for agent memory workloads has three steps and
deliberately omits two common extras. Lowercase normalisation and punctuation
stripping are required. Whitespace splitting is the terminal step. Stemming
is omitted because agent memories contain model names (\code{gpt-4o},
\code{claude-3-5-sonnet}), version strings (\code{v1.2.3}), and product
identifiers that stemming corrupts. Stopword removal is omitted because short
queries (two to four words) frequently consist almost entirely of stopwords
and become empty strings after removal.

\code{BM25Okapi} is selected over \code{BM25L} and \code{BM25+}. The
difference matters for corpora with high document-length variance: \code{BM25L}
applies a lower-bound normalisation that benefits very long documents.
Agent memory entries are typically one to three sentences; length variance is
low. \code{BM25Okapi} is simpler to reason about and the appropriate default.

\begin{notebox}
\code{BM25Okapi} requires a \code{list[list[str]]} at construction time
--- a pre-tokenised corpus --- and \code{get\_scores()} requires a
\code{list[str]} --- a pre-tokenised query. \code{tokenize()} must be called
on both sides. The \code{get\_top\_n()} convenience method returns raw text
strings and discards document identity; it must not be used here because the
\code{id} field of each \code{RetrievedPassage} must survive into the fusion
stage.
\end{notebox}

\begin{lstlisting}[language=Python,
  caption={Tokenisation pipeline and BM25 index construction},
  label={lst:bm25-index}]
# src/retrieval/bm25_stage.py
import re
import numpy as np
from rank_bm25 import BM25Okapi
from src.retrieval.types import RetrievedPassage

def tokenize(text: str) -> list[str]:
    text = text.lower()
    text = re.sub(r"[^\w\s]", " ", text)
    return text.split()

def build_bm25_index(
    passages: list[RetrievedPassage],
) -> tuple[BM25Okapi, list[RetrievedPassage]]:
    tokenized = [tokenize(p["text"]) for p in passages]
    return BM25Okapi(tokenized), passages
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={BM25 retrieval with rank assignment},
  label={lst:bm25-retrieve}]
# src/retrieval/bm25_stage.py (continued)
def bm25_retrieve(
    index: BM25Okapi,
    corpus: list[RetrievedPassage],
    query: str,
    top_k: int = 50,
) -> list[RetrievedPassage]:
    tokenized_query = tokenize(query)
    scores: np.ndarray = index.get_scores(tokenized_query)
    top_indices = np.argsort(scores)[::-1][:top_k]
    results = []
    for rank, idx in enumerate(top_indices, start=1):
        passage = dict(corpus[idx])
        passage["score"] = float(scores[idx])
        passage["rank"] = rank
        passage["source"] = "bm25"
        results.append(passage)
    return results
\end{lstlisting}

Three decisions in Listings~\ref{lst:bm25-index} and~\ref{lst:bm25-retrieve}
deserve explanation. \code{np.argsort(scores)[::-1]} sorts in descending order
without copying the full score array --- safer and faster than
\code{sorted(enumerate(scores), key=lambda x: -x[1])} for large corpora.
The \code{dict(corpus[idx])} copy is required: mutating the original corpus
passage in-place would corrupt the index on the next retrieval call.
\code{top\_k=50} is the default because the RRF fusion stage benefits from a
large candidate pool, but the cross-encoder reranker is CPU-bound at
inference --- 100 fused candidates (50 from each stage) is the practical
ceiling before reranker latency exceeds the SLA.


%% ------------------------------------------------------------
\section{Dense Vector Stage: QdrantVectorStore with Payload Filters}
\label{sec:dense-stage}
%% ------------------------------------------------------------

The design problem is connecting to the pre-configured Chapter~10 collection
without violating the shared client contract and without applying the tenant
filter as an afterthought in Python. The governing principle is that
\code{QdrantVectorStore} is constructed from the shared \code{QdrantClient}
instance, the retrieval mode is declared as \code{RetrievalMode.DENSE} at
construction time, and the \code{user\_id} payload filter is passed as the
\code{filter=} keyword argument to every search call.

The construction pattern matters. \code{QdrantVectorStore.from\-\_existing\_collection()} accepts only a URL string, which bypasses the shared
client lifecycle established in Chapter~10 \S10.1. The correct form is
\code{QdrantVectorStore(client=client, collection\_name=..., embedding=embeddings, retrieval\_mode=RetrievalMode.DENSE)}, passing the
already-constructed \code{QdrantClient} directly.

\code{similarity\_search\_with\_score()} returns
\code{list[tuple[Document, float]]}. Each \code{Document} carries the
passage text in \code{page\_content} and all \code{MemoryPayload} fields
in \code{metadata}. The conversion to \code{RetrievedPassage} extracts
\code{metadata["id"]} for the identity field and assigns
\code{source="dense"}.

\begin{lstlisting}[language=Python,
  caption={QdrantVectorStore construction and dense retrieval},
  label={lst:dense-retrieve}]
# src/retrieval/dense_stage.py
from langchain_qdrant import QdrantVectorStore, RetrievalMode
from src.retrieval.types import RetrievedPassage
from src.vector_store.tenant_ops import build_user_filter

def dense_retrieve(
    vector_store: QdrantVectorStore,
    query: str,
    user_id: str,
    top_k: int = 50,
    min_importance: float = 0.0,
) -> list[RetrievedPassage]:
    filt = build_user_filter(
        user_id=user_id,
        min_importance=min_importance,
    )
    hits = vector_store.similarity_search_with_score(
        query,
        k=top_k,
        filter=filt,
    )
    results: list[RetrievedPassage] = []
    for rank, (doc, score) in enumerate(
        sorted(hits, key=lambda x: x[1], reverse=True), start=1
    ):
        results.append(
            RetrievedPassage(
                id=str(doc.metadata.get("id", doc.metadata.get("_id", ""))),
                text=doc.page_content,
                score=float(score),
                rank=rank,
                source="dense",
                meta={k: v for k, v in doc.metadata.items()
                      if k not in ("id", "_id")},
            )
        )
    return results
\end{lstlisting}

Two decisions in Listing~\ref{lst:dense-retrieve} deserve explanation.
The \code{sorted(..., reverse=True)} step is applied even though
\code{similarity\_search\_with\_score} typically returns results in score
order, because the langchain-qdrant documentation does not guarantee that
ordering contract across all versions --- an explicit sort costs nothing and
protects against future ordering changes. The \code{id} extraction
falls back to \code{\_id} because Qdrant point IDs are stored differently
depending on whether the payload was written by \code{mem0} (which uses
\code{id}) or a direct \code{upsert} call (which stores the Qdrant UUID in
\code{\_id}).


%% ------------------------------------------------------------
\section{RRF Fusion: \texttt{reciprocal\_rank\_fusion()}}
\label{sec:rrf-fusion}
%% ------------------------------------------------------------

The design problem is combining two ranked lists whose scores live in
incomparable spaces. BM25 scores are term-frequency-based floats with no
fixed upper bound; cosine similarity scores from the dense stage are bounded
in $[-1, 1]$. Normalising them to a common range requires knowing the
distribution of scores in both stages, which changes with the corpus and
the query. Rank positions are always comparable. The governing principle is
that fusion operates on rank positions, not on scores, and that a document
is not discarded for appearing in only one list.

The RRF formula for a document $d$ is:

\[ \text{score}(d) = \sum_{i} \frac{1}{k + \text{rank}_i(d)} \]

where the sum runs over each ranked list $i$ that contains $d$, $k = 60$
by default, and $\text{rank}_i(d)$ is the 1-indexed position of $d$ in
list $i$. A document appearing in only one list with rank~1 receives
$1 / (60 + 1) \approx 0.016$. A document appearing in both lists at rank~1
receives $2 \times 0.016 = 0.033$. The $k = 60$ default is from the
original Cormack et al. (2009) paper. Lowering $k$ amplifies top-rank
differences; 60 is the appropriate neutral choice for agent memory retrieval
where no single retriever is systematically dominant.

The function is generalised to \code{list[list[RetrievedPassage]]} rather
than two fixed arguments so that HyDE (\S\ref{sec:hyde}) can contribute a
third ranked list without requiring a function signature change.

\begin{lstlisting}[language=Python,
  caption={Reciprocal Rank Fusion over N ranked lists},
  label={lst:rrf-fusion}]
# src/retrieval/fusion.py
from collections import defaultdict
from src.retrieval.types import RetrievedPassage

def reciprocal_rank_fusion(
    ranked_lists: list[list[RetrievedPassage]],
    k: int = 60,
    top_k: int = 100,
) -> list[RetrievedPassage]:
    scores: dict[str, float] = defaultdict(float)
    registry: dict[str, RetrievedPassage] = {}
    for ranked in ranked_lists:
        for passage in ranked:
            pid = passage["id"]
            scores[pid] += 1.0 / (k + passage["rank"])
            if pid not in registry:
                registry[pid] = passage
    sorted_ids = sorted(scores, key=lambda pid: scores[pid], reverse=True)
    fused: list[RetrievedPassage] = []
    for rank, pid in enumerate(sorted_ids[:top_k], start=1):
        merged = dict(registry[pid])
        merged["score"] = scores[pid]
        merged["rank"] = rank
        merged["source"] = "fused"
        fused.append(merged)
    return fused
\end{lstlisting}

Three decisions in Listing~\ref{lst:rrf-fusion} deserve explanation.
\code{defaultdict(float)} accumulates scores across all lists without
requiring an existence check; a passage that appears in only one list simply
receives one contribution. The \code{registry} dict stores the first
encountered version of each passage --- its \code{text} and \code{meta} are
identical across lists since they come from the same underlying corpus; the
duplicate detection is on \code{id}. The \code{top\_k=100} ceiling exists
because the cross-encoder reranker in the next section is $O(\text{candidates}
\times \text{query length})$ at inference time; passing more than 100
candidates to a CPU cross-encoder reliably exceeds a 200\,ms latency budget.


%% ------------------------------------------------------------
\section{flashrank Cross-Encoder Reranker}
\label{sec:reranker}
%% ------------------------------------------------------------

The design problem is that bi-encoder retrieval scores the query and each
passage independently. The query embedding and the passage embedding are
computed separately and combined only at the very end with a dot product.
This is fast enough to run over millions of passages, but it misses
fine-grained interaction between query terms and passage terms. A
cross-encoder receives the full (query, passage) pair and produces a joint
score, capturing term-level interactions at the cost of being too slow
to run over a full corpus. The governing principle is that the cross-encoder
is a precision stage over a bounded pool, never a retrieval stage over an
unbounded corpus.

\code{flashrank} provides a cross-encoder reranker that runs on CPU without
a GPU dependency. The official README identifies two models for production
use. \code{ms-marco-TinyBERT-L-2-v2} (\textasciitilde{}4\,MB) is the library
default: fast, low-memory. \code{ms-marco-MiniLM-L-12-v2}
(\textasciitilde{}34\,MB) is labelled ``Best Cross-encoder'' in the
official documentation and is the recommended production default for agent
memory workloads where latency budget allows 100--200\,ms for reranking 100
candidates.

\begin{warningbox}
\code{Ranker} loads the cross-encoder model into memory at
\code{\_\_init\_\_} time. Instantiating it inside a per-request function
reloads the model on every call, adding seconds of latency and exhausting
memory. Instantiate \code{Ranker} once at process startup --- at module
level or via dependency injection --- and pass the singleton to all
calling code.
\end{warningbox}

flashrank passages are plain Python dicts with keys \code{id}, \code{text},
and optionally \code{meta}. flashrank exports no \code{Passage} class.
The listing below defines a local \code{FlashPassage} \code{TypedDict}
as an author-defined adapter type to enforce structure before the
\code{RerankRequest} call; it is not part of the flashrank API.

\begin{notebox}
flashrank \code{Ranker.rerank()} output scores are raw cross-encoder logit
outputs --- for example, \code{0.0168}, \code{0.0116}, \code{0.00081}.
They are \emph{not} softmax probabilities bounded in $(0, 1)$. Sort
descending and take the top $K$; do not interpret the absolute values as
confidence scores.
\end{notebox}

\begin{lstlisting}[language=Python,
  caption={FlashPassage adapter, Ranker factory, and rerank\_passages()},
  label={lst:reranker}]
# src/retrieval/reranker.py
from typing import Any
from typing_extensions import TypedDict
from flashrank import Ranker, RerankRequest
from src.retrieval.types import RetrievedPassage

# Author-defined adapter; NOT a flashrank exported type.
class FlashPassage(TypedDict):
    id: int | str
    text: str
    meta: dict[str, Any]

def build_ranker(
    model_name: str = "ms-marco-MiniLM-L-12-v2",
    max_length: int = 256,
    cache_dir: str = "/opt/flashrank",
) -> Ranker:
    return Ranker(model_name=model_name, max_length=max_length,
                 cache_dir=cache_dir)

def rerank_passages(
    ranker: Ranker,
    query: str,
    candidates: list[RetrievedPassage],
    top_k: int = 5,
) -> list[RetrievedPassage]:
    flash_passages: list[FlashPassage] = [
        FlashPassage(id=p["id"], text=p["text"], meta=p["meta"])
        for p in candidates
    ]
    request = RerankRequest(query=query, passages=flash_passages)
    reranked = ranker.rerank(request)
    id_to_original = {p["id"]: p for p in candidates}
    results: list[RetrievedPassage] = []
    for rank, item in enumerate(
        sorted(reranked, key=lambda x: x["score"], reverse=True)[:top_k],
        start=1,
    ):
        base = dict(id_to_original[item["id"]])
        base["score"] = float(item["score"])
        base["rank"] = rank
        base["source"] = "reranked"
        results.append(base)
    return results
\end{lstlisting}

Three decisions in Listing~\ref{lst:reranker} deserve explanation.
\code{id\_to\_original} maps flash passage IDs back to the original
\code{RetrievedPassage} dicts so that all metadata fields survive the
reranking round-trip --- flashrank returns only \code{id}, \code{text},
and \code{score}. The \code{sorted(..., reverse=True)[:top\_k]} slice is
applied to the reranker output rather than relying on the library's
output ordering, because the official flashrank README does not guarantee
that \code{rerank()} returns results in score-descending order. The
\code{max\_length=256} default is appropriate for agent memory entries of
one to three sentences; increasing it to 512 for short passages doubles
attention computation without improving scores.

The base pipeline is now complete. Sections~\ref{sec:hyde}
and~\ref{sec:crag} are opt-in extensions.


%% ------------------------------------------------------------
\section{HyDE: Hypothetical Answer Embedding}
\label{sec:hyde}
%% ------------------------------------------------------------

The design problem is that a short or ambiguous query embeds into a vector
space region that is not close to the relevant passages, even though those
passages clearly answer the query. The dense encoder learned a representation
from passages, and a one-sentence question looks structurally very different
from a one-paragraph answer in that space. HyDE closes this gap by generating
a hypothetical answer --- a passage that \emph{looks like} the answer the
corpus would provide --- and embedding that answer in place of the original
query. The governing principle is that the hypothetical answer is only used
as the dense query vector; BM25 always receives the original query because
exact term matching benefits from the user's own vocabulary, not a generated
paraphrase.

\begin{warningbox}
HyDE adds one full LLM round-trip before dense retrieval. Expected added
latency is 200--500\,ms depending on model and load. It must be activated
explicitly via \code{use\_hyde=True} and should be restricted to queries
with \code{len(query.split()) < 10}. It is disabled by default. The HyDE
call uses a separate, cheaper model (\code{gpt-4o-mini} or
\code{claude-haiku-3}) --- never the primary agent reasoning model.
\end{warningbox}

The hypothetical answer is validated with a \code{HypotheticalAnswer}
Pydantic model before its embedding is computed. A degenerate one-word
output would embed near the centre of the space and return worse results
than the original query. On \code{ValidationError}, the function falls back
silently to the original query, degrading to standard dense retrieval rather
than raising.

\begin{lstlisting}[language=Python,
  caption={HyDE hypothetical answer prompt (hyde\_v1.jinja2)},
  label={lst:hyde-prompt}]
{# prompts/hyde_v1.jinja2 #}
{# Versioned per Chapter 9 §9.1 rules. #}
You are generating a passage for a memory retrieval system.
Write a single paragraph (3–5 sentences) that directly answers the
following query as if it were an entry in the agent's memory corpus.
Do not hedge or qualify. Write the answer, not a description of the answer.

Query: {{ query }}

Answer:
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={HypotheticalAnswer model and generation function},
  label={lst:hyde-generate}]
# src/retrieval/hyde.py
import litellm
import instructor
from pydantic import BaseModel, model_validator

client = instructor.from_litellm(litellm.completion)

class HypotheticalAnswer(BaseModel):
    text: str

    @model_validator(mode="after")
    def must_be_substantive(self) -> "HypotheticalAnswer":
        if len(self.text.split()) < 20:
            raise ValueError(
                "hypothetical answer too short for useful embedding"
            )
        return self

def generate_hypothetical_answer(
    query: str,
    model: str = "gpt-4o-mini",
) -> str:
    try:
        result = client.chat.completions.create(
            model=model,
            messages=[{"role": "user",
                       "content": f"Query: {query}\n\nAnswer:"}],
            response_model=HypotheticalAnswer,
        )
        return result.text
    except Exception:
        return query  # fallback to original query
\end{lstlisting}

Two decisions in Listing~\ref{lst:hyde-generate} deserve explanation.
The broad \code{except Exception} fallback is intentional here: HyDE is an
optional enhancement, and any failure --- network error, validation failure,
or model refusal --- should degrade gracefully to standard dense retrieval
rather than propagating an exception that would abort the entire retrieval
pipeline. The \code{instructor.from\_litellm(litellm.completion)} pattern
patches the LiteLLM completion function to extract and validate the
\code{HypotheticalAnswer} structured output; it does not require a separate
OpenAI or Anthropic client and is therefore consistent with the LiteLLM
abstraction established in Chapter~5.


%% ------------------------------------------------------------
\section{CRAG: Retrieval Grader and Re-Retrieval Branch}
\label{sec:crag}
%% ------------------------------------------------------------

The design problem is that neither the base pipeline nor HyDE can detect
when their output is insufficient. A query whose relevant memories were
never stored, or whose context requires a wider importance threshold, will
silently produce a low-quality context window. The governing principle is
that relevance judgment must be a first-class node in the graph, and that
a negative judgment triggers a second retrieval attempt with widened
parameters before the system falls back to a conservative answer.

The grader is a regular LangGraph node that uses the Pydantic-validated
structured output pattern from Chapter~4 \S4.3. It calls one LiteLLM binary
classification prompt, extracts a \code{GradeResult} via \code{instructor},
and writes \code{retrieval\_grade} into state. The conditional edge reads
that field: \code{"RELEVANT"} routes to generation; \code{"NOT\_RELEVANT"}
routes to re-retrieval. A cycle counter\code{crag\_attempts} caps the loop
at two iterations before the graph forces generation regardless of grade,
preventing infinite re-retrieval loops.

The re-retrieval node calls the same pipeline functions as the first pass
but widens \code{top\_k} from 50 to 80 and relaxes \code{min\_importance}
to 0.0, recovering memories that would have been filtered out by a stricter
threshold.

\begin{lstlisting}[language=Python,
  caption={GradeResult model and retrieval grader node},
  label={lst:crag-grader}]
# src/retrieval/crag_nodes.py
from typing import Literal
from typing_extensions import TypedDict
from pydantic import BaseModel
import litellm
import instructor
from src.retrieval.types import RetrievedPassage

client = instructor.from_litellm(litellm.completion)

class GradeResult(BaseModel):
    grade: Literal["RELEVANT", "NOT_RELEVANT"]
    reason: str  # logged to Langfuse as span metadata (Ch. 23)

class CRAGState(TypedDict):
    query: str
    user_id: str
    final_passages: list[RetrievedPassage]
    retrieval_grade: Literal["RELEVANT", "NOT_RELEVANT"] | None
    crag_attempts: int

GRADER_MODEL = "gpt-4o-mini"

def retrieval_grader_node(state: CRAGState) -> dict:
    passages_text = "\n".join(
        f"{i+1}. {p['text'][:200]}"
        for i, p in enumerate(state["final_passages"])
    )
    result = client.chat.completions.create(
        model=GRADER_MODEL,
        messages=[{
            "role": "user",
            "content": (
                f"Query: {state['query']}\n\nPassages:\n{passages_text}\n\n"
                "Are these passages relevant to the query? "
                "Respond RELEVANT or NOT_RELEVANT with a one-sentence reason."
            ),
        }],
        response_model=GradeResult,
    )
    return {"retrieval_grade": result.grade}
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={CRAG StateGraph with cycle-bounded conditional edge},
  label={lst:crag-graph}]
# src/retrieval/crag_graph.py
from langgraph.graph import StateGraph
from src.retrieval.crag_nodes import CRAGState, retrieval_grader_node
from src.retrieval.crag_nodes import GradeResult

MAX_CRAG_ATTEMPTS = 2

def re_retrieval_node(state: CRAGState) -> dict:
    # wider top_k, relaxed importance — same pipeline functions
    from src.retrieval.bm25_stage import bm25_retrieve
    from src.retrieval.dense_stage import dense_retrieve
    from src.retrieval.fusion import reciprocal_rank_fusion
    from src.retrieval.reranker import rerank_passages
    # caller injects index, vector_store, ranker via context_schema
    raise NotImplementedError("inject via context_schema — see Ch. 11 §11.7")

def should_re_retrieve(
    state: CRAGState,
) -> Literal["re_retrieve", "generate"]:
    if state["crag_attempts"] >= MAX_CRAG_ATTEMPTS:
        return "generate"
    if state["retrieval_grade"] == "NOT_RELEVANT":
        return "re_retrieve"
    return "generate"

def build_crag_graph() -> StateGraph:
    builder = StateGraph(state_schema=CRAGState)
    builder.add_node("retrieval_grader", retrieval_grader_node)
    builder.add_node("re_retrieval", re_retrieval_node)
    builder.set_entry_point("retrieval_grader")
    builder.add_conditional_edges(
        "retrieval_grader",
        should_re_retrieve,
        {"re_retrieve": "re_retrieval", "generate": "__end__"},
    )
    builder.add_edge("re_retrieval", "retrieval_grader")
    return builder.compile()
\end{lstlisting}

Three decisions in Listings~\ref{lst:crag-grader} and~\ref{lst:crag-graph}
deserve explanation. \code{should\_re\_retrieve} checks
\code{crag\_attempts >= MAX\_CRAG\_ATTEMPTS} \emph{before} checking the
grade, ensuring the cycle limit is enforced even if the grader consistently
returns \code{NOT\_RELEVANT}. The \code{reason} field on \code{GradeResult}
is included not for programmatic use but for observability: Chapter~23
attaches it as span metadata in Langfuse, making grader decisions inspectable
after the fact without re-running the query. \code{StateGraph(state\_schema=CRAGState)} is the correct constructor; the \code{config\_schema}
parameter is deprecated as of LangGraph v0.6.0 --- use \code{context\_schema=}
if immutable per-run context (such as the injected \code{Ranker} singleton)
needs to be passed into nodes.


%% ------------------------------------------------------------
\section{Quality Metrics as pytest Assertions}
\label{sec:metrics}
%% ------------------------------------------------------------

The final design problem is proving that the pipeline actually works. Teams
frequently declare retrieval ``good enough'' after passing a handful of manual
spot checks, which proves nothing about systematic behaviour across a
diverse query set. The governing principle is that retrieval quality is
expressed as machine-checkable thresholds over a committed fixture dataset,
not as aspirational guidelines, and that the tests call the full pipeline
end-to-end rather than mocking individual stages.

Three metrics are measured. \textbf{Precision@5} is the fraction of the
five returned passages that are relevant: $\text{P@5} = |\text{relevant}
\cap \text{top5}| / 5$. \textbf{Recall@5} is the fraction of all relevant
passages that appear in the top five: $\text{R@5} = |\text{relevant} \cap
\text{top5}| / |\text{relevant}|$. \textbf{MRR} is the mean reciprocal
rank of the first relevant passage across all queries; a query with no
relevant passage in the top five contributes 0. Each metric is averaged over
all fixture queries before the threshold comparison --- a single difficult
query does not break the suite, but systematic degradation does.

The fixture dataset requirements are non-negotiable constraints:

\begin{itemize}
  \item Minimum 20 \code{EvalQuery} records committed to
    \code{tests/fixtures/retrieval\_eval.json} --- never generated at test
    runtime.
  \item Each query has 2--5 explicitly labelled \code{relevant\_ids}.
  \item Corpus size \(\_\geq\)\, 200 passages. A 20-passage corpus makes
    Recall@5 trivially achievable; 200 makes it a real constraint.
  \item Dense retrieval uses \code{QdrantClient(":memory:")} with
    Chapter~10's \code{create\_agent\_memory\_collection()} and
    \code{setup\_payload\_indexes()} called in the fixture setup --- CI
    uses the same collection schema as production.
\end{itemize}

\begin{lstlisting}[language=Python,
  caption={EvalQuery model and metric functions},
  label={lst:metrics}]
# src/retrieval/metrics.py
from pydantic import BaseModel

class EvalQuery(BaseModel):
    query_id: str
    query: str
    relevant_ids: list[str]  # len in [2, 5]

def precision_at_k(
    retrieved_ids: list[str], relevant_ids: list[str], k: int = 5
) -> float:
    top_k = retrieved_ids[:k]
    hits = len(set(top_k) & set(relevant_ids))
    return hits / k

def recall_at_k(
    retrieved_ids: list[str], relevant_ids: list[str], k: int = 5
) -> float:
    if not relevant_ids:
        return 0.0
    top_k = retrieved_ids[:k]
    hits = len(set(top_k) & set(relevant_ids))
    return hits / len(relevant_ids)

def mean_reciprocal_rank(
    retrieved_ids: list[str], relevant_ids: list[str]
) -> float:
    rel_set = set(relevant_ids)
    for i, pid in enumerate(retrieved_ids, start=1):
        if pid in rel_set:
            return 1.0 / i
    return 0.0
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={pytest threshold assertions over the full pipeline},
  label={lst:quality-tests}]
# tests/retrieval/test_retrieval_quality.py
import json, statistics, pytest
from qdrant_client import QdrantClient
from src.retrieval.metrics import (
    EvalQuery, precision_at_k, recall_at_k, mean_reciprocal_rank
)
from src.vector_store.collection_setup import create_agent_memory_collection
from src.vector_store.index_setup import setup_payload_indexes

P5_THRESHOLD  = 0.7
R5_THRESHOLD  = 0.8
MRR_THRESHOLD = 0.6

@pytest.fixture(scope="module")
def eval_queries() -> list[EvalQuery]:
    raw = json.loads(
        open("tests/fixtures/retrieval_eval.json").read()
    )
    return [EvalQuery(**r) for r in raw]

@pytest.fixture(scope="module")
def qdrant_memory():
    client = QdrantClient(":memory:")
    create_agent_memory_collection(client, dims=1536)
    setup_payload_indexes(client, "agent_memory")
    return client

def run_pipeline(query: str, qdrant_client) -> list[str]:
    """Returns top-5 passage IDs from full pipeline."""
    # bm25_retrieve + dense_retrieve + reciprocal_rank_fusion
    # + rerank_passages — abbreviated here; full impl in src/
    raise NotImplementedError

def test_precision_at_5(eval_queries, qdrant_memory):
    scores = [
        precision_at_k(run_pipeline(q.query, qdrant_memory),
                       q.relevant_ids)
        for q in eval_queries
    ]
    assert statistics.mean(scores) >= P5_THRESHOLD

def test_recall_at_5(eval_queries, qdrant_memory):
    scores = [
        recall_at_k(run_pipeline(q.query, qdrant_memory),
                    q.relevant_ids)
        for q in eval_queries
    ]
    assert statistics.mean(scores) >= R5_THRESHOLD

def test_mrr(eval_queries, qdrant_memory):
    scores = [
        mean_reciprocal_rank(run_pipeline(q.query, qdrant_memory),
                             q.relevant_ids)
        for q in eval_queries
    ]
    assert statistics.mean(scores) >= MRR_THRESHOLD
\end{lstlisting}

Two decisions in Listing~\ref{lst:quality-tests} deserve explanation.
\code{scope="module"} on both fixtures ensures that the in-memory Qdrant
instance and the eval dataset are constructed once per test module run,
not once per test function --- construction is not cheap and the fixtures
are read-only within the test module. The \code{statistics.mean(scores)}
aggregation is intentional: the threshold applies to the \emph{mean} across
all 20+ queries, not to each query individually. A single hard query with
zero recall is informative but not a build-breaker; if the mean drops below
threshold, the pipeline has regressed systematically.

Chapter~12 extends this pipeline by adding a Neo4j knowledge graph retrieval
stage that contributes a third ranked list to \code{reciprocal\_rank\_fusion}.
The \code{RetrievedPassage} type, the RRF function signature, and the
quality test fixture are all designed to accommodate that extension without
modification.
