%% ============================================================
%%  Chapter 16 — MCP: Hosts, Clients & Servers
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%%  Final revised version — June 2026
%% ============================================================

\chapter{MCP: Hosts, Clients, and Servers}
\label{chap:mcp}

The failure mode that motivates this chapter is invisible until it is
expensive: an agent that embeds HTTP logic, credential management, and
schema definitions directly inside its own graph.
When a second agent needs the same capability, the implementation is
copied. When the schema changes, every copy breaks independently.
When the service moves behind authentication, every caller is patched
in isolation. The result is not a platform --- it is a distributed ball
of mud whose failure surface grows with every new agent added to the
system.

The Model Context Protocol (MCP) resolves this through a three-layer
capability-delivery architecture. A \emph{server} owns a capability
domain and publishes it as a versioned, schema-carrying interface.
A \emph{client} speaks the MCP wire protocol, opens sessions against
servers, and translates those capabilities into objects the agent runtime
can consume. A \emph{host} owns the reasoning loop and never touches the
wire protocol; it receives typed tool objects and nothing else.
Each layer changes independently: servers scale horizontally, clients
swap transports, and the host graph evolves without knowledge of either.

The sections in this chapter follow the dependency order imposed by those
layers. Section~\ref{sec:mcp-roles} draws the role boundaries before any
code exists, because conflating host and client is the single most common
mistake in MCP integrations. Section~\ref{sec:mcp-decorators} builds the
server-side capability surface. Section~\ref{sec:mcp-transport} explains
why the transport that works on a developer's laptop fails under production
load. Section~\ref{sec:mcp-toolkit} converts server capabilities into
LangGraph tool nodes. Section~\ref{sec:mcp-multi} extends that wiring to
N~servers. Section~\ref{sec:mcp-oauth} ensures credentials never appear
in source. Section~\ref{sec:mcp-capstone} assembles all six layers into a
single runnable system.

\begin{table}[h]
\centering
\caption{MCP domain terms introduced in this chapter}
\label{tab:mcp-terms}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
Host &
  The process that owns the LLM reasoning loop.
  In this book, always a compiled LangGraph \code{StateGraph}.
  Consumes tool objects; never touches MCP wire format. \\
Client &
  The in-process adapter that speaks MCP wire format, opens
  \code{ClientSession} objects, and delivers a \code{list[BaseTool]}
  to the host. \\
Server &
  An independently-deployed capability provider built with FastMCP.
  Has no knowledge of the host graph topology. \\
Tool &
  A callable published by a server.
  Its JSON Schema is generated from the function's Python signature;
  its natural-language description comes from the function's docstring. \\
Resource &
  A data endpoint at a URI template; fetched by the client on demand
  via \code{resources/read}, never pushed by the server. \\
Prompt &
  A server-side message template fetched via \code{prompts/get};
  keeps prompt logic versioned alongside the capability it describes. \\
Streamable HTTP &
  The MCP transport that exposes a \code{POST /mcp} endpoint.
  The only transport suitable for horizontally-scaled
  production deployments. \\
\bottomrule
\end{tabularx}
\end{table}

\begin{lstlisting}[language=bash,
                   caption={Chapter~16 source layout},
                   label={lst:ch16-layout}]
ch16/
├── servers/
│   ├── inventory_server.py   # @mcp.tool() x2, uvicorn :8000
│   ├── pricing_server.py     # @mcp.tool() x2, uvicorn :8001
│   └── orders_server.py      # @mcp.tool() x2, uvicorn :8002
├── auth/
│   └── bearer_provider.py    # MCPAuthSettings + BearerTokenProvider
└── agent/
    └── capstone_agent.py     # MultiServerMCPClient + StateGraph
\end{lstlisting}


%% ------------------------------------------------------------------
\section{The Three Roles Cannot Share a Process Boundary}
\label{sec:mcp-roles}
%% ------------------------------------------------------------------

The design problem every MCP integration faces first is attribution:
which piece of code is responsible for which concern.
The naive approach is to give the agent class a method that opens an
HTTP connection, parses a JSON response, and presents the result to
the LLM. This works for one tool on one server and breaks the moment
either side changes independently --- which in production they always do.
The principle is strict role separation: each MCP actor has exactly
one responsibility, and the interfaces between them are the only stable
surface.

Table~\ref{tab:mcp-terms} defined the three roles. Their runtime
boundaries are best understood as a data-flow sequence, reproduced
below as an annotated ASCII diagram.

\begin{lstlisting}[language=bash,
                   caption={MCP role boundaries and data flow},
                   label={lst:mcp-boundary}]
┌──────────────────────────────────────────────┐
│  HOST: LangGraph StateGraph                  │
│  Owns the LLM reasoning loop.                │
│  Receives list[BaseTool] — nothing else.     │
│  Never imports mcp.* wire-protocol modules.  │
└───────────────────┬──────────────────────────┘
                    │  list[BaseTool]
┌───────────────────▼──────────────────────────┐
│  CLIENT: MultiServerMCPClient                │
│  Speaks MCP wire protocol.                   │
│  Opens one ClientSession per server.         │
│  Runs initialize() capability handshake.     │
│  Fetches tools/list → populates BaseTool.    │
└──────┬──────────────────────┬────────────────┘
       │  streamable_http      │  stdio
┌──────▼──────────┐   ┌───────▼──────────────┐
│ FastMCP Server A│   │ FastMCP Server B      │
│ POST /mcp :8000 │   │ subprocess (dev only) │
└─────────────────┘   └──────────────────────┘
\end{lstlisting}

Two properties of this diagram must be understood before any code is
written. First, the host never imports anything from the MCP wire-protocol
layer. It receives a \code{list[BaseTool]} from the client, passes it to
\code{ToolNode} and \code{model.bind\_tools()}, and proceeds. From the
host's perspective, those tools could have come from any source. Replacing
the transport --- from stdio to HTTP, or rotating authentication schemes
--- requires no change to the graph. Second, the client holds all session
state. The host never holds a reference to a \code{ClientSession} object.

\begin{notebox}
The MCP session lifecycle has three mandatory phases.
\texttt{initialize()} performs the capability handshake and negotiates
protocol version. \texttt{tools/list} retrieves the tool manifest.
Subsequent \texttt{tools/call} messages dispatch individual invocations.
\code{MultiServerMCPClient} executes all three phases transparently;
the host sees only the output of the second: a \code{list[BaseTool]}
with \code{name}, \code{description}, and \code{args\_schema} populated
from the server manifest.
\end{notebox}

The three imports below express the role boundary in code.
No logic appears; the point is that each import belongs to exactly one
layer and the layers share no modules.

\begin{lstlisting}[language=Python,
                   caption={One import per MCP role},
                   label={lst:mcp-role-imports}]
# Server role — SDK variant (mcp >= 1.2)
from mcp.server.fastmcp import FastMCP

# Client role — langchain-mcp-adapters
from langchain_mcp_adapters.client import MultiServerMCPClient

# Host role — LangGraph
from langgraph.graph import StateGraph
\end{lstlisting}

Two decisions in Listing~\ref{lst:mcp-role-imports} deserve explanation.
\code{FastMCP} is imported from \code{mcp.server.fastmcp}, which is the
official MCP Python SDK (\code{pip install mcp}); FastMCP~1.0 was
incorporated into this SDK in 2024.
The standalone \code{fastmcp} package (\code{pip install fastmcp}) uses
\code{@mcp.tool} without parentheses, while the SDK uses \code{@mcp.tool()}
with parentheses; this chapter uses the SDK variant throughout because it
is the package \code{langchain-mcp-adapters} itself depends on, confirmed
in its own quickstart documentation.
The \code{MultiServerMCPClient} is the only client-layer symbol the host
ever references; it abstracts transport, session, and protocol negotiation
behind a single \code{await client.get\_tools()} call.

Understanding what each role does \emph{not} own is as important as
understanding what it does. The server owns no knowledge of the agent
topology. The client owns no business logic. The host owns no transport
configuration. With those boundaries established, the next question is
what the server must publish for the client to produce a useful tool list.

%% ------------------------------------------------------------------
\section{A Tool's Docstring Is Its Public Contract}
\label{sec:mcp-decorators}
%% ------------------------------------------------------------------

The design problem in server construction is information asymmetry: the
LLM must decide which tool to call based solely on the tool's description,
but that description is generated from the same source the developer writes
for human readers. The naive approach is terse docstrings that satisfy a
linter but carry no semantic content --- \emph{``Search products.''} instead
of a precise description of what the function returns, in what format,
and when it should be called relative to other tools. This produces
incorrect LLM tool selection under any input that could plausibly match
more than one tool. The principle is that the docstring is a public API
contract between the server author and every LLM that will ever consume
this tool; it must be written to that standard.

FastMCP's \code{@mcp.tool()} decorator serialises the decorated
function's docstring verbatim into the MCP \code{tools/list} response.
That response populates the \code{description} field of each
\code{BaseTool} object. There is no secondary documentation mechanism:
a missing docstring produces an empty \code{description}; a vague one
produces incorrect routing.

\begin{warningbox}
FastMCP generates JSON Schema from Python type annotations through Pydantic.
If a parameter has no \code{Annotated[T,~Field(description="...")]} wrapper,
its schema entry carries no description and the LLM receives no guidance on
what value to supply. Missing parameter descriptions are the most common
cause of tool-call argument errors in production.
Every parameter of every \code{@mcp.tool()} function must carry an
\code{Annotated} field with a non-empty \code{description}.
\end{warningbox}

\begin{lstlisting}[language=Python,
                   caption={FastMCP server with \texttt{@mcp.tool()},
                            \texttt{@mcp.resource()}, and \texttt{@mcp.prompt()}},
                   label={lst:mcp-decorators}]
from mcp.server.fastmcp import FastMCP
from pydantic import Field
from typing import Annotated

mcp = FastMCP(
    "inventory",
    instructions="Inventory management: search products, fetch specs.",
)

@mcp.tool()
async def search_products(
    query: Annotated[str, Field(
        description="Full-text product search query")],
    limit: Annotated[int, Field(
        description="Max results to return, 1-50")] = 10,
) -> list[dict]:
    """Search the product catalogue and return matching SKUs with stock.

    Returns a list of dicts with keys: sku (str), name (str),
    stock (int), price_usd (float). Returns an empty list when
    no products match. Call this first to confirm a SKU exists
    before calling get_product_spec.
    """
    ...  # implementation omitted

@mcp.resource("inventory://products/{sku}")
async def get_product_spec(sku: str) -> str:
    """Return the full product specification as Markdown text.
    Raises ValueError if the SKU does not exist."""
    ...

@mcp.prompt()
def low_stock_alert(threshold: int = 5) -> list[dict]:
    """Instruct the agent to prioritise SKUs with low stock."""
    return [{"role": "system",
             "content": f"Flag any SKU with stock <= {threshold}."}]
\end{lstlisting}

Three decisions in Listing~\ref{lst:mcp-decorators} deserve explanation.
The \code{search\_products} docstring contains three distinct pieces of
information: what the function returns, the exact schema of the return
value, and the calling precondition relative to another tool in the same
server. All three are necessary for correct multi-step LLM tool sequencing;
none of them is available from the function signature alone.
The \code{@mcp.resource()} decorator takes a URI template; the client
fetches resources on demand via \code{resources/read}, making them
pull-only --- the server never pushes resource content to the client.
The \code{@mcp.prompt()} decorator returns a \code{list[dict]} matching
the MCP \code{PromptMessage} schema (\code{role}/\code{content} keys);
it is fetched via \code{prompts/get}, which means prompt logic is
versioned and deployed alongside the capability it describes rather
than scattered across host graphs.

\begin{notebox}
The \code{FastMCP} constructor's \code{instructions} parameter becomes
the server-level system prompt sent to the client during the
\code{initialize()} capability handshake.
It scopes all tools on this server in the LLM's context window.
Keep it to one sentence naming the domain; longer instructions compete
with tool descriptions for the model's attention.
\end{notebox}

A completed server definition is necessary but not sufficient for
production deployment. The transport the server uses to receive requests
determines whether it can survive a real traffic pattern.

%% ------------------------------------------------------------------
\section{The Transport That Works Locally Will Fail in Production}
\label{sec:mcp-transport}
%% ------------------------------------------------------------------

The design problem is that the most convenient server transport couples
the server's lifecycle to the host's process. When the client starts an
MCP server as a subprocess using the stdio transport, the server cannot
exist without the host, cannot be restarted independently, cannot be
load-balanced, and cannot expose a health-check endpoint.
The naive approach is to keep this transport in production because it
requires no network configuration. This is a deployment anti-pattern:
any failure of the host process --- a crash, a rolling update, a
scaling event --- kills all in-flight tool calls on every server it
was running. The principle is process isolation: a production server
must be an independently-addressable service whose lifecycle the host
does not control.

The stdio transport initialises with \code{mcp.run(transport="stdio")}.
The client starts the server as a child process and communicates over
pipes via \code{StdioServerParameters}. This is zero-configuration
and correct for local development and CI, but the server cannot
outlive the client process, cannot accept connections from more than
one client simultaneously, and cannot sit behind a reverse proxy.

\emph{Streamable HTTP} is the transport the MCP specification recommends
for production. The server exposes a single \code{POST /mcp} endpoint
(\code{mcp.run(transport="streamable-http")}).
Each session is a stateless HTTP exchange, enabling horizontal
replication and independent restarts without disrupting other sessions.

\begin{notebox}
The transport name string differs between the server and the client.
The server's \code{mcp.run()} call uses the hyphenated form
\code{"streamable-http"} as specified by the MCP protocol.
The \code{MultiServerMCPClient} config dict uses the underscored form
\code{"streamable\_http"} as required by the \code{langchain-mcp-adapters}
library. Passing the hyphenated form to \code{MultiServerMCPClient}
silently falls back to the default transport rather than raising an error.
Keep the two forms straight: hyphens on the server, underscores on the
client.
\end{notebox}

For production, wrapping the FastMCP server in a FastAPI application
adds three capabilities the bare \code{mcp.run()} launcher cannot
provide: a \code{/health} endpoint for load-balancer probes, a standard
middleware stack for logging and error handling, and proper lifespan
management for resource acquisition and release.
The FastMCP server exposes its ASGI app through
\code{mcp.get\_asgi\_app()}, which can be mounted into any
ASGI-compatible framework.

\begin{lstlisting}[language=Python,
                   caption={FastMCP server mounted into FastAPI, served by uvicorn},
                   label={lst:mcp-production-server}]
# ch16/servers/inventory_server.py  (transport + launch section)
import uvicorn
from fastapi import FastAPI
from mcp.server.fastmcp import FastMCP

mcp = FastMCP("inventory")
# @mcp.tool() definitions from Listing 16.2 go here

# mcp.get_asgi_app() returns the raw ASGI callable.
# Mounting it into FastAPI adds /health and middleware.
api = FastAPI(title="Inventory MCP Server")
api.mount("/mcp", mcp.get_asgi_app())

@api.get("/health")
async def health() -> dict:
    return {"status": "ok", "server": mcp.name}

if __name__ == "__main__":
    # String "module:app" is required for workers > 1.
    # Passing the object directly disables multi-process mode.
    uvicorn.run(
        "inventory_server:api",
        host="0.0.0.0",
        port=8000,
        workers=4,
        log_level="info",
    )
\end{lstlisting}

Three decisions in Listing~\ref{lst:mcp-production-server} deserve
explanation. \code{mcp.get\_asgi\_app()} extracts the ASGI callable
from the FastMCP server; mounting it at \code{"/mcp"} on the FastAPI
application preserves the \code{POST /mcp} path the client expects.
The \code{uvicorn.run()} call uses a string reference
(\code{"inventory\_server:api"}) rather than the object directly;
this is required for the \code{workers=4} argument to take effect
because uvicorn's multi-worker mode forks new processes and needs a
string-importable entry point --- passing the object restricts the
server to a single worker regardless of the \code{workers} parameter,
silently ignoring the scaling intent. The \code{/health} route exists
specifically to satisfy Kubernetes readiness probes and load-balancer
health checks, neither of which can reach an MCP endpoint without first
receiving a 200 response on a non-MCP path.

\begin{warningbox}
Do not use \code{mcp.run()} in production server files.
It is a development convenience wrapper with no worker model, no
graceful shutdown hooks, and no integration with process supervisors
(systemd, Docker, Kubernetes). The only visual difference between
a working and a broken production server file is often this single
method call.
\end{warningbox}

The client-side transport configuration for a server deployed with this
pattern uses \code{"streamable\_http"} (underscore) and points to the
mounted path.

\begin{lstlisting}[language=Python,
                   caption={Client transport configuration for a production server},
                   label={lst:mcp-transport-config}]
# This dict is the only change needed when moving
# a server from stdio to streamable_http.
# Note: transport key uses underscore (client convention),
# not hyphen (server/MCP-spec convention).
server_config: dict = {
    "transport": "streamable_http",   # underscore, not hyphen
    "url": "http://localhost:8000/mcp",
    "headers": {
        "Authorization": "Bearer <token>",  # replaced in Section 16.6
    },
}
\end{lstlisting}

Two decisions in Listing~\ref{lst:mcp-transport-config} deserve
explanation. The comment on the \code{transport} line is not decorative;
the hyphen/underscore asymmetry between the server and client APIs
is a live source of silent misconfiguration in the ecosystem, and
making it visible at the definition site is cheaper than debugging a
silent transport fallback. The \code{headers} field is valid only for
\code{streamable\_http} and \code{sse} transports; passing it with a
\code{stdio} config silently has no effect rather than raising an error,
so the placeholder comment anchors where the real token injection of
Section~\ref{sec:mcp-oauth} will appear.

With the server deployed and the transport configured, the remaining
question is how the client converts the server's capability manifest
into objects the host can consume.

%% ------------------------------------------------------------------
\section{The Client Is the Only Layer That Speaks Protocol}
\label{sec:mcp-toolkit}
%% ------------------------------------------------------------------

The design problem at the host-client boundary is a type mismatch:
a LangGraph \code{StateGraph} operates on \code{BaseTool} objects and
knows nothing about MCP sessions, capability manifests, or wire-protocol
messages. The naive approach is to have agent code call
\code{ClientSession} methods directly, embedding protocol concerns
inside the host graph. This dissolves the role boundary established in
Section~\ref{sec:mcp-roles}: the host becomes responsible for session
lifecycle, and the client layer becomes an unnecessary abstraction.
The principle is that \code{MultiServerMCPClient.get\_tools()} is the
single crossing point between the protocol layer and the host layer;
everything the host needs crosses that boundary, and nothing else does.

\code{MultiServerMCPClient} accepts a config dict mapping logical server
names to transport configurations. Calling
\code{await client.get\_tools()} opens a \code{ClientSession} per
configured server, performs the \code{initialize()} handshake, fetches
each server's \code{tools/list} response, and returns a flat
\code{list[BaseTool]} whose \code{name}, \code{description}, and
\code{args\_schema} fields are populated from the server manifests.

\begin{warningbox}
\code{client.get\_tools()} is a coroutine and must be awaited.
Writing \code{tools = client.get\_tools()} without \code{await}
assigns a coroutine object, not a list.
\code{ToolNode(coroutine)} accepts this without raising at construction
time, but the node will fail on its first invocation with a type error
that points to dispatch logic rather than to this line.
Always verify the return type with \code{assert isinstance(tools, list)}
during development.
\end{warningbox}

The tool list returned by \code{get\_tools()} is passed to two places
in the host graph: \code{ToolNode(tools)} creates the dispatch node,
and \code{model.bind\_tools(tools)} attaches the tool schemas to the
model so it knows which tools are available. Both must receive the
same list; an inconsistency between them produces a model that requests
tools the dispatch node cannot route.

\begin{lstlisting}[language=Python,
                   caption={Wiring \code{MultiServerMCPClient} to a LangGraph \code{StateGraph}},
                   label={lst:mcp-toolkit-wiring}]
import asyncio
from langchain_mcp_adapters.client import MultiServerMCPClient
from langgraph.graph import StateGraph, MessagesState, START
from langgraph.prebuilt import ToolNode, tools_condition
from langchain.chat_models import init_chat_model

async def build_graph() -> StateGraph:
    client = MultiServerMCPClient({
        "inventory": {
            "transport": "streamable_http",
            "url": "http://localhost:8000/mcp",
        }
    })
    tools = await client.get_tools()   # coroutine — must await

    model = init_chat_model("openai:gpt-4.1")

    def call_model(state: MessagesState) -> dict:
        # bind_tools inside the node, not at graph construction time
        return {"messages": [
            model.bind_tools(tools).invoke(state["messages"])
        ]}

    builder = StateGraph(MessagesState)
    builder.add_node("call_model", call_model)
    builder.add_node("tools", ToolNode(tools))  # "tools" — exact key
    builder.add_edge(START, "call_model")
    builder.add_conditional_edges("call_model", tools_condition)
    builder.add_edge("tools", "call_model")
    return builder.compile()
\end{lstlisting}

Three decisions in Listing~\ref{lst:mcp-toolkit-wiring} deserve
explanation. \code{model.bind\_tools(tools)} is called inside
\code{call\_model} rather than at graph construction time; this ensures
that if the tool list is refreshed between invocations (for example,
after a server upgrade publishes new capabilities), the model always
receives the current schema without requiring a graph rebuild.
The \code{add\_node} key for the dispatch node is the string
\code{"tools"}, not a variable; \code{tools\_condition} from
\code{langgraph.prebuilt} is hardcoded to route to this exact string,
and a different key raises a \code{KeyError} at runtime that points
to the conditional-edge evaluation rather than to the
\code{add\_node} call, making it difficult to trace.
The \code{build\_graph()} function is \code{async} because
\code{get\_tools()} is a coroutine; the entire graph construction
is deferred to the \code{asyncio.run()} call at the entry point,
which is the correct pattern for async LangGraph programs.

\begin{notebox}
\code{langchain-mcp-adapters} also exposes a lower-level path:
\code{from langchain\_mcp\_adapters.tools import load\_mcp\_tools}.
This function accepts an already-open \code{ClientSession} and returns
tools from that session.
Use it with the \code{async with client.session("server-name") as session}
context manager when you need explicit session lifetime control --- for
example, to pin a long-lived session and avoid the per-call session
overhead on latency-sensitive paths.
For the common case, \code{get\_tools()} is the correct entry point.
\end{notebox}

A single-server graph is straightforward. The next problem is that
production systems have multiple capability domains, and the host graph
must not require modification when a new one is added.

%% ------------------------------------------------------------------
\section{The Host Graph Must Not Know How Many Servers Exist}
\label{sec:mcp-multi}
%% ------------------------------------------------------------------

The design problem in multi-server integration is coupling between
operational topology and graph logic. If the host graph must be modified
every time a new server is added, the graph becomes a registration point
for infrastructure decisions. The naive approach is to call
\code{get\_tools()} once per server and concatenate the lists in the
agent file. This requires the host to enumerate servers, which is the
same coupling expressed differently. The principle is that
\code{MultiServerMCPClient} is the federation layer: N~server configs
enter, one flat \code{list[BaseTool]} exits, and the host graph built
in Section~\ref{sec:mcp-toolkit} is identical whether N is one or twenty.

Extending from one server to N requires only extending the config dict.
The client opens sessions in parallel, fetches tool manifests from all
configured servers, and returns a single concatenated list. The
\code{StateGraph} does not change.

\begin{lstlisting}[language=Python,
                   caption={Three heterogeneous servers in one \code{MultiServerMCPClient}},
                   label={lst:mcp-multi-client}]
from langchain_mcp_adapters.client import MultiServerMCPClient

client = MultiServerMCPClient({
    "inventory": {
        "transport": "streamable_http",
        "url": "http://inventory-svc:8000/mcp",
        "headers": {"Authorization": "Bearer <inv_token>"},
    },
    "pricing": {
        "transport": "streamable_http",
        "url": "http://pricing-svc:8001/mcp",
        "headers": {"Authorization": "Bearer <price_token>"},
    },
    "legacy_sku": {
        "transport": "stdio",
        "command": "python",
        "args": ["/opt/legacy/sku_server.py"],
        # headers not supported on stdio transport — omit entirely
    },
})

# Single await returns a flat list from all three servers.
# The StateGraph from Listing 16.4 is unchanged.
tools = await client.get_tools()
\end{lstlisting}

Two decisions in Listing~\ref{lst:mcp-multi-client} deserve explanation.
The stdio and \code{streamable\_http} transports coexist in the same
config dict; \code{MultiServerMCPClient} handles transport negotiation
per-server, so a legacy subprocess-based tool can be federated alongside
HTTP microservices without requiring the legacy tool to be ported.
The \code{headers} key is absent from the \code{legacy\_sku} config
and the comment explains why; adding a \code{headers} key to a stdio
config silently has no effect rather than raising an error, and the
explicit omission with a comment prevents a future maintainer from
adding it and spending time debugging why their tokens are not being
sent.

\begin{warningbox}
If two servers publish a tool with the same \code{name}, the second
registration silently shadows the first in the flat tool list.
\code{MultiServerMCPClient} raises no error on name collision.
In a multi-team environment, two servers will eventually register a
tool named \code{search} or \code{list}.
Enforce a naming convention at the server level (e.g., prefix every
\code{@mcp.tool()} function name with the server's domain:
\code{inventory\_search\_products}, \code{pricing\_get\_price})
and document this convention in the server's
\code{FastMCP(..., instructions=...)} string.
\end{warningbox}

\begin{tipbox}
For graceful degradation when a server may be temporarily unreachable
at startup, use the lower-level \code{load\_mcp\_tools} path per server
inside individual try/except blocks, accumulate the available tools, and
log which servers failed rather than raising. The host graph receives
a reduced tool list but continues to function.
Use \code{async with client.session("server-name") as session} to
acquire the session for each server, then call
\code{await load\_mcp\_tools(session)} inside the try block.
\end{tipbox}

The multi-server config above contains placeholder token strings.
Hardcoded credentials are a security vulnerability; they also cannot
rotate without a code deployment. The next section replaces them with
a credential management layer.

%% ------------------------------------------------------------------
\section{Credentials Must Not Live in Code}
\label{sec:mcp-oauth}
%% ------------------------------------------------------------------

The design problem is that every token string in a config dict has a
lifecycle the dict cannot express: tokens expire, are revoked, and
must be scoped to the minimum permissions required.
The naive approach is to read a token from an environment variable
at startup and embed it in the config dict. This fails because the
token is evaluated once and frozen for the lifetime of the process;
a token valid for one hour causes all tool calls to fail silently after
that hour with a 401 from the MCP server, with no indication in the
host graph of what went wrong.
The principle is that credential acquisition must be lazy and renewable:
a token is fetched at the moment it is needed and refreshed before it
expires, without restarting the process.

A \emph{bearer token provider} is a callable object that encapsulates
the OAuth2 client-credentials grant and caches the result until
expiry. Using a callable object rather than a plain function allows
the token cache to live as instance state, and allows the provider to
be injected into any component that needs an \code{Authorization} header.
The credentials themselves are read from the environment via
\code{pydantic\_settings.BaseSettings}, which ensures they are never
present in source and are masked in logs through \code{SecretStr}.

\begin{lstlisting}[language=Python,
                   caption={\code{MCPAuthSettings} and \code{BearerTokenProvider}},
                   label={lst:mcp-auth-provider}]
import time
import httpx
from pydantic import SecretStr
from pydantic_settings import BaseSettings, SettingsConfigDict

class MCPAuthSettings(BaseSettings):
    # Reads MCP_CLIENT_ID, MCP_CLIENT_SECRET, MCP_TOKEN_URL,
    # MCP_SCOPE from environment or .env file.
    model_config = SettingsConfigDict(
        env_file=".env",
        secrets_dir="/run/secrets",
        env_prefix="MCP_",
    )
    client_id: str
    client_secret: SecretStr
    token_url: str
    scope: str = "mcp:tools"

class BearerTokenProvider:
    def __init__(self, settings: MCPAuthSettings) -> None:
        self._s = settings
        self._token: str | None = None
        self._expires_at: float = 0.0

    def __call__(self) -> str:
        """Return a valid Bearer token, refreshing before expiry."""
        if self._token and time.monotonic() < self._expires_at - 30:
            return self._token
        self._refresh()
        return self._token  # type: ignore[return-value]

    def _refresh(self) -> None:
        resp = httpx.post(self._s.token_url, data={
            "grant_type": "client_credentials",
            "client_id": self._s.client_id,
            "client_secret": self._s.client_secret.get_secret_value(),
            "scope": self._s.scope,
        }, timeout=10)
        resp.raise_for_status()
        body = resp.json()
        self._token = body["access_token"]
        self._expires_at = time.monotonic() + body["expires_in"]
\end{lstlisting}

Three decisions in Listing~\ref{lst:mcp-auth-provider} deserve
explanation. \code{time.monotonic()} is used for expiry comparison
rather than \code{time.time()}; monotonic time is immune to system
clock adjustments and NTP corrections, both of which can cause a
wall-clock-based expiry check to skip a refresh or trigger one
prematurely on cloud instances that frequently synchronise clocks.
The 30-second safety margin (\code{self.\_expires\_at - 30}) accounts
for network round-trip time between the token refresh and the first
MCP request that uses the new token; a token refreshed at exactly
\\(T\\) may expire before reaching a server if \code{\_refresh()} takes
29~seconds under load. \code{SecretStr} prevents the credential from
appearing in \code{print()}, \code{logging}, or Pydantic's
\code{model\_dump()}; the raw value is accessible only via
\code{.get\_secret\_value()}, called exclusively at the point of the
HTTP request.

The server-side counterpart validates incoming tokens before any tool
call is dispatched. A FastAPI dependency provides this as a reusable
guard attached to the application that wraps the FastMCP ASGI mount.

\begin{lstlisting}[language=Python,
                   caption={Server-side Bearer token validation dependency},
                   label={lst:mcp-verify-bearer}]
from fastapi import FastAPI, HTTPException, Security
from fastapi.security import HTTPAuthorizationCredentials, HTTPBearer
from mcp.server.fastmcp import FastMCP

bearer_scheme = HTTPBearer()

async def verify_bearer(
    creds: HTTPAuthorizationCredentials = Security(bearer_scheme),
) -> str:
    """Validate the Bearer token; raise 401 if invalid."""
    token = creds.credentials
    if not await _introspect_token(token):   # your auth-server call
        raise HTTPException(status_code=401, detail="Invalid token")
    return token

mcp = FastMCP("inventory")
api = FastAPI(dependencies=[Security(verify_bearer)])
api.mount("/mcp", mcp.get_asgi_app())

@api.get("/health")
async def health() -> dict:
    return {"status": "ok"}
\end{lstlisting}

Two decisions in Listing~\ref{lst:mcp-verify-bearer} deserve
explanation. Token validation delegates to \code{\_introspect\_token},
which represents a call to the authorisation server's token introspection
endpoint or local JWKS validation; inline JWT verification without key
rotation support is omitted intentionally because key rotation is a
deployment-specific concern that falls outside the MCP integration layer.
Applying \code{Security(verify\_bearer)} as a global FastAPI dependency
(\code{FastAPI(dependencies=[...])}) means every request to every
route --- including \code{POST /mcp} and all future routes --- must
pass token validation; the \code{/health} endpoint is deliberately
excluded by mounting it outside the secured router if unauthenticated
probes are required.

\begin{warningbox}
The \code{BearerTokenProvider.\_\_call\_\_()} in
Listing~\ref{lst:mcp-auth-provider} is evaluated \emph{once per client
construction} when the return value is embedded in the headers dict:
\code{\{"Authorization": f"Bearer \{provider()\}"\}}.
For long-running agents where the session outlives the token's validity,
call \code{build\_graph()} on a schedule to reconstruct the client with
a fresh evaluation rather than relying solely on the provider's cache.
The safe production pattern is to construct the graph at the start of
each agent invocation rather than caching the compiled graph across
invocation boundaries when token lifetimes are shorter than agent
session lifetimes.
\end{warningbox}

\begin{tipbox}
For least-privilege scoping, instantiate one \code{BearerTokenProvider}
per server with the minimum scope for that server's operations:
\code{scope="inventory:read inventory:write"} for the inventory server,
\code{scope="pricing:read"} for a read-only pricing server.
Each provider maintains its own token cache independently, so a
pricing-server token expiry does not affect the inventory-server cache.
\end{tipbox}

With role separation established, servers built, transport configured,
the client wired to the host, N-server federation understood, and
credentials managed, all six layers are ready to be assembled into
a single runnable system.

%% ------------------------------------------------------------------
\section{The Complete System: Three Servers, One Agent}
\label{sec:mcp-capstone}
%% ------------------------------------------------------------------

The design problem this section solves is integration under constraints:
the preceding sections established each layer in isolation, but a
production agent requires them to operate coherently, with failures in
any layer producing diagnosable errors rather than silent misbehaviour.
The naive integration approach is a single script that imports from
every layer and wires them sequentially without preserving the role
boundaries. This produces a system that works on the first run and
becomes unmaintainable the moment any layer is updated. The principle
is that integration code should be as thin as the layer boundaries
permit: each layer has already solved its problem, and the capstone
only composes those solutions.

The system consists of three FastMCP servers, each owning one business
domain. The inventory server exposes product search and specification
retrieval. The pricing server exposes price lookup and discount
calculation. The orders server exposes draft order creation and order
status queries. All three are deployed as independent services under
uvicorn using the FastAPI-wrapper pattern from
Section~\ref{sec:mcp-transport}. The agent is a LangGraph
\code{StateGraph} that federates all three through a single
\code{MultiServerMCPClient} with per-server OAuth2 scoping from
Section~\ref{sec:mcp-oauth}.

All three server files share the same structure. The inventory server
is shown in full; pricing and orders follow the identical pattern with
different tool names and ports.

\begin{lstlisting}[language=Python,
                   caption={Inventory server (ch16/servers/inventory\_server.py)},
                   label={lst:capstone-inventory}]
import uvicorn
from fastapi import FastAPI
from mcp.server.fastmcp import FastMCP
from pydantic import Field
from typing import Annotated

mcp = FastMCP("inventory",
              instructions="Search products and retrieve specifications.")

@mcp.tool()
async def inventory_search_products(
    query: Annotated[str, Field(description="Full-text product search")],
    limit: Annotated[int, Field(description="Max results, 1-50")] = 10,
) -> list[dict]:
    """Search the product catalogue. Returns list of dicts with keys:
    sku (str), name (str), stock (int), price_usd (float).
    Returns empty list if no match. Call before
    inventory_get_product_spec to confirm the SKU exists."""
    ...   # implementation

@mcp.tool()
async def inventory_get_product_spec(
    sku: Annotated[str, Field(description="Product SKU identifier")],
) -> str:
    """Return full product specification as Markdown.
    Raises ValueError if the SKU does not exist."""
    ...   # implementation

api = FastAPI(title="Inventory MCP Server")
api.mount("/mcp", mcp.get_asgi_app())

@api.get("/health")
async def health() -> dict:
    return {"status": "ok"}

if __name__ == "__main__":
    uvicorn.run("inventory_server:api",
                host="0.0.0.0", port=8000, workers=4)
\end{lstlisting}

Two decisions in Listing~\ref{lst:capstone-inventory} deserve
explanation. Every tool function name is prefixed with its domain
(\code{inventory\_}) rather than being a short generic name such
as \code{search}; this is the server-level naming convention
recommended in Section~\ref{sec:mcp-multi} to prevent silent tool
shadowing when multiple servers are federated. Each docstring names
the return schema, the empty-result contract, and the calling
precondition relative to the other tool on this server; omitting the
precondition causes the LLM to call \code{inventory\_get\_product\_spec}
without first verifying the SKU exists, producing a \code{ValueError}
during tool dispatch that the host graph cannot recover from without a
dedicated error-handling node.

The capstone agent wires all layers. Its length across two listings
is justified because it is the integration layer; no individual concern
within it is new.

\begin{lstlisting}[language=Python,
                   caption={Capstone agent: graph construction
                            (ch16/agent/capstone\_agent.py, part~1)},
                   label={lst:capstone-build}]
import asyncio
from langchain_mcp_adapters.client import MultiServerMCPClient
from langgraph.graph import StateGraph, MessagesState, START
from langgraph.prebuilt import ToolNode, tools_condition
from langchain.chat_models import init_chat_model
from langchain_core.messages import HumanMessage
from ch16.auth.bearer_provider import BearerTokenProvider, MCPAuthSettings

async def build_graph() -> StateGraph:
    inv   = BearerTokenProvider(MCPAuthSettings(scope="inventory:read inventory:write"))
    price = BearerTokenProvider(MCPAuthSettings(scope="pricing:read"))
    orders = BearerTokenProvider(MCPAuthSettings(scope="orders:write orders:read"))

    client = MultiServerMCPClient({
        "inventory": {"transport": "streamable_http",
                      "url": "http://localhost:8000/mcp",
                      "headers": {"Authorization": f"Bearer {inv()}"}},
        "pricing":   {"transport": "streamable_http",
                      "url": "http://localhost:8001/mcp",
                      "headers": {"Authorization": f"Bearer {price()}"}},
        "orders":    {"transport": "streamable_http",
                      "url": "http://localhost:8002/mcp",
                      "headers": {"Authorization": f"Bearer {orders()}"}},
    })
    tools = await client.get_tools()
    model = init_chat_model("openai:gpt-4.1")

    def call_model(state: MessagesState) -> dict:
        return {"messages": [
            model.bind_tools(tools).invoke(state["messages"])
        ]}

    builder = StateGraph(MessagesState)
    builder.add_node("call_model", call_model)
    builder.add_node("tools", ToolNode(tools))
    builder.add_edge(START, "call_model")
    builder.add_conditional_edges("call_model", tools_condition)
    builder.add_edge("tools", "call_model")
    return builder.compile()
\end{lstlisting}

Three decisions in Listing~\ref{lst:capstone-build} deserve explanation.
One \code{BearerTokenProvider} is instantiated per server with the
minimum scope for that server's operations; scope separation ensures
that a compromised pricing token cannot authorise order creation.
Each provider call (\code{inv()}, \code{price()}, \code{orders()})
is evaluated at \code{build\_graph()} time and embedded in the headers
dict; as the warning in Section~\ref{sec:mcp-oauth} noted, calling
\code{build\_graph()} at the start of each agent invocation ensures
a fresh token evaluation when token lifetimes are shorter than session
lifetimes. The graph topology is identical to Listing~\ref{lst:mcp-toolkit-wiring}
with three servers instead of one; the host graph did not change when
the server count tripled, validating the design principle from
Section~\ref{sec:mcp-multi}.

\begin{lstlisting}[language=Python,
                   caption={Capstone agent: entry point and invocation
                            (ch16/agent/capstone\_agent.py, part~2)},
                   label={lst:capstone-run}]
async def main() -> None:
    graph = await build_graph()

    result = await graph.ainvoke({
        "messages": [HumanMessage(content=(
            "Find SKU-42, get its current price, "
            "then create a draft order for 5 units."
        ))]
    })

    for msg in result["messages"]:
        # msg.type: "human" | "ai" | "tool"
        print(f"[{msg.type:6s}] {str(msg.content)[:120]}")

if __name__ == "__main__":
    asyncio.run(main())
\end{lstlisting}

Two decisions in Listing~\ref{lst:capstone-run} deserve explanation.
The invocation prompt is deliberately multi-step and cross-server:
\emph{find SKU-42} dispatches \code{inventory\_search\_products};
\emph{get its current price} dispatches \code{pricing\_get\_price}
using the SKU returned by the first tool call; \emph{create a draft
order} dispatches \code{orders\_create\_draft\_order} using the SKU
and confirmed price from the preceding calls. This sequence exercises
the \code{tools\,$\to$\,call\_model} cycle three times and validates
that context accumulated across tool calls is preserved in
\code{MessagesState}. The \code{msg.type} field in the output loop
exposes the full call sequence --- human input, AI reasoning steps,
tool responses, and final AI answer --- which is the minimal observable
needed to confirm multi-server cross-domain dispatch is working;
in production this loop is replaced by LangFuse trace instrumentation,
which Chapter~\ref{chap:observability} adds to the \code{call\_model}
node without modifying any other part of the graph.

\begin{notebox}
The explicit \code{StateGraph} construction in
Listing~\ref{lst:capstone-build} is intentional.
\code{create\_react\_agent} from \code{langgraph.prebuilt} produces
an equivalent graph in fewer lines, but with a fixed topology that
cannot be extended with the tracing nodes, checkpointing, and guardrail
nodes added in later chapters. Always build the graph explicitly when
you know it will be modified downstream.
\end{notebox}

Chapter~\ref{chap:a2a} addresses the next unsolved problem this
system exposes: the capstone agent can consume capabilities from MCP
servers, but it cannot itself be discovered and invoked as a capability
by another agent. The Agent-to-Agent (A2A) protocol resolves this by
promoting the host graph to an addressable peer; doing so requires
no changes to the MCP layer built here.
