%% ============================================================
%%  Chapter 15 --- ReAct: Tool Execution Loops
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%%  Dependencies: langgraph >= 0.2.0,
%%                langchain-core >= 0.3.0,
%%                langchain-litellm >= 0.3.0,
%%                pydantic >= 2.7.0,
%%                pydantic-settings >= 2.3.0,
%%                langfuse >= 2.35.0,
%%                pytest >= 8.0.0
%% ============================================================

\chapterquote{The prebuilt works until it doesn't. Know what it hides before you ship it.}{anon}

The production failure this chapter prevents is silent and compounding. An
agent calls a tool, the tool returns a JSON blob with a missing required
field, and the model receives that blob as a \code{ToolMessage} with its
structural defect intact. No exception is raised. No validation fires. The
model reasons from broken data, produces a plausible-sounding next step,
invokes another tool based on that step, and the second tool amplifies the
original error into the third. The graph completes with a status of
\texttt{success}. The answer is wrong in ways that take a human minutes to
detect and hours to trace back to the original malformed \code{ToolMessage}.
The cost is not a single bad response --- it is the debugging time, the
eroded trust, and the absence of the one record that would make the failure
traceable: a per-tool trace span.

The solution this chapter builds has four layers. The first is
\emph{conceptual}: understanding what \code{create\_react\_agent} actually
compiles to, and why it is a convenience shortcut over the explicit graph
architecture the book established in Chapter~1, not a replacement for it.
The second is \emph{structural}: using \code{get\_graph()} and
\code{get\_subgraphs()} to read the compiled topology before trusting it,
and recognising the three conditions that require abandoning the prebuilt
for a hand-built \code{StateGraph}. The third is \emph{mechanical}: the
\code{ToolMessage} protocol --- its fields, their contracts, and the
\code{add\_messages} reducer that governs how each result enters the message
chain. The fourth is \emph{defensive}: a \code{ValidationNode} gate that
intercepts every tool result before the model reads it, backed by a
Langfuse span that records the input, output, and latency of every tool
call.

The sections are ordered by dependency. Section~\ref{sec:react-convenience}
establishes the conceptual frame before any code appears, so that the
reader understands the trade-off before committing to the shortcut.
Section~\ref{sec:react-internals} teaches graph introspection and defines
the switch-to-custom triggers; it requires the conceptual frame from
Section~\ref{sec:react-convenience} to motivate the inspection.
Section~\ref{sec:react-toolmessage} examines reducer mechanics;
understanding them is a prerequisite for Section~\ref{sec:react-validation},
which exploits the reducer's replace-by-id behaviour to implement the
validation gate. Section~\ref{sec:react-langfuse} adds observability and has
no upstream code dependency, but is placed last because instrumentation
without structural understanding produces data no one can interpret.

This chapter introduces five terms defined in Table~\ref{tab:ch15-terms}.

\begin{table}[h]
\centering
\caption{Domain terms introduced in Chapter~15}
\label{tab:ch15-terms}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
ReAct loop &
  A two-node graph topology (\code{agent} $\leftrightarrow$ \code{tools})
  that encodes the Reason--Act--Observe cycle; compiled to a
  \code{CompiledStateGraph} by \code{create\_react\_agent}. \\
\code{ToolMessage} &
  The \code{BaseMessage} subtype that carries a tool's serialised return
  value back into \code{MessagesState}; its \code{tool\_call\_id} must
  exactly match the originating \code{AIMessage.tool\_calls[i]["id"]}. \\
\code{add\_messages} reducer &
  The function assigned to the \code{messages} field in
  \code{MessagesState}; appends messages by default, but replaces
  in-place when an incoming message shares an \code{id} with an
  existing one. \\
\code{ValidationNode} &
  A dedicated graph node (implemented in Chapter~4 \S4.3) that
  validates the last \code{ToolMessage} against a Pydantic schema
  before the agent node executes. \\
\code{DrawableGraph} &
  The return type of \code{CompiledStateGraph.get\_graph()}; a
  serialisable topology representation with an \code{.print\_ascii()}
  method for terminal rendering. \\
\bottomrule
\end{tabularx}
\end{table}

\medskip
\noindent\textbf{Source layout for this chapter:}
\begin{verbatim}
ch15/
  01a_react_minimal.py          # §15.1 — minimal create_react_agent
  01b_explicit_equivalent.py    # §15.1 — hand-rolled two-node equivalent
  02_inspect_react_graph.py     # §15.2 — topology introspection
  03_tool_message_history.py    # §15.3 — ToolMessage chain & stream_mode
  04_validation_node_react.py   # §15.4 — ValidationNode gate
  05_langfuse_tool_traces.py    # §15.5 — @observe per tool call
\end{verbatim}


%% ------------------------------------------------------------
\section{The Prebuilt Is a Shortcut, Not a Starting Point}
\label{sec:react-convenience}
%% ------------------------------------------------------------

The design problem is architectural capture: an engineer who reaches for
\code{create\_react\_agent} before understanding what it compiles to will,
sooner or later, hit one of its hard boundaries --- no domain state beyond
\code{messages}, no routing logic beyond \emph{``did the model request a
tool?''}, no in-flight state editing --- and will discover that migrating
from the prebuilt to a hand-built graph at that point requires rewriting
code that is already in production. The naive approach is to start with the
prebuilt because it reduces the code to fit on half a screen, which it does;
the failure is that its boundaries are invisible until violated. The
governing principle is: understand the topology the prebuilt compiles to
before writing the first production line, and treat every prebuilt as a
named point in the design space rather than the origin of that space.

The \emph{ReAct pattern} originates in Yao et al.\ (2022) as a prompting
strategy: ask the model to interleave \texttt{Thought}, \texttt{Action}, and
\texttt{Observation} steps in its output, and use each \texttt{Observation}
to ground the next \texttt{Thought}. LangGraph encodes only the structural
consequence of that strategy: a loop where the model runs, optionally
requests tools, the tools run, and control returns to the model.
The prompt that elicits interleaved reasoning is the engineer's
responsibility; the library provides only the loop topology.

What \code{create\_react\_agent} produces is a \code{CompiledStateGraph}
containing exactly two application nodes: \code{agent}, which invokes
the model, and \code{tools}, which is a \code{ToolNode} that dispatches
the \code{tool\_calls} present in the most recent \code{AIMessage}.
A single conditional edge governed by a \code{has\_tool\_calls} check routes
either back to \code{agent} (if the last message contains tool calls) or
to \code{\_\_end\_\_} (if it does not). This is the same \code{CompiledStateGraph}
runtime that Chapter~1 produced by calling \code{StateGraph.compile()} ---
no different execution engine, no different checkpointing behaviour, no
different streaming API. The prebuilt saves approximately thirty lines of
wiring code and hides three things in exchange: the conditional edge
function, the \code{ToolNode} construction, and the state schema
restriction to \code{MessagesState}.

\begin{table}[h]
\centering
\caption{Prebuilt vs.\ explicit \code{StateGraph}: trade-offs}
\label{tab:ch15-tradeoffs}
\begin{tabularx}{\textwidth}{lXX}
\toprule
\textbf{Dimension}        & \textbf{\code{create\_react\_agent}} & \textbf{Explicit \code{StateGraph}} \\
\midrule
Lines to running agent    & $\sim$10 & $\sim$45 \\
State schema              & \code{MessagesState} (default) & Any \code{TypedDict} \\
Domain state fields       & Via \code{state\_schema} override (constrained) & Native, any reducer \\
Routing logic             & \code{has\_tool\_calls} only & Arbitrary callable \\
In-flight state editing   & Constrained (interrupt list only) & Full \code{update\_state()} control \\
Topology auditability     & Hidden until \code{get\_graph()} called & Explicit and readable \\
\bottomrule
\end{tabularx}
\end{table}

The listings below make the equivalence concrete by showing both forms
side by side. Listing~\ref{lst:react-minimal} is the prebuilt form.
Listing~\ref{lst:react-explicit} is the hand-built equivalent.

\begin{lstlisting}[language=Python,
  caption={Minimal \texttt{create\_react\_agent} with two arithmetic tools},
  label={lst:react-minimal}]
# ch15/01a_react_minimal.py
from langchain_litellm import ChatLiteLLM
from langchain_core.tools import tool
from langchain_core.messages import HumanMessage
from langgraph.prebuilt import create_react_agent

@tool
def add(a: int, b: int) -> int:
    """Add two integers and return the result."""
    return a + b

@tool
def multiply(a: int, b: int) -> int:
    """Multiply two integers and return the result."""
    return a * b

model  = ChatLiteLLM(model="gpt-4o-mini")
agent  = create_react_agent(model=model, tools=[add, multiply])
result = agent.invoke(
    {"messages": [HumanMessage(content="What is (3 + 4) * 5?")]}
)
print(result["messages"][-1].content)
\end{lstlisting}

Two decisions in Listing~\ref{lst:react-minimal} deserve explanation.
The tools are plain functions decorated with \code{@tool}; \code{create\_react\_agent}
wraps them in a \code{ToolNode} internally and calls
\code{model.bind\_tools([add, multiply])} before constructing the agent node,
so no explicit schema registration is needed. The input is a plain \code{dict}
matching the \code{MessagesState} schema (\code{\{"messages": [...]\}}); this
is the correct call-site form because \code{invoke()} accepts any value that
satisfies the state schema by key name, and \code{MessagesState} is a
\code{TypedDict} whose only required key is \code{messages}.

\begin{lstlisting}[language=Python,
  caption={Hand-rolled two-node equivalent of the prebuilt loop},
  label={lst:react-explicit}]
# ch15/01b_explicit_equivalent.py
import operator
from typing import Annotated, Literal
from typing_extensions import TypedDict
from langchain_core.messages import BaseMessage, AIMessage
from langchain_core.tools import tool
from langchain_litellm import ChatLiteLLM
from langgraph.graph import StateGraph, START, END
from langgraph.graph.message import add_messages
from langgraph.prebuilt import ToolNode

class State(TypedDict):
    messages: Annotated[list[BaseMessage], add_messages]

@tool
def add(a: int, b: int) -> int:
    """Add two integers and return the result."""
    return a + b

@tool
def multiply(a: int, b: int) -> int:
    """Multiply two integers and return the result."""
    return a * b

TOOLS = [add, multiply]
model  = ChatLiteLLM(model="gpt-4o-mini")
bound  = model.bind_tools(TOOLS)

def agent_node(state: State) -> dict:
    return {"messages": [bound.invoke(state["messages"])]}

def should_continue(state: State) -> Literal["tools", "__end__"]:
    last = state["messages"][-1]
    if isinstance(last, AIMessage) and last.tool_calls:
        return "tools"
    return "__end__"

builder = StateGraph(State)
builder.add_node("agent", agent_node)
builder.add_node("tools", ToolNode(TOOLS))
builder.add_edge(START, "agent")
builder.add_conditional_edges("agent", should_continue)
builder.add_edge("tools", "agent")
graph = builder.compile()
\end{lstlisting}

Three decisions in Listing~\ref{lst:react-explicit} deserve explanation.
\code{model.bind\_tools(TOOLS)} is called once at module level rather than
inside \code{agent\_node}; binding at construction time ensures the tool
schemas are attached to the model before any graph execution begins and
avoids re-binding on every node invocation, which would re-serialise the
schemas on every call. \code{should\_continue} returns the string literal
\code{"\_\_end\_\_"} rather than the imported \code{END} sentinel as its
direct return value; \code{add\_conditional\_edges} treats string return
values as node names, and \code{"\_\_end\_\_"} is the internal node name
that \code{END} resolves to when no \code{path\_map} is provided. When
\code{END} (the sentinel object, not the string) is one of the routing
targets, a \code{path\_map} is required --- the absence of a \code{path\_map}
here is correct because both targets are strings. \code{ToolNode(TOOLS)}
receives the same list as \code{bind\_tools} to guarantee that the tool
schemas the model learns and the tools the node dispatches are derived from
the same source.

\begin{notebox}
\code{create\_react\_agent} accepts \code{interrupt\_before} and
\code{interrupt\_after} parameters that place LangGraph \code{interrupt()}
calls at node boundaries without a custom graph. This covers the common
HITL case of pausing before or after tool execution. The trigger for
moving to a custom graph is not HITL itself but \emph{in-flight state
editing}: if a human reviewer needs to call \code{graph.update\_state()}
to modify state between the interrupt and the resume, the explicit graph
structure is required to place that edit at the correct node boundary.
Chapter~21 \S21.4 covers this pattern in full.
\end{notebox}

The prebuilt topology is now understood; the question is how to verify at
runtime that a compiled graph built from \code{create\_react\_agent} matches
the topology described here, and how to confirm the switch-to-custom
triggers before they become blocking production issues.


%% ------------------------------------------------------------
\section{Trust Requires Inspection, Not Assumption}
\label{sec:react-internals}
%% ------------------------------------------------------------

The design problem is unverified assumption: accepting a compiled graph from
a factory function and proceeding directly to \code{invoke()} is structurally
equivalent to deploying code whose call graph has never been read. The
happy path exercises only one route through the topology; the
\code{agent~$\to$~tools} back-edge and the terminal condition are structural
features that observed behaviour alone cannot confirm. The naive approach is
to run the agent against a test prompt and verify the output --- which
confirms only that the happy path works. The governing principle is: a
compiled graph must be readable before it is trusted, and LangGraph provides
\code{get\_graph()} and \code{get\_subgraphs()} specifically for that purpose.

\code{CompiledStateGraph.get\_graph(xray=True)} returns a \code{DrawableGraph},
a serialisable representation of the compiled topology. When \code{xray} is
\code{False} (the default), composite nodes such as \code{ToolNode} appear
as single opaque boxes. When \code{xray} is \code{True}, LangGraph expands
subgraphs recursively, revealing the internal nodes and edges that composite
nodes encapsulate. \code{DrawableGraph.print\_ascii()} renders the result as
a terminal diagram using only standard library output.

\code{get\_subgraphs(recurse=True)} returns an iterator of
\code{(name,~subgraph)} pairs for every embedded compiled graph inside the
top-level graph. \code{ToolNode} is itself a compiled \code{Pregel}
execution unit; iterating with \code{recurse=True} and reading
\code{subgraph.nodes.keys()} reveals its internal node names. This
attribute is stable across LangGraph versions and is safe to assert on in
\code{pytest} tests that verify expected topology.

\begin{lstlisting}[language=Python,
  caption={Inspecting compiled topology with \texttt{get\_graph} and \texttt{get\_subgraphs}},
  label={lst:react-inspect}]
# ch15/02_inspect_react_graph.py
from langchain_litellm import ChatLiteLLM
from langchain_core.tools import tool
from langgraph.prebuilt import create_react_agent

@tool
def web_search(query: str) -> str:
    """Search the web and return a summary of results."""
    return f"Top result for '{query}': (stub)"

model = ChatLiteLLM(model="gpt-4o-mini")
agent = create_react_agent(model=model, tools=[web_search])

# ── 1. Opaque top-level view ─────────────────────────────────
print("=== top-level (xray=False) ===")
agent.get_graph(xray=False).print_ascii()

# ── 2. Expanded view — ToolNode internals visible ────────────
print("\n=== expanded (xray=True) ===")
agent.get_graph(xray=True).print_ascii()

# ── 3. Enumerate all embedded compiled subgraphs ─────────────
print("\n=== subgraph inventory ===")
for name, subgraph in agent.get_subgraphs(recurse=True):
    node_names = list(subgraph.nodes.keys())
    print(f"  {name!r:30s}  nodes={node_names}")
\end{lstlisting}

Two decisions in Listing~\ref{lst:react-inspect} deserve explanation.
\code{get\_graph} is called twice --- once with \code{xray=False} and once
with \code{xray=True} --- and the comparison is deliberate: the first call
shows the graph as the prebuilt presents it (what is being trusted), and
the second shows what it actually contains (what is running). Seeing both
in a single terminal session establishes the habit of checking before
trusting. \code{subgraph.nodes.keys()} reads from the \code{PregelProtocol}
object returned by the iterator; the \code{nodes} attribute on that object
is the internal node registry and its keys are the names that appear in the
\code{get\_graph()} ASCII diagram, making it possible to assert in a
\code{pytest} fixture that the expected topology is present before a
deployment ships.

\begin{warningbox}
\code{get\_graph(xray=True)} does not mutate the compiled graph and is safe
to call in production for diagnostics. \code{get\_subgraphs(recurse=True)}
returns an iterator that yields live \code{PregelProtocol} objects; do not
modify the objects it yields. Both methods reflect the topology at the time
\code{compile()} was called; adding nodes or edges after compilation is not
supported by the LangGraph API and will produce undefined behaviour.
\end{warningbox}

The three conditions that mandate switching to a custom \code{StateGraph}
follow directly from what inspection reveals.

\begin{enumerate}
  \item \textbf{Custom routing.} The prebuilt's conditional edge calls
    \code{has\_tool\_calls} and nothing else. Any routing logic beyond
    that binary check --- dispatching to different tool subsets by tool
    class, branching on a confidence score, routing to a human-approval
    node for specific tool names --- requires an explicit callable passed
    to \code{add\_conditional\_edges}. There is no parameter on
    \code{create\_react\_agent} to replace this function.

  \item \textbf{Domain state.} Any key beyond \code{messages} in the
    graph state requires a custom \code{TypedDict} with explicit
    reducers. Examples: a \code{tool\_errors} counter with
    \code{operator.add}, a \code{token\_budget} with a cap-and-truncate
    reducer, a \code{cycle\_count} for the termination policy from
    Chapter~2. The \code{state\_schema} parameter on
    \code{create\_react\_agent} accepts a custom schema, but the
    prebuilt's internal routing logic still assumes \code{messages} is the
    only operationally meaningful field.

  \item \textbf{In-flight state editing.} \code{interrupt\_before} and
    \code{interrupt\_after} on the prebuilt support pausing and resuming.
    The trigger for leaving the prebuilt is needing to call
    \code{graph.update\_state()} between an interrupt and a resume ---
    for example, to inject a corrected \code{ToolMessage} or to modify an
    accumulated counter before the agent node re-executes.
    Chapter~21 \S21.4 covers this pattern.
\end{enumerate}

Inspection reveals the topology; it does not reveal the protocol that
governs what travels through the back-edge after every tool call. That
protocol --- the \code{ToolMessage} contract and the reducer that merges
it into state --- is what the next section establishes.


%% ------------------------------------------------------------
\section{The Message Chain Is a Protocol, Not a Log}
\label{sec:react-toolmessage}
%% ------------------------------------------------------------

The design problem is hidden coupling: the \code{ToolMessage} that carries
a tool's result back into state must match the specific \code{tool\_call\_id}
of the \code{AIMessage} that triggered the call, or the model receives a
message chain that is structurally incoherent at the OpenAI message
protocol level. The naive approach is to return a string from a tool
function and trust LangGraph to handle the bookkeeping --- which it does
correctly when the tool is dispatched through \code{ToolNode}, but fails
silently when a developer constructs a \code{ToolMessage} manually in a
test fixture, a retry handler, or a custom node without matching the
\code{tool\_call\_id}. The governing principle is: \code{ToolMessage} is a
protocol artefact with three mandatory fields that each carry an
explicit contract, and that contract must be understood before any manual
construction or state patching is attempted.

\code{ToolMessage} carries three mandatory fields. \code{content} is the
serialised tool result as a string; structured data must be
\code{json.dumps()}-ed before being placed here because the message
protocol boundary is text. \code{tool\_call\_id} must be an exact copy of
the \code{id} field from the corresponding entry in
\code{AIMessage.tool\_calls}; this is the identifier the model uses to
associate each result with the request that produced it, and a mismatch
causes the model to receive a history it cannot reconcile. \code{name} is
the tool's function name as a string; it appears in Langfuse traces and in
the model's context and must match the \code{@tool}-decorated function name
exactly to avoid ambiguity in multi-tool runs.

The \code{add\_messages} reducer, introduced as the \code{messages} field
reducer in Chapter~7 \S7.1, governs how each incoming message merges into
the existing list. Its default path is append: if the incoming message's
\code{id} does not appear in the existing list, the message is added at the
end. Its exception path is replace: if the incoming message's \code{id}
matches an existing message's \code{id}, the new message replaces the old
one in-place. This replace-by-id behaviour is not an edge case --- it is the
mechanism that enables the validation gate in Section~\ref{sec:react-validation}
to overwrite a malformed \code{ToolMessage} with a corrected one, without
the model ever seeing the original.

\begin{warningbox}
When the model issues multiple tool calls in a single \code{AIMessage},
\code{AIMessage.tool\_calls} contains one entry per call, each with a
distinct \code{id}. \code{ToolNode} dispatches all calls and appends one
\code{ToolMessage} per call, each bearing the matching \code{id}.
If you construct \code{ToolMessage} objects manually --- in a test fixture,
in a retry handler, or in a custom tool node --- iterate
\code{AIMessage.tool\_calls} and bind each result to its originating
\code{id}. A \code{ToolMessage} whose \code{tool\_call\_id} does not
appear in the preceding \code{AIMessage.tool\_calls} will cause
the model to return an error or silently ignore that result on the
next reasoning step.
\end{warningbox}

The shape of the message chain after two sequential tool calls has a
specific structure that every node in the graph receives on its next
execution:

\begin{verbatim}
HumanMessage(content="What is (3+4)*5?")
AIMessage(tool_calls=[{id:"c1", name:"add",      args:{a:3, b:4}}])
ToolMessage(content="7",  tool_call_id="c1", name="add")
AIMessage(tool_calls=[{id:"c2", name:"multiply", args:{a:7, b:5}}])
ToolMessage(content="35", tool_call_id="c2", name="multiply")
AIMessage(content="The answer is 35.")
\end{verbatim}

The model receives this entire list on every reasoning step. After ten
tool calls the list contains twenty-two messages, and after twenty it
contains forty-two. Chapter~7 \S7.2 covers \code{trim\_messages()} for
windowing this growth; Chapter~7 \S7.3 covers the summarisation node that
compresses history into a single \code{SystemMessage} when the list exceeds
a token budget.

\begin{lstlisting}[language=Python,
  caption={Streaming updates to observe each \texttt{ToolMessage} as it enters state},
  label={lst:react-toolmsg}]
# ch15/03_tool_message_history.py
import json
from langchain_litellm import ChatLiteLLM
from langchain_core.tools import tool
from langchain_core.messages import HumanMessage, AIMessage, ToolMessage
from langchain_core.runnables import RunnableConfig
from langgraph.prebuilt import create_react_agent
from langgraph.checkpoint.memory import InMemorySaver

@tool
def add(a: int, b: int) -> int:
    """Add two integers and return the result."""
    return a + b

@tool
def multiply(a: int, b: int) -> int:
    """Multiply two integers and return the result."""
    return a * b

model       = ChatLiteLLM(model="gpt-4o-mini")
checkpointer = InMemorySaver()
agent       = create_react_agent(
    model=model, tools=[add, multiply], checkpointer=checkpointer
)
config: RunnableConfig = {"configurable": {"thread_id": "ch15-demo-1"}}

for chunk in agent.stream(
    {"messages": [HumanMessage(content="What is (3+4)*5?")]},
    config=config,
    stream_mode="updates",
):
    node_name, updates = next(iter(chunk.items()))
    for msg in updates.get("messages", []):
        tc = getattr(msg, "tool_call_id", None)
        print(f"[{node_name}] {type(msg).__name__}"
              + (f"  tool_call_id={tc}" if tc else ""))

snapshot = agent.get_state(config)
print("\n--- accumulated message chain ---")
for msg in snapshot.values["messages"]:
    preview = str(msg.content)[:55].replace("\n", " ")
    print(f"  {type(msg).__name__:15s}  {preview}")
\end{lstlisting}

Three decisions in Listing~\ref{lst:react-toolmsg} deserve explanation.
\code{stream\_mode="updates"} is used rather than \code{stream\_mode="values"};
with \code{"updates"} each chunk is a \code{dict} keyed by the node name
that produced it (e.g.,
\code{\{"tools": \{"messages": [ToolMessage(...)]\}\}})
and contains only the delta for that step, making it straightforward to
observe which node injected each \code{ToolMessage} and in which order.
\code{stream\_mode="values"} emits the full accumulated state after every
node execution --- for a two-tool interaction that is six full copies of the
growing message list, all containing the same \code{HumanMessage} repeated.
A \code{checkpointer} is passed to \code{create\_react\_agent} because
\code{get\_state()} requires a checkpointer to retrieve the accumulated
state after streaming completes; without one, the call raises a
\code{ValueError}. The \code{snapshot.values} attribute on the returned
\code{StateSnapshot} is a \code{dict} of the complete state at the final
checkpoint; \code{snapshot.values["messages"]} is the full accumulated list,
not a delta.

\begin{notebox}
To inject a synthetic \code{ToolMessage} into a running graph --- for
testing, for manual retry, or for state patching after a human interrupt
--- use \code{graph.update\_state(config, \{"messages": [ToolMessage(...)]\},
as\_node="tools")}. The \code{as\_node} parameter tells the runtime which
node is conceptually producing the update, which determines which outgoing
edges are evaluated next. Without \code{as\_node}, the state update is
applied without triggering any routing logic. Chapter~21 \S21.4 shows
this pattern in the context of HITL state editing.
\end{notebox}

Knowing how the message chain accumulates makes the next failure mode
visible: the model reads every \code{ToolMessage} in that chain without
any schema check, and a structurally invalid payload is indistinguishable
from a valid one at the message layer. Section~\ref{sec:react-validation}
intercepts that payload before the model sees it.


%% ------------------------------------------------------------
\section{Malformed Tool Output Must Never Reach the Model}
\label{sec:react-validation}
%% ------------------------------------------------------------

The design problem is epistemic contamination: the model cannot detect
that a \code{ToolMessage} contains structurally invalid data, because the
message layer carries content as a string with no schema enforcement.
A JSON blob with a missing required field, a numeric value encoded as a
string, or an empty object where a populated record was expected will all
be consumed by the model and incorporated into its next reasoning step
as though they were correct. The naive approach --- catching exceptions
inside the tool function --- fails because most structural violations are
not exceptions: they are valid JSON that does not match the expected schema.
The governing principle is: every tool result must pass a Pydantic
validation gate before the model's next reasoning step, and that gate must
be a graph node with its own routing authority, not an inline check inside
the tool.

Chapter~4 \S4.3 defined \code{ValidationNode} as a reusable graph node
that reads the last \code{ToolMessage} from state, calls
\code{OutputSchema.model\_validate(json.loads(msg.content))}, and on failure
returns a replacement \code{ToolMessage} bearing the \emph{original}
\code{tool\_call\_id} and a \code{VALIDATION\_ERROR:~...} prefix in
\code{content}. The \code{add\_messages} reducer's replace-by-id behaviour,
established in Section~\ref{sec:react-toolmessage}, ensures that this
replacement message overwrites the invalid original; the model's next step
sees the error description, not the malformed data. This chapter imports
\code{ValidationNode} without any modification --- its implementation
belongs to Chapter~4 and is not reproduced here.

Inserting \code{ValidationNode} requires a custom \code{StateGraph} rather
than the prebuilt, because the prebuilt's fixed topology has no slot between
\code{tools} and \code{agent}. The custom graph adds one state field beyond
\code{messages}: \code{tool\_errors}, an integer that accumulates across
the run using \code{operator.add} as its reducer. Without a ceiling on this
count, a tool that consistently returns malformed output drives the graph
into an infinite validation-retry loop; the ceiling converts that loop
into a deterministic escalation to \code{END}.

\begin{lstlisting}[language=Python,
  caption={\texttt{AgentState} and routing logic for the validation gate},
  label={lst:react-validation-state}]
# ch15/04_validation_node_react.py
import operator, json
from typing import Annotated, Literal
from typing_extensions import TypedDict
from langchain_core.messages import BaseMessage, AIMessage, ToolMessage
from langchain_core.tools import tool
from langchain_litellm import ChatLiteLLM
from langgraph.graph import StateGraph, START, END
from langgraph.graph.message import add_messages
from langgraph.prebuilt import ToolNode
from pydantic import BaseModel, ValidationError
from ch04.validation_node import ValidationNode   # Ch.4 §4.3 — do not redefine

MAX_TOOL_ERRORS: int = 3

class WeatherOutput(BaseModel):
    temperature_c: float
    condition:     str

class AgentState(TypedDict):
    messages:    Annotated[list[BaseMessage], add_messages]
    tool_errors: Annotated[int, operator.add]

def route_after_validation(
    state: AgentState,
) -> Literal["agent", "tools"] | type[END]:
    if state["tool_errors"] >= MAX_TOOL_ERRORS:
        return END
    last = state["messages"][-1]
    if (isinstance(last, ToolMessage)
            and last.content.startswith("VALIDATION_ERROR")):
        return "tools"
    return "agent"
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={Graph wiring: \texttt{agent} → \texttt{tools} → \texttt{validate} → routing},
  label={lst:react-validation-graph}]
# ch15/04_validation_node_react.py  (continued)
@tool
def get_weather(city: str) -> str:
    """Return weather JSON for a city.

    Returns: JSON with temperature_c (float) and condition (str).
    """
    # Deliberately broken payload: missing required 'condition' field
    return json.dumps({"temperature_c": 22.0})

model         = ChatLiteLLM(model="gpt-4o-mini")
bound         = model.bind_tools([get_weather])
validate_node = ValidationNode(schema=WeatherOutput)

def agent_node(state: AgentState) -> dict:
    return {"messages": [bound.invoke(state["messages"])]}

def should_call_tools(state: AgentState) -> Literal["tools", "__end__"]:
    last = state["messages"][-1]
    if isinstance(last, AIMessage) and last.tool_calls:
        return "tools"
    return "__end__"

builder = StateGraph(AgentState)
builder.add_node("agent",    agent_node)
builder.add_node("tools",    ToolNode([get_weather]))
builder.add_node("validate", validate_node)

builder.add_edge(START, "agent")
builder.add_conditional_edges("agent", should_call_tools)
builder.add_edge("tools", "validate")
builder.add_conditional_edges(
    "validate",
    route_after_validation,
    path_map={"agent": "agent", "tools": "tools", END: END},
)
graph = builder.compile()
\end{lstlisting}

Three decisions across Listings~\ref{lst:react-validation-state}
and~\ref{lst:react-validation-graph} deserve explanation.
\code{tool\_errors} uses \code{operator.add} as its reducer, which means
any node returning \code{\{"tool\_errors": 1\}} increments the counter
and any node that does not return the key leaves it unchanged.
\code{ValidationNode} increments by returning \code{\{"tool\_errors": 1\}}
only on failure, making the counter an accurate running total of
structurally invalid results across the entire run without any explicit
counter-management logic in the routing function. The \code{path\_map}
argument to \code{add\_conditional\_edges} is required here because
\code{END} (the imported sentinel object) is one of the possible return
values of \code{route\_after\_validation}; without an explicit
\code{path\_map}, LangGraph cannot resolve the sentinel to a valid edge
target and raises a graph compilation error. The \code{agent\_node}
function is a named function rather than a lambda so that its name
appears correctly in the \code{get\_graph()} ASCII output and in Langfuse
node-level traces.

\begin{notebox}
\code{ValidationNode} (Chapter~4 \S4.3) operates on the last
\code{ToolMessage} in \code{state["messages"]}. When \code{ToolNode}
dispatches multiple tool calls in one step it appends multiple
\code{ToolMessage} objects before returning. The single-message pattern
shown here is correct for the common case where the model issues one tool
call per reasoning step. If your agent issues batched tool calls and each
result must be independently validated, extend \code{ValidationNode} to
iterate all trailing \code{ToolMessage} objects since the last
\code{AIMessage}, as described in Chapter~4 \S4.3.
\end{notebox}

The validation gate ensures the model never reads a structurally invalid
tool result. What remains invisible after the gate is in place is the cost
and latency of every tool call --- information that is essential for
diagnosing production incidents and for attribution. Section~\ref{sec:react-langfuse}
makes that information explicit.


%% ------------------------------------------------------------
\section{Every Tool Call Needs Its Own Trace Span}
\label{sec:react-langfuse}
%% ------------------------------------------------------------

The design problem is observability granularity: when a run spanning five
tool calls takes fourteen seconds and costs \$0.06, the production question
is which specific tool consumed nine of those seconds and which call
sequence drove the cost. Without per-tool spans, the Langfuse trace shows
one block of latency and one block of token usage with no attribution to
individual tools or reasoning steps. The naive approach wraps the entire
\code{agent.invoke()} call in a single \code{@observe} span, which produces
exactly this problem: one blob of data, no decomposition. The governing
principle is: observability granularity must match operational
accountability granularity, and in a tool-calling agent that granularity
is per-tool-call.

Langfuse's \code{@observe} decorator, introduced in Chapter~23 \S23.1,
instruments any Python callable as a trace span. The \code{as\_type}
parameter controls how Langfuse classifies the span: \code{"generation"}
activates token-cost attribution and is reserved for LLM calls;
\code{"span"} is the correct type for all other units of work, including
tool calls. Using \code{"generation"} on a tool function is not merely
a labelling error --- Langfuse will attempt to extract token usage from
the tool's return value, find nothing, and record zero cost for what may
be the most expensive operation in the run.

When \code{@observe}-decorated functions are called inside an already-active
\code{@observe} context, Langfuse infers the parent-child relationship from
a Python context variable the decorator maintains. No explicit trace ID or
parent span ID needs to be threaded through function arguments. In a
LangGraph agent, wrapping the LLM call inside the agent node with
\code{@observe(as\_type="generation")} and wrapping each tool with
\code{@observe(as\_type="span")} produces a trace hierarchy where every tool
span is a child of the reasoning step that triggered it --- without any
modification to the graph wiring.

\begin{lstlisting}[language=Python,
  caption={Tool functions instrumented with \texttt{@observe(as\_type="span")}},
  label={lst:react-langfuse-tools}]
# ch15/05_langfuse_tool_traces.py
from langfuse.decorators import observe, langfuse_context
from langfuse import flush_langfuse
from langchain_core.tools import tool
from langchain_core.messages import HumanMessage, AIMessage, BaseMessage
from langchain_litellm import ChatLiteLLM
from langgraph.prebuilt import create_react_agent
from pydantic_settings import BaseSettings

class Settings(BaseSettings):
    langfuse_secret_key: str
    langfuse_public_key: str
    langfuse_host:       str = "https://cloud.langfuse.com"
    model_config = {"env_file": ".env"}

settings = Settings()  # validates env vars at import time

@tool
@observe(name="web_search", as_type="span")
def web_search(query: str) -> str:
    """Search the web and return a summary."""
    result = f"Top result for '{query}': (stub)"
    langfuse_context.update_current_observation(
        input={"query": query},
        output={"result": result},
        metadata={"tool": "web_search"},
    )
    return result

@tool
@observe(name="calculator", as_type="span")
def calculator(expression: str) -> str:
    """Evaluate a safe arithmetic expression and return the result."""
    # ast.literal_eval is safe for arithmetic expressions
    import ast
    result = str(ast.literal_eval(expression))
    langfuse_context.update_current_observation(
        input={"expression": expression},
        output={"result": result},
    )
    return result
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={Agent runner instrumented as a generation span; trace URL captured inside context},
  label={lst:react-langfuse-agent}]
# ch15/05_langfuse_tool_traces.py  (continued)
model = ChatLiteLLM(model="gpt-4o-mini")

@observe(name="run_agent", as_type="generation")
def run_agent(question: str) -> tuple[str, str]:
    """Run the agent and return (answer, trace_url)."""
    agent  = create_react_agent(model=model,
                                tools=[web_search, calculator])
    result = agent.invoke(
        {"messages": [HumanMessage(content=question)]}
    )
    answer    = result["messages"][-1].content
    trace_url = langfuse_context.get_current_trace_url()
    langfuse_context.update_current_observation(
        input={"question": question},
        output={"answer": answer},
    )
    return answer, trace_url

if __name__ == "__main__":
    try:
        answer, url = run_agent(
            "Search for the Eiffel Tower height and compute 324 * 3.281."
        )
        print(f"Answer: {answer}")
        print(f"Trace:  {url}")
    finally:
        flush_langfuse()
\end{lstlisting}

Three decisions across Listings~\ref{lst:react-langfuse-tools}
and~\ref{lst:react-langfuse-agent} deserve explanation.
\code{@tool} is the outer decorator and \code{@observe} is the inner
decorator on each tool function; the order is significant because
\code{@tool} must wrap the Langfuse-instrumented function, not the
\code{StructuredTool} that \code{@tool} produces. Reversing the order
would instrument LangChain's dispatch machinery rather than the tool's
own logic, producing a span that measures serialisation overhead instead of
business logic latency. \code{langfuse\_context.get\_current\_trace\_url()}
is called \emph{inside} the \code{@observe}-decorated \code{run\_agent}
function, before it returns, because this method is only valid within an
active observation context; calling it after the function exits raises
\code{RuntimeError} because the context variable has been cleared. The URL
is captured as a return value so the caller can log or surface it after
\code{run\_agent} completes. \code{flush\_langfuse()} is called in a
\code{finally} block rather than after \code{run\_agent}: Langfuse's SDK
sends spans over a background thread, and if the Python process exits
before that thread drains its queue, all spans produced in that process are
silently lost. The \code{finally} guarantee preserves the trace even if
\code{run\_agent} raises.

\begin{tipbox}
In long-running services --- FastAPI workers, Celery tasks, persistent
LangGraph Platform deployments --- the SDK background thread has time to
drain between requests and \code{flush\_langfuse()} is unnecessary. Call
\code{flush\_langfuse()} only in short-lived scripts, in \code{pytest}
fixtures that each produce one trace, and in CLI invocations where the
process exits immediately after the agent completes. Chapter~23 \S23.1
covers the full Langfuse integration pattern for production service
deployments, including the \code{LangfuseCallbackHandler} that instruments
every LangGraph node automatically.
\end{tipbox}

The trace produced by these two listings mirrors, in Langfuse's timeline,
the message chain in \code{MessagesState}: one \code{generation} span per
\code{AIMessage} reasoning step, one \code{span} per \code{ToolMessage}
tool call, all nested under the \code{run\_agent} root span. When a
production incident occurs, the trace identifies which tool span had
anomalous latency or which step produced a \code{VALIDATION\_ERROR}
message, connecting the observability record directly to the graph
mechanism that produced it. Chapter~16 extends the tool surface outward:
where this chapter assumes tools are Python callables in the same process,
the next chapter considers tools that are MCP servers in separate
processes, reachable over HTTP, and discoverable at runtime --- a topology
that introduces transport failure, schema negotiation, and OAuth token
scoping as new failure surfaces.
