%% ============================================================
%%  CHAPTER 28 — Testing & CI/CD
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%%  Author: Haythem Aber
%% ============================================================

\chapterquote{Every production incident is a test you never wrote.}{anon}


An agent that fails its first production deploy does not fail quietly.
It fails with a silent wrong answer---a hallucinated fact delivered
confidently, a routing edge that was never exercised in development, a
state field that serialises fine in memory but raises \code{TypeError}
the moment it touches a real PostgreSQL checkpointer.
By the time monitoring surfaces the problem, hundreds of users may have
received corrupted responses, and the fix requires a full rollback plus
a post-mortem.
The production cost of an untested agent is not measured in bugs; it is
measured in the cost of the rollback, the time to isolate which node
broke, and the trust deficit that follows.

The solution this chapter builds has four layers.
The first layer---\emph{unit tests}---validates every node in complete
isolation, substituting all network dependencies with typed async mocks;
a node that passes unit tests is provably correct given correct inputs.
The second layer---\emph{integration tests}---compiles the full graph
against a real SQLite checkpointer and exercises routing, interrupt, and
resume, catching serialisation failures that mocks cannot surface.
The third layer---\emph{end-to-end tests}---runs the whole agent against
real Qdrant, Redis, and PostgreSQL instances provisioned by Docker
Compose, with a sandboxed Ollama LLM replacing vendor API calls.
The fourth layer---\emph{the regression gate}---attaches quantitative
thresholds to each node: error rate, latency p99, and RAGAS faithfulness
score; a deploy is blocked unless every node passes every threshold.

The sections appear in dependency order.
Section~\ref{sec:unit-isolation} establishes the shared mock fixtures
that every subsequent tier reuses.
Section~\ref{sec:integration-topology} builds on those fixtures by
adding a real checkpointer and asserting state persistence.
Section~\ref{sec:e2e-sandboxed} replaces the in-process checkpointer
with Docker-provisioned services, making the environment production-like
without touching real vendor APIs.
Section~\ref{sec:replay-fixtures} introduces replay fixtures that let
CI reproduce a real run deterministically without Docker.
Section~\ref{sec:regression-matrix} defines the quantitative gate that
converts passing tests into passing deploys.
Section~\ref{sec:github-actions} wires all five tiers into a single
GitHub Actions pipeline where each job is a prerequisite of the next.

\begin{table}[htbp]
\centering
\caption{Domain terms introduced in this chapter.}
\label{tab:ch28-terms}
\begin{tabularx}{\linewidth}{@{} l P{5cm} P{5.5cm} @{}}
\toprule
\textbf{Term} & \textbf{Definition} & \textbf{Where used} \\
\midrule
Unit test       & Test of one node in isolation; all external dependencies replaced with typed async mocks & \S28.1 \\
Integration test & Test of the compiled graph with a real checkpointer; no external network calls & \S28.2 \\
End-to-end test  & Test against real services (Qdrant, Redis, PostgreSQL) running in Docker Compose & \S28.3 \\
Replay fixture   & JSON snapshot of a real run's state sequence, replayed deterministically in CI & \S28.4 \\
Regression gate  & Per-node threshold check on error rate, p99 latency, and RAGAS score; blocks deploy on breach & \S28.5 \\
RAGAS score      & Faithfulness and answer relevancy metrics computed by the RAGAS library against ground-truth references & \S28.5 \\
\bottomrule
\end{tabularx}
\end{table}

\begin{sourcelayout}
\dirtree{%
  .1 tests/.\
  .2 conftest.py \DTcomment{MockLiteLLM, MockMemory, make\_config, make\_context}.\
  .2 unit/.\
  .3 test\_llm\_primary\_node.py \DTcomment{\S28.1}.\
  .2 integration/.\
  .3 test\_graph\_routing.py \DTcomment{\S28.2}.\
  .2 e2e/.\
  .3 conftest.py \DTcomment{Docker service fixtures}.\
  .3 test\_full\_agent\_run.py \DTcomment{\S28.3}.\
  .2 fixtures/.\
  .3 helpers.py \DTcomment{load\_fixture, save\_fixture}.\
  .3 schema.py \DTcomment{FixtureEntry Pydantic v2 model}.\
  .3 happy\_path\_baseline.json \DTcomment{recorded state sequence}.\
  .2 regression/.\
  .3 regression\_matrix.py \DTcomment{NodeMetrics, evaluate\_metrics, write\_csv}.\
  .3 ragas\_eval.py \DTcomment{RAGAS EvaluationDataset builder}.\
  .3 test\_replay.py \DTcomment{\S28.4}.\
  .3 test\_regression\_gate.py \DTcomment{\S28.5}.\
  .3 regression\_thresholds.yaml \DTcomment{per-node thresholds}.\
  .1 docker-compose.test.yml \DTcomment{Qdrant, Redis, PostgreSQL, Ollama}.\
  .1 .github/.\
  .2 workflows/.\
  .3 ci.yml \DTcomment{\S28.6 --- lint, unit, integration, eval, deploy}.\
}
\end{sourcelayout}

%% ─────────────────────────────────────────────────────────────
\section{Nodes Must Be Testable in Isolation}
\label{sec:unit-isolation}
%% ─────────────────────────────────────────────────────────────

The design problem is that a LangGraph node reaches out to three or more
external systems---an LLM router, a memory client, a vector store---and
any one of them can fail for reasons unrelated to the logic inside the
node itself.
The naive approach is to call the node from a compiled graph and mock at
the HTTP boundary, but that couples the test to graph topology; when a
routing edge changes, unrelated node tests break.
The principle is simpler: a node is a pure Python async function with a
typed signature, and it can be called directly with typed arguments,
bypassing the graph entirely.

\begin{warningbox}
Using \code{unittest.mock.MagicMock} as a stand-in for an async
dependency raises \code{TypeError: object MagicMock can't be used in
'await' expression} at runtime.
Every mock that replaces an \code{async def} method must itself be an
\code{async def} method.
A bare \code{AsyncMock(spec=Router)} from the standard library is
valid, but a hand-written class that mirrors the exact method signatures
is easier to type-check and easier to read in failure output.
\end{warningbox}

The shared fixtures live in \code{tests/conftest.py}, which pytest
discovers automatically for every test in the \code{tests/} tree.
Two fixture factories---\code{mock\_llm} and \code{mock\_memory}---and two
builder helpers---\code{make\_config} and \code{make\_context}---form the
complete surface that all three test tiers import.

\begin{lstlisting}[language=Python, caption={Shared async mock fixtures
  and config helpers (\texttt{tests/conftest.py}).},
  label={lst:conftest}]
import uuid
import pytest
from typing import Any

class MockLiteLLM:
    def __init__(self, response_content: str = "mock response") -> None:
        self.calls: list[dict[str, Any]] = []
        self._content = response_content

    async def acompletion(self, **kwargs: Any) -> dict[str, Any]:
        self.calls.append(kwargs)
        return {
            "choices": [
                {"message": {"role": "assistant",
                             "content": self._content}}
            ],
            "usage": {"prompt_tokens": 10, "completion_tokens": 5},
        }

class MockMemory:
    async def add(self, messages: list[dict],
                  user_id: str, **kwargs: Any) -> dict[str, str]:
        return {"id": "mock-mem-id"}
    async def search(self, query: str,
                     user_id: str, **kwargs: Any) -> list[dict]:
        return []
    async def get_all(self, user_id: str,
                      **kwargs: Any) -> list[dict]:
        return []

@pytest.fixture
def mock_llm() -> MockLiteLLM:
    return MockLiteLLM()

@pytest.fixture
def mock_memory() -> MockMemory:
    return MockMemory()

def make_config(thread_id: str | None = None) -> dict[str, Any]:
    return {"configurable": {"thread_id": thread_id or str(uuid.uuid4())}}

def make_context(llm: MockLiteLLM | None = None,
                 memory: MockMemory | None = None) -> dict[str, Any]:
    ctx: dict[str, Any] = {}
    if llm:    ctx["llm_router"]    = llm
    if memory: ctx["memory_client"] = memory
    return ctx
\end{lstlisting}

Three decisions in Listing~\ref{lst:conftest} deserve explanation.
First, \code{make\_config} and \code{make\_context} are separated into
two helpers rather than one because they represent two different
concerns: \code{configurable} carries the \code{thread\_id} which is
required by the checkpointer, while \code{context} carries
runtime-injected dependencies that LangGraph~v0.6 routes through the
\code{context\_schema} mechanism---merging them would couple checkpointer
state to LLM provisioning, forcing every test to supply a
\code{thread\_id} even when no checkpoint is written.
Second, the fixtures are \code{scope="function"} (the pytest default),
not \code{scope="session"}; because \code{MockLiteLLM.calls} accumulates
per-test call counts, a session-scoped fixture would carry counts from
previous tests into assertions about the current test, making count-based
assertions non-deterministic.
Third, \code{MockMemory} implements all three methods (\code{add},
\code{search}, \code{get\_all}); a partial mock that omits any one of
them causes an \code{AttributeError} that looks like a node failure but
is actually a test configuration error.

\begin{notebox}
LangGraph~v0.6 introduced two related but distinct mechanisms for
dependency injection.
At the call site, \code{graph.ainvoke(input, config=cfg, context=\{...\})}
passes a context dict that the runtime maps to the graph's declared
\code{context\_schema}.
Inside the node itself, the context is accessed via a typed
\code{Runtime} parameter: \code{from langgraph.types import Runtime} and
then \code{async def my\_node(state, runtime: Runtime[MyContext])},
where \code{runtime.context} holds the injected values.
Unit tests that call the node function directly must construct and pass
a \code{Runtime} instance, or the node must be written to accept an
optional \code{context=} kwarg for testability---both are valid
patterns; the chapter examples use the direct kwarg form for clarity.
\end{notebox}

\begin{tipbox}
Add the following snippet to \texttt{pyproject.toml} to eliminate the
\code{@pytest.mark.asyncio} decorator from every test function.
Without it, async tests silently pass without executing when
\code{pytest-asyncio} is installed but \code{asyncio\_mode} is not set.
\begin{lstlisting}[language=Python]
# pyproject.toml
[tool.pytest.ini_options]
asyncio_mode = "auto"
markers = [
    "integration: requires SqliteSaver",
    "e2e: requires Docker daemon",
    "regression: RAGAS gate",
]
\end{lstlisting}
\end{tipbox}

With fixtures defined, a unit test calls the node function directly,
passes a minimal \code{AgentState}, and asserts on the returned state
fragment---no graph compilation, no I/O.

\begin{lstlisting}[language=Python, caption={Unit tests for the primary
  LLM node (\texttt{tests/unit/test\_llm\_primary\_node.py}).},
  label={lst:unit-node}]
import pytest
from unittest.mock import AsyncMock
from langchain_core.messages import HumanMessage
from litellm.exceptions import RateLimitError
from src.graph.nodes import llm_primary_node
from src.graph.state import AgentState
from tests.conftest import MockLiteLLM, make_config, make_context

async def test_happy_path(mock_llm: MockLiteLLM) -> None:
    state: AgentState = {"messages": [HumanMessage(content="hello")]}
    result = await llm_primary_node(
        state,
        config=make_config(),
        context=make_context(llm=mock_llm),
    )
    assert result["messages"]
    assert mock_llm.calls

async def test_sets_soft_error_on_rate_limit(
    mock_llm: MockLiteLLM,
) -> None:
    mock_llm.acompletion = AsyncMock(
        side_effect=RateLimitError(
            "rate limit", model="gpt-4", llm_provider="openai"
        )
    )
    state: AgentState = {"messages": [HumanMessage(content="hello")]}
    result = await llm_primary_node(
        state,
        config=make_config(),
        context=make_context(llm=mock_llm),
    )
    assert result.get("error_class") == "soft"
\end{lstlisting}

Two decisions in Listing~\ref{lst:unit-node} deserve explanation.
The \code{RateLimitError} constructor requires \code{model} and
\code{llm\_provider} as keyword arguments from LiteLLM~v1.x; omitting
them raises a \code{TypeError} inside the mock setup, not inside the
node under test---a confusing failure that implies the node is broken
when the fixture is broken.
The test for \code{error\_class} asserts on the returned state fragment,
not on whether the mock was called with specific arguments; this
separates the node's observable output contract from its internal
implementation, so the node can be refactored without breaking the test.

Unit tests confirm that each node behaves correctly given correct inputs
and correctly classified errors, but they cannot confirm that the graph
routes between nodes correctly or that state survives a checkpoint cycle;
that requires a compiled graph with a real persisting checkpointer.

%% ─────────────────────────────────────────────────────────────
\section{Graph Topology Must Be Validated Against a Real Checkpointer}
\label{sec:integration-topology}
%% ─────────────────────────────────────────────────────────────

The design problem is that a compiled graph can have routing edges that
are logically correct in code but that fail at runtime because the state
they write cannot be serialised to the checkpointer's storage format.
The naive approach is to use \code{InMemorySaver} for integration tests;
it is zero-configuration and fast, but it holds Python objects in memory
directly and never calls the serialisation codec, so a \code{datetime}
field without timezone info, a custom class, or a \code{HumanMessage}
stored as a raw object will all pass \code{InMemorySaver} and fail the
moment the real checkpointer is used.
The principle is that integration tests must exercise the full
serialisation round-trip; the cheapest checkpointer that does so is
\code{SqliteSaver}.

\begin{warningbox}
\code{SqliteSaver} is a test-only checkpointer.
It has no connection pooling, no concurrent-write safety, and no
migration tooling.
Any fixture that instantiates \code{SqliteSaver} must carry a comment
reading \texttt{\# TEST ONLY --- use AsyncPostgresSaver in production}
so that the pattern is never promoted to a deployment configuration.
\end{warningbox}

\code{SqliteSaver} is constructed with a DSN string; the
\code{":memory:"} DSN creates an in-process SQLite database that is
owned by the \code{SqliteSaver} instance for its lifetime.
This makes it safe to scope the fixture to the test module: the
in-memory database is isolated from other modules and disappears when
the fixture is torn down.

\begin{lstlisting}[language=Python, caption={Integration tests for
  routing and checkpoint resumption
  (\texttt{tests/integration/test\_graph\_routing.py}).},
  label={lst:integration}]
import pytest
from unittest.mock import AsyncMock
from langgraph.checkpoint.sqlite import SqliteSaver
from litellm.exceptions import RateLimitError
from src.graph.agent import build_resilient_graph
from tests.conftest import MockLiteLLM, make_config, make_context

# TEST ONLY -- SqliteSaver has no connection pooling; use AsyncPostgresSaver in prod
@pytest.fixture(scope="module")
def saver() -> SqliteSaver:
    return SqliteSaver.from_conn_string(":memory:")

@pytest.fixture(scope="module")
def graph(saver: SqliteSaver):
    return build_resilient_graph(checkpointer=saver)

@pytest.mark.integration
async def test_checkpoint_roundtrip(graph) -> None:
    cfg = make_config()
    await graph.ainvoke(
        {"messages": [{"role": "user", "content": "hello"}]},
        config=cfg,
        context=make_context(llm=MockLiteLLM()),
    )
    history = list(graph.get_state_history(cfg))
    assert len(history) >= 2

@pytest.mark.integration
async def test_interrupt_and_resume(graph) -> None:
    mock = MockLiteLLM()
    mock.acompletion = AsyncMock(
        side_effect=RateLimitError(
            "rl", model="gpt-4", llm_provider="openai"
        )
    )
    cfg = make_config()
    await graph.ainvoke(
        {"messages": [{"role": "user", "content": "hello"}]},
        config=cfg,
        context=make_context(llm=mock),
        interrupt_before=["llm_cheap"],
    )
    resumed = await graph.ainvoke(
        None, config=cfg,
        context=make_context(llm=MockLiteLLM()),
    )
    assert resumed is not None
\end{lstlisting}

Three decisions in Listing~\ref{lst:integration} deserve explanation.
\code{get\_state\_history} is synchronous and returns an iterator
ordered newest-first; calling \code{list()} on it in an async test
is valid because the method does not block the event loop---it reads
from the in-memory SQLite connection synchronously.
The assertion \code{len(history) >= 2} is deliberately loose: it
confirms that the checkpointer wrote at least a START snapshot and one
node snapshot, without tying the test to the exact number of nodes the
graph may add or remove in future.
The resume call passes \code{None} as input; this is the correct
LangGraph pattern for resuming from an interrupt---the graph reads the
persisted state from the checkpointer and continues from the last
saved position, so providing new input would overwrite the saved state.

Integration tests confirm routing and persistence but cannot confirm
that the agent behaves correctly when talking to a real Qdrant
collection, a real Redis cache, or a real PostgreSQL checkpointer with
its schema migrations applied; that boundary is the responsibility of
end-to-end tests.

%% ─────────────────────────────────────────────────────────────
\section{External Services Must Be Sandboxed, Not Mocked}
\label{sec:e2e-sandboxed}
%% ─────────────────────────────────────────────────────────────

The design problem is that the interaction between the agent and its
external services involves timing, connection pooling, schema
versioning, and network serialisation that no mock can faithfully
reproduce.
The naive approach is to point E2E tests at shared staging
environments, but that introduces flakiness from concurrent test
runs, requires credential rotation on every CI runner, and makes the
test environment non-reproducible across developer machines and CI
agents.
The principle is that every service the agent depends on must be
provisioned locally and reproducibly for the duration of the test
session, and torn down completely afterwards.

Docker Compose is the correct tool for this: a single
\code{docker-compose.test.yml} file declares four services---Qdrant,
Redis, PostgreSQL, and an Ollama sidecar---and the CI job starts them
before tests run and stops them after.
The \code{pytest-docker} plugin provides a session-scoped
\code{docker\_services} fixture that wraps this lifecycle and exposes
a \code{wait\_until\_responsive} helper that polls a readiness check
function until the service accepts connections.

\begin{lstlisting}[language=Python, caption={E2E session fixtures with
  Docker service readiness checks
  (\texttt{tests/e2e/conftest.py}).},
  label={lst:e2e-conftest}]
import pytest
import psycopg2
import requests
from langgraph.checkpoint.postgres.aio import AsyncPostgresSaver

DSN = "postgresql://agent:agent@localhost:5432/agentdb"

@pytest.fixture(scope="session")
def docker_compose_file(pytestconfig):
    return str(pytestconfig.rootdir / "docker-compose.test.yml")

def _pg_ready() -> bool:
    try:
        psycopg2.connect(DSN).close()
        return True
    except Exception:
        return False

def _qdrant_ready() -> bool:
    try:
        return (requests.get("http://localhost:6333/health",
                             timeout=2)
                .json().get("status") == "ok")
    except Exception:
        return False

@pytest.fixture(scope="session")
async def pg_saver(docker_services):
    docker_services.wait_until_responsive(
        check=_pg_ready, timeout=60, pause=1
    )
    docker_services.wait_until_responsive(
        check=_qdrant_ready, timeout=60, pause=1
    )
    saver = await AsyncPostgresSaver.from_conn_string(DSN)
    await saver.setup()   # idempotent -- creates tables if absent
    return saver
\end{lstlisting}

Two decisions in Listing~\ref{lst:e2e-conftest} deserve explanation.
The readiness functions catch all exceptions and return \code{False}
rather than re-raising them; during the startup window, connection
refused, DNS resolution failure, and timeout are all expected and
transient---re-raising would abort the fixture instead of retrying.
\code{AsyncPostgresSaver.setup()} is idempotent by design; it uses
\code{CREATE TABLE IF NOT EXISTS} internally, so calling it on a
database that was previously initialised is safe and produces no
error---this matters because the same Docker Compose volume may persist
across CI runs if the runner caches Docker volumes.

\begin{lstlisting}[language=Python, caption={Full-agent E2E test
  (\texttt{tests/e2e/test\_full\_agent\_run.py}).},
  label={lst:e2e-test}]
import pytest
from src.graph.agent import build_resilient_graph
from src.graph.llm import build_ollama_router
from tests.conftest import make_config

@pytest.mark.e2e
async def test_full_run_no_dead_letter(pg_saver) -> None:
    graph  = build_resilient_graph(checkpointer=pg_saver)
    cfg    = make_config()
    result = await graph.ainvoke(
        {"messages": [{"role": "user", "content": "What is 2+2?"}]},
        config=cfg,
        context={"llm_router": build_ollama_router()},
    )
    assert result["messages"][-1].content
    assert not result.get("dead_letter")
\end{lstlisting}

Two decisions in Listing~\ref{lst:e2e-test} deserve explanation.
The \code{import uuid} line was deliberately omitted: \code{make\_config()}
generates the \code{thread\_id} UUID internally, so importing \code{uuid}
directly in the test file would be an unused import that mypy and ruff
both flag as an error.
The assertion \code{assert not result.get("dead\_letter")} uses a falsy
check rather than \code{is None} because \code{dead\_letter} may be
typed as a list that defaults to \code{[]} in \code{AgentState}; an
empty list is falsy, so the assertion correctly passes when no
dead-letter entries were recorded regardless of whether the field is
absent or present and empty.

The Ollama router constructed by \code{build\_ollama\_router()} points
LiteLLM at \code{http://localhost:11434} with model string
\code{"ollama/llama3.2:1b"}; because Ollama is a Docker Compose
sidecar rather than a vendor API, there are no API keys to rotate, no
rate limits to hit, and no cost per token---making the E2E suite safe to
run on every pull request.

E2E tests confirm that the agent completes a real request without
routing to the dead-letter node, but they do not provide a
\emph{reproducible} regression baseline: the Ollama model's output
varies across model versions and the services may return different data
across runs.
Reproducibility requires a separate mechanism that records and replays
a specific state sequence.

%% ─────────────────────────────────────────────────────────────
\section{Regressions Are Only Detectable Against a Recorded Baseline}
\label{sec:replay-fixtures}
%% ─────────────────────────────────────────────────────────────

The design problem is that mocks assert on how the code calls its
dependencies---the arguments, the call count---but they cannot assert
on what the code produces, because the mock defines the production.
A regression in prompt logic, a change to a routing condition, or a
new state field that overwrites an older one will all pass mock-based
tests while silently changing agent behaviour in production.
The principle is that the output of a verified real run must be
committed to the repository as a fixture, and CI must fail if a
subsequent run diverges from that output.

A \emph{replay fixture} is a JSON array where each element represents
one \code{StateSnapshot} from \code{get\_state\_history}: a \code{values}
dict (the full state at that checkpoint), a \code{next\_node} list, and
a \code{step} integer.
The fixtures are stored in \code{tests/fixtures/} as
\code{\{name\}.json} files and committed to version control alongside
the code they validate.
Two helpers---\code{save\_fixture} and \code{load\_fixture}---are the
entire interface between real runs and replay tests.

\begin{lstlisting}[language=Python, caption={Fixture save and load
  helpers (\texttt{tests/fixtures/helpers.py}).},
  label={lst:fixture-helpers}]
from __future__ import annotations
import json
from pathlib import Path
from typing import Any

FIXTURE_DIR = Path(__file__).parent

def load_fixture(name: str) -> list[dict[str, Any]]:
    path = FIXTURE_DIR / f"{name}.json"
    if not path.exists():
        raise FileNotFoundError(f"Fixture not found: {path}")
    data = json.loads(path.read_text())
    if not isinstance(data, list):
        raise ValueError(
            f"Fixture '{name}' must be a JSON array, got {type(data)}"
        )
    return data

async def save_fixture(name: str, graph, config: dict) -> None:
    entries: list[dict[str, Any]] = []
    i = 0
    async for snap in graph.aget_state_history(config):
        entries.append({
            "values":    snap.values,
            "next_node": list(snap.next),
            "step":      i,
        })
        i += 1
    entries.reverse()   # aget_state_history yields newest-first
    path = FIXTURE_DIR / f"{name}.json"
    path.write_text(json.dumps(entries, indent=2, default=str))
\end{lstlisting}

Two decisions in Listing~\ref{lst:fixture-helpers} deserve explanation.
\code{save\_fixture} uses \code{aget\_state\_history} (the async
iterator variant) rather than the synchronous \code{get\_state\_history},
because \code{save\_fixture} is typically called from an async test or
an async CLI tool; using the sync variant inside a running event loop
would require wrapping it in \code{asyncio.to\_thread}, adding
unnecessary complexity.
The \code{json.dumps(default=str)} fallback serialises
\code{datetime}, \code{UUID}, and \code{BaseMessage} objects as their
string representations; on reload, these fields are plain strings, not
typed objects, so the replay test must reconstruct messages explicitly
from the fixture data rather than assuming the values dict contains
\code{HumanMessage} instances.

The replay test references \code{FixtureEntry.validate\_list}, which
validates each raw fixture dict against a Pydantic~v2 schema before the
test runs; a schema mismatch raises a \code{ValidationError} that
identifies exactly which field drifted, turning what would otherwise be
an obscure \code{KeyError} in the test into an actionable error message.

\begin{lstlisting}[language=Python, caption={Fixture entry schema with
  Pydantic~v2 \texttt{model\_validate}
  (\texttt{tests/fixtures/schema.py}).},
  label={lst:fixture-schema}]
from __future__ import annotations
from typing import Any
from pydantic import BaseModel

class FixtureEntry(BaseModel):
    values:    dict[str, Any]
    next_node: list[str]
    step:      int

    @classmethod
    def validate_list(
        cls, data: list[dict[str, Any]]
    ) -> list["FixtureEntry"]:
        # model_validate is the Pydantic v2 replacement for parse_obj
        return [cls.model_validate(entry) for entry in data]
\end{lstlisting}

One decision in Listing~\ref{lst:fixture-schema} deserves explanation.
\code{model\_validate} is the Pydantic~v2 classmethod that replaced the
removed \code{parse\_obj}; calling \code{parse\_obj} on a Pydantic~v2
model raises \code{AttributeError} at import time, not at validation
time, which can produce a confusing ``fixture module failed to import''
error in pytest output that obscures the real cause.

\begin{lstlisting}[language=Python, caption={Replay test that verifies
  a baseline state sequence
  (\texttt{tests/regression/test\_replay.py}).},
  label={lst:replay-test}]
import pytest
from langgraph.checkpoint.memory import InMemorySaver
from langchain_core.messages import HumanMessage
from tests.fixtures.helpers import load_fixture
from tests.fixtures.schema import FixtureEntry
from src.graph.agent import build_resilient_graph
from tests.conftest import MockLiteLLM, make_config, make_context

@pytest.mark.regression
async def test_replay_happy_path() -> None:
    raw     = load_fixture("happy_path_baseline")
    entries = FixtureEntry.validate_list(raw)

    saver   = InMemorySaver()
    graph   = build_resilient_graph(checkpointer=saver)
    cfg     = make_config()

    first    = entries[0].values["messages"][0]
    seed     = HumanMessage(content=first["content"])
    expected = entries[-1].values["messages"][-1]["content"]

    mock   = MockLiteLLM(response_content=expected)
    result = await graph.ainvoke(
        {"messages": [seed]},
        config=cfg,
        context=make_context(llm=mock),
    )
    assert result["messages"][-1].content == expected
\end{lstlisting}

The replay test uses \code{InMemorySaver} rather than
\code{SqliteSaver} because the goal here is to verify output
correctness, not serialisation fidelity; the integration tier already
handles serialisation.
The \code{expected} value is pinned from the fixture's last snapshot,
then injected into \code{MockLiteLLM} as the canned response; this
makes the test deterministic---the mock will always return the same
string that was recorded in the fixture, so any divergence between the
recorded output and the current output indicates a logic regression,
not an LLM non-determinism problem.

Replay tests confirm that the agent's observable output has not changed,
but they do not answer the question of \emph{how much} it has
degraded---a response that technically matches the baseline string may
have regressed in faithfulness if the retrieval context changed.
Quantitative metrics are needed for that boundary.

%% ─────────────────────────────────────────────────────────────
\section{Quality Degradation Must Be Caught Before It Reaches Users}
\label{sec:regression-matrix}
%% ─────────────────────────────────────────────────────────────

The design problem is that a test suite produces a binary result---pass
or fail---while LLM quality degrades continuously: faithfulness might
drop from 0.92 to 0.71 across three model updates, with no single
change large enough to break a deterministic assertion.
The naive approach is to add numerical assertions directly into test
functions, but scattered threshold literals across dozens of test files
cannot be reviewed, versioned, or adjusted without touching code.
The principle is that thresholds are configuration, not logic: they
belong in a YAML file that can be reviewed and adjusted independently
of the test code that enforces them.

The \emph{regression matrix} is a CSV artifact produced by CI: one row
per node, five metric columns (\code{node\_name}, \code{error\_rate},
\code{p99\_ms}, \code{ragas\_faithfulness}, \code{ragas\_answer\_relevancy}),
and a \code{passed} column that is \code{True} only when all applicable
thresholds are met.
The deploy job reads this CSV and aborts the rollout if any row has
\code{passed=False}.

\begin{notebox}
The RAGAS library changed its dataset schema in v0.2.
The legacy \code{datasets.Dataset} with columns \code{ground\_truth}
and \code{contexts} is no longer accepted.
The current API uses \code{EvaluationDataset} populated with
\code{SingleTurnSample} objects whose fields are \code{user\_input},
\code{response}, \code{retrieved\_contexts} (a list of strings), and
\code{reference} (the ground-truth string).
Metric classes must be instantiated---\code{Faithfulness()} and
\code{AnswerRelevancy()}---not passed as bare class objects;
passing the class raises a \code{TypeError} at \code{evaluate()} time.
\end{notebox}

\begin{lstlisting}[language=Python, caption={Per-node metrics schema
  and threshold enforcement
  (\texttt{tests/regression/regression\_matrix.py}).},
  label={lst:regression-matrix}]
from __future__ import annotations
import csv, dataclasses, yaml
from dataclasses import dataclass
from typing import Optional

@dataclass
class NodeMetrics:
    node_name:              str
    error_rate:             float
    p99_ms:                 float
    ragas_faithfulness:     Optional[float]
    ragas_answer_relevancy: Optional[float]
    passed:                 bool = True

def load_thresholds(path: str = "tests/regression_thresholds.yaml") -> dict:
    with open(path) as f:
        return yaml.safe_load(f)["thresholds"]

def evaluate_metrics(metrics: list[NodeMetrics],
                     thresholds: dict) -> list[NodeMetrics]:
    for m in metrics:
        t = thresholds.get(m.node_name, {})
        min_faith = t.get("min_ragas_faithfulness")
        m.passed = (
            m.error_rate <= t.get("max_error_rate", 1.0)
            and m.p99_ms  <= t.get("max_p99_ms", float("inf"))
            and (min_faith is None
                 or (m.ragas_faithfulness or 0.0) >= min_faith)
        )
    return metrics

def write_csv(metrics: list[NodeMetrics],
              path: str = "regression_matrix.csv") -> None:
    fieldnames = [f.name for f in dataclasses.fields(NodeMetrics)]
    with open(path, "w", newline="") as fh:
        writer = csv.DictWriter(fh, fieldnames=fieldnames)
        writer.writeheader()
        for m in metrics:
            writer.writerow(dataclasses.asdict(m))
\end{lstlisting}

Two decisions in Listing~\ref{lst:regression-matrix} deserve explanation.
\code{write\_csv} derives \code{fieldnames} from
\code{dataclasses.fields(NodeMetrics)} rather than from
\code{NodeMetrics.\_\_dataclass\_fields\_\_}: the latter returns a dict
of \code{Field} objects, not strings, so passing it directly to
\code{DictWriter} produces column headers like
\code{Field(name='node\_name',\ldots)} instead of \code{node\_name}.
\code{evaluate\_metrics} treats a \code{None} threshold
(\code{min\_ragas\_faithfulness: null} in YAML) as ``not applicable''
and skips the check for non-LLM nodes like \code{static\_fallback};
this avoids a false-positive failure when a node that never calls an
LLM is evaluated against a RAGAS threshold that has no meaning for it.

\begin{lstlisting}[language=Python, caption={RAGAS evaluation helper
  using the v0.2 \texttt{EvaluationDataset} API
  (\texttt{tests/regression/ragas\_eval.py}).},
  label={lst:ragas-eval}]
from ragas import evaluate
from ragas.dataset_schema import EvaluationDataset, SingleTurnSample
from ragas.metrics import Faithfulness, AnswerRelevancy

def run_ragas_eval(question: str, answer: str,
                   contexts: list[str], reference: str) -> dict[str, float]:
    sample = SingleTurnSample(
        user_input=question,
        response=answer,
        retrieved_contexts=contexts,
        reference=reference,
    )
    dataset = EvaluationDataset(samples=[sample])
    result  = evaluate(
        dataset=dataset,
        metrics=[Faithfulness(), AnswerRelevancy()],
    )
    df = result.to_pandas()
    return {
        "faithfulness":     float(df["faithfulness"].iloc[0]),
        "answer_relevancy": float(df["answer_relevancy"].iloc[0]),
    }
\end{lstlisting}

The helper returns a plain \code{dict[str, float]} rather than the
\code{Result} object directly; this decouples the test code from the
RAGAS internal API, which changed its schema twice between v0.1 and
v0.2.
If RAGAS changes its column names again, only this one helper needs to
change, and every test that calls \code{run\_ragas\_eval} remains
unaffected.

With quantitative thresholds defined and enforced by test code, the
remaining question is how to sequence, parallelise, and enforce these
four tiers automatically on every push and pull request---which is the
responsibility of the CI pipeline.

%% ─────────────────────────────────────────────────────────────
\section{Each Test Tier Must Be an Independent CI Gate}
\label{sec:github-actions}
%% ─────────────────────────────────────────────────────────────

The design problem is that a single CI job running all test tiers
sequentially is both slow---unit tests must wait for Docker to start---
and uninformative---when integration tests fail, it is unclear whether
they failed because of a logic regression or because a Docker service
did not start in time.
The naive approach is to put everything in one job with a long timeout;
the consequence is that a lint error costs the same CI minutes as a
full E2E run, and every developer waits for Docker even when only
fixing a comment.
The principle is that each tier is an independent job with its own
\code{needs:} dependency; the cheapest tier runs first, and a failure
at any tier aborts the chain before the more expensive tiers begin.

The five jobs are: \code{lint} (ruff + mypy, no services, approximately
30 seconds), \code{unit} (no services, approximately 60 seconds),
\code{integration} (three Docker services via the \code{services:}
block), \code{eval} (four Docker services via Docker Compose, includes
the RAGAS gate and artifact upload), and \code{deploy} (downloads the
artifact, asserts no threshold breaches, then runs \code{kubectl
rollout restart}).

\begin{warningbox}
The GitHub Actions \code{services:} block starts individual containers
and does not accept a Docker Compose file.
The \code{--health-cmd} value in \code{options:} is passed directly to
\code{docker run} as a shell command that must exist inside the
container image.
For \code{postgres:16-alpine}, use \code{pg\_isready -U agent -d agentdb}.
For \code{qdrant/qdrant:latest}, use \code{wget -qO- http://localhost:6333/health}.
Using \code{curl} in a Qdrant health check will fail silently because
the official Qdrant image does not ship \code{curl}; it ships \code{wget}.
\end{warningbox}

\begin{lstlisting}[language=yaml, caption={CI pipeline: lint, unit, and
  integration jobs (\texttt{.github/workflows/ci.yml}, Part~1).},
  label={lst:ci-part1}]
name: Agent CI
on:
  push:   { branches: [main] }
  pull_request:

jobs:
  lint:
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4
      - uses: actions/setup-python@v5
        with: { python-version: "3.12" }
      - run: pip install uv && uv sync --frozen
      - run: uv run ruff check . && uv run mypy src/

  unit:
    needs: lint
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4
      - uses: actions/setup-python@v5
        with: { python-version: "3.12" }
      - run: pip install uv && uv sync --frozen
      - run: |
          uv run pytest tests/unit/ tests/conftest.py \
            -m "not integration and not e2e and not regression" \
            -x -q

  integration:
    needs: unit
    runs-on: ubuntu-latest
    services:
      postgres:
        image: postgres:16-alpine
        env:
          POSTGRES_DB:       agentdb
          POSTGRES_USER:     agent
          POSTGRES_PASSWORD: agent
        ports: ["5432:5432"]
        options: >-
          --health-cmd "pg_isready -U agent -d agentdb"
          --health-interval 5s --health-timeout 3s --health-retries 10
      redis:
        image: redis:7-alpine
        ports: ["6379:6379"]
        options: >-
          --health-cmd "redis-cli ping"
          --health-interval 5s --health-retries 5
      qdrant:
        image: qdrant/qdrant:latest
        ports: ["6333:6333"]
        options: >-
          --health-cmd "wget -qO- http://localhost:6333/health"
          --health-interval 5s --health-timeout 3s --health-retries 10
    steps:
      - uses: actions/checkout@v4
      - uses: actions/setup-python@v5
        with: { python-version: "3.12" }
      - run: pip install uv && uv sync --frozen
      - run: uv run pytest tests/integration/ -m integration -x -q
\end{lstlisting}

\begin{lstlisting}[language=yaml, caption={CI pipeline: eval gate and
  deploy jobs (\texttt{.github/workflows/ci.yml}, Part~2).},
  label={lst:ci-part2}]
  eval:
    needs: integration
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4
      - uses: actions/setup-python@v5
        with: { python-version: "3.12" }
      - run: pip install uv && uv sync --frozen
      - name: Start test services with Ollama
        run: docker compose -f docker-compose.test.yml up -d
      - name: Wait for Ollama and pull model
        run: |
          timeout 120 bash -c \
            'until curl -sf http://localhost:11434/api/tags; do sleep 2; done'
          docker exec \
            $(docker compose -f docker-compose.test.yml ps -q ollama) \
            ollama pull llama3.2:1b
      - name: Run regression gate
        run: uv run pytest tests/regression/ -m regression -v
      - uses: actions/upload-artifact@v4
        with:
          name: regression-matrix
          path: regression_matrix.csv
      - name: Tear down services
        if: always()
        run: docker compose -f docker-compose.test.yml down -v

  deploy:
    needs: [eval]
    if: github.ref == 'refs/heads/main' && success()
    runs-on: ubuntu-latest
    steps:
      - uses: actions/download-artifact@v4
        with: { name: regression-matrix }
      - name: Assert no threshold breaches
        run: |
          python - <<'EOF'
          import csv, sys
          rows = list(csv.DictReader(open("regression_matrix.csv")))
          fails = [r for r in rows if r["passed"] != "True"]
          if fails:
              for f in fails: print(f"BREACH: {f}", file=sys.stderr)
              sys.exit(1)
          print(f"All {len(rows)} node(s) within thresholds.")
          EOF
      - name: Rollout to Kubernetes
        env:
          KUBECONFIG_DATA: ${{ secrets.KUBECONFIG_DATA }}
        run: |
          echo "$KUBECONFIG_DATA" | base64 -d > /tmp/kubeconfig
          KUBECONFIG=/tmp/kubeconfig \
            kubectl rollout restart deployment/agent-api
\end{lstlisting}

Three decisions across Listings~\ref{lst:ci-part1}
and~\ref{lst:ci-part2} deserve explanation.
\code{uv sync --frozen} uses the committed lockfile exactly as it was
written, preventing implicit dependency upgrades during CI; a
\code{uv sync} without \code{--frozen} would silently upgrade packages
whose version constraints allow a new release, potentially changing
agent behaviour between the developer's machine and the CI run.
The \code{if: always()} condition on the Docker Compose teardown step
ensures that services are stopped and volumes deleted even when the
regression test fails; without it, a failed test leaves Docker
containers running on the CI runner, consuming memory and eventually
causing port conflicts on subsequent runs.
The \code{KUBECONFIG\_DATA} secret is base64-encoded and decoded inline
rather than written to a persistent file; the decoded file is written
to \code{/tmp/kubeconfig} and referenced only via the \code{KUBECONFIG}
environment variable scoped to the single \code{kubectl} step, so the
raw cluster credentials are never accessible to subsequent steps in the
same job.
