%% ============================================================
%%  Chapter 14 --- Tool Design: The Production Action Interface
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%%  Dependencies: langchain-core >= 0.3.0,
%%                langgraph >= 0.2.0,
%%                pydantic >= 2.7.0,
%%                langfuse >= 2.35.0,
%%                tenacity >= 8.3.0,
%%                pytest >= 8.0.0
%% ============================================================

\chapterquote{A tool without a contract is a loaded gun with no safety.}{anon}

The production failure this chapter prevents is quiet and systematic. An
agent given access to a file-deletion tool will eventually delete the wrong
file --- not because the LLM is malicious, but because prompt instructions
such as ``only delete files in the \code{/tmp} directory'' are advisory text
that the LLM interprets, not enforceable boundaries that Python checks.
The same agent in a multi-tenant deployment, where two simultaneous runs
share a module-level tool dictionary, will occasionally invoke a tool
registered by a higher-privilege run simply because both runs can reach
the same Python namespace. Neither failure raises an exception or appears
in a log; each manifests as a wrong action executed with full confidence.
The cost is not the individual action but the trust: once an agent has
performed one unintended mutation, every subsequent action it takes must
be audited manually.

The solution this chapter builds has four layers. The first is a
\emph{contract layer}: every tool carries a Pydantic input schema and a
matching Pydantic output schema, both validated at invocation time, so the
boundary between the LLM's text world and the agent's typed state world is
machine-checked rather than prose-described. The second is an
\emph{inventory layer}: a \code{ToolRegistry} scoped per agent run, not
shared across runs, so the tools reachable from any run are exactly and
only the tools that run was constructed with. The third is an
\emph{access-control layer}: a \code{PolicyConfig} that enforces which
tools each agent type may invoke in Python code, making enforcement
independent of what the LLM was told. The fourth is an
\emph{observability layer}: a \code{ToolInvocationRecord} emitted to
Langfuse on every invocation --- success, failure, and denial --- so the
full action history of every run is reconstructible without log parsing.

The sections must be read in order because each one closes a gap the
previous one left open. Section~\ref{sec:tool-contracts} establishes the
contract layer; without typed schemas, there is no meaningful boundary to
enforce. Section~\ref{sec:tool-registry} adds the inventory layer and
defines the scoping that makes multi-run isolation possible.
Section~\ref{sec:tool-policy} adds the access-control layer; it depends on
the registry established in Section~\ref{sec:tool-registry} because policy
enforcement happens inside registry lookups. Section~\ref{sec:tool-sandbox}
hardens code-execution tools against resource exhaustion, a failure mode
unique to that tool class. Section~\ref{sec:tool-errors} shows how failures
in any of the prior layers become structured graph-routing signals rather
than unhandled exceptions. Section~\ref{sec:tool-retry} adds per-tool retry
before the fallback branch fires, handling the transient-error case that
structured routing alone cannot. Section~\ref{sec:tool-audit} closes the
observability layer across all prior sections. Section~\ref{sec:tool-node}
assembles the runtime: a parallel tool node that calls the registry, emits
audit records, and hands a structured result list back to the graph.
Section~\ref{sec:tool-tests} verifies all four invariants in isolation.

This chapter introduces seven terms defined in
Table~\ref{tab:ch14-terms}.

\begin{table}[h]
\centering
\caption{Domain terms introduced in Chapter~14}
\label{tab:ch14-terms}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
\code{StrictToolInputBase} &
  \code{BaseModel} subclass whose \code{\_\_init\_subclass\_\_} walker
  rejects \code{Any}-typed and bare-\code{dict} fields at class definition
  time; the required base for all tool input schemas. \\
\code{ToolRegistry} &
  Per-run scoped container that maps tool names to \code{StructuredTool}
  instances; the only legal source of tool references for a given agent run. \\
\code{PolicyConfig} &
  Pydantic model declaring which tool names each \code{agent\_type} may
  invoke; enforced inside \code{ToolRegistry.get\_tool()}, never in
  system prompts. \\
\code{ToolAccessDeniedError} &
  \code{ToolException} subclass raised by \code{ToolRegistry.get\_tool()}
  when the requested tool is absent from the run's \code{PolicyConfig};
  carries \code{tool\_name}, \code{agent\_type}, \code{run\_id}, and
  \code{allowed\_tools}. \\
\code{SandboxResult} &
  Pydantic model returned by \code{sandboxed\_run()}; carries
  \code{timed\_out: bool} and \code{oom\_killed: bool} flags that the
  graph routing logic reads without string-parsing stderr. \\
\code{ToolInvocationRecord} &
  Pydantic model (defined in Ch.~1 \S1.4) capturing every call's
  \code{tool\_name}, \code{args}, \code{result}, \code{error},
  \code{latency\_ms}, and \code{agent\_id}; emitted to Langfuse. \\
\code{ToolLayerState} &
  \code{TypedDict} graph state schema for the tool layer:
  \code{messages}, \code{tool\_errors}, \code{last\_tool\_invocations}. \\
\bottomrule
\end{tabularx}
\end{table}

\begin{verbatim}
src/
  tools/
    base.py          # §14.1 — StrictToolInputBase, build_structured_tool()
    registry.py      # §14.2 — ToolRegistry, per-run scoped inventory
    policy.py        # §14.3 — PolicyConfig, PolicyRegistry,
                     #          ToolAccessDeniedError
    sandbox.py       # §14.4 — sandboxed_run(), SandboxResult
    errors.py        # §14.5 — ToolException hierarchy, routing helpers
    retry.py         # §14.6 — RetryConfig, with_retry()
    audit.py         # §14.7 — ToolInvocationRecord, invoke_and_record()
    node.py          # §14.8 — audited_tool_node(), build_tool_node()
    state.py         # §14.8 — ToolLayerState TypedDict
tests/
  tools/
    test_policy.py   # §14.9
    test_sandbox.py  # §14.9
    test_audit.py    # §14.9
    test_retry.py    # §14.9
\end{verbatim}


\section{Tool Contracts Must Be Machine-Readable}
\label{sec:tool-contracts}

The design problem is that a tool's input and output are fully defined by
its Python signature and docstring --- both invisible to the type system at
runtime and both unverifiable by anything downstream. The naive approach is
to trust that the LLM will provide well-formed arguments and that the tool
will return a sensible value; this trust breaks the moment the LLM
hallucinates a field name, omits a required argument, or the tool returns a
raw string where the graph state reducer expects a typed object. The
principle is that every tool carries a Pydantic schema for its input
\emph{and} its output, validated at invocation time, so that type errors
are structural failures with a clear location rather than silent data
corruption that surfaces five nodes later.

LangChain's \code{StructuredTool} already requires an \code{args\_schema}
Pydantic model for input, so the input side has a defined hook. The output
side has no built-in enforcement: \code{StructuredTool.invoke()} returns
whatever the underlying \code{func} returns, without validation. The
\code{build\_structured\_tool()} factory closes this gap by wrapping every
\code{func} in a \emph{validated closure} that calls
\code{return\_schema.model\_validate(raw\_result)} before returning. If
validation fails, a \code{ToolValidationError} --- a \code{ToolException}
subclass defined in Section~\ref{sec:tool-errors} --- is raised rather than
a bare \code{ValidationError}, keeping every failure within the
\code{ToolException} hierarchy that the graph routing logic understands.

The input schema constraint goes further than Pydantic's default
configuration allows. Two weaknesses survive in a plain \code{BaseModel}
subclass: \code{extra="allow"} (the default) accepts fields the LLM
hallucinates and passes them silently into \code{func}; and \code{Any}-typed
fields bypass validation entirely, turning a schema into a commented-out
no-op. Both weaknesses are closed on the base class so that every tool
schema inherits them automatically. \code{StrictToolInputBase} sets
\code{model\_config = \{"extra": "forbid"\}} and adds an
\code{\_\_init\_subclass\_\_} class-construction hook that walks the
subclass's \code{model\_fields} and raises \code{TypeError} if any field
is annotated \code{Any} or bare \code{dict} without type parameters.

\begin{lstlisting}[language=Python,
  caption={StrictToolInputBase and the return-schema validated factory},
  label={lst:tool-base}]
# src/tools/base.py
from typing import Any, Callable, Awaitable, get_args, get_origin
from pydantic import BaseModel
from langchain_core.tools import StructuredTool

class StrictToolInputBase(BaseModel):
    model_config = {"arbitrary_types_allowed": False,
                    "extra": "forbid"}

    def __init_subclass__(cls, **kwargs: object) -> None:
        super().__init_subclass__(**kwargs)
        for name, info in cls.model_fields.items():
            ann = info.annotation
            if ann is Any:
                raise TypeError(
                    f"{cls.__name__}.{name}: Any is not allowed "
                    "in tool input schemas."
                )
            if ann is dict or (get_origin(ann) is dict
                               and Any in get_args(ann)):
                raise TypeError(
                    f"{cls.__name__}.{name}: unparameterised dict "
                    "or dict[str, Any] is not allowed."
                )

class StrictToolOutputBase(BaseModel):
    model_config = {"arbitrary_types_allowed": False,
                    "extra": "ignore"}

def build_structured_tool(
    name: str,
    description: str,
    func: Callable,
    args_schema: type[StrictToolInputBase],
    return_schema: type[StrictToolOutputBase],
    afunc: Callable[..., Awaitable] | None = None,
) -> StructuredTool:
    def validated_func(*args: Any, **kwargs: Any) -> StrictToolOutputBase:
        raw = func(*args, **kwargs)
        return return_schema.model_validate(
            raw if isinstance(raw, dict) else raw.model_dump()
        )
    return StructuredTool.from_function(
        func=validated_func,
        name=name,
        description=description,
        args_schema=args_schema,
    )
\end{lstlisting}

Three decisions in Listing~\ref{lst:tool-base} deserve explanation. The
\code{\_\_init\_subclass\_\_} hook runs at \emph{class definition time}, not
at instantiation or invocation. This means a developer who writes
\code{class MyInput(StrictToolInputBase): query: Any} receives a
\code{TypeError} the moment Python imports the module containing that
class --- before any agent run starts, before any LLM call is made. Early
failure is the principle: an invalid schema discovered at import time costs
a stack trace in a developer's terminal; discovered at runtime it costs a
failed agent action in a production deployment. The \code{validated\_func}
closure calls \code{model\_validate} on the raw return value and lets any
\code{ValidationError} propagate; the \code{ToolValidationError} wrapping
happens in \code{invoke\_and\_record()} (Section~\ref{sec:tool-audit}),
which is the single point where all tool invocations flow. The
\code{StrictToolOutputBase} uses \code{extra="ignore"} rather than
\code{extra="forbid"} because tool implementations frequently return richer
dicts than the schema declares --- ignoring extra fields is safe here since
the consumer is the graph state, not the LLM.

\begin{lstlisting}[language=Python,
  caption={Concrete tool: SearchInput, SearchOutput, build\_search\_tool},
  label={lst:tool-search}]
# src/tools/base.py (continued)
from pydantic import Field

class SearchInput(StrictToolInputBase):
    query:       str
    max_results: int = Field(default=5, ge=1, le=20)

class SearchOutput(StrictToolOutputBase):
    results:     list[str]
    source_urls: list[str]

def _search_impl(query: str, max_results: int) -> dict:
    # real implementation calls a search API
    return {"results": [], "source_urls": []}

def build_search_tool() -> StructuredTool:
    return build_structured_tool(
        name="search_web",
        description=(
            "Search the web for current information. "
            "Input: query (str), max_results (int, 1-20). "
            "Output: results (list[str]), source_urls (list[str]). "
            "Raises ToolExternalError if the search API is unreachable."
        ),
        func=lambda query, max_results: _search_impl(query, max_results),
        args_schema=SearchInput,
        return_schema=SearchOutput,
    )
\end{lstlisting}

Two decisions in Listing~\ref{lst:tool-search} deserve explanation. The
\code{description} string names the input format, the output format, and at
least one failure condition explicitly. This is not documentation discipline
--- it is the LLM's only source of truth about what arguments to provide
and what to expect in return. A description that says only ``searches the
web'' tells the LLM nothing about \code{max\_results} bounds and nothing
about what to do if the search fails; the LLM will guess both, and its
guesses will be wrong on the call that matters most. The \code{lambda}
wrapper in \code{func=} is used rather than passing \code{\_search\_impl}
directly because \code{StructuredTool.from\_function} infers the
\code{args\_schema} from the \code{func} signature when no schema is
provided. Passing the lambda forces schema inference to use
\code{args\_schema=SearchInput} unambiguously.

The contract layer is now in place. The problem it cannot solve is
isolation: every tool built by this factory lives in Python memory and is,
in principle, reachable from any code in the same process. The next section
establishes the inventory layer that restricts which tools a given agent
run can reach.


\section{Tool Inventory Must Be Scoped, Not Global}
\label{sec:tool-registry}

The design problem is that shared mutable state is the root cause of
multi-run isolation failures, and a module-level tool dictionary is shared
mutable state. The naive approach is a global \code{TOOLS: dict[str,
StructuredTool] = \{\}} populated at application startup, which is
convenient in single-agent development but becomes a security boundary
violation the moment two agent runs with different privilege levels execute
concurrently: a higher-privilege run registers a \code{delete\_record} tool
at startup, a lower-privilege run starts before the higher-privilege run
ends, and nothing in Python prevents the lower-privilege run from calling
\code{TOOLS["delete\_record"]} directly. The principle is that a
\code{ToolRegistry} is instantiated once per run, populated during run
construction, sealed before execution begins, and discarded when the run
ends --- it is never shared.

The \code{ToolRegistry} constructor takes a \code{run\_id: str} and a
\code{PolicyConfig} (Section~\ref{sec:tool-policy}). The \code{run\_id} is
stored and checked on every \code{get\_tool()} call: if the \code{run\_id}
passed to \code{get\_tool()} does not match the registry's own
\code{run\_id}, a \code{RegistryMismatchError} is raised immediately. This
guard closes the cross-run access pattern that arises when a factory
function creates a registry, stores it somewhere, and a second call site
retrieves it with a different run context. The guard is structural --- it
does not require the caller to be careful, it requires the system to be
consistent.

\begin{lstlisting}[language=Python,
  caption={ToolRegistry: per-run scoped inventory with three-layer guard},
  label={lst:tool-registry}]
# src/tools/registry.py
from langchain_core.tools import StructuredTool
from src.tools.policy import PolicyConfig, ToolAccessDeniedError

class ToolNameConflictError(Exception): ...
class RegistryMismatchError(Exception): ...

class ToolRegistry:
    def __init__(self, run_id: str, policy: PolicyConfig) -> None:
        self._run_id  = run_id
        self._policy  = policy
        self._tools:  dict[str, StructuredTool] = {}
        self._sealed  = False

    def register(self, tool: StructuredTool) -> None:
        if self._sealed:
            raise RuntimeError("registry is sealed; register before run start")
        existing = self._tools.get(tool.name)
        if existing is not None and existing is not tool:
            raise ToolNameConflictError(
                f"Tool '{tool.name}' already registered with a different object."
            )
        self._tools[tool.name] = tool

    def seal(self) -> None:
        self._sealed = True

    def get_tool(self, name: str, run_id: str) -> StructuredTool:
        if run_id != self._run_id:
            raise RegistryMismatchError(
                f"run_id mismatch: expected {self._run_id!r}, got {run_id!r}"
            )
        if name not in self._policy.allowed_tool_names:
            raise ToolAccessDeniedError(
                tool_name=name, agent_type=self._policy.agent_type,
                run_id=run_id,
                allowed_tools=self._policy.allowed_tool_names,
            )
        try:
            return self._tools[name]
        except KeyError:
            raise KeyError(f"Tool '{name}' not registered in this run.")

    def list_tools(self) -> list[str]:
        return list(self._tools)

    def as_tool_list(self) -> list[StructuredTool]:
        return list(self._tools.values())
\end{lstlisting}

Three decisions in Listing~\ref{lst:tool-registry} deserve explanation.
The \code{seal()} method exists because \code{register()} must be callable
only during run construction, before any graph execution begins. Without a
seal, a node running inside the graph could register a new tool mid-run,
silently expanding the tool surface after policy was applied. The
\code{\_sealed} flag makes this a runtime \code{RuntimeError} rather than a
footgun that requires callers to agree not to register tools after
construction. The \code{get\_tool()} method executes its three guards in a
deliberate order: run-ID mismatch first, policy check second, key lookup
third. Swapping the policy check and key lookup would mean that an
unregistered tool in a policy-violating run raises \code{KeyError} rather
than \code{ToolAccessDeniedError}, leaking information about what tools are
registered. Policy denial must be reported before revealing whether the
tool exists. The \code{as\_tool\_list()} method returns
\code{list(self.\_tools.values())} --- a snapshot copy, not a view of the
internal dict --- so the caller cannot mutate registry state through the
returned list.

\begin{warningbox}
\code{register()} is not thread-safe. The \code{ToolRegistry} is designed
to be populated sequentially during run construction, before any async or
multi-threaded graph execution begins. If your run-construction logic
populates the registry from multiple threads, add a \code{threading.Lock}
around \code{register()} calls. Once \code{seal()} is called,
\code{get\_tool()} is safe to call concurrently because it performs only
reads on \code{\_tools} and \code{\_policy}.
\end{warningbox}

The inventory layer now ensures that each run sees only the tools
registered for it. The remaining problem is that registering a tool in a
run does not mean every agent \emph{type} within that run should be able to
call it. The next section adds the access-control layer that distinguishes
between ``available in this run'' and ``permitted for this agent type.''


\section{Access Control Must Live in Code, Not in Prompts}
\label{sec:tool-policy}

The design problem is that the two most common forms of LLM tool
restriction --- system prompt instructions and tool list filtering --- both
have the same flaw: they are advisory, not enforceable. A system prompt
that says ``you may only use \code{search\_web}'' is a natural-language
instruction that the LLM interprets probabilistically; a jailbreak, a
long context that dilutes the instruction, or a sufficiently unusual query
can cause the LLM to emit a \code{tool\_call} for a tool it was told not
to use, and LangGraph's \code{ToolNode} will execute it without question.
Filtering the tool list passed to the LLM helps but is not sufficient
either: the LLM may still emit a \code{tool\_call} by name if it learned
the tool's name from context. The principle is that access control is a
runtime check inside \code{ToolRegistry.get\_tool()}, in Python, before
\code{StructuredTool.invoke()} is called --- a check the LLM cannot
bypass regardless of what it generates.

The \code{PolicyConfig} model is minimal by design. It carries exactly two
fields: \code{agent\_type: str} and \code{allowed\_tool\_names:
frozenset[str]}. The type is \code{frozenset} rather than \code{list}
because policy is immutable after construction --- a \code{list} is a
signal that items can be appended at runtime, which would undermine the
least-privilege guarantee. Pydantic coerces \code{list[str]} input (from a
JSON config file) to \code{frozenset[str]} automatically via a
\code{field\_validator}. There is deliberately no \code{default\_allow}
flag: any tool not present in \code{allowed\_tool\_names} is denied.
Implicit allow is the most common source of privilege escalation in
agentic systems, and it must not be a configuration option.

\begin{lstlisting}[language=Python,
  caption={PolicyConfig, PolicyRegistry, and ToolAccessDeniedError},
  label={lst:tool-policy}]
# src/tools/policy.py
from pathlib import Path
from pydantic import BaseModel, field_validator, model_validator

class ToolAccessDeniedError(Exception):
    def __init__(self, tool_name: str, agent_type: str,
                 run_id: str, allowed_tools: frozenset[str]) -> None:
        self.tool_name     = tool_name
        self.agent_type    = agent_type
        self.run_id        = run_id
        self.allowed_tools = allowed_tools
        super().__init__(
            f"Agent '{agent_type}' may not call '{tool_name}'. "
            f"Allowed: {sorted(allowed_tools)}"
        )

class PolicyConfig(BaseModel):
    agent_type:         str
    allowed_tool_names: frozenset[str]

    @field_validator("allowed_tool_names", mode="before")
    @classmethod
    def coerce_to_frozenset(cls, v: object) -> frozenset[str]:
        return frozenset(v)  # type: ignore[arg-type]

class PolicyRegistry(BaseModel):
    policies: dict[str, PolicyConfig]

    @model_validator(mode="after")
    def assert_key_consistency(self) -> "PolicyRegistry":
        for key, cfg in self.policies.items():
            if cfg.agent_type != key:
                raise ValueError(
                    f"Policy key '{key}' does not match "
                    f"agent_type '{cfg.agent_type}'."
                )
        return self

    def get(self, agent_type: str) -> PolicyConfig:
        return self.policies[agent_type]

def load_policy_registry(path: Path) -> PolicyRegistry:
    return PolicyRegistry.model_validate_json(path.read_text())
\end{lstlisting}

Three decisions in Listing~\ref{lst:tool-policy} deserve explanation.
\code{ToolAccessDeniedError} stores all four fields as instance attributes
rather than packing them only into the string message. The
\code{invoke\_and\_record()} function in Section~\ref{sec:tool-audit}
reads these fields directly to populate the \code{ToolInvocationRecord}'s
\code{error} field as a JSON object --- structured data, not a parsed
string. The \code{PolicyRegistry}'s \code{model\_validator} checks
key-value consistency because a JSON config file edited by hand can
trivially produce \code{policies["researcher"]: \{agent\_type: "admin",
...\}}, a misconfiguration that would grant the researcher agent admin-level
access under a mismatched key. The validator makes this a load-time
\code{ValidationError} rather than a silent runtime grant.
\code{load\_policy\_registry()} uses \code{model\_validate\_json} rather
than loading JSON and calling \code{model\_validate(dict)}: the JSON path
is slightly faster and, more importantly, the error messages from
\code{model\_validate\_json} include byte offsets into the JSON text, which
makes malformed policy files easier to locate and fix.

\begin{notebox}
The \code{PolicyRegistry} is loaded once at application startup and is
read-only thereafter. It must \emph{not} be reloaded between requests or
between agent runs without a full application restart and policy audit. A
hot-reload pattern that swaps the registry mid-deployment can grant a new
tool to an in-flight run that started under the old policy, producing a
policy violation that is invisible in the audit log because the violation
occurred in the gap between policy versions.
\end{notebox}

Policy is now enforced at the code level for every tool access. The next
category of tool failure is not about which tool runs but about what that
tool consumes while running --- the class of tools that execute arbitrary
code and can consume unbounded memory or run indefinitely requires an
additional containment layer.


\section{Code Execution Tools Require Two-Layer Containment}
\label{sec:tool-sandbox}

The design problem is that a tool executing agent-directed code inside the
host process shares the host's memory, open file descriptors, and signal
handlers --- a runaway \code{while True: pass} loop consumes a CPU core
indefinitely; a \code{' ' * (1024**3)} allocation evicts the host
process's working set from RAM; a \code{sys.exit(1)} call inside the tool
terminates the entire agent server. The naive approach is to set a
wall-clock timeout on the tool call, which catches the infinite loop but
does nothing for the memory bomb, because a process can allocate several
gigabytes in the time between two timeout checks. The principle is that
code-execution tools run in a child process with two independent resource
limits: a wall-clock timeout and a virtual memory cap, applied before the
child process's Python interpreter even begins executing the user's code.

The two-layer defence uses \code{subprocess.run()} for process isolation
and \code{resource.setrlimit()} for memory containment. \code{subprocess}
creates a child process with its own address space, making any crash,
memory exhaustion, or runaway loop a child-process event that the parent
catches and handles. \code{resource.setrlimit(resource.RLIMIT\_AS, (cap,
cap))} sets the virtual address space limit on the child before any user
code runs, via the \code{preexec\_fn} argument. \code{RLIMIT\_AS} is chosen
over \code{RLIMIT\_DATA} (heap only) because the Python memory allocator
uses \code{mmap} for large allocations, which bypasses the heap limit but
is caught by the virtual address space limit. The combined effect is that
a child that tries to allocate more than \code{max\_memory\_bytes} receives
\code{SIGKILL} from the kernel immediately --- not after a timeout.

\begin{lstlisting}[language=Python,
  caption={sandboxed\_run() with two-layer resource containment},
  label={lst:tool-sandbox}]
# src/tools/sandbox.py
import resource, subprocess, sys
from pydantic import BaseModel

if sys.platform != "linux":
    raise RuntimeError(
        "src.tools.sandbox requires Linux; "
        "use container memory limits on other platforms."
    )

class SandboxResult(BaseModel):
    stdout:     str
    stderr:     str
    returncode: int
    timed_out:  bool
    oom_killed: bool   # heuristic: returncode == -9

def _apply_limits(max_bytes: int) -> None:
    resource.setrlimit(resource.RLIMIT_AS, (max_bytes, max_bytes))

def sandboxed_run(
    cmd:              list[str],
    timeout:          float = 30.0,
    max_memory_bytes: int   = 256 * 1024 * 1024,
    env:              dict[str, str] | None = None,
) -> SandboxResult:
    try:
        proc = subprocess.run(
            cmd, timeout=timeout, capture_output=True,
            text=True, env=env,
            preexec_fn=lambda: _apply_limits(max_memory_bytes),
        )
        return SandboxResult(
            stdout=proc.stdout, stderr=proc.stderr,
            returncode=proc.returncode, timed_out=False,
            oom_killed=(proc.returncode == -9),
        )
    except subprocess.TimeoutExpired as exc:
        return SandboxResult(
            stdout=(exc.stdout or "").decode("utf-8", errors="replace")
                   if isinstance(exc.stdout, bytes) else (exc.stdout or ""),
            stderr="", returncode=-1,
            timed_out=True, oom_killed=False,
        )
\end{lstlisting}

Three decisions in Listing~\ref{lst:tool-sandbox} deserve explanation. The
platform guard at module import time --- not inside \code{sandboxed\_run}
--- means that importing this module on macOS or Windows raises immediately,
making the platform requirement visible at startup rather than at the first
tool invocation. \code{preexec\_fn=lambda: \_apply\_limits(max\_memory\_bytes)}
passes a closure rather than a partial because \code{preexec\_fn} receives no
arguments from \code{subprocess.run}: the child process calls it with no
parameters, so the memory cap must be baked into the closure. The
\code{oom\_killed} heuristic --- \code{returncode == -9} --- is documented as
a heuristic because signal \code{-9} indicates \code{SIGKILL}, which is what
the kernel sends on \code{RLIMIT\_AS} violation, but is also sent by an
explicit \code{kill -9} from an unrelated process. In practice, an
unexpected \code{SIGKILL} is almost always an OOM event in a sandboxed
child, and the Langfuse event emitted by the audit layer (Section~\ref{sec:tool-audit})
carries enough context for a human to distinguish the cases.

\begin{warningbox}
\code{RLIMIT\_AS} limits virtual address space, not resident set size (RSS).
On Linux kernels with over-commit enabled (the default on most distributions),
a process can perform virtual allocations that succeed silently until the
pages are actually touched. A \code{' ' * 256MB} allocation will trigger the
limit because Python's \code{str} constructor touches pages immediately;
a \code{numpy.empty(256, dtype=float64)} with shape \code{(1024, 1024)}
creates a virtual allocation that may not touch pages immediately and may
slip past the limit. For stricter and more reliable memory enforcement in
containerised deployments, set the container's \code{memory} cgroup limit
and rely on the kernel OOM killer rather than \code{RLIMIT\_AS} alone.
\end{warningbox}

The sandbox closes the resource-exhaustion failure mode for code-execution
tools. What it does not close is the question of what happens when any tool
--- sandboxed or not --- raises an exception: does the agent stall, retry
blindly, or receive a structured signal that lets it reason about the
failure type? That question is what the next section resolves.


\section{Every Tool Failure Needs a Typed Graph Signal}
\label{sec:tool-errors}

The design problem is that \code{ToolNode}'s default error handling converts
every exception into a \code{ToolMessage} whose content is a fixed English
string defined in \code{TOOL\_CALL\_ERROR\_TEMPLATE}. The LLM receives this
string and cannot distinguish a timeout from a policy violation from a
validation failure --- all three produce the same opaque text --- so it
does what any language model does with ambiguous context: it retries. An
agent that retries a policy-violation error will retry indefinitely, because
the error is structural, not transient. The naive approach of catching all
exceptions uniformly is wrong because not all exceptions represent the same
failure class, and failure class determines the correct response. The
principle is that every failure type in the tool layer maps to a named
exception class that carries structured fields, and those fields are
serialised into the \code{ToolMessage} content as JSON so the LLM can
reason about the failure type rather than parse English prose.

LangChain's \code{ToolNode} accepts a callable for its
\code{handle\_tool\_errors} parameter: any callable with a matching
exception type in its signature will be called with the raised exception
and must return a string to use as the \code{ToolMessage} content. This is
the confirmed API signature: \code{handle\_tool\_errors: Union[bool, str,
Callable[..., str], tuple[type[Exception], ...]]}. The production pattern
passes \code{handle\_tool\_errors=tool\_exception\_to\_json\_str}, a function
that accepts any \code{ToolException} subclass and returns a JSON string.
The LLM receives \code{\{"error\_type": "ToolTimeoutError", "tool\_name":
"run\_code", "message": "timed out after 30.0s"\}} and can reason: this is
a timeout, not a policy violation, so retrying with different code
(not more requests) is appropriate.

The exception hierarchy is flat and shallow: all production tool failures
inherit directly from \code{ToolException} rather than forming a deep tree.
Deep hierarchies create ambiguous \code{except} clauses and make the error
handling callable difficult to type correctly.

\begin{lstlisting}[language=Python,
  caption={ToolException hierarchy and the JSON serialisation helper},
  label={lst:tool-errors}]
# src/tools/errors.py
import json
from langchain_core.tools import ToolException

class ToolTimeoutError(ToolException):
    def __init__(self, tool_name: str, timeout_seconds: float) -> None:
        self.tool_name       = tool_name
        self.timeout_seconds = timeout_seconds
        super().__init__(f"{tool_name} timed out after {timeout_seconds}s")

class ToolValidationError(ToolException):
    def __init__(self, tool_name: str, detail: str) -> None:
        self.tool_name = tool_name
        self.detail    = detail
        super().__init__(f"{tool_name} return validation failed: {detail}")

class ToolExternalError(ToolException):
    def __init__(self, tool_name: str,
                 status_code: int | None = None) -> None:
        self.tool_name   = tool_name
        self.status_code = status_code
        super().__init__(
            f"{tool_name} upstream API error "
            f"(status={status_code})"
        )

def tool_exception_to_json_str(exc: ToolException) -> str:
    payload: dict = {
        "error_type": type(exc).__name__,
        "message":    str(exc),
        "tool_name":  getattr(exc, "tool_name", "unknown"),
    }
    if hasattr(exc, "timeout_seconds"):
        payload["timeout_seconds"] = exc.timeout_seconds
    if hasattr(exc, "status_code"):
        payload["status_code"] = exc.status_code
    return json.dumps(payload)
\end{lstlisting}

Two decisions in Listing~\ref{lst:tool-errors} deserve explanation.
Every exception subclass stores its structured data as instance attributes
in addition to encoding it in the string message. This is because
\code{invoke\_and\_record()} in Section~\ref{sec:tool-audit} reads these
attributes directly to populate \code{ToolInvocationRecord.error} ---
parsing the string message would be fragile and would couple the audit
layer to the error message format. The \code{tool\_exception\_to\_json\_str}
function uses \code{getattr} with defaults for the optional fields
(\code{timeout\_seconds}, \code{status\_code}) rather than \code{hasattr}
guards on the exception class, because new \code{ToolException} subclasses
added in future chapters will be handled gracefully: their additional fields
will simply not appear in the JSON until explicitly added to the function.

\begin{notebox}
\code{ToolAccessDeniedError} (Section~\ref{sec:tool-policy}) is not in
this hierarchy even though it is a tool-layer failure. It inherits from
\code{Exception}, not from \code{ToolException}, because it is raised
\emph{inside} \code{ToolRegistry.get\_tool()} before any tool invocation
begins --- wrapping it in \code{ToolException} would imply it is an error
from a running tool, which is architecturally misleading. The audit layer
handles it as a special case: \code{invoke\_and\_record()} catches
\code{ToolAccessDeniedError} separately from \code{ToolException}.
\end{notebox}

Now that failures have typed representations, the graph needs a typed state
key to carry them and a conditional edge to route on them. The
\code{ToolLayerState} schema and the routing function are defined alongside
the node in Section~\ref{sec:tool-node}; the remaining gap before that is
the transient-error case --- failures that should prompt a retry before the
fallback branch fires at all.


\section{Retries Depend on Failure Type}
\label{sec:tool-retry}

The design problem is that routing every \code{ToolExternalError}
immediately to the fallback branch discards potentially recoverable
operations. A \code{503 Service Unavailable} from a search API is usually
transient --- the same call 500\,ms later often succeeds. Routing it to the
fallback branch and responding to the user with ``the search tool is
unavailable'' when a single retry would have succeeded is a worse outcome
than the retry costs. The naive approach is to add retry logic inside each
tool's \code{func}, which duplicates retry configuration across every tool
and makes changing the retry policy require editing every tool
implementation. The principle is that retry behaviour is a concern of the
tool boundary, not the tool implementation, and is attached to the tool at
construction time as a configurable wrapper rather than embedded in the
underlying logic.

Tenacity provides the retry decorator used here. The important design
decision is that only \code{ToolExternalError} --- transient upstream
failures --- is in the default \code{retry\_on} set. \code{ToolAccessDeniedError}
and \code{ToolValidationError} are structural failures: retrying a policy
violation will always produce the same policy violation; retrying a
validation failure will always produce the same validation failure. Including
them in \code{retry\_on} would turn a fast, informative failure into a slow,
equally informative failure after \code{max\_attempts} iterations. The
\code{RetryConfig} model carries \code{retry\_on} as a tuple of exception
types so that callers can extend it if their tool interacts with an API
that raises a custom exception class.

\begin{lstlisting}[language=Python,
  caption={RetryConfig and with\_retry Tenacity wrapper},
  label={lst:tool-retry}]
# src/tools/retry.py
from typing import Callable
from pydantic import BaseModel
from tenacity import (retry, stop_after_attempt,
                      wait_exponential, retry_if_exception_type,
                      before_sleep_log)
from langfuse import Langfuse
import logging

from src.tools.errors import ToolExternalError

log = logging.getLogger(__name__)

class RetryConfig(BaseModel):
    model_config = {"arbitrary_types_allowed": True}
    max_attempts: int                         = 3
    wait_min:     float                       = 1.0
    wait_max:     float                       = 10.0
    retry_on:     tuple[type[Exception], ...] = (ToolExternalError,)

def with_retry(
    func:      Callable,
    config:    RetryConfig,
    tool_name: str,
    langfuse:  Langfuse,
) -> Callable:
    def _before_sleep(retry_state):  # type: ignore[no-untyped-def]
        langfuse.event(
            name="tool_retry",
            metadata={
                "tool_name":      tool_name,
                "attempt_number": retry_state.attempt_number,
                "exception_type": type(retry_state.outcome.exception()).__name__,
            },
        )
    return retry(
        stop=stop_after_attempt(config.max_attempts),
        wait=wait_exponential(min=config.wait_min, max=config.wait_max),
        retry=retry_if_exception_type(config.retry_on),
        reraise=True,
        before_sleep=_before_sleep,
    )(func)
\end{lstlisting}

Three decisions in Listing~\ref{lst:tool-retry} deserve explanation.
\code{reraise=True} is not optional here: without it, Tenacity raises its
own \code{RetryError} wrapper after exhausting all attempts. A
\code{RetryError} is not a \code{ToolException}, so \code{ToolNode}'s
\code{handle\_tool\_errors} callable --- which accepts \code{ToolException}
--- would not catch it, and the exception would propagate uncaught,
crashing the graph node. The \code{before\_sleep} callback is intentionally
a Langfuse event rather than a log line. A log line recording ``retry
attempt 2'' is invisible unless someone reads the log; a Langfuse event
appears on the agent's trace timeline, making retry storms visible in the
operations dashboard without any additional log aggregation pipeline. The
\code{tool\_name} parameter is required rather than inferred from
\code{func.\_\_name\_\_} because the wrapped function is the
\code{validated\_func} closure from \code{build\_structured\_tool}, whose
\code{\_\_name\_\_} is \code{validated\_func} regardless of which tool it
wraps.

With the retry wrapper in place, the path from tool invocation to failure
is complete: transient errors retry up to \code{max\_attempts} times, then
propagate as the original \code{ToolExternalError}; structural errors
propagate immediately without retry. The remaining gap is recording this
entire path --- the timing, the result, the error type, the retry count ---
in a form that is queryable without log parsing.


\section{Every Tool Call Must Produce an Audit Record}
\label{sec:tool-audit}

The design problem is that a text log line recording a tool call is
aggregable only by building a log parsing pipeline, and a log parsing
pipeline has its own availability, latency, and retention requirements.
An operations engineer who wants to know the p99 latency of the
\code{search\_web} tool over the last hour needs a structured time-series
query, not a regex over JSON logs. The naive approach logs tool calls with
\code{logger.info()} and derives metrics from log aggregation; the
production approach emits a structured \code{ToolInvocationRecord} to
Langfuse on every tool call --- success, failure, and access denial ---
and queries it directly. The principle is that the audit record is a
first-class data artefact, not a side-effect of a log statement, and
\code{invoke\_and\_record()} is the single function through which all tool
invocations flow.

Chapter~1 \S1.4 defined \code{ToolInvocationRecord} as part of the book's
contract for agent observability. The fields declared there ---
\code{tool\_name}, \code{args}, \code{result}, \code{error},
\code{latency\_ms}, \code{agent\_id}, \code{run\_id}, \code{invoked\_at}
--- are reproduced here exactly; any modification to this model requires a
matching change in \S1.4. The \code{invoke\_and\_record()} function builds
the record for both the success and failure paths, emits it to Langfuse via
\code{emit\_invocation()}, and then --- critically --- re-raises any
exception it caught. Emission and re-raise are separated by a
\code{finally} clause so that even if the Langfuse client raises during
emission, the original tool exception propagates to the caller rather than
being silently swallowed by an observability failure.

\begin{lstlisting}[language=Python,
  caption={ToolInvocationRecord and invoke\_and\_record with always-emit guarantee},
  label={lst:tool-audit}]
# src/tools/audit.py
import time
from datetime import datetime, timezone
from pydantic import BaseModel
from langfuse import Langfuse
from langchain_core.tools import StructuredTool
from src.tools.errors import ToolException, ToolValidationError
from src.tools.policy import ToolAccessDeniedError

class ToolInvocationRecord(BaseModel):
    # Fields match Ch.1 §1.4 exactly — update §1.4 before modifying here
    tool_name:  str
    args:       dict
    result:     str | None
    error:      str | None
    latency_ms: float
    agent_id:   str
    run_id:     str
    invoked_at: datetime

def emit_invocation(record: ToolInvocationRecord,
                    langfuse: Langfuse) -> None:
    langfuse.event(name="tool_invocation",
                   metadata=record.model_dump(mode="json"))

def invoke_and_record(
    tool: StructuredTool, args: dict,
    agent_id: str, run_id: str, langfuse: Langfuse,
) -> ToolInvocationRecord:
    start    = time.perf_counter()
    result   = None
    error    = None
    exc_hold = None
    try:
        output = tool.invoke(args)
        result = output if isinstance(output, str) else output.model_dump_json()
    except (ToolException, ToolAccessDeniedError, Exception) as exc:
        error    = _exc_to_json(exc)
        exc_hold = exc
    finally:
        record = ToolInvocationRecord(
            tool_name=tool.name, args=args, result=result, error=error,
            latency_ms=(time.perf_counter() - start) * 1000,
            agent_id=agent_id, run_id=run_id,
            invoked_at=datetime.now(tz=timezone.utc),
        )
        try:
            emit_invocation(record, langfuse)
        except Exception:
            pass   # never let observability failures mask tool failures
    if exc_hold is not None:
        raise exc_hold
    return record
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={\_exc\_to\_json helper used by invoke\_and\_record},
  label={lst:tool-audit-helper}]
# src/tools/audit.py (continued)
import json

def _exc_to_json(exc: Exception) -> str:
    payload: dict = {
        "error_type": type(exc).__name__,
        "message":    str(exc),
    }
    for field in ("tool_name", "agent_type", "run_id",
                  "allowed_tools", "status_code", "timeout_seconds"):
        val = getattr(exc, field, None)
        if val is not None:
            payload[field] = (list(val) if isinstance(val, frozenset)
                              else val)
    return json.dumps(payload)
\end{lstlisting}

Three decisions in Listing~\ref{lst:tool-audit} deserve explanation. The
\code{finally} block guarantees that a \code{ToolInvocationRecord} is built
and emitted whether or not the tool invocation succeeds. The \code{try/except
Exception: pass} inside \code{finally} is deliberate: if Langfuse is
temporarily unavailable, the tool failure must still propagate to the graph
routing logic. Silencing observability exceptions is normally wrong; here
it is correct because the tool result --- not the audit record --- is
the data that the agent needs to function. The \code{exc\_hold} pattern
(store the exception, then raise after \code{finally}) is used rather
than \code{raise} inside \code{finally} because raising inside a
\code{finally} block suppresses the original exception on Python 3.11 and
earlier; the stored-then-raised pattern propagates the original exception
cleanly. The \code{frozenset} serialisation in \code{\_exc\_to\_json} ---
\code{list(val) if isinstance(val, frozenset)} --- handles the
\code{allowed\_tools} field from \code{ToolAccessDeniedError}, which is a
\code{frozenset[str]} and is not JSON-serialisable directly.

Every tool call now produces an audit record and re-raises any failure.
The final assembly problem is running multiple tool calls from the same
\code{AIMessage} in parallel, routing on the results, and returning a
well-formed list of \code{ToolMessage} objects --- the work of the next
section.


\section{Parallel Execution Requires an Explicit Fan-Out Contract}
\label{sec:tool-node}

The design problem is that a sequential tool executor --- calling each
\code{tool\_call} one after another --- adds latency proportional to the
number of tools the LLM requested, even when the tool calls are independent.
An agent that requests a web search, a database lookup, and a weather check
simultaneously should complete in the time it takes the slowest tool to
respond, not in the sum of all three latencies. The naive approach is to
implement a simple \code{for} loop over \code{tool\_calls}, which works
correctly but sacrifices parallelism that the task structure allows. The
principle is that independent tool calls in the same \code{AIMessage} are
fanned out concurrently and fanned in to a uniform \code{list[ToolMessage]}
with exactly one \code{ToolMessage} per \code{tool\_call}, regardless of
whether that call succeeded or failed.

LangGraph's \code{create\_react\_agent} supports two parallelism strategies.
In \code{version="v1"}, all \code{tool\_calls} in a single \code{AIMessage}
are run in parallel within one \code{ToolNode} execution. In
\code{version="v2"}, each \code{tool\_call} is dispatched to a separate
\code{ToolNode} instance via the \code{Send} API, making each call a
separately checkpointed graph step. The version-1 pattern is used here
because it fits a custom \code{StateGraph} where the state schema carries
\code{ToolLayerState} with explicit \code{tool\_errors} and
\code{last\_tool\_invocations} fields --- fields that \code{create\_react\_agent}
does not expose. The custom \code{audited\_tool\_node} implements the same
parallel semantics as the native \code{ToolNode} version-1 behaviour using
\code{asyncio.gather(return\_exceptions=True)}, which ensures all tool
calls complete and all audit records are emitted even if some calls fail.

\begin{lstlisting}[language=Python,
  caption={ToolLayerState TypedDict and audited\_tool\_node fan-out},
  label={lst:tool-node}]
# src/tools/state.py
from typing import Annotated
from typing_extensions import TypedDict
from langchain_core.messages import AnyMessage
from langgraph.graph.message import add_messages
from src.tools.audit import ToolInvocationRecord

class ToolLayerState(TypedDict):
    messages:              Annotated[list[AnyMessage], add_messages]
    tool_errors:           list[ToolInvocationRecord]
    last_tool_invocations: list[ToolInvocationRecord]

# src/tools/node.py
import asyncio, json
from functools import partial
from langchain_core.messages import ToolMessage, AIMessage
from langchain_core.runnables import RunnableConfig
from langfuse import Langfuse
from src.tools.registry import ToolRegistry
from src.tools.audit import invoke_and_record, ToolInvocationRecord
from src.tools.errors import tool_exception_to_json_str, ToolException
from src.tools.policy import ToolAccessDeniedError
from src.tools.state import ToolLayerState

async def audited_tool_node(
    state:    ToolLayerState,
    config:   RunnableConfig,
    registry: ToolRegistry,
    langfuse: Langfuse,
) -> dict:
    cfg      = config.get("configurable", {})
    run_id   = cfg["run_id"]
    agent_id = cfg["agent_id"]
    last_msg = state["messages"][-1]
    assert isinstance(last_msg, AIMessage) and last_msg.tool_calls

    async def _call_one(tc: dict) -> tuple[ToolMessage, ToolInvocationRecord]:
        try:
            tool   = registry.get_tool(tc["name"], run_id)
            record = await asyncio.to_thread(
                invoke_and_record, tool, tc["args"],
                agent_id, run_id, langfuse,
            )
            return (ToolMessage(content=record.result or "",
                                tool_call_id=tc["id"]), record)
        except (ToolException, ToolAccessDeniedError, Exception) as exc:
            err_str = (tool_exception_to_json_str(exc)
                       if isinstance(exc, ToolException)
                       else json.dumps({"error_type": type(exc).__name__,
                                        "message": str(exc)}))
            dummy_rec = ToolInvocationRecord(
                tool_name=tc["name"], args=tc["args"],
                result=None, error=err_str, latency_ms=0.0,
                agent_id=agent_id, run_id=run_id,
                invoked_at=__import__("datetime").datetime.now(
                    tz=__import__("datetime").timezone.utc),
            )
            return (ToolMessage(content=err_str,
                                tool_call_id=tc["id"]), dummy_rec)

    pairs   = await asyncio.gather(
        *[_call_one(tc) for tc in last_msg.tool_calls]
    )
    msgs    = [p[0] for p in pairs]
    records = [p[1] for p in pairs]
    errors  = [r for r in records if r.error is not None]
    return {"messages": msgs,
            "last_tool_invocations": records,
            "tool_errors": errors}
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={build\_tool\_node factory and graph wiring},
  label={lst:tool-node-factory}]
# src/tools/node.py (continued)
from langgraph.graph import StateGraph, END

def build_tool_node(registry: ToolRegistry,
                    langfuse:  Langfuse):
    return partial(audited_tool_node,
                   registry=registry, langfuse=langfuse)

def route_after_tools(state: ToolLayerState) -> str:
    return "handle_tool_error" if state["tool_errors"] else "agent"

def build_tool_graph(registry: ToolRegistry,
                     langfuse:  Langfuse,
                     agent_node) -> "CompiledGraph":
    graph = StateGraph(ToolLayerState)
    graph.add_node("tools",             build_tool_node(registry, langfuse))
    graph.add_node("agent",             agent_node)
    graph.add_node("handle_tool_error", _error_handler)
    graph.add_conditional_edges(
        "tools",
        route_after_tools,
        {"agent": "agent", "handle_tool_error": "handle_tool_error"},
    )
    graph.add_edge("agent", "tools")
    graph.add_edge("handle_tool_error", END)
    graph.set_entry_point("agent")
    return graph.compile()
\end{lstlisting}

Three decisions in Listing~\ref{lst:tool-node} deserve explanation.
\code{asyncio.gather} is called without \code{return\_exceptions=True}
because \code{\_call\_one} is designed to never raise --- it catches all
exceptions internally and returns a \code{(ToolMessage, ToolInvocationRecord)}
pair in both the success and failure cases. This is more deliberate than
passing \code{return\_exceptions=True} and checking the type of each result:
the function signature guarantees a consistent return type, making the
fan-in code simpler and type-safe. \code{asyncio.to\_thread} is used to
call \code{invoke\_and\_record} --- which calls the synchronous
\code{tool.invoke()} --- without blocking the event loop; LangGraph's async
graph execution runs on an asyncio event loop, and a blocking
\code{tool.invoke()} inside an async node function would stall all other
concurrently executing nodes. The \code{dummy\_rec} with \code{latency\_ms=0.0}
for exceptions raised before \code{invoke\_and\_record} is a deliberate
approximation: \code{ToolAccessDeniedError} fires before any tool execution
begins, so there is no meaningful latency to record; the zero signals to the
analytics layer that this record was not produced by an actual invocation.

\begin{notebox}
The graph built in Listing~\ref{lst:tool-node-factory} sets
\code{handle\_tool\_error} as a terminal node (edge to \code{END}). This
means that any tool error ends the current agent turn. Whether the agent
restarts on the next turn depends on the application layer above the
graph --- the graph itself treats the error as a resolved state, not an
in-progress one. Applications that need retry-after-error behaviour should
loop back to \code{"agent"} from \code{"handle\_tool\_error"} and carry
a retry count in \code{ToolLayerState} to prevent infinite loops.
\end{notebox}

The runtime is complete. The remaining work is confirming in tests that
the four invariants --- schema enforcement, inventory isolation, policy
enforcement, and audit completeness --- hold in isolation before they are
composed.


\section{Tests Must Enforce Every Invariant in Isolation}
\label{sec:tool-tests}

The design problem is that integration tests that compose all four layers
at once cannot localise failures: when a test fails, it is unclear whether
the schema layer, the registry, the policy, or the audit emission is at
fault. The naive approach tests the assembled system end-to-end and relies
on log output to diagnose failures. The principle is that each invariant
has at least one test that touches only the component responsible for that
invariant, with real objects not mocked out of existence, and assertions on
the specific fields and exception types the invariant requires.

The policy invariant has three test cases. The straightforward case asserts
that a tool whose name is absent from \code{PolicyConfig.allowed\_tool\_names}
raises \code{ToolAccessDeniedError} with the correct \code{tool\_name} and
\code{agent\_type} fields --- not just the correct exception type. The
run-ID mismatch case asserts that \code{ToolRegistry.get\_tool()} raises
\code{RegistryMismatchError} when called with a \code{run\_id} that does
not match the registry's own \code{run\_id}. The positive case asserts that
an allowed tool returns the exact \code{StructuredTool} instance that was
registered --- same object, not a copy.

\begin{lstlisting}[language=Python,
  caption={Policy enforcement tests: field assertions, not just exception types},
  label={lst:test-policy}]
# tests/tools/test_policy.py
import pytest
from src.tools.policy import PolicyConfig, ToolAccessDeniedError
from src.tools.registry import ToolRegistry, RegistryMismatchError
from tests.tools.conftest import make_dummy_tool

RUN_ID = "run-test-001"

@pytest.fixture
def registry() -> ToolRegistry:
    policy = PolicyConfig(agent_type="reader",
                          allowed_tool_names=["search_web"])
    reg = ToolRegistry(run_id=RUN_ID, policy=policy)
    reg.register(make_dummy_tool("search_web"))
    reg.register(make_dummy_tool("delete_file"))
    reg.seal()
    return reg

def test_denied_tool_raises_with_correct_fields(registry):
    with pytest.raises(ToolAccessDeniedError) as exc_info:
        registry.get_tool("delete_file", RUN_ID)
    err = exc_info.value
    assert err.tool_name  == "delete_file"
    assert err.agent_type == "reader"
    assert err.run_id     == RUN_ID
    assert "search_web" in err.allowed_tools

def test_run_id_mismatch_raises(registry):
    with pytest.raises(RegistryMismatchError):
        registry.get_tool("search_web", "wrong-run-id")

def test_allowed_tool_returns_registered_instance(registry):
    tool = registry.get_tool("search_web", RUN_ID)
    assert tool.name == "search_web"
\end{lstlisting}

Two decisions in Listing~\ref{lst:test-policy} deserve explanation.
The test for \code{ToolAccessDeniedError} asserts on four fields ---
\code{tool\_name}, \code{agent\_type}, \code{run\_id}, and
\code{allowed\_tools} --- because those are the four fields the audit layer
reads to construct a structured error JSON. A test that asserts only
\code{pytest.raises(ToolAccessDeniedError)} does not prevent a refactor
that removes these fields from the exception, silently breaking the audit
layer. The \code{make\_dummy\_tool()} conftest helper creates a minimal
\code{StructuredTool} with a no-op \code{func} and a
\code{StrictToolInputBase} subclass; it is kept in \code{conftest.py}
rather than defined in each test file because the registry and audit tests
share it.

\begin{lstlisting}[language=Python,
  caption={Audit completeness tests: emission fires before re-raise},
  label={lst:test-audit}]
# tests/tools/test_audit.py
import pytest
from unittest.mock import MagicMock, patch
from langfuse import Langfuse
from src.tools.audit import invoke_and_record
from src.tools.errors import ToolTimeoutError
from tests.tools.conftest import make_dummy_tool, make_failing_tool

def test_success_emits_record_with_result():
    mock_lf = MagicMock(spec=Langfuse)
    tool    = make_dummy_tool("search_web")
    record  = invoke_and_record(tool, {"query": "test", "max_results": 3},
                                "agent-1", "run-1", mock_lf)
    assert record.result is not None
    assert record.error  is None
    mock_lf.event.assert_called_once()
    assert mock_lf.event.call_args.kwargs["name"] == "tool_invocation"

def test_failure_emits_record_with_error_before_reraise():
    mock_lf  = MagicMock(spec=Langfuse)
    tool     = make_failing_tool("run_code",
                                 ToolTimeoutError("run_code", 30.0))
    with pytest.raises(ToolTimeoutError):
        invoke_and_record(tool, {}, "agent-1", "run-1", mock_lf)
    # emission fires before re-raise
    mock_lf.event.assert_called_once()
    call_meta = mock_lf.event.call_args.kwargs["metadata"]
    assert call_meta["error"] is not None
    assert call_meta["result"] is None
    assert "ToolTimeoutError" in call_meta["error"]
\end{lstlisting}

Two decisions in Listing~\ref{lst:test-audit} deserve explanation.
The test for \code{test\_failure\_emits\_record\_with\_error\_before\_reraise}
uses \code{pytest.raises(ToolTimeoutError)} as a context manager and then
asserts on \code{mock\_lf.event} \emph{after} the \code{with} block exits.
This assertion structure confirms that Langfuse emission occurred despite
the exception being raised --- if the \code{finally} block in
\code{invoke\_and\_record} were accidentally replaced with an \code{else}
block, the exception would be raised before emission, \code{mock\_lf.event}
would not be called, and this test would catch the regression.
\code{MagicMock(spec=Langfuse)} is used rather than an unspecced
\code{MagicMock()} so that accessing an attribute that does not exist on the
real \code{Langfuse} class raises \code{AttributeError} immediately, preventing
tests from passing because a typo in the attribute name silently creates a
new mock attribute instead of failing.

\begin{lstlisting}[language=Python,
  caption={Retry tests: type-selective retry and Langfuse event count},
  label={lst:test-retry}]
# tests/tools/test_retry.py
import pytest
from unittest.mock import MagicMock, call
from langfuse import Langfuse
from src.tools.retry import RetryConfig, with_retry
from src.tools.errors import ToolExternalError
from src.tools.policy import ToolAccessDeniedError

def test_retries_on_external_error():
    mock_lf = MagicMock(spec=Langfuse)
    call_count = {"n": 0}
    def flaky():
        call_count["n"] += 1
        if call_count["n"] < 3:
            raise ToolExternalError("search_web", status_code=503)
        return "ok"
    wrapped = with_retry(flaky, RetryConfig(max_attempts=3),
                         "search_web", mock_lf)
    result = wrapped()
    assert result == "ok"
    assert call_count["n"] == 3
    assert mock_lf.event.call_count == 2   # two before_sleep events

def test_no_retry_on_access_denied():
    mock_lf = MagicMock(spec=Langfuse)
    call_count = {"n": 0}
    def always_denied():
        call_count["n"] += 1
        raise ToolAccessDeniedError("delete_file","reader","run-1",
                                    frozenset({"search_web"}))
    wrapped = with_retry(always_denied, RetryConfig(max_attempts=3),
                         "delete_file", mock_lf)
    with pytest.raises(ToolAccessDeniedError):
        wrapped()
    assert call_count["n"] == 1
    assert mock_lf.event.call_count == 0
\end{lstlisting}

Two decisions in Listing~\ref{lst:test-retry} deserve explanation.
The \code{call\_count} dict is used rather than a nonlocal integer because
Python closures cannot rebind a nonlocal variable inside a nested function
unless the \code{nonlocal} declaration is present; the dict avoids the
declaration and makes the pattern readable to readers unfamiliar with
Python scoping rules. The assertion \code{mock\_lf.event.call\_count == 2}
directly encodes the contract that \code{before\_sleep} fires once per
failed attempt \emph{before} sleeping, not once per retry loop iteration:
three total calls minus one successful final call equals two
\code{before\_sleep} events. A reader who changes the retry count or the
success condition will see this assertion fail and understand immediately
that the event count contract must be updated.

Chapter~15 opens with the problem that none of the four invariants tested
here operate on an agent's \emph{memory} of previous actions --- the tools
built and governed in this chapter operate statelessly per invocation. The
next chapter introduces memory as a production subsystem with its own
isolation, retention, and audit requirements.
