
\chapterquote{Coupling is not a design smell. Invisible coupling is.}{anon}

The production failure this chapter prevents appears the moment one agent
must call another and the two teams have no shared, versioned contract for
identity, task state, or authentication scope. The naive approach is a
private HTTP endpoint and an informal document describing the payloads,
which works until the remote agent changes a field, adds a new pause state,
or tightens authentication and the caller discovers the change only after
shipping. The cost is not one failed request but a network of agents whose
integration surface grows faster than the number of agents themselves.

The solution has four layers. The first is \emph{identity}: every agent
publishes a machine-readable \code{AgentCard} describing its skills,
transport, and security expectations. The second is \emph{lifecycle}:
every delegated unit of work moves through one of nine canonical
\code{TaskState} values, with each transition written to durable state
before it is observed. The third is \emph{transport}: the official SDK
standardises request dispatch and SSE streaming so the caller consumes a
typed event stream rather than inventing its own wire protocol. The fourth
is \emph{delegation}: the parent agent mints short-lived, scope-narrowed
credentials for a single child call, preventing privilege escalation when
work crosses process boundaries.

The sections are ordered by dependency. Section~\ref{sec:a2a-lifecycle}
defines the task state machine first because every subsequent section
creates, observes, or routes on these states. Section~\ref{sec:a2a-agent-card}
builds the public identity document next, because the caller cannot safely
construct a client before it validates the remote agent's card.
Section~\ref{sec:a2a-sdk} introduces the official SDK on both the server
and caller sides, because the remaining sections assume the transport layer
is already solved. Section~\ref{sec:a2a-delegation} narrows delegated
authority with short-lived JWTs, and Section~\ref{sec:a2a-errors} maps
subagent failures to retry-or-escalate decisions while preserving partial
artifacts. Section~\ref{sec:a2a-integration} assembles the complete system
and hands the identity problem outward to the decentralised discovery model
of Chapter~18.

This chapter introduces eight terms defined in Table~\ref{tab:ch17-terms}.

\begin{table}[h]
\centering
\caption{Domain terms introduced in Chapter~17}
\label{tab:ch17-terms}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
\code{TaskState} &
  A \code{Literal} type alias of exactly nine lowercase string values:
  \code{"submitted"}, \code{"working"}, \code{"input-required"},
  \code{"completed"}, \code{"canceled"}, \code{"failed"},
  \code{"rejected"}, \code{"auth-required"}, and \code{"unknown"}. \\
\code{TaskStatus} &
  A wrapper object containing \code{state: TaskState}, an optional
  \code{message: Message}, and an optional \code{timestamp: str}. \\
\code{AgentCard} &
  The well-known JSON document served from
  \code{/.well-known/agent.json} that declares an agent's identity,
  capabilities, skills, and security requirements. \\
\code{AgentSkill} &
  A single named capability within an \code{AgentCard}; it is the unit
  of discovery and the granularity at which delegation tokens scope
  access. \\
Pause state &
  A non-terminal state that suspends execution pending caller action;
  in A2A these are \code{"input-required"} and
  \code{"auth-required"}. \\
\code{A2ATaskEvent} &
  A Pydantic model used in this chapter to validate every lifecycle
  transition before it is sent to Langfuse. \\
\code{TaskHandler} &
  The server-side SDK abstraction whose \code{on\_message()} method
  translates an incoming A2A request into \code{TaskStatusUpdateEvent}
  and \code{TaskArtifactUpdateEvent} objects. \\
Delegation token &
  A short-lived RS256 JWT minted by a parent agent and scoped to one
  or more child \code{AgentSkill.id} values. \\
\bottomrule
\end{tabularx}
\end{table}

\medskip
\noindent\textbf{Source layout for this chapter:}
\begin{verbatim}
chapter17/
  a2a_lifecycle.py          # §17.1 — lifecycle + Langfuse events
  agent_card.py             # §17.2 — AgentCard factory + routes
  a2a_server_base.py        # §17.3 — SDK server + client streaming
  delegation_tokens.py      # §17.4 — scoped JWT delegation
  error_propagation.py      # §17.5 — retry vs ESCALATE
  orchestrator.py           # §17.6 — parent LangGraph orchestrator
  subagent_summarizer.py    # §17.6 — summarizer subagent
  subagent_translator.py    # §17.6 — translator subagent
\end{verbatim}


%% ------------------------------------------------------------
\section{A Task Has a Finite Number of Valid States}
\label{sec:a2a-lifecycle}
%% ------------------------------------------------------------

The design problem is state drift: if two agents use different internal
labels for the same phase of work, the caller cannot reliably decide
whether to wait, retry, or escalate. The naive approach is a free-form
\code{status} string or a local \code{Enum} whose members are convenient
for one codebase but meaningless to every other one. The governing
principle is that the lifecycle vocabulary is shared protocol, not local
implementation detail: \code{TaskState} is an exhaustive set of nine
lowercase string literals, and every transition is an atomic
store-then-observe write.

The nine states split naturally into three families. \code{"submitted"}
and \code{"working"} indicate forward progress. \code{"input-required"}
and \code{"auth-required"} are pause states: they suspend execution while
keeping the task open, because the correct next action belongs to the
caller rather than the callee. \code{"completed"}, \code{"canceled"},
\code{"failed"}, \code{"rejected"}, and \code{"unknown"} are terminal
states; once a task reaches one of them, a second transition call must be
a no-op. This is also why \code{Task.contextId} matters alongside
\code{Task.id}: the \code{id} identifies one task record, while the
\code{contextId} ties together the caller's follow-up messages when a task
enters a pause state and later resumes.

Observability follows the state machine rather than defining it. The store
is written first, then a validated \code{A2ATaskEvent} is emitted to
Langfuse. Reversing that order would let an observability failure corrupt
the system of record: if the event is emitted first and the store write
fails second, the trace claims work happened that the store cannot prove.
Chapter~23~\S23.1 introduces Langfuse spans as a higher-level tracing
mechanism; here the event is deliberately lower level because transitions
must remain valid even when the rest of the trace pipeline is degraded.

\begin{warningbox}
Terminal writes must be idempotent. Without a guard that checks the
current state before writing, a slow \code{"working"} update can arrive
after a fast \code{"completed"} update and silently resurrect a task that
should already be closed.
\end{warningbox}

\begin{lstlisting}[language=Python,
  caption={Task lifecycle: atomic transition with validated event emission},
  label={lst:a2a-lifecycle}]
# chapter17/a2a_lifecycle.py
from datetime import datetime, timezone
from typing import Any, Protocol
from pydantic import BaseModel, ConfigDict
from a2a.types import Task, TaskState, TaskStatus

TERMINAL: frozenset[TaskState] = frozenset({
    "completed", "canceled", "failed", "rejected", "unknown",
})

class TaskStore(Protocol):
    async def get_task(self, task_id: str) -> Task: ...
    async def save_task(self, task: Task) -> None: ...

class A2ATaskEvent(BaseModel):
    model_config = ConfigDict(use_enum_values=False)
    task_id:    str
    context_id: str
    agent_id:   str
    from_state: TaskState | None
    to_state:   TaskState
    timestamp:  datetime
    metadata:   dict[str, Any] = {}

def emit_transition(lf_trace, event: A2ATaskEvent) -> None:
    payload = event.model_dump(mode="json")
    lf_trace.event(name=f"task.{event.to_state}", input=payload)

async def transition(
    store: TaskStore,
    task_id: str,
    to_state: TaskState,
    lf_trace,
    agent_id: str,
    metadata: dict[str, Any] | None = None,
) -> Task:
    task = await store.get_task(task_id)
    if task.status.state in TERMINAL:
        return task
    from_state = task.status.state
    task.status = TaskStatus(state=to_state)
    await store.save_task(task)
    emit_transition(lf_trace, A2ATaskEvent(
        task_id=task.id,
        context_id=task.contextId,
        agent_id=agent_id,
        from_state=from_state,
        to_state=to_state,
        timestamp=datetime.now(timezone.utc),
        metadata=metadata or {},
    ))
    return task
\end{lstlisting}

Three decisions in Listing~\ref{lst:a2a-lifecycle} deserve explanation.
\code{TaskState} is used directly as a string literal type instead of
being wrapped in a Python \code{Enum}; that keeps the wire format and the
in-memory representation identical, which is exactly what a protocol type
should do. The \code{TaskStore} is expressed as a \code{Protocol} rather
than a concrete class because both \code{InMemoryTaskStore} for
development and a persistent store from Chapter~22~\S22.3 can satisfy the
same contract without sharing an implementation. The write happens before
\code{emit\_transition()}; this is the chapter's core invariant, because
a failed Langfuse event must never roll back the task's actual state.

Pause states need one additional convention: they must carry a structured
prompt in \code{TaskStatus.message} so the caller knows what information is
missing. The next section addresses where the caller discovers that such a
pause is even possible, and which skills on the remote agent can trigger
it.


%% ------------------------------------------------------------
\section{An Agent's Public Identity Is a Contract, Not Documentation}
\label{sec:a2a-agent-card}
%% ------------------------------------------------------------

The design problem is discovery without validation: a caller that learns
about a remote agent from prose documentation has no machine-checkable
proof that the document is current, complete, or even written for the same
version of the service. The naive approach is to hard-code the remote
agent's URL and payload shape in the caller, which works exactly until the
remote team changes either one. The governing principle is that identity
must be served as a typed contract: \code{AgentCard} is the single public
document a caller trusts before it sends its first task.

The well-known path matters as much as the document shape.
\code{/.well-known/agent.json} is intentionally unauthenticated: the
caller must be able to fetch the card before it knows what security scheme
the task endpoint expects. This is not a convenience exception; it is the
security boundary that prevents a bootstrap deadlock. The task endpoint
itself remains protected, and the card advertises that protection through
\code{securitySchemes} and \code{security}, which follow the OpenAPI
Security Scheme Object pattern rather than the older
\code{authentication: dict} style that no longer matches the spec.

\code{defaultInputModes} and \code{defaultOutputModes} are required, even
if every skill on the agent also sets its own mode overrides. They define
the agent-level defaults that a generic client can assume before it reads
individual skills. \code{AgentSkill.id} must be unique within one agent,
because the delegation token in Section~\ref{sec:a2a-delegation} scopes
authority to skill IDs rather than to route names. \code{AgentSkill.tags}
is also required; Chapter~18~\S18.4 uses these tags as discovery facets
when a registry needs to find a suitable remote agent by capability rather
than by hard-coded URL.

\begin{warningbox}
Do not protect \code{/.well-known/agent.json} with the same Bearer-token
dependency as \code{/a2a}. Doing so turns the card into a secret rather
than an identity document and makes generic client resolution impossible.
\end{warningbox}

\begin{lstlisting}[language=Python,
  caption={Spec-compliant \texttt{AgentCard} factory and route split},
  label={lst:a2a-agent-card}]
# chapter17/agent_card.py
from a2a.types import (
    AgentCard, AgentCapabilities, AgentSkill, SecurityScheme,
)
from fastapi import APIRouter, Depends
from pydantic import ConfigDict

router = APIRouter()

summarize_skill = AgentSkill(
    id="summarize-text-v1",
    name="Text Summarizer",
    description="Summarise text to a target length.",
    tags=["nlp", "summarization"],
    inputModes=["text/plain"],
    outputModes=["text/plain"],
)

def build_summarizer_card(base_url: str) -> AgentCard:
    return AgentCard(
        name="Summarizer Agent",
        description="Condenses documents for downstream processing.",
        url=f"{base_url}/a2a",
        version="1.0.0",
        capabilities=AgentCapabilities(
            streaming=True,
            pushNotifications=False,
        ),
        securitySchemes={
            "bearerAuth": SecurityScheme(
                type="http", scheme="bearer"
            )
        },
        security=[{"bearerAuth": []}],
        defaultInputModes=["text/plain", "application/json"],
        defaultOutputModes=["text/plain"],
        skills=[summarize_skill],
    )

@router.get("/.well-known/agent.json", response_model=AgentCard)
async def get_agent_card() -> AgentCard:
    return build_summarizer_card("http://localhost:8001")

@router.post("/a2a")
async def a2a_endpoint(token: str = Depends(verify_bearer)):
    ...
\end{lstlisting}

Two decisions in Listing~\ref{lst:a2a-agent-card} deserve explanation.
\code{securitySchemes} and \code{security} are included even in this small
example because leaving them out would hide the core architectural point of
the route split: the card is public, but it still describes a protected
task endpoint. The skill ID uses kebab-case with a version suffix
(\code{"summarize-text-v1"}) rather than a Pythonic identifier such as
\code{"summarize\_text"} because this value is not code --- it is public
protocol, and versioned protocol names should remain stable even if the
server-side function name changes.

A valid card tells the caller what the remote agent claims to support. The
next problem is turning that claim into a real network session without
reimplementing the transport or the SSE protocol by hand.


%% ------------------------------------------------------------
\section{The SDK Owns the Wire; Your Code Owns the Logic}
\label{sec:a2a-sdk}
%% ------------------------------------------------------------

The design problem is protocol reimplementation: if every team builds its
own \code{httpx} wrapper around A2A, the organisation ends up with several
slightly different interpretations of the same protocol and no reliable way
to reason about failure. The naive approach is to write the transport
layer directly because it appears to save a dependency, but in practice it
turns your application code into a part-time SDK maintainer. The governing
principle is separation of concerns: \code{A2AClient} and
\code{A2AStarletteApplication} own the wire protocol, and the application
owns only task logic.

The caller side starts with \code{A2ACardResolver}. This matters because
it validates the remote \code{AgentCard} before the client is even
constructed, which means the caller can fail fast if the card is missing
required fields or claims to support streaming when it does not. Skipping
resolution and building a client from a URL string discards the only
pre-flight validation step in the protocol.

The server side centers on \code{TaskHandler.on\_message()}. That method
does not write raw SSE frames or manage sockets; it places typed events
into an \code{EventQueue}, and the SDK serialises them to the client.
There are only two event families the caller must handle in the streaming
loop: \code{TaskStatusUpdateEvent} for lifecycle changes and
\code{TaskArtifactUpdateEvent} for output chunks. Treating artifacts as
status or status as output is not a cosmetic mistake; it loses either
progress semantics or data.

\begin{lstlisting}[language=Python,
  caption={Server-side \texttt{TaskHandler} with typed status and artifact events},
  label={lst:a2a-sdk-server}]
# chapter17/a2a_server_base.py
from uuid import uuid4
from a2a.server.apps import A2AStarletteApplication
from a2a.server.events import EventQueue
from a2a.server.request_handlers import DefaultRequestHandler
from a2a.server.tasks import InMemoryTaskStore, TaskHandler
from a2a.types import (
    Artifact, TaskArtifactUpdateEvent,
    TaskStatus, TaskStatusUpdateEvent, TextPart,
)
from langchain_litellm import ChatLiteLLM

llm = ChatLiteLLM(model="anthropic/claude-sonnet-4-8")

class SummarizerHandler(TaskHandler):
    async def on_message(self, context, event_queue: EventQueue) -> None:
        await event_queue.put(TaskStatusUpdateEvent(
            status=TaskStatus(state="submitted"), final=False,
        ))
        await event_queue.put(TaskStatusUpdateEvent(
            status=TaskStatus(state="working"), final=False,
        ))
        text = context.message.parts[0].text
        summary = llm.invoke(
            f"Summarise in <= 50 words:\n{text}"
        ).content
        await event_queue.put(TaskArtifactUpdateEvent(
            artifact=Artifact(
                artifactId=uuid4().hex,
                name="summary",
                parts=[TextPart(kind="text", text=summary)],
            )
        ))
        await event_queue.put(TaskStatusUpdateEvent(
            status=TaskStatus(state="completed"), final=True,
        ))

app = A2AStarletteApplication(
    agent_card=build_summarizer_card("http://localhost:8001"),
    http_handler=DefaultRequestHandler(
        agent_executor=SummarizerHandler(),
        task_store=InMemoryTaskStore(),
    ),
).build()
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={Client-side streaming loop with typed event dispatch},
  label={lst:a2a-sdk-client}]
# chapter17/a2a_server_base.py  (client section)
from uuid import uuid4
import httpx
from a2a.client import A2AClient, A2ACardResolver
from a2a.types import (
    Message, Role, SendMessageRequest, TextPart,
    TaskArtifactUpdateEvent, TaskStatusUpdateEvent,
)

async def stream_summary(text: str) -> str:
    async with httpx.AsyncClient() as http:
        resolver = A2ACardResolver(http)
        card = await resolver.get_agent_card(
            "http://localhost:8001/.well-known/agent.json"
        )
        client = A2AClient(http, card)
        request = SendMessageRequest(
            message=Message(
                messageId=uuid4().hex,
                kind="message",
                role=Role.user,
                parts=[TextPart(kind="text", text=text)],
            )
        )
        chunks: list[str] = []
        async for event in client.send_message_streaming(request):
            match event:
                case TaskArtifactUpdateEvent():
                    chunks.extend(
                        part.text for part in event.artifact.parts
                    )
                case TaskStatusUpdateEvent():
                    if event.status.state == "failed":
                        raise RuntimeError("remote task failed")
        return "".join(chunks)
\end{lstlisting}

Three decisions across Listings~\ref{lst:a2a-sdk-server}
and~\ref{lst:a2a-sdk-client} deserve explanation.
The server emits \code{final=True} only on the terminal status event
because stream closure is a lifecycle fact, not an artifact fact; closing
the stream on an artifact would make partial-output workflows impossible.
The client uses pattern matching on event types rather than trying to
infer semantics from fields, which keeps the dispatch logic aligned with
the protocol's type system instead of with fragile string inspection.
\code{A2ACardResolver} and \code{A2AClient} share the same
\code{httpx.AsyncClient}, which means card resolution and task streaming
reuse the same connection pool rather than opening two independent network
stacks to the same host.

\begin{notebox}
\code{A2AClient.send\_message()} is appropriate when the caller wants only
the final task result. \code{send\_message\_streaming()} is required when
the caller must surface progress, pause on \code{"input-required"}, or
retain partial artifacts from a task that eventually fails.
\end{notebox}

The SDK solves transport and typing, but it does not answer a more subtle
question: how a parent agent proves it may delegate a specific skill to a
specific child without forwarding all of its own authority.


%% ------------------------------------------------------------
\section{Delegated Authority Must Be Narrower Than the Source}
\label{sec:a2a-delegation}
%% ------------------------------------------------------------

The design problem is privilege escalation by convenience: if the parent
agent simply forwards its own Bearer token to the child, the child gains
every right the parent possesses for the full lifetime of that token. The
naive approach is attractive because it requires no additional crypto or
token plumbing, but it turns delegation into impersonation. The governing
principle is least authority: every child call receives a short-lived token
whose claims are bound to one audience, one parent task, and one or more
explicit \code{AgentSkill.id} values.

RS256 is a good fit here because signature verification and signing
authority are separated. The parent keeps the RSA private key; the child
receives only the signed token and verifies it against the parent's public
JWKS. This prevents key material from crossing service boundaries and
allows rotation via the JWT \code{kid} header. The \code{scope} claim is a
space-separated list of skill IDs, not route names or arbitrary labels,
because the agent card already defines skill IDs as the public capability
surface. A five-minute TTL is short enough to make replay expensive and
long enough to survive a normal subagent call.

Enforcement belongs at the HTTP boundary, before any task state is created.
A FastAPI dependency that validates the token and checks
\code{required\_scope in claims.scope.split()} can reject unauthorised
requests with \code{403} before the request ever reaches
\code{TaskHandler.on\_message()}. That avoids creating ghost tasks whose
only lifecycle event would have been a rejected credential.

\begin{warningbox}
Do not validate delegation scope inside \code{TaskHandler.on\_message()}.
By the time that method runs, the request has already crossed the
application boundary and may already have caused task allocation or metric
increments that should never exist for an unauthorised caller.
\end{warningbox}

\begin{lstlisting}[language=Python,
  caption={Scoped delegation token: mint, validate, inject},
  label={lst:a2a-delegation}]
# chapter17/delegation_tokens.py
from typing import Generator
import httpx
from jose import JWTError, jwt
from jose.constants import ALGORITHMS
from pydantic import BaseModel

class DelegationTokenClaims(BaseModel):
    sub:            str
    aud:            str
    scope:          str
    parent_task_id: str
    exp:            int
    iat:            int
    iss:            str

def mint_delegation_token(
    claims: DelegationTokenClaims,
    private_key: str,
    kid: str,
) -> str:
    return jwt.encode(
        claims.model_dump(),
        private_key,
        algorithm=ALGORITHMS.RS256,
        headers={"kid": kid},
    )

def validate_delegation_token(
    token: str,
    jwks_url: str,
    expected_aud: str,
    required_scope: str,
) -> DelegationTokenClaims:
    header = jwt.get_unverified_header(token)
    jwks = httpx.get(jwks_url, timeout=5).json()
    key = next(k for k in jwks["keys"] if k["kid"] == header["kid"])
    claims = jwt.decode(
        token, key, algorithms=[ALGORITHMS.RS256],
        audience=expected_aud,
    )
    if required_scope not in claims["scope"].split():
        raise JWTError("delegation scope missing")
    return DelegationTokenClaims(**claims)

class ScopedBearerAuth(httpx.Auth):
    def __init__(self, token: str) -> None:
        self._token = token
    def auth_flow(
        self, request: httpx.Request,
    ) -> Generator[httpx.Request, httpx.Response, None]:
        request.headers["Authorization"] = f"Bearer {self._token}"
        yield request
\end{lstlisting}

Three decisions in Listing~\ref{lst:a2a-delegation} deserve explanation.
The \code{aud} claim is the child agent's public URL rather than its name,
because audience is a property of where the token may be presented, not of
who minted it. \code{validate\_delegation\_token()} checks scope after
signature verification and audience validation, which preserves the normal
JWT trust chain: claim-level authorisation only matters once the token has
already been proven authentic and intended for this service.
\code{ScopedBearerAuth} uses \code{httpx.Auth} instead of manual headers so
every request emitted by the client --- including internal retries or
redirects handled by \code{httpx} --- carries the same credential path.

Delegated identity narrows authority, but it does not decide what to do
when the remote side fails. The next section classifies those failures and
connects them to LangGraph routing.


%% ------------------------------------------------------------
\section{Not All Failures Deserve a Retry}
\label{sec:a2a-errors}
%% ------------------------------------------------------------

The design problem is undifferentiated error handling: a caller that
treats every subagent failure as retryable converts permanent request bugs
into delayed permanent request bugs. The naive approach is a blanket retry
decorator around the A2A call, which makes transient failures look robust
while obscuring the difference between bad input and bad infrastructure.
The governing principle is failure typing before recovery: the caller must
inspect the SDK error, classify it as retryable or terminal, preserve any
partial artifacts, and route explicitly to retry or escalation.

\code{A2AError} is the SDK base class, while \code{ServerError} carries a
JSON-RPC \code{code} and \code{message}. The mapping is operational rather
than theoretical. \code{-32600}, \code{-32601}, and \code{-32602} are
request-shape errors and therefore non-retryable. \code{-32603} and
\code{-32000} represent internal or server-side failure and are often
transient, so they can justify bounded retry. Partial artifacts from a
failed task are still valuable; they may contain a summary generated before
the translation step failed, or the first half of a streamed answer that a
human can still inspect. Losing them makes escalation harder, not cleaner.

The routing contract in this section relies on \code{next} as a state
field. On success, a node writes the name of the next business node. On
failure, it writes the name of the node that failed, so the error handler
knows which operation would be retried if the error proves transient. This
keeps retry count inside graph state rather than in a decorator's hidden
closure, which is why the section uses \code{AsyncRetrying} conceptually
but records retry budget explicitly in the graph state.

\begin{lstlisting}[language=Python,
  caption={Error classification with preserved artifacts and explicit routing},
  label={lst:a2a-errors}]
# chapter17/error_propagation.py
from typing import Any, TypedDict
from a2a.client import A2AClient
from a2a.errors import A2AError, ServerError
from a2a.types import (
    SendMessageRequest,
    TaskArtifactUpdateEvent,
    TaskStatusUpdateEvent,
)
from langgraph.graph import END
from langgraph.types import Command

RETRYABLE_CODES: frozenset[int] = frozenset({-32603, -32000})
MAX_RETRIES: int = 3

class OrchestratorState(TypedDict):
    input_text:        str
    summary:           str
    translation:       str
    last_error:        A2AError | None
    retry_count:       int
    partial_artifacts: list[dict[str, Any]]
    next:              str

async def safe_delegate(
    client: A2AClient,
    request: SendMessageRequest,
    state: OrchestratorState,
) -> OrchestratorState:
    artifacts: list[dict[str, Any]] = []
    try:
        async for event in client.send_message_streaming(request):
            match event:
                case TaskArtifactUpdateEvent():
                    artifacts.append(event.artifact.model_dump())
                case TaskStatusUpdateEvent():
                    if event.status.state == "failed":
                        raise ServerError(-32603, "subagent failed")
    except A2AError as exc:
        return {**state, "last_error": exc,
                "partial_artifacts": artifacts}
    return {**state, "last_error": None,
            "partial_artifacts": artifacts}

def handle_subagent_error(state: OrchestratorState) -> OrchestratorState:
    err = state["last_error"]
    retryable = (
        isinstance(err, ServerError)
        and err.code in RETRYABLE_CODES
        and state["retry_count"] < MAX_RETRIES
    )
    return {
        **state,
        "retry_count": state["retry_count"] + (1 if retryable else 0),
        "next": state["next"] if retryable else "ESCALATE",
    }

def escalate_node(state: OrchestratorState) -> Command:
    return Command(goto=END)
\end{lstlisting}

Three decisions in Listing~\ref{lst:a2a-errors} deserve explanation.
\code{safe\_delegate()} collects artifacts on the success path and the
failure path alike; this keeps the node interface simple, and the caller
can decide later whether those artifacts are complete output or merely
partial salvage. \code{handle\_subagent\_error()} does not itself choose a
business node such as \code{"summarize"} or \code{"translate"}; it trusts
the failing node to have already written its own identity into
\code{state["next"]}, which preserves locality of responsibility. The
escalation path returns \code{Command(goto=END)} instead of raising an
exception because escalation is a valid terminal outcome, not an unhandled
error in the graph runtime.

\begin{notebox}
Chapter~1~\S1.5 introduced \code{ReflectNode} as the place where an agent
concludes it should stop autonomous action and hand control to a more
authoritative decision maker. The \code{ESCALATE} branch here is the
protocol-level equivalent: the graph stops cleanly, but it preserves the
error object and artifacts needed for a human or supervisor to continue.
\end{notebox}

The individual pieces are now correct in isolation. The remaining problem
is composition: how to wire two subagents and one parent graph together
without collapsing the boundaries the chapter has spent five sections
establishing.


%% ------------------------------------------------------------
\section{Integration Must Preserve Boundaries}
\label{sec:a2a-integration}
%% ------------------------------------------------------------

The design problem is boundary erosion during composition: a system whose
parts are individually well designed can still fail once integration logic
starts copying credentials, parsing wire events in the wrong layer, or
smearing business state across transport code. The naive approach is to
build the capstone as a monolithic script and then pull pieces apart
later, which guarantees the first working version has the weakest
architecture. The governing principle is that the capstone should make
boundaries more visible, not less: each subagent stays a standalone A2A
server, and the parent orchestrator remains a LangGraph graph that talks
only through the SDK.

The translator subagent demonstrates why pause states matter. If the
source language is missing, translation is not failed work; it is
incomplete work whose next action belongs to the caller. The correct
behaviour is therefore \code{"input-required"} with a structured
\code{DataPart} prompt, not \code{"failed"}. In a real system this would
often route to a human approval or clarification step; Chapter~21~\S21.2
introduces \code{interrupt()} as the LangGraph mechanism for inserting
that pause before autonomous work continues.

\begin{lstlisting}[language=Python,
  caption={Subagent A: summarizer server},
  label={lst:capstone-summarizer}]
# chapter17/subagent_summarizer.py
import uvicorn
from uuid import uuid4
from a2a.server.apps import A2AStarletteApplication
from a2a.server.events import EventQueue
from a2a.server.request_handlers import DefaultRequestHandler
from a2a.server.tasks import InMemoryTaskStore, TaskHandler
from a2a.types import (
    AgentCard, AgentCapabilities, AgentSkill,
    Artifact, TaskArtifactUpdateEvent,
    TaskStatus, TaskStatusUpdateEvent, TextPart,
)
from langchain_litellm import ChatLiteLLM

llm = ChatLiteLLM(model="anthropic/claude-sonnet-4-8")

class SummarizerHandler(TaskHandler):
    async def on_message(self, ctx, q: EventQueue) -> None:
        for state in ("submitted", "working"):
            await q.put(TaskStatusUpdateEvent(
                status=TaskStatus(state=state), final=False))
        text = ctx.message.parts[0].text
        summary = llm.invoke(f"Summarise <= 50 words:\n{text}").content
        await q.put(TaskArtifactUpdateEvent(artifact=Artifact(
            artifactId=uuid4().hex,
            name="summary",
            parts=[TextPart(kind="text", text=summary)],
        )))
        await q.put(TaskStatusUpdateEvent(
            status=TaskStatus(state="completed"), final=True))

card = AgentCard(
    name="Summarizer Agent",
    description="Summarises raw text.",
    url="http://localhost:8001/a2a",
    version="1.0.0",
    capabilities=AgentCapabilities(streaming=True),
    defaultInputModes=["text/plain"],
    defaultOutputModes=["text/plain"],
    skills=[AgentSkill(id="summarize-text-v1",
        name="Text Summarizer",
        description="Summarise text to <= 50 words.",
        tags=["nlp", "summarization"])],
)
app = A2AStarletteApplication(
    agent_card=card,
    http_handler=DefaultRequestHandler(
        agent_executor=SummarizerHandler(),
        task_store=InMemoryTaskStore(),
    ),
).build()

if __name__ == "__main__":
    uvicorn.run("subagent_summarizer:app", host="0.0.0.0", port=8001)
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={Subagent B: translator with \texttt{input-required} pause},
  label={lst:capstone-translator}]
# chapter17/subagent_translator.py
import uvicorn
from uuid import uuid4
from a2a.server.apps import A2AStarletteApplication
from a2a.server.events import EventQueue
from a2a.server.request_handlers import DefaultRequestHandler
from a2a.server.tasks import InMemoryTaskStore, TaskHandler
from a2a.types import (
    AgentCard, AgentCapabilities, AgentSkill,
    Artifact, DataPart, Message, Role,
    TaskArtifactUpdateEvent, TaskStatus,
    TaskStatusUpdateEvent, TextPart,
)
from langchain_litellm import ChatLiteLLM

llm = ChatLiteLLM(model="anthropic/claude-sonnet-4-8")

class TranslatorHandler(TaskHandler):
    async def on_message(self, ctx, q: EventQueue) -> None:
        src = next((p.data.get("source_language")
                    for p in ctx.message.parts if hasattr(p, "data")), None)
        if not src:
            await q.put(TaskStatusUpdateEvent(
                status=TaskStatus(
                    state="input-required",
                    message=Message(
                        messageId=uuid4().hex,
                        kind="message",
                        role=Role.agent,
                        parts=[DataPart(kind="data",
                            data={"required_field": "source_language"})],
                    ),
                ),
                final=False,
            ))
            return
        await q.put(TaskStatusUpdateEvent(
            status=TaskStatus(state="working"), final=False))
        text = next(p.text for p in ctx.message.parts if hasattr(p, "text"))
        out = llm.invoke(f"Translate from {src} to English:\n{text}").content
        await q.put(TaskArtifactUpdateEvent(artifact=Artifact(
            artifactId=uuid4().hex,
            name="translation",
            parts=[TextPart(kind="text", text=out)],
        )))
        await q.put(TaskStatusUpdateEvent(
            status=TaskStatus(state="completed"), final=True))
\end{lstlisting}

Two decisions across Listings~\ref{lst:capstone-summarizer}
and~\ref{lst:capstone-translator} deserve explanation.
The translator emits \code{"input-required"} instead of guessing the source
language because guessing is a model convenience, not a protocol right;
the whole point of the pause state is to preserve correctness when caller
input is missing. Both subagents keep their \code{AgentCard} definition
inside the server file rather than centralising cards in the orchestrator,
because the published identity belongs to the service that owns the skill,
not to the caller that consumes it.

\begin{lstlisting}[language=Python,
  caption={Parent orchestrator: request builders and delegation nodes},
  label={lst:capstone-orchestrator-a}]
# chapter17/orchestrator.py  (part 1)
from uuid import uuid4
import os, time, httpx
from typing import Any
from a2a.client import A2AClient, A2ACardResolver
from a2a.types import (
    DataPart, Message, Role, SendMessageRequest, TextPart,
)
from langfuse import Langfuse
from chapter17.delegation_tokens import (
    DelegationTokenClaims, ScopedBearerAuth, mint_delegation_token,
)
from chapter17.error_propagation import OrchestratorState, safe_delegate

PRIVATE_KEY = os.environ["PARENT_RSA_PRIVATE_KEY"]
KID = os.environ["PARENT_KEY_ID"]
lf = Langfuse()

def build_request(text: str, extra: dict[str, Any] | None = None):
    parts = [TextPart(kind="text", text=text)]
    if extra:
        parts.append(DataPart(kind="data", data=extra))
    return SendMessageRequest(message=Message(
        messageId=uuid4().hex,
        kind="message",
        role=Role.user,
        parts=parts,
    ))

def artifact_text(artifacts: list[dict[str, Any]]) -> str:
    return "".join(
        part["text"]
        for art in artifacts
        for part in art.get("parts", [])
        if part.get("kind") == "text"
    )

async def summarize_node(state: OrchestratorState) -> OrchestratorState:
    now = int(time.time())
    token = mint_delegation_token(DelegationTokenClaims(
        sub="Summarizer Agent",
        aud="http://localhost:8001/a2a",
        scope="summarize-text-v1",
        parent_task_id=uuid4().hex,
        exp=now + 300, iat=now, iss="http://orchestrator:8000",
    ), PRIVATE_KEY, KID)
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={Parent orchestrator: delegation nodes and graph wiring},
  label={lst:capstone-orchestrator-b}]
# chapter17/orchestrator.py  (part 2)
    async with httpx.AsyncClient(auth=ScopedBearerAuth(token)) as http:
        card = await A2ACardResolver(http).get_agent_card(
            "http://localhost:8001/.well-known/agent.json")
        client = A2AClient(http, card)
        span = lf.span(name="summarize_delegate")
        updated = await safe_delegate(
            client, build_request(state["input_text"]),
            {**state, "next": "summarize"},
        )
        span.end(output={"artifacts": len(updated["partial_artifacts"])})
    if updated["last_error"] is not None:
        return updated
    return {**updated, "summary": artifact_text(
        updated["partial_artifacts"]), "partial_artifacts": [],
        "next": "translate"}

async def translate_node(state: OrchestratorState) -> OrchestratorState:
    now = int(time.time())
    token = mint_delegation_token(DelegationTokenClaims(
        sub="Translator Agent",
        aud="http://localhost:8002/a2a",
        scope="translate-text-v1",
        parent_task_id=uuid4().hex,
        exp=now + 300, iat=now, iss="http://orchestrator:8000",
    ), PRIVATE_KEY, KID)
    async with httpx.AsyncClient(auth=ScopedBearerAuth(token)) as http:
        card = await A2ACardResolver(http).get_agent_card(
            "http://localhost:8002/.well-known/agent.json")
        client = A2AClient(http, card)
        updated = await safe_delegate(
            client, build_request(state["summary"],
            {"source_language": "French"}),
            {**state, "next": "translate"},
        )
    if updated["last_error"] is not None:
        return updated
    return {**updated, "translation": artifact_text(
        updated["partial_artifacts"]), "partial_artifacts": [],
        "next": "__end__"}
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={Graph routing for success, retry, and escalation},
  label={lst:capstone-orchestrator-c}]
# chapter17/orchestrator.py  (part 3)
from langgraph.graph import END, START, StateGraph
from chapter17.error_propagation import (
    handle_subagent_error, escalate_node,
)

def route_after_summarize(state: OrchestratorState) -> str:
    return "handle_error" if state["last_error"] else "translate"

def route_after_translate(state: OrchestratorState) -> str:
    return "handle_error" if state["last_error"] else "__end__"

builder = StateGraph(OrchestratorState)
builder.add_node("summarize", summarize_node)
builder.add_node("translate", translate_node)
builder.add_node("handle_error", handle_subagent_error)
builder.add_node("escalate", escalate_node)
builder.add_edge(START, "summarize")
builder.add_conditional_edges("summarize", route_after_summarize,
    {"translate": "translate", "handle_error": "handle_error"})
builder.add_conditional_edges("translate", route_after_translate,
    {"__end__": END, "handle_error": "handle_error"})
builder.add_conditional_edges("handle_error", lambda s: s["next"], {
    "summarize": "summarize",
    "translate": "translate",
    "ESCALATE": "escalate",
})
graph = builder.compile()
\end{lstlisting}

Three decisions across Listings~\ref{lst:capstone-orchestrator-a},
\ref{lst:capstone-orchestrator-b}, and
\ref{lst:capstone-orchestrator-c} deserve explanation.
The parent mints one token per subagent call rather than per process start,
which binds delegated authority to a single task boundary and prevents
token reuse across unrelated work. Each node writes its own identity into
\code{state["next"]} before failure handling occurs, which lets the shared
\code{handle\_subagent\_error()} function decide whether to retry the same
node or route to \code{ESCALATE} without knowing which business operation
failed. The graph uses conditional edges after both delegation nodes
instead of a single global error node attached at the end, because failure
classification is meaningful only while the failing node's context is still
present in state.

The three-file system in this section is intentionally small, but the
boundaries are already the production boundaries: each subagent is its own
A2A service, the parent remains a LangGraph orchestrator, and delegation
tokens are minted at the moment authority crosses that line. Chapter~18
takes the next step outward by replacing the known-URL assumption in this
chapter with registry-backed, decentralised agent discovery and handshake.
