%% ============================================================
%% Chapter 26 — Deployment, Serving & Configuration
%% Production-Grade Agentic AI: LangGraph + Python Frameworks
%% ============================================================


\chapterquote{The only thing worse than a bad prompt in production
is the inability to roll it back.}{anon}


A LangGraph graph that works perfectly on a laptop will fail in production
the moment two requests arrive simultaneously, a pod restarts mid-run, an
engineer pushes a bad prompt at 2~a.m., or the model vendor changes a
pricing tier. These are not theoretical risks: teams running agentic
workloads consistently report that the gap between a working local
prototype and a stable production service is wider for AI agents than for
conventional APIs, precisely because an agent's correctness depends on
external state---LLM weights, prompt text, checkpointer data---that changes
independently of the code. The deployment layer exists to close that gap by
making the agent's environment explicit, versioned, and observable.

The solution in this chapter is organised into four layers. The
\emph{serving layer} translates graph invocations into HTTP semantics,
distinguishing synchronous responses from streaming ones. The
\emph{infrastructure layer} packages the serving layer into a reproducible,
non-root container and operates it on Kubernetes with health-aware traffic
management. The \emph{configuration layer} externalises every value that
changes between environments so that the same image runs in staging and
production without modification. The \emph{experiment layer} makes model
and prompt choices runtime decisions rather than code changes, enabling
zero-downtime rollouts and controlled A/B comparisons.

The sections are ordered by dependency: you cannot containerise code you
have not written (\S\ref{sec:serving-contract} precedes
\S\ref{sec:infra-layer}), you cannot externalise secrets you have not
identified (\S\ref{sec:config-must-not-live} follows the serving and
infrastructure sections where those secrets first appear), and you cannot
roll out a new prompt atomically without both a config layer and a running
service (\S\ref{sec:zero-downtime-rollout} follows~\S\ref{sec:config-must-not-live}).
The chapter then pivots to the LangGraph Platform managed path
(\S\ref{sec:platform-json}--\S\ref{sec:ab-routing}), which replaces the OSS
infrastructure layer with a single configuration file but uses the same
configuration and experiment layers.

\begin{table}[h]
\centering
\caption{Domain terms introduced in this chapter}
\label{tab:ch26-terms}
\begin{tabular}{lp{9.5cm}}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
\emph{Server-Sent Events (SSE)} &
  A unidirectional HTTP streaming protocol in which the server pushes
  newline-delimited \texttt{data:} frames to a client over a single
  long-lived connection; distinguished from WebSockets by being
  half-duplex and HTTP/1.1-compatible. \\
\emph{liveness probe} &
  A Kubernetes health check that determines whether a container's process is
  alive; failure causes the kubelet to restart the container. \\
\emph{readiness probe} &
  A Kubernetes health check that determines whether a container is ready to
  accept traffic; failure removes the pod from the Service's endpoint list
  without restarting it. \\
\emph{multi-stage build} &
  A Docker build pattern in which intermediate stages (e.g.\ dependency
  installation) are discarded and only their outputs are copied into the
  final image, minimising image size and attack surface. \\
\emph{configurable dict} &
  The \texttt{RunnableConfig["configurable"]} sub-dictionary that LangGraph
  threads through every node call; the canonical channel for per-request
  runtime parameters that do not belong in agent state. \\
\emph{double-texting} &
  The condition in which a client sends a second request to the same thread
  before the first run has completed; the Platform's
  \texttt{multitask\_strategy} field specifies how to handle it. \\
\emph{LiteLLM Router} &
  A proxy object from the \texttt{litellm} library that presents a unified
  OpenAI-compatible completion API in front of multiple model deployments,
  with configurable routing strategies, retries, and cooldown. \\
\bottomrule
\end{tabular}
\end{table}

\medskip
\noindent\textbf{Source layout for this chapter:}

\begin{lstlisting}[language=bash, caption={Chapter 26 source tree},
  label={lst:ch26-tree}]
src/
  serve/
    app.py          # FastAPI application (S26.1, S26.4)
  config/
    settings.py     # Pydantic Settings (S26.3)
  graph/
    factory.py      # graph builder + prompt pull (S26.4)
    nodes.py        # LLM node with variant routing (S26.6)
  llm/
    router.py       # LiteLLM Router (S26.6)
  platform/
    cron_setup.py   # Platform cron + webhook helpers (S26.5)
Dockerfile          # multi-stage build (S26.2)
k8s/
  deployment.yaml   # Kubernetes manifest (S26.2)
langgraph.json      # Platform config (S26.5)
.env.example        # documented environment template
\end{lstlisting}

%%------------------------------------------------------------
%% PART A — OSS Runtime (Self-Hosted)
%%------------------------------------------------------------
\noindent\rule{\linewidth}{0.8pt}
\begin{center}
  \textbf{Part A\quad OSS Runtime\quad FastAPI\,\textperiodcentered\,Docker\,\textperiodcentered\,Kubernetes\,\textperiodcentered\,Config}
\end{center}
\noindent\rule{\linewidth}{0.8pt}

%% -----------------------------------------------------------
\section{An Agent Needs Two Distinct HTTP Contracts}
\label{sec:serving-contract}

The design problem here is that a LangGraph graph produces two qualitatively
different outputs depending on the caller's needs: a complete, atomic result
when the caller wants to process the answer as a unit, and a stream of
incremental token events when the caller is a chat UI that must update
character-by-character to feel responsive. Serving both from the same
endpoint is tempting---return streaming always and let the client
buffer---but it conflates transport with semantics and breaks every
JSON-consuming downstream consumer that expects a single response body.
The principle is that synchronous invocations and streaming invocations are
different contracts and must be expressed as different endpoints.

FastAPI's default \code{JSONResponse} handles the synchronous case directly:
the graph runs to completion via \code{ainvoke}, and the result is serialised
once. The streaming case requires \emph{Server-Sent Events (SSE)}, a protocol
in which the server sends a sequence of \texttt{data: \{json\}\textbackslash
n\textbackslash n} frames over a long-lived HTTP connection. SSE is
preferable to WebSockets for agent token streaming because it is
half-duplex (the client never needs to send mid-stream), is automatically
reconnected by browsers, and traverses load balancers without special
configuration. The \code{sse-starlette} library provides
\code{EventSourceResponse}, which wraps any async generator and handles
frame formatting, keep-alive pings, and connection teardown.

For token-by-token streaming, LangGraph's \code{astream} method accepts a
\code{stream\_mode} parameter. Passing \code{stream\_mode="messages"} causes
the graph to yield \code{(AIMessageChunk, metadata)} tuples as each token
arrives from the LLM, rather than waiting for the node to finish. This is the
correct and officially supported mechanism for token streaming from a
\code{CompiledStateGraph}; it does not require LangChain's LCEL
\code{astream\_events} pipeline protocol, which is designed for chain
composition rather than graph execution.

The graph is expensive to construct---it compiles the state machine, wires
the checkpointer, and in \S\ref{sec:zero-downtime-rollout} will also pull a
remote prompt. Constructing it inside a request handler would add hundreds of
milliseconds to every request and spawn redundant network calls. FastAPI's
\emph{lifespan} context manager, introduced in FastAPI~0.93 and now the
canonical alternative to deprecated \code{on\_event} hooks, solves this: the
graph is built once at startup and stored on \code{app.state} where every
request handler can reach it without a global variable.

\begin{lstlisting}[language=Python,
  caption={FastAPI lifespan, \texttt{/invoke}, \texttt{/stream}, and
    \texttt{/health} --- \texttt{src/serve/app.py}},
  label={lst:fastapi-app}]
from contextlib import asynccontextmanager
from typing import Any, AsyncGenerator, Literal
import threading
from datetime import datetime, timezone

from fastapi import FastAPI
from fastapi.responses import JSONResponse
from sse_starlette.sse import EventSourceResponse
from pydantic import BaseModel
from langchain_core.runnables import RunnableConfig

from src.config.settings import get_settings
from src.graph.factory import build_graph

_cycle_count: int = 0
_count_lock = threading.Lock()

class InvokeRequest(BaseModel):
    messages: list[dict[str, str]]
    thread_id: str
    configurable: dict[str, Any] = {}

@asynccontextmanager
async def lifespan(app: FastAPI) -> AsyncGenerator[None, None]:
    settings = get_settings()
    app.state.graph, app.state.prompt_meta = build_graph(settings)
    yield

app = FastAPI(lifespan=lifespan)

@app.post("/invoke")
async def invoke(req: InvokeRequest) -> JSONResponse:
    global _cycle_count
    cfg: RunnableConfig = {
        "configurable": {"thread_id": req.thread_id, **req.configurable}
    }
    result = await app.state.graph.ainvoke(
        {"messages": req.messages}, config=cfg
    )
    with _count_lock:
        _cycle_count += 1
    return JSONResponse({"output": result})

@app.post("/stream", response_class=EventSourceResponse)
async def stream(req: InvokeRequest) -> EventSourceResponse:
    cfg: RunnableConfig = {
        "configurable": {"thread_id": req.thread_id, **req.configurable}
    }
    async def event_generator() -> AsyncGenerator[dict, None]:
        # stream_mode="messages" yields (AIMessageChunk, metadata) tuples
        async for chunk, _meta in app.state.graph.astream(
            {"messages": req.messages},
            config=cfg,
            stream_mode="messages",
        ):
            token = getattr(chunk, "content", "")
            if token:
                yield {"data": token}
    return EventSourceResponse(event_generator(), ping=15)

@app.get("/health")
def health() -> JSONResponse:           # sync: avoids event-loop deadlock
    meta = getattr(app.state, "prompt_meta", {})
    return JSONResponse({
        "status": "ok",
        "cycle_count": _cycle_count,
        "prompt_commit": meta.get("commit", "unknown"),
        "prompt_loaded_at": meta.get("loaded_at", ""),
    })
\end{lstlisting}

Four decisions in Listing~\ref{lst:fastapi-app} deserve explanation. First,
\code{astream(stream\_mode="messages")} is the correct LangGraph API for
token streaming: it yields \code{(AIMessageChunk, metadata)} pairs as each
token arrives from any LLM node in the graph. This is distinct from
LangChain's \code{astream\_events} method, which operates on LCEL chains
(sequences of runnables), not on compiled state graphs; using
\code{astream\_events} on a \code{CompiledStateGraph} would yield node-level
events rather than per-token chunks and is not the intended API for this
use case. Second, \code{health} is a plain synchronous function, not an
\code{async def}: a liveness probe that awaits anything can deadlock if the
event loop is saturated with graph invocations, causing the probe to fail
and the pod to restart exactly when it is busiest. Third,
\code{\_cycle\_count} is guarded by a \code{threading.Lock} rather than an
\code{asyncio.Lock} because both the async \code{invoke} handler and the
synchronous \code{health} endpoint read or write the counter; mixing asyncio
and threading locks causes deadlocks. Fourth, \code{EventSourceResponse} is
constructed with \code{ping=15}: the 15-second keep-alive ping prevents load
balancers and proxies with idle-connection timeouts shorter than a typical
agent run from closing the SSE connection mid-stream.

\begin{notebox}
  \textbf{NodeInterrupt vs.\ GraphRecursionError.}\quad
  LangGraph raises \code{NodeInterrupt} when a node calls \code{interrupt()}
  to pause for human input; the graph state is safely checkpointed and the
  run can be resumed. This is not an error condition and should return
  HTTP~202 with a checkpoint reference. \code{GraphRecursionError} signals a
  genuinely broken graph that has exceeded its recursion limit and should
  return HTTP~422. Treating both as HTTP~500 will cause your monitoring to
  page on normal human-in-the-loop pauses.
\end{notebox}

With the HTTP interface defined, the next problem is making it reproducible
across machines and environments---the subject of the infrastructure layer.

%% -----------------------------------------------------------
\section{The Infrastructure Layer Must Protect Traffic in Transit}
\label{sec:infra-layer}

The design problem here is that deploying a new pod version without
dropping in-flight requests requires the infrastructure layer to know two
independent things: whether the process is alive at all, and whether it is
ready to accept new traffic. The naive approach---a single health check
that governs both---creates a false equivalency: a pod can be alive (the
process is running) without being ready (the graph is not yet compiled, the
checkpointer connection is not yet established). Conflating them means either
that live traffic hits an unready pod, or that a pod is unnecessarily
restarted when it is merely warming up. The principle is to decouple
liveness from readiness at both the image and the Kubernetes manifest levels.

\subsection*{The Image Must Not Run as Root}

Before the Kubernetes manifest can be correct, the image it runs must be
safe. A process running as root inside a container can, if an attacker
achieves code execution, write to mounted secrets, modify \code{/etc/hosts},
or escape to the host via certain kernel CVEs. The multi-stage Docker build
pattern addresses this in two steps: a \emph{builder stage} installs all
Python dependencies using \code{uv}, whose \code{--frozen} flag guarantees
the lockfile governs the install; a \emph{runtime stage} copies only the
virtual environment and application source into a minimal image and creates a
dedicated non-root user before setting the entrypoint.

\begin{lstlisting}[language=bash,
  caption={Multi-stage Dockerfile with non-root runtime --- \texttt{Dockerfile}},
  label={lst:dockerfile}]
# --- builder ---
FROM ghcr.io/astral-sh/uv:0.5-python3.12-bookworm-slim AS builder
WORKDIR /app
COPY pyproject.toml uv.lock ./
RUN uv sync --frozen --no-cache --compile-bytecode
COPY src/ ./src/

# --- runtime ---
FROM python:3.12-slim-bookworm AS runtime
WORKDIR /app
RUN addgroup --system appgroup \
 && adduser --system --ingroup appgroup \
            --no-create-home appuser
COPY --from=builder /app/.venv ./.venv
COPY --from=builder /app/src  ./src
ENV PATH="/app/.venv/bin:$PATH"
USER appuser
ENTRYPOINT ["python", "-m", "uvicorn", "src.serve.app:app", \
            "--host", "0.0.0.0", "--port", "8080", "--workers", "1"]
\end{lstlisting}

Three decisions in Listing~\ref{lst:dockerfile} deserve explanation. First,
\code{--compile-bytecode} in the builder stage pre-compiles all \code{.py}
files to \code{.pyc} bytecode during the build rather than at first import;
this shaves 200--400\,ms from pod startup time when the virtual environment
contains large packages such as \code{langchain-core}. Second, the runtime
base image is plain \code{python:3.12-slim-bookworm}, not the \code{uv}
image; the runtime stage has no need for \code{uv} itself, and including it
adds roughly 20\,MB and an additional binary with network capabilities.
Third, the \code{ENTRYPOINT} uses array form (exec form) rather than shell
form: in exec form, \code{uvicorn} is PID~1 and receives \code{SIGTERM}
directly, enabling graceful shutdown; in shell form, the shell is PID~1 and
\code{SIGTERM} may never reach \code{uvicorn}, leaving in-flight requests
orphaned.

\begin{warningbox}
  \textbf{One worker per pod.}\quad
  The \code{--workers~1} flag is intentional. LangGraph graphs run on
  \code{asyncio} event loops; multiple Gunicorn workers each have a separate
  memory space, meaning \code{app.state.graph} is not shared. If you need
  horizontal scale, add Kubernetes replicas---each pod gets one event loop and
  one graph instance. Running multiple workers per pod while holding
  event-loop-bound state (checkpointer connections, the compiled graph) causes
  subtle race conditions that are nearly impossible to reproduce locally.
\end{warningbox}

\subsection*{Kubernetes Needs Two Distinct Health Checks}

With the image defined, the Kubernetes manifest must wire the two health
check semantics correctly. The \code{livenessProbe} asks: ``is the container
still functioning?'' Its \code{initialDelaySeconds} must be long enough for
the graph to finish compiling and the checkpointer connection to be
established---30\,s is a safe baseline for typical dependency trees. The
\code{readinessProbe} asks: ``should this pod receive traffic?'' It fires
earlier (10\,s) and with a lower failure threshold (2 versus 3) because
removing a pod from rotation is cheap and reversible, while kubelet restarts
are disruptive. Both probes hit the same \code{GET /health} endpoint from
Listing~\ref{lst:fastapi-app}, which is synchronous and fast.

\begin{lstlisting}[language=yaml,
  caption={Kubernetes Deployment with rolling strategy ---
    \texttt{k8s/deployment.yaml} (key fields)},
  label={lst:k8s-deployment}]
apiVersion: apps/v1
kind: Deployment
metadata:
  name: langgraph-agent
spec:
  replicas: 3
  minReadySeconds: 15
  strategy:
    type: RollingUpdate
    rollingUpdate:
      maxSurge: 1
      maxUnavailable: 0
  selector:
    matchLabels: {app: langgraph-agent}
  template:
    metadata:
      labels: {app: langgraph-agent}
    spec:
      containers:
        - name: agent
          image: registry.example.com/langgraph-agent:$(TAG)
          ports: [{containerPort: 8080}]
          envFrom:
            - secretRef:    {name: langgraph-secrets}
            - configMapRef: {name: langgraph-config}
          resources:
            limits:   {cpu: "2", memory: 4Gi}
            requests: {cpu: "500m", memory: 1Gi}
          livenessProbe:
            httpGet: {path: /health, port: 8080}
            initialDelaySeconds: 30
            periodSeconds: 20
            failureThreshold: 3
          readinessProbe:
            httpGet: {path: /health, port: 8080}
            initialDelaySeconds: 10
            periodSeconds: 10
            failureThreshold: 2
\end{lstlisting}

Three decisions in Listing~\ref{lst:k8s-deployment} deserve explanation.
First, \code{minReadySeconds:~15} sits at the \code{spec} level of the
\code{Deployment}, not inside \code{rollingUpdate}; it is a separate field
that requires a pod to remain ready for 15\,s before the rollout controller
considers it ``available'' and proceeds to replace the next old pod. Without
it, a pod that passes its first readiness check but crashes two seconds later
is counted as available, and the rollout continues into a degraded cluster.
Second, \code{maxUnavailable:~0} combined with \code{maxSurge:~1} means
there are always at least \code{replicas} pods serving traffic; the surge pod
is created, passes readiness, waits \code{minReadySeconds}, and only then is
an old pod terminated. Third, \code{envFrom.secretRef} references a
Kubernetes \code{Secret} object by name---the manifest itself contains no
credential literals, satisfying audit requirements and enabling secret
rotation without image rebuilds.

Knowing which environment variables belong in which \code{Secret} or
\code{ConfigMap} requires a principled inventory of all configuration values,
which is the responsibility of the configuration layer.

%% -----------------------------------------------------------
\section{Configuration Must Not Live in Code}
\label{sec:config-must-not-live}

The design problem here is that configuration values scattered across source
files make it impossible to run the same artefact in multiple environments.
A model name hardcoded in a node function, a Redis URL hardcoded in the
checkpointer factory, an API key string-interpolated into a client
constructor: each of these ties a specific environment to a specific image
tag, defeating the purpose of containerisation. The naive approach of
\code{os.environ.get("KEY", "default")} is marginally better but does not
validate types, does not document what values are expected, and silently
accepts malformed input. The principle is that every external value must be
declared, typed, validated at startup, and injected from the environment.

\emph{pydantic-settings} extends Pydantic's validation to environment
variables and \code{.env} files. A class inheriting from \code{BaseSettings}
declares fields exactly as a normal Pydantic model, but pydantic-settings
populates those fields by reading the environment first, then falling back to
\code{.env} files, then to declared defaults. The key feature for this
architecture is \emph{nested sub-models}: by setting
\code{env\_nested\_delimiter="\_\_"}, the environment variable
\code{GRAPH\_\_RECURSION\_LIMIT=25} automatically maps to
\code{settings.graph.recursion\_limit}, grouping related values and avoiding
a flat namespace of thirty top-level variables.

\begin{lstlisting}[language=Python,
  caption={Typed, validated settings with nested sub-models ---
    \texttt{src/config/settings.py}},
  label={lst:settings}]
from functools import lru_cache
from typing import Literal
from pydantic import BaseModel, SecretStr, field_validator
from pydantic_settings import BaseSettings, SettingsConfigDict
import re

class LLMConfig(BaseModel):
    provider: str = "anthropic"
    model: str = "claude-sonnet-4-8"
    temperature: float = 0.0
    api_key: SecretStr

class GraphConfig(BaseModel):
    recursion_limit: int = 25
    checkpointer_backend: Literal["memory", "redis"] = "redis"

class LangSmithConfig(BaseModel):
    api_key: SecretStr
    hub_repo_id: str
    project_name: str

    @field_validator("hub_repo_id")
    @classmethod
    def _validate_repo_id(cls, v: str) -> str:
        if not re.fullmatch(r"[a-zA-Z0-9_-]+/[a-zA-Z0-9_-]+", v):
            raise ValueError("hub_repo_id must be 'owner/name'")
        return v

class Settings(BaseSettings):
    model_config = SettingsConfigDict(
        env_file=(".env", ".env.prod"),
        env_file_encoding="utf-8",
        env_nested_delimiter="__",
        case_sensitive=False,
    )
    llm: LLMConfig
    graph: GraphConfig
    langsmith: LangSmithConfig
    redis_url: str = "redis://localhost:6379"
    prompt_variant: Literal["v1", "v2"] = "v1"

    @field_validator("redis_url")
    @classmethod
    def _validate_redis_url(cls, v: str) -> str:
        if not v.startswith(("redis://", "rediss://")):
            raise ValueError("redis_url must use redis:// or rediss://")
        return v

@lru_cache
def get_settings() -> Settings:
    return Settings()  # type: ignore[call-arg]
\end{lstlisting}

Four decisions in Listing~\ref{lst:settings} deserve explanation. First,
\code{SecretStr} is used for all API keys: Pydantic's \code{SecretStr} type
masks the value in \code{\_\_repr\_\_} and \code{\_\_str\_\_} output, so an
accidentally logged settings object will show \code{'**********'} rather than
the live key. The raw value is accessed only at client construction time via
\code{.get\_secret\_value()}. Second, \code{AnyUrl} (Pydantic v1 style) is
intentionally absent: in Pydantic v2, assigning a plain string default to an
\code{AnyUrl} field raises a validation error because \code{AnyUrl} no longer
coerces strings automatically. A plain \code{str} with a custom
\code{@field\_validator} is the correct pattern. Third, \code{env\_file}
accepts a tuple: \code{(".env", ".env.prod")} means the prod file is loaded
after the base file and its values win---a clean multi-environment override
without templating tools. Fourth, \code{@lru\_cache} on \code{get\_settings}
makes the \code{Settings} object a per-process singleton; in tests,
\code{get\_settings.cache\_clear()} invalidates the cache between test cases
so that each case can inject its own environment variables without
monkeypatching.

\begin{notebox}
  \code{SettingsConfigDict} and \code{BaseSettings} are both imported from
  \code{pydantic\_settings}, \emph{not} from \code{pydantic}. They are
  provided by the separate \code{pydantic-settings} package, which must be
  listed as an explicit dependency. Importing \code{BaseSettings} from
  \code{pydantic} directly was deprecated in Pydantic v2 and will raise an
  \code{ImportError}.
\end{notebox}

With all values externalised, it is now possible to change a configuration
value---including the identity of the prompt---without modifying or
redeploying the image, which is the mechanism that enables the next
capability: atomic prompt rollout.

%% -----------------------------------------------------------
\section{Prompt Changes Must Be Atomic and Reversible}
\label{sec:zero-downtime-rollout}

The design problem here is that a LangGraph agent's behaviour is determined
by two independent artefacts: the code and the prompt. Deploying a new prompt
by releasing a new Docker image is safe but slow---a typical image build and
rollout takes five to fifteen minutes, far too long to recover from a bad
prompt in production. Storing the prompt inside the codebase ties two
independent change velocities together: engineers deploy code on one schedule
and prompt engineers iterate on another. The principle is that the prompt must
be a versioned external artefact, swappable at runtime, with an instant
rollback path.

LangSmith Hub provides a versioned store for \code{ChatPromptTemplate}
objects. A prompt pushed to Hub is identified by a fully qualified reference
of the form \code{owner/name:commit-hash}. The \code{langsmith.hub.pull()}
function retrieves the prompt by reference and returns a \code{Runnable} (in
practice a \code{ChatPromptTemplate}) that can be composed into any LangGraph
node. Pinning the commit hash makes the pull deterministic: the same hash
will always return the same prompt bytes regardless of subsequent pushes to
the same repo. The floating \code{:latest} reference is appropriate only in
CI/CD pipelines where the intent is always to test the newest prompt.

A typed graph state is required by \code{StateGraph}: passing a plain
\code{dict} is not a valid schema. The correct pattern is to define a
\code{TypedDict} with the state fields the graph reads and writes. The
\code{config\_schema} parameter to \code{StateGraph} was deprecated in
LangGraph v0.6.0 and removed in the v0.6 release cycle; runtime context that
was previously passed via \code{config\_schema} now belongs in
\code{context\_schema}, which makes immutable per-run data (such as a
\code{user\_id} or database connection) available to nodes via the
\code{Runtime} object without polluting agent state.

\begin{lstlisting}[language=Python,
  caption={Graph factory with typed state, prompt pull, and feature-flag ---
    \texttt{src/graph/factory.py}},
  label={lst:graph-factory}]
from typing import Annotated
from typing_extensions import TypedDict
from langsmith import hub
from langchain_core.runnables import Runnable
from langchain_core.messages import BaseMessage
from langgraph.graph import StateGraph, START, END
from langgraph.graph.message import add_messages
from langgraph.checkpoint.redis import RedisSaver
from src.config.settings import Settings

class AgentState(TypedDict):
    messages: Annotated[list[BaseMessage], add_messages]

def pull_prompt(repo_id: str, commit: str | None = None) -> Runnable:
    ref = f"{repo_id}:{commit}" if commit else repo_id
    return hub.pull(ref)

def build_graph(settings: Settings) -> tuple:
    """Returns (CompiledStateGraph, prompt_meta dict)."""
    from datetime import datetime, timezone
    commit = getattr(settings, "prompt_commit", None)
    prompt = pull_prompt(settings.langsmith.hub_repo_id, commit)
    prompt_meta = {
        "commit": commit or "latest",
        "loaded_at": datetime.now(timezone.utc).isoformat(),
        "variant": settings.prompt_variant,
    }
    checkpointer = RedisSaver.from_conn_string(settings.redis_url)
    builder = StateGraph(AgentState)
    builder.add_node("llm", _make_llm_node(prompt, settings))
    builder.add_edge(START, "llm")
    builder.add_edge("llm", END)
    compiled = builder.compile(checkpointer=checkpointer)
    return compiled, prompt_meta
\end{lstlisting}

Three decisions in Listing~\ref{lst:graph-factory} deserve explanation.
First, \code{hub.pull} is imported from \code{langsmith}, not from
\code{langchain\_hub} or \code{langchain}: as of LangChain~0.2, the Hub
client was migrated to the \code{langsmith} package, and importing from
\code{langchain} raises a deprecation warning that will become an
\code{ImportError}. Second, \code{AgentState} is a \code{TypedDict} with a
\code{messages} field annotated with the \code{add\_messages} reducer: this
is the correct and required pattern for a \code{StateGraph} schema;
\code{add\_messages} ensures that new messages are appended rather than
replacing the entire list on each node return. Third, the function returns a
tuple of the compiled graph and a \code{prompt\_meta} dictionary: the
metadata---commit hash, load timestamp, variant---is stored on
\code{app.state.prompt\_meta} and served by \code{/health}, enabling
LangSmith traces to be correlated with the exact prompt version that produced
them.

Hot-swapping the prompt without restarting the pod requires a mechanism to
trigger \code{build\_graph} after startup. On UNIX systems, \code{SIGUSR1}
is the conventional ``reload configuration'' signal: it is not used by the
Python runtime itself and is safe to repurpose. The signal handler schedules
a reload coroutine on the running event loop using
\code{loop.add\_signal\_handler}, which is the thread-safe way to hand work
to an asyncio loop from a synchronous OS signal context.

\begin{lstlisting}[language=Python,
  caption={SIGUSR1 hot-swap handler added to lifespan ---
    \texttt{src/serve/app.py} (additions)},
  label={lst:sigusr1}]
import asyncio, signal, sys
from src.graph.factory import build_graph
from src.config.settings import get_settings

_reload_lock = asyncio.Lock()

async def _reload_graph(app) -> None:
    async with _reload_lock:          # prevents concurrent reloads
        settings = get_settings()
        new_graph, new_meta = build_graph(settings)
        app.state.graph = new_graph   # atomic single-assignment swap
        app.state.prompt_meta = new_meta

def _install_reload_handler(app, loop: asyncio.AbstractEventLoop) -> None:
    if sys.platform == "win32":
        # Windows has no SIGUSR1; expose POST /admin/reload instead
        @app.post("/admin/reload", include_in_schema=False)
        async def admin_reload():
            await _reload_graph(app)
            return {"reloaded": True}
        return
    loop.add_signal_handler(
        signal.SIGUSR1,
        lambda: asyncio.ensure_future(_reload_graph(app))
    )

# Call inside lifespan, after graph is first built:
# loop = asyncio.get_event_loop()
# _install_reload_handler(app, loop)
\end{lstlisting}

Two decisions in Listing~\ref{lst:sigusr1} deserve explanation. First,
\code{\_reload\_lock} is an \code{asyncio.Lock} rather than a
\code{threading.Lock}: the reload coroutine runs on the event loop, so the
async lock is correct; a threading lock here would block the event loop
thread. The lock prevents a second \code{SIGUSR1} from starting a new Hub
pull while the first is still in flight, which would cause two concurrent
\code{build\_graph} calls racing to write \code{app.state.graph}. Second,
the Windows path is not a workaround---it is the correct design for that
platform: UNIX signals and Windows signals have fundamentally different
semantics, and a \code{POST /admin/reload} endpoint behind network-policy
restrictions is the idiomatic Windows-compatible alternative.

\begin{warningbox}
  \textbf{Multi-worker GIL caveat.}\quad
  The \code{app.state.graph = new\_graph} swap is a single Python attribute
  assignment and is atomic under CPython's GIL within a single process.
  It is \emph{not} safe across multiple Gunicorn workers, because each worker
  is a separate OS process with its own memory space. If you must use multiple
  workers per pod, store the active prompt commit hash in Redis and rebuild
  the graph from it on each request---at the cost of per-request latency.
\end{warningbox}

The OSS runtime is now fully operational: requests are served, the image is
safe, configuration is externalised, and prompts roll out atomically. The
next two sections cover the LangGraph Platform managed path, which replaces
the infrastructure layer with a single configuration file while preserving
the same configuration and experiment primitives.

%%------------------------------------------------------------
%% PART B — LangGraph Platform (Managed Service)
%%------------------------------------------------------------
\noindent\rule{\linewidth}{0.8pt}
\begin{center}
  \textbf{Part B\quad LangGraph Platform (Managed)\quad
    langgraph.json\,\textperiodcentered\,Cron\,\textperiodcentered\,Webhooks\,\textperiodcentered\,Studio\,\textperiodcentered\,A/B}
\end{center}
\noindent\rule{\linewidth}{0.8pt}

%% -----------------------------------------------------------
\section{A Single File Replaces the Infrastructure Layer}
\label{sec:platform-json}

The design problem here is that the OSS infrastructure path requires
maintaining a Dockerfile, a Kubernetes manifest, an ingress configuration,
and a rolling-deploy pipeline---artefacts whose combined complexity rivals
the agent code itself. For teams whose core competency is agent behaviour
rather than container orchestration, this maintenance tax is a recurring
source of production incidents unrelated to the agent's logic. The LangGraph
Platform managed runtime internalises all of this infrastructure and exposes
a single declarative configuration file, \code{langgraph.json}, as the only
interface between the application code and the deployment system.

\code{langgraph.json} has two required keys. \code{graphs} is a dictionary
mapping an agent name to a Python module path and variable (or zero-argument
factory function) in the form \code{"./path/to/file.py:variable"}. A
zero-argument callable is detected automatically and called at startup,
making it the correct form when the graph constructor reads from
\code{get\_settings()} and therefore requires environment variables to be
populated first. \code{dependencies} is a list of pip package names or
relative local paths (e.g.\ \code{"."} to install the project root as a
package) that the Platform runtime installs before starting the server.
All other keys---\code{env}, \code{dockerfile\_lines}, \code{http},
\code{auth}---are optional.

\begin{lstlisting}[language=json,
  caption={Complete \texttt{langgraph.json} with required and key optional fields},
  label={lst:langgraph-json}]
{
  "graphs": {
    "research_agent": "./src/graph/agent.py:graph",
    "summariser":     "./src/graph/summariser.py:make_graph"
  },
  "dependencies": [
    ".",
    "langchain_anthropic",
    "langchain_openai",
    "qdrant-client",
    "litellm"
  ],
  "env": "./.env",
  "dockerfile_lines": [
    "RUN apt-get update && apt-get install -y --no-install-recommends poppler-utils && rm -rf /var/lib/apt/lists/*"
  ],
  "http": {
    "app": "./src/serve/middleware.py:app"
  }
}
\end{lstlisting}

Three decisions in Listing~\ref{lst:langgraph-json} deserve explanation.
First, \code{"env":~"./.env"} is present for local development with
\code{langgraph dev} but must be omitted or treated as inert in cloud and
self-hosted deployments, where the Platform control plane injects environment
variables directly into the container; relying on a committed \code{.env}
file in production is a credentials-leak risk. Second, \code{make\_graph}
(the \code{summariser} entry) points to a factory function rather than a
compiled graph variable: the Platform detects zero-argument callables and
calls them at startup, which is necessary when the graph constructor reads
from \code{get\_settings()} and therefore requires the environment to be
populated first. Third, \code{dockerfile\_lines} injects raw \code{RUN}
instructions into the Platform's generated Dockerfile; the
\code{--no-install-recommends} and cache-cleanup flags are not optional
niceties---they prevent the runtime image from ballooning with documentation
packages and APT cache files.

\begin{notebox}
  \textbf{Four deployment modes, one config file.}\quad
  The LangGraph Platform supports four deployment modes: Cloud SaaS
  (one-click deploy from a GitHub repo), Self-Hosted Data Plane (your cloud,
  LangChain's control plane), Self-Hosted Control Plane (fully air-gapped),
  and Standalone Container (single Docker container for on-premise). The
  \code{langgraph.json} file is identical across all four modes; the
  distinction lies in where \code{langgraph build} pushes its output and who
  manages the serving infrastructure.
\end{notebox}

The Platform's managed runtime also provides primitives for time-based
scheduling and asynchronous run management that would otherwise require
Celery or APScheduler in the OSS path.

\subsection*{Cron Triggers and Webhook Callbacks}

Many production agents are not request-driven: a daily digest agent, a
nightly data-enrichment pipeline, or an hourly anomaly detector all need
to fire on a schedule. In the OSS path this requires a separate Celery worker,
a beat scheduler, and careful coordination of database connections. The
Platform provides \emph{cron triggers} as a first-class primitive: a cron job
is registered once via the SDK and thereafter fires runs against a specified
assistant on the platform's scheduler, with no additional infrastructure.

A related problem is long-running background runs. A research agent that
takes three minutes to complete should not hold an HTTP connection open for
that duration. The Platform's \code{webhook} parameter on run creation allows
a caller to fire a run and receive the result asynchronously: the Platform
POSTs the complete run output as a JSON payload to a URL the caller provides
when the run finishes.

\begin{lstlisting}[language=Python,
  caption={Platform cron registration and webhook run creation ---
    \texttt{src/platform/cron\_setup.py}},
  label={lst:cron-setup}]
from langgraph_sdk import get_client
from typing import Any

async def register_daily_digest(
    base_url: str,
    assistant_id: str,
) -> dict[str, Any]:
    client = get_client(url=base_url)
    return await client.crons.create(
        assistant_id=assistant_id,
        schedule="0 9 * * 1-5",           # weekdays at 09:00 UTC
        input={"task": "daily_digest"},
        metadata={"created_by": "cron_setup"},
    )

async def create_webhook_run(
    base_url: str,
    thread_id: str,
    assistant_id: str,
    payload: dict[str, Any],
    webhook_url: str,
) -> dict[str, Any]:
    client = get_client(url=base_url)
    return await client.runs.create(
        thread_id=thread_id,
        assistant_id=assistant_id,
        input=payload,
        webhook=webhook_url,
        multitask_strategy="rollback",
    )
\end{lstlisting}

Two decisions in Listing~\ref{lst:cron-setup} deserve explanation. First,
\code{crons.create} takes \code{assistant\_id}, not \code{graph\_id}: in
the Platform's data model, an \emph{assistant} is a deployed instance of a
graph with a specific configuration; the cron is attached to the assistant,
not the graph definition, so that different assistants built from the same
graph can have independent schedules and inputs. Second,
\code{multitask\_strategy="rollback"} on \code{runs.create} is the correct
choice for most interactive agents: if a second message arrives on a thread
while a run is in progress, \code{"rollback"} cancels the in-flight run and
starts a fresh one with the new input, preserving checkpointed state up to the
last completed node. The alternatives---\code{"enqueue"} (queue behind the
current run) and \code{"interrupt"} (cancel without rollback)---suit batch
pipelines and stateless agents respectively.

\subsection*{Studio Debugging}

The Platform's \emph{Studio} interface connects to any LangGraph Server---local
\code{langgraph dev} or hosted---and renders a real-time timeline of node
activations, state values at each step, and the full message history. For
debugging, the most useful Studio feature is \emph{graph-level breakpoints}:
by compiling a graph with \code{interrupt\_before=["node\_name"]} or
\code{interrupt\_after=["node\_name"]}, every run pauses at that node and
exposes the current state for inspection. A human operator can then modify
the state via the compiled graph's \code{aupdate\_state()} method (or the
SDK's \code{client.threads.update\_state()}) and resume the run with an
empty-input \code{runs.create}, which is the same mechanism used for
human-in-the-loop approval flows in production.

\begin{tipbox}
  During development, pass \code{interrupt\_before=["llm"]} to the
  \code{compile()} call to pause before every LLM invocation. Studio will
  show you the exact messages being sent to the model---far more informative
  than reading a trace after the fact. Remove the interrupt before deploying
  to production.
\end{tipbox}

With the Platform's infrastructure concerns solved by configuration, the
remaining problem shared by both the OSS and Platform paths is how to make
model selection a runtime decision, enabling controlled experiments without
code changes.

%% -----------------------------------------------------------
\section{Model Choice Must Be a Runtime Parameter}
\label{sec:ab-routing}

The design problem here is that choosing between two models---to measure
quality, latency, or cost differences---requires the model identity to be
a runtime variable rather than a compile-time constant. The naive approach
of a simple \code{if} statement inside a node function couples the routing
logic to the graph code, requires a redeployment to change routing weights,
and makes it impossible to run both variants simultaneously on real traffic.
The principle is that the caller assigns a variant label, the label travels
through the request without touching agent state, and the serving layer
resolves the label to a model deployment at call time.

The \code{configurable} dictionary in \code{RunnableConfig} is the correct
carrier for this label. Unlike agent state---which is part of the graph's
domain model and persisted to the checkpointer---\code{configurable} is a
per-request transport channel: it is threaded through every node call without
being stored or returned in the graph's output. A node reads the variant label
with \code{config["configurable"].get("model\_variant",~"control")} and uses
it to select a model deployment. The caller assigns the variant before
invoking the graph:
\code{config=\{"configurable":\{"thread\_id":~...,~"model\_variant":~"treatment"\}\}}.

The variant label is resolved to a specific model deployment by a
\code{LiteLLM Router}. The Router holds a list of model deployments, each
identified by a \code{model\_name} alias. For A/B testing, each alias maps to
exactly one \code{litellm\_params.model} entry---there is no load balancing
within an alias, because load balancing would randomise which model each
request hits and destroy experimental attribution. LiteLLM's recommended and
default routing strategy is \code{simple-shuffle}: with one deployment per
alias it is effectively deterministic, always returning the single entry that
matches the requested \code{model\_name}.

\begin{lstlisting}[language=Python,
  caption={LiteLLM Router with per-alias model deployments ---
    \texttt{src/llm/router.py}},
  label={lst:litellm-router}]
import os
from litellm import Router

ROUTER_MODEL_LIST: list[dict] = [
    {
        "model_name": "control",          # alias passed as model= arg
        "litellm_params": {
            "model": "anthropic/claude-sonnet-4-8",
            "api_key": os.environ["ANTHROPIC_API_KEY"],
        },
        "model_info": {"id": "control-sonnet-4-8"},
    },
    {
        "model_name": "treatment",
        "litellm_params": {
            "model": "openai/gpt-5-3-instant",
            "api_key": os.environ["OPENAI_API_KEY"],
        },
        "model_info": {"id": "treatment-gpt-5-3"},
    },
]

def build_router() -> Router:
    return Router(
        model_list=ROUTER_MODEL_LIST,
        routing_strategy="simple-shuffle",  # default; deterministic at 1 dep/alias
        num_retries=2,
        allowed_fails=3,
        cooldown_time=30,
    )
\end{lstlisting}

Two decisions in Listing~\ref{lst:litellm-router} deserve explanation. First,
each alias (\code{"control"}, \code{"treatment"}) maps to exactly one entry.
LiteLLM's \code{simple-shuffle} strategy picks randomly among all entries
sharing the same \code{model\_name}; with one entry per name the selection is
deterministic, which is what A/B experiments require---every request labelled
\code{"treatment"} must hit the same model. Second, \code{allowed\_fails:~3}
combined with \code{cooldown\_time:~30} means that after three consecutive
failures, a deployment is removed from the active pool for 30\,s. During an
outage this means your traffic split becomes temporarily 100/0---a known
tradeoff that should be documented in the experiment plan so the data science
team can exclude the outage window from significance calculations.

With the Router built, the node that calls the model needs only to read the
variant from \code{configurable} and pass it as the \code{model} argument:

\begin{lstlisting}[language=Python,
  caption={LLM node reading variant from \texttt{configurable} ---
    \texttt{src/graph/nodes.py}},
  label={lst:llm-node}]
from langchain_core.runnables import RunnableConfig
from langchain_core.messages import BaseMessage
from litellm import Router

async def llm_node(
    state: dict,
    config: RunnableConfig,
) -> dict:
    configurable = config.get("configurable", {})
    variant: str = configurable.get("model_variant", "control")
    router: Router = configurable["llm_router"]   # injected via lifespan
    messages: list[BaseMessage] = state["messages"]
    # Convert LangChain messages to OpenAI dicts for LiteLLM
    oai_msgs = [
        {"role": m.type if m.type != "human" else "user",
         "content": m.content}
        for m in messages
    ]
    response = await router.acompletion(
        model=variant,    # "control" or "treatment" matches model_name
        messages=oai_msgs,
    )
    usage = response.usage
    return {
        "messages": [{"role": "assistant",
                      "content": response.choices[0].message.content}],
        "model_variant": variant,
        "prompt_tokens": usage.prompt_tokens,
        "completion_tokens": usage.completion_tokens,
    }
\end{lstlisting}

Three decisions in Listing~\ref{lst:llm-node} deserve explanation. First,
\code{llm\_router} is injected through \code{configurable} rather than
imported as a module-level global: this makes the node testable in isolation
by passing a mock router in \code{config["configurable"]["llm\_router"]}, and
avoids the startup-order dependency that arises when a module-level import
triggers \code{Router()} before environment variables are populated. Second,
LangChain \code{BaseMessage} objects must be converted to OpenAI-compatible
dictionaries before passing to \code{router.acompletion}, which speaks the
OpenAI messages format; the \code{m.type} field uses \code{"human"} where
OpenAI expects \code{"user"}, requiring the explicit mapping shown. Third,
\code{variant} and token counts are returned as state fields: downstream nodes
(such as a Langfuse logging node) can read these values from state without
re-accessing the response object, which may no longer be in scope by the time
the logging node runs.

\begin{tipbox}
  To assign variants consistently across a session, hash the \code{user\_id}
  modulo 2:
  \code{variant = "control" if int(hashlib.md5(user\_id.encode()).hexdigest(), 16) \% 2 == 0 else "treatment"}.
  This produces a stable assignment---the same user always sees the same
  variant---without requiring a feature-flag service. Use a proper flag
  service (LaunchDarkly, Flagsmith) only when you need real-time override
  capability or multi-variant experiments beyond binary splits.
\end{tipbox}

The Router's \code{allowed\_fails} mechanism, the lifespan injection pattern,
and the \code{configurable} channel are all load-bearing primitives that
reappear in Chapter~27's authentication middleware (which injects identity
claims into the same \code{configurable} dict) and in Chapter~29's Langfuse
integration (which groups traces by the \code{model\_variant} tag to produce
the statistical comparison that closes the A/B experiment loop).
