%% ============================================================
%%  Chapter 6 — Prompt Architecture & Context Assembly
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%% ============================================================

\chapterquote{A context window is not a notepad. It is a budget. Spend it
like one.}{anon}

By the end of Chapter~5 the agent has a robust LLM infrastructure layer:
routing, retries, circuit breaking, cost tracking, and externalised
configuration. Every call to the model is protected. What is not yet
protected is what the model receives. In a long-horizon agent loop the
context window fills incrementally with history, retrieved evidence,
tool schemas, and goal state. Left unmanaged, it overflows. The model
receives a truncated, mid-sentence input. The output degrades. The goal
drifts. The budget burns on cycles that are reasoning over noise.

The production failure this chapter prevents is silent context pollution:
the condition where the assembled prompt is technically within the context
limit but carries stale history, low-relevance evidence, over-wide tool
schemas, and a goal statement that has been crowded to the edge of the
window or evicted entirely. That condition produces hallucination,
goal drift, and KV-cache misses that double latency on every cycle ---
all without raising an exception or triggering a retry.

The solution is to treat the context window as a typed, budgeted data
structure rather than a string. This chapter builds that structure in
six layers. The first layer (\S\ref{sec:slot-budget}) defines the schema:
six named slots with fixed priorities, independent trim policies, and a
Pydantic model that makes the budget explicit and testable. The second
layer (\S\ref{sec:tokenizer}) enforces the budget: a \code{tiktoken}
measurement and trim pipeline that drops content before overflow, never
after, using strategies calibrated to each slot's semantics. The third
layer (\S\ref{sec:context-assembly-node}) composes the prompt as a pure
graph node, making context assembly observable, testable, and replaceable
without touching the Reason node. The fourth layer
(\S\ref{sec:kv-cache}) adds a structural constraint: the stable prefix
rule that preserves KV-cache reuse across cycles. The fifth layer
(\S\ref{sec:prompt-hub}) externalises the prompt template itself into a
version-pinned LangSmith Prompt Hub entry, eliminating silent regression
from template drift. The sixth layer (\S\ref{sec:goal-summary}) anchors
the agent's purpose: a structured \code{GoalSummary} object re-injected
into the \code{ACTIVE\_GOAL} slot on every cycle so the model never loses
sight of what it is trying to accomplish.

This chapter introduces six slot names and three error types that appear
throughout the rest of the book. Table~\ref{tab:slot-taxonomy} maps each
term to its role before you build any of the components.

\begin{table}[ht]
\centering
\caption{Slot taxonomy, trim priority, and error types introduced in Chapter~6}
\label{tab:slot-taxonomy}
\begin{tabular}{llll}
\toprule
\textbf{Slot / Term} & \textbf{Default tokens} & \textbf{Trim priority} & \textbf{Trimmable} \\
\midrule
\code{ROLE}         & 512   & ---   & No \\
\code{ACTIVE\_GOAL} & 256   & ---   & No \\
\code{EVIDENCE}     & 4\,096 & 2nd   & Yes \\
\code{TOOLS}        & 1\,024 & 3rd   & Yes \\
\code{HISTORY}      & 8\,192 & 1st   & Yes \\
\code{HEADROOM}     & 512   & ---   & No (sentinel) \\
\midrule
\code{SlotOverflowError}          & \multicolumn{3}{l}{Raised when a non-trimmable slot exceeds its budget} \\
\code{ContextWindowExceededError} & \multicolumn{3}{l}{Raised when the assembled prompt overflows after all trimming} \\
\code{CachePrefixUnstableError}   & \multicolumn{3}{l}{Raised when the system-message prefix is not byte-identical across renders} \\
\bottomrule
\end{tabular}
\end{table}

The sections are ordered by dependency: you cannot trim a slot until you
have a schema to measure it against (\S\ref{sec:slot-budget}); you cannot
assemble the prompt until you have a trimmer
(\S\ref{sec:tokenizer}); you cannot enforce the cache rule until the
assembly node exists (\S\ref{sec:context-assembly-node},
\S\ref{sec:kv-cache}); and the goal summary cannot be rendered without
the slot budget and tokenizer already in place
(\S\ref{sec:goal-summary}).

The source layout for this chapter is:

\begin{verbatim}
src/
  agent/
    nodes/
      context_assembly.py    # §6.3
    prompt/
      slot_budget.py         # §6.1
      tokenizer.py           # §6.2
      cache_guard.py         # §6.4
      hub.py                 # §6.5
      goal_summary.py        # §6.6
\end{verbatim}

%% ------------------------------------------------------------
\section{A Prompt Is a Typed, Budgeted Data Structure}
\label{sec:slot-budget}
%% ------------------------------------------------------------

The fundamental problem with treating a prompt as a free-form string is
that a string has no allocation policy. When the context window fills,
there is no principled answer to the question: what should be dropped,
and in what order? In practice, the answer defaults to whatever happens
to be at the end of the string --- which is usually the most recent
observations, exactly the content that should be preserved. The result
is that history and goal state are lost while stale role preamble
occupies the top of the window.

The correct model is a budget with named slots. Each slot has a fixed
allocation, a trim priority, and a policy for whether trimming is
permissible at all. The \code{ROLE} slot --- the system message ---
contains the agent's behavioural policy and output contract; it must never
be truncated. The \code{ACTIVE\_GOAL} slot carries the current goal
state; losing it mid-run is catastrophic. The \code{HISTORY} slot is the
largest trimmable slot and should be shed first. The \code{HEADROOM}
slot is a sentinel: it is never filled with content but reserves tokens
to prevent the off-by-one overflows that occur when a tokeniser estimate
and a provider's actual counter disagree by a few tokens.

The \code{SlotBudget} Pydantic model enforces this allocation statically.
It is \code{frozen=True} because a budget that can be mutated after
construction is not a budget --- it is a variable. The
\code{total\_fits\_context} validator ensures that the sum of all
allocations never exceeds the model's context window, catching
configuration errors at import time rather than at the moment the first
call overflows. Section~\ref{sec:context-assembly-node} reads
\code{SlotBudget} from \code{RunnableConfig} so that each graph
invocation can carry its own budget without mutating shared state.

\begin{lstlisting}[language=Python,
  caption={SlotBudget model with validators and SlotName enum},
  label={lst:slot-budget}]
# src/agent/prompt/slot_budget.py
from __future__ import annotations
from enum import Enum
from pydantic import BaseModel, model_validator


class SlotName(str, Enum):
    ROLE        = "role"
    ACTIVE_GOAL = "active_goal"
    EVIDENCE    = "evidence"
    TOOLS       = "tools"
    HISTORY     = "history"
    HEADROOM    = "headroom"


class SlotBudget(BaseModel):
    model_config = {"frozen": True}
    context_window: int = 128_000
    role:           int = 512
    active_goal:    int = 256
    evidence:       int = 4_096
    tools:          int = 1_024
    history:        int = 8_192
    headroom:       int = 512

    @model_validator(mode="after")
    def total_fits_context(self) -> "SlotBudget":
        allocated = (
            self.role + self.active_goal + self.evidence
            + self.tools + self.history + self.headroom
        )
        if allocated > self.context_window:
            raise ValueError(
                f"Slot allocations sum to {allocated} tokens but "
                f"context_window is {self.context_window}."
            )
        return self


DEFAULT_BUDGET = SlotBudget()
\end{lstlisting}

Three decisions in Listing~\ref{lst:slot-budget} deserve explanation.

First, \code{SlotName} inherits from both \code{str} and \code{Enum}.
This means a \code{SlotName} value serialises as a plain string in
Langfuse metadata and JSON logs without any custom encoder, while still
being a closed enumeration that the type checker validates at every call
site.

Second, \code{frozen=True} is set at the class level, not per-field.
Pydantic's \code{frozen} config makes every field immutable after
construction and makes the model hashable. The hashability is not
cosmetic: \code{SlotBudget} is passed as an argument to
\code{@lru\_cache}-wrapped tokeniser functions in
\S\ref{sec:tokenizer}, where hashability is a prerequisite for
caching.

Third, the validator sums all six slots against \code{context\_window}.
The headroom allocation is included in that sum intentionally. If a
calculation elsewhere treats headroom as free space rather than a
reserved allocation, that assumption will surface as a validation failure
here rather than as a silent overflow at runtime. Chapter~9's
frozen-\code{SlotBudget} tests rely on this validator to catch
misconfigured test fixtures before any node runs.

With the budget schema defined, the next question is how to measure
whether any piece of content fits within its slot's allocation --- and
what to drop when it does not.

%% ------------------------------------------------------------
\section{Trim Before Overflow, Never After}
\label{sec:tokenizer}
%% ------------------------------------------------------------

The canonical mistake in context management is to assemble the full
prompt and then truncate it to the context limit. Truncation at the
assembly boundary cuts wherever the string ends, which means the model
receives a sentence fragment, a broken JSON structure, or half a tool
schema. The output that follows is unpredictable. The correct approach
is the inverse: measure each slot's content against its allocation
before assembly, drop excess content according to a per-slot strategy,
and only then compose the final prompt. The assembled prompt is always
valid by construction.

The tokeniser module has three responsibilities. It provides a cached
encoder instance so the tiktoken binary is loaded once per process, not
per call. It provides a \code{count\_tokens} function that measures
content accurately. It provides a \code{trim\_to\_budget} function that
applies one of two strategies: \code{oldest\_first} for \code{HISTORY}
(which preserves recent exchanges --- the most contextually relevant),
and \code{lowest\_score} for \code{EVIDENCE} (which preserves
high-relevance chunks, dropping low-signal retrieved content first).
Chapter~8's \code{EvidenceChunk.score} field provides the relevance
signal consumed by the \code{lowest\_score} strategy.

\begin{warningbox}
\code{cl100k\_base} is the correct default encoder for GPT-4o and most
OpenAI-family models. It is also a reasonable approximation for Anthropic
models. However, the exact token count for a given string differs
slightly between providers. The \code{HEADROOM} slot exists precisely to
absorb those discrepancies. Do not remove it and do not set it below 256
tokens.
\end{warningbox}

\begin{lstlisting}[language=Python,
  caption={Cached encoder, token counter, and SlotOverflowError},
  label={lst:tokenizer}]
# src/agent/prompt/tokenizer.py
import logging
from functools import lru_cache
from typing import Literal
import tiktoken
from langchain_core.messages import BaseMessage
from src.agent.prompt.slot_budget import SlotName

logger = logging.getLogger(__name__)


@lru_cache(maxsize=8)
def get_encoder(name: str = "cl100k_base") -> tiktoken.Encoding:
    return tiktoken.get_encoding(name)


def count_tokens(text: str, encoder: tiktoken.Encoding) -> int:
    return len(encoder.encode(text))


class SlotOverflowError(ValueError):
    def __init__(self, slot: SlotName, count: int, budget: int) -> None:
        super().__init__(
            f"Slot {slot.value!r} contains {count} tokens "
            f"but budget is {budget}."
        )
        self.slot   = slot
        self.count  = count
        self.budget = budget
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={trim\_to\_budget: oldest-first and lowest-score strategies},
  label={lst:trim-strategy}]
# src/agent/prompt/tokenizer.py (continued)

def trim_to_budget(
    messages: list[BaseMessage],
    budget:   int,
    encoder:  tiktoken.Encoding,
    strategy: Literal["oldest_first", "lowest_score"] = "oldest_first",
) -> tuple[list[BaseMessage], int]:
    """Return (trimmed_messages, tokens_used). Never raises."""
    if strategy == "oldest_first":
        while messages:
            total = sum(count_tokens(m.content, encoder) for m in messages)
            if total <= budget:
                break
            # Drop the oldest user+assistant pair (atomic unit).
            drop     = 2 if len(messages) >= 2 else 1
            dropped  = messages[:drop]
            messages = messages[drop:]
            logger.debug(
                "trim_to_budget: dropped %d msg(s), ~%d tokens",
                drop,
                sum(count_tokens(m.content, encoder) for m in dropped),
            )
    else:  # lowest_score — used for EVIDENCE
        messages = sorted(
            messages,
            key=lambda m: getattr(m, "score", 0.0),
            reverse=True,
        )
        kept: list[BaseMessage] = []
        used = 0
        for msg in messages:
            t = count_tokens(msg.content, encoder)
            if used + t <= budget:
                kept.append(msg)
                used += t
        messages = kept

    tokens_used = sum(count_tokens(m.content, encoder) for m in messages)
    return messages, tokens_used
\end{lstlisting}

Two decisions in Listings~\ref{lst:tokenizer} and~\ref{lst:trim-strategy}
deserve explanation.

First, the \code{oldest\_first} strategy drops messages in pairs of two,
not individually. A lone assistant message without its preceding user
message is meaningless context --- it answers a question the model can
no longer see. Dropping in pairs preserves coherence across the remaining
history. If the list has an odd length (which can occur after a
mid-session recovery), the strategy drops a single message at the
boundary rather than raising.

Second, \code{trim\_to\_budget} returns \code{(trimmed, tokens\_used)}
rather than just the trimmed list. The \code{tokens\_used} return value
is written into \code{\_slot\_token\_counts} in the assembly node
(\S\ref{sec:context-assembly-node}), which is then passed to Langfuse as
span metadata. Without the per-slot count at the call site, the Langfuse
span would require a second full encoding pass to reconstruct it.

With measurement and trimming in place, the final question is where in
the graph these operations live and who calls them in what order.

%% ------------------------------------------------------------
\section{Prompt Construction Is a Graph Node}
\label{sec:context-assembly-node}
%% ------------------------------------------------------------

The tempting implementation is to build the prompt inside the Reason
node: read state, format a string, call the model. That works in
development. In production it creates an untestable coupling between
context assembly logic and model invocation. When the assembled prompt
is wrong --- wrong slot order, incorrect trim, a goal that was evicted
--- there is no way to test the assembly in isolation without also
invoking the model. There is no Langfuse span for the assembly step, so
there is no per-slot token count in the trace. There is no way to swap
the assembly strategy without touching the model call.

The correct design is a dedicated \code{context\_assembly\_node}: a pure
\code{(state, config) -> dict} function that reads six sources from
\code{AgentState}, fills six slots in trim-priority order, enforces
per-slot budgets, assembles a \code{ChatPromptTemplate}, and writes the
result back to \code{AgentState.assembled\_prompt}. The Reason node reads
\code{assembled\_prompt} and never calls the template directly. Assembly
and reasoning are independently testable, independently observable, and
independently replaceable.

\begin{notebox}
Trim order is structural, not conventional. \code{HISTORY} is trimmed
first because it is the largest slot and the least semantically critical
for the current cycle. \code{EVIDENCE} is trimmed second because
relevance scoring can guide what to drop. \code{TOOLS} is trimmed third
because the tool list rarely changes mid-run. \code{ROLE} and
\code{ACTIVE\_GOAL} are never trimmed. Changing the order changes the
agent's behaviour under pressure; treat this ordering as a design
decision that must be documented and tested.
\end{notebox}

\begin{lstlisting}[language=Python,
  caption={ContextAssemblyNode: pure state transform with Langfuse instrumentation},
  label={lst:context-assembly}]
# src/agent/nodes/context_assembly.py
from langchain_core.prompts import ChatPromptTemplate
from langfuse.decorators import observe
from langgraph.types import RunnableConfig
from src.agent.state import AgentState
from src.agent.prompt.slot_budget import SlotName, DEFAULT_BUDGET
from src.agent.prompt.tokenizer import get_encoder, trim_to_budget, SlotOverflowError
from src.agent.prompt.goal_summary import GoalSummary


class ContextWindowExceededError(RuntimeError):
    """Raised after all permissible trimming still leaves the prompt
    over budget. This is a configuration error, not a retriable condition.
    """


@observe(name="context_assembly_node")
def context_assembly_node(state: AgentState, config: RunnableConfig) -> dict:
    budget  = config["configurable"].get("slot_budget") or DEFAULT_BUDGET
    encoder = get_encoder()
    counts: dict[str, int] = {}

    role_text   = state["role_prompt"]
    role_tokens = len(encoder.encode(role_text))
    if role_tokens > budget.role:
        raise SlotOverflowError(SlotName.ROLE, role_tokens, budget.role)
    counts[SlotName.ROLE.value] = role_tokens

    goal_text = GoalSummary(**state["goal_summary"]).render(
        budget=budget.active_goal, encoder=encoder
    )
    counts[SlotName.ACTIVE_GOAL.value] = len(encoder.encode(goal_text))

    history, counts[SlotName.HISTORY.value] = trim_to_budget(
        list(state.get("messages", [])), budget.history, encoder, "oldest_first"
    )
    evidence, counts[SlotName.EVIDENCE.value] = trim_to_budget(
        state.get("evidence_messages", []), budget.evidence, encoder, "lowest_score"
    )
    tools_text = state.get("tools_schema", "")
    tools_msgs = [m for m in [tools_text] if m]
    tools_msgs, counts[SlotName.TOOLS.value] = trim_to_budget(
        tools_msgs, budget.tools, encoder, "oldest_first"
    )

    total = sum(counts.values())
    if total > budget.context_window - budget.headroom:
        raise ContextWindowExceededError(
            f"Prompt uses {total} tokens; limit after headroom is "
            f"{budget.context_window - budget.headroom}."
        )

    prompt = ChatPromptTemplate.from_messages([
        ("system", role_text),
        ("human",  goal_text),
        *[(m.type, m.content) for m in evidence],
        *[(m.type, m.content) for m in tools_msgs],
        *[(m.type, m.content) for m in history],
    ])
    return {
        "assembled_prompt":   prompt.format_messages(),
        "_slot_token_counts": counts,
    }
\end{lstlisting}

Three decisions in Listing~\ref{lst:context-assembly} deserve
explanation.

First, the \code{slot\_budget} is read from
\code{config["configurable"]} rather than imported as a module-level
constant. This means a test can pass a \code{SlotBudget} with a
2\,000-token context window and 100-token slots without patching any
global state. Chapter~9's isolation tests exercise this exact pattern to
verify trim behaviour deterministically on small fixture states.

Second, \code{ContextWindowExceededError} is a \code{RuntimeError}, not
a \code{ValueError}. It signals a misconfiguration --- the operator has
set slot allocations that cannot fit in the declared model context window
--- that should crash the graph at startup or on first invocation, not
be silently retried. A retry would produce the same error on every cycle.

Third, the \code{\_slot\_token\_counts} dict is written to state with a
leading underscore by convention. Chapter~23's Langfuse instrumentation
reads this field from the node's output dict and attaches it as span
metadata automatically, without any additional encoding pass inside the
observability layer.

The assembly node composes the prompt correctly. The next constraint is
less obvious: the order of content within the prompt has latency and
cost implications at the provider level, because of how the KV cache
operates across consecutive calls.

%% ------------------------------------------------------------
\section{Stable Prefixes Preserve KV-Cache Reuse}
\label{sec:kv-cache}
%% ------------------------------------------------------------

Every major LLM provider caches the key-value representations of prompt
tokens across calls. When the first \emph{N} tokens of two consecutive
calls are byte-identical --- the same characters in the same order ---
the provider can reuse the cached KV representation for those tokens
rather than recomputing them. The latency saving is proportional to the
size of the reused prefix: a 512-token system message that is reused
saves roughly 512 token-generations of compute on every cycle, which at
scale translates directly to latency reduction and cost reduction.

KV-cache reuse is broken by a single character change in the prefix. If
a timestamp, a run ID, a session variable, or a username is
interpolated into the system message, the prefix changes on every call
and the cache is useless. This is a common mistake because system
prompts are often written as templates with dynamic fields, and the
temptation to personalise the role message is strong. The engineering
rule is absolute: the \code{ROLE} slot must be a compile-time constant.
All dynamic content --- goal state, history, evidence --- lives in the
human message, after the cache boundary.

\begin{warningbox}
Tool descriptions injected into the \code{ROLE} slot are a frequent
source of cache instability. Tool descriptions belong in the
\code{TOOLS} slot, which is in the human message. A role message that
includes dynamically-selected tool descriptions will change its hash on
every cycle where the available tool set changes, destroying cache reuse
across the entire prefix.
\end{warningbox}

\begin{notebox}
The stability rule applies to the \emph{rendered} string, not to the
template source. A \code{ChatPromptTemplate} that contains only static
fields in its system message is safe. A template that interpolates any
\code{AgentState} field into the system message violates the rule, even
if the interpolated value happens to be identical across cycles in
practice. The only safe test is to render the system message twice with
different state instances and assert byte equality.
\end{notebox}

\begin{lstlisting}[language=Python,
  caption={assert\_stable\_prefix with difflib diagnostic and CachePrefixUnstableError},
  label={lst:cache-guard}]
# src/agent/prompt/cache_guard.py
from __future__ import annotations
import difflib
from src.agent.prompt.slot_budget import SlotName


class CachePrefixUnstableError(AssertionError):
    def __init__(self, diff_at_char: int, slot: SlotName) -> None:
        super().__init__(
            f"Cache prefix for slot {slot.value!r} diverges at "
            f"character offset {diff_at_char}. Dynamic content has "
            f"leaked into the stable prefix."
        )
        self.diff_at_char = diff_at_char
        self.slot         = slot


def assert_stable_prefix(
    render_a: str,
    render_b: str,
    slot: SlotName = SlotName.ROLE,
) -> None:
    """Raise CachePrefixUnstableError if the two renders differ.

    Called inside context_assembly_node only when LLMSettings.debug=True
    so that the overhead of two render passes is never paid in production.
    Always called by Chapter 28's CI template-change regression test.
    """
    if render_a == render_b:
        return
    for opcode in difflib.SequenceMatcher(None, render_a, render_b).get_opcodes():
        if opcode[0] != "equal":
            raise CachePrefixUnstableError(diff_at_char=opcode[1], slot=slot)
\end{lstlisting}

Two decisions in Listing~\ref{lst:cache-guard} deserve explanation.

First, the \code{difflib.SequenceMatcher} diagnostic reports the exact
character offset where the two renders diverge. Without it, a
\code{CachePrefixUnstableError} is a binary pass/fail with no
information about what leaked into the prefix. With it, the engineer
sees immediately that the divergence is at character offset 412 ---
which maps to the timestamp field that was incorrectly interpolated into
the role template --- and can fix the template without a guessing loop.

Second, \code{assert\_stable\_prefix} is guarded by
\code{LLMSettings.debug} in the assembly node. The check requires two
render passes of the system message. In production, that overhead is
paid on every cycle for zero benefit; the system message does not change
between cycles. In CI, the check is always executed by the
template-change test described in Chapter~28, and by the linting step
in Chapter~13 that runs on every template modification.

The system message is now stable. The remaining question is where the
content of that system message comes from, and how to ensure it does not
change silently between deployments.

%% ------------------------------------------------------------
\section{A Template Version Is a Dependency: Pin It}
\label{sec:prompt-hub}
%% ------------------------------------------------------------

Pulling a prompt template from a registry without pinning its version is
the same mistake as writing \code{pip install langchain} without a
version specifier. The template can change between deployments with no
code diff, no review, and no rollback path. A model tuned against one
template version silently receives a different one at the next deploy.
The regression appears in evaluation metrics weeks later, after many
confounding changes have been made, and the root cause is expensive to
isolate.

LangSmith Prompt Hub supports version pinning through commit hashes.
Calling \code{hub.pull("prod-agent-role:a3f9c2d")} always returns the
exact template at that commit, regardless of what the \code{latest}
pointer resolves to. The commit hash must live in
\code{LLMSettings.prompt\_commit\_hash} --- not hardcoded in the call
site. A call site that embeds the hash directly is impossible to update
across environments without a code change. For A/B testing, the
\code{configurable} field in \code{RunnableConfig} carries a
\code{prompt\_variant} hash that selects between two pinned template
versions at invocation time, with no code change and no global state
mutation.

\begin{lstlisting}[language=Python,
  caption={PromptPinConfig, load\_pinned\_prompt with hub fallback, and A/B variant selection},
  label={lst:prompt-hub}]
# src/agent/prompt/hub.py
from __future__ import annotations
from functools import lru_cache
from pathlib import Path
from pydantic import BaseModel, SecretStr
from langchain import hub
from langchain_core.prompts import ChatPromptTemplate


class PromptPinConfig(BaseModel):
    model_config = {"frozen": True}
    name:          str
    commit_hash:   str
    fallback_path: Path | None = None


@lru_cache(maxsize=16)
def load_pinned_prompt(
    config:            PromptPinConfig,
    langsmith_api_key: SecretStr,
) -> ChatPromptTemplate:
    """Load a version-pinned prompt from LangSmith Hub.

    Falls back to config.fallback_path if the Hub is unreachable.
    lru_cache ensures the network round-trip is made once per process.
    """
    ref = f"{config.name}:{config.commit_hash}"
    try:
        return hub.pull(ref, api_key=langsmith_api_key.get_secret_value())
    except Exception:
        if config.fallback_path and config.fallback_path.exists():
            return ChatPromptTemplate.from_file(str(config.fallback_path))
        raise


def select_prompt_variant(
    base:         PromptPinConfig,
    variant_hash: str | None,
) -> PromptPinConfig:
    """Return base with commit_hash replaced if variant_hash is set.

    Reads from config["configurable"].get("prompt_variant") in the
    assembly node. Returns base unchanged when variant_hash is None.
    """
    if variant_hash is None:
        return base
    return base.model_copy(update={"commit_hash": variant_hash})
\end{lstlisting}

Three decisions in Listing~\ref{lst:prompt-hub} deserve explanation.

First, \code{load\_pinned\_prompt} is wrapped in \code{@lru\_cache}.
The arguments are \code{PromptPinConfig} (a frozen Pydantic model,
therefore hashable) and \code{SecretStr} (also hashable). The cache
ensures the LangSmith Hub is queried once at graph construction time;
subsequent invocations in the same process pay no network cost. In tests,
\code{load\_pinned\_prompt.cache\_clear()} resets the cache between
cases that use different commit hashes.

Second, the fallback path is a local \code{.langchain/} file checked
into the repository. Its purpose is to allow the graph to start up
during a LangSmith outage without blocking the deployment pipeline. The
fallback template contains no secrets --- only the prompt text --- so
checking it in is safe. Chapter~13's CI step validates that
\code{prompt\_commit\_hash} in \code{.env.production} resolves to a real
Hub commit before any deployment proceeds, catching stale pins before
they reach production.

Third, \code{select\_prompt\_variant} returns a new
\code{PromptPinConfig} rather than mutating the base config. The base
config is a shared constant loaded at graph construction time; the
variant config is a short-lived object that exists only for the duration
of a single graph invocation. Chapter~23's Langfuse instrumentation
records the \code{commit\_hash} from whichever config was used so that
offline analysis can attribute quality differences to specific template
versions with precision.

The template is now stable, versioned, and loaded efficiently. The final
structural question is whether the model, across many cycles, continues
to work on the right goal --- or whether the original intent has drifted
out of the context window's effective attention range.

%% ------------------------------------------------------------
\section{The Goal Must Survive Every Cycle}
\label{sec:goal-summary}
%% ------------------------------------------------------------

The most insidious form of context degradation is not a token overflow
that raises an exception. It is silent goal drift: the original objective
is technically present in the context window but has been pushed so far
back by accumulated history and evidence that the model's attention
effectively ignores it. The agent continues taking actions, calling
tools, and producing outputs --- all coherent locally, all disconnected
from the original intent. By the time the drift is visible in outputs,
many cycles and significant budget have been wasted.

The solution is not to make the system prompt longer. System prompt
verbosity does not guarantee that a long-context model anchors on goal
state at the end of a 40-cycle run. The solution is structural: a
compact, deterministic \code{GoalSummary} object is re-injected into the
\code{ACTIVE\_GOAL} slot on every cycle, positioned immediately after the
system message at the highest-priority position available in the human
message. It is not history; it is not evidence; it is the agent's
mandate, restated precisely at the beginning of every cycle's variable
content.

\code{GoalSummary} is a Pydantic model with four fields.
\code{original\_goal} is never truncated. \code{current\_step} is
updated by the Reflect node after each cycle; Chapter~1's
\code{reflect\_decision} maps to a step label and calls
\code{update\_step} before writing state back. \code{constraints} is a
\code{list[str]} where each entry is one sentence; the \code{render()}
method trims this list by priority to fit the slot. \code{priority} is a
\code{Literal["critical","high","normal"]} that governs whether the last
constraint is partially truncated or silently dropped when the slot is
tight. Chapter~2 shows how to include the termination policy as a
constraint: \code{"Cost limit: \$2.00 maximum per run"} becomes a
first-class constraint that the model sees on every cycle.

\begin{lstlisting}[language=Python,
  caption={GoalSummary with priority-aware render() and update\_step},
  label={lst:goal-summary}]
# src/agent/prompt/goal_summary.py
from __future__ import annotations
from typing import Literal
from pydantic import BaseModel
import tiktoken
from src.agent.prompt.tokenizer import count_tokens, SlotOverflowError
from src.agent.prompt.slot_budget import SlotName


class GoalSummary(BaseModel):
    original_goal: str
    current_step:  str
    constraints:   list[str] = []
    priority:      Literal["critical", "high", "normal"] = "normal"

    def render(self, budget: int, encoder: tiktoken.Encoding) -> str:
        base        = f"Goal: {self.original_goal}\nStep: {self.current_step}"
        base_tokens = count_tokens(base, encoder)
        if base_tokens > budget:
            raise SlotOverflowError(SlotName.ACTIVE_GOAL, base_tokens, budget)
        remaining = budget - base_tokens
        kept: list[str] = []
        for constraint in self.constraints:
            t = count_tokens(constraint, encoder)
            if remaining - t >= 0:
                kept.append(constraint)
                remaining -= t
            elif self.priority == "critical":
                # Partial truncation: a visible fragment is better than silence.
                words, partial = constraint.split(), ""
                for word in words:
                    candidate = (partial + " " + word).strip()
                    if count_tokens(candidate, encoder) <= remaining:
                        partial = candidate
                    else:
                        break
                if partial:
                    kept.append(partial + " [truncated]")
                break
            else:
                break   # drop remaining constraints for high/normal priority
        parts = [base]
        if kept:
            parts.append("Constraints:\n" + "\n".join(f"- {c}" for c in kept))
        return "\n".join(parts)

    def update_step(self, new_step: str) -> "GoalSummary":
        return self.model_copy(update={"current_step": new_step})
\end{lstlisting}

Three decisions in Listing~\ref{lst:goal-summary} deserve explanation.

First, \code{original\_goal} is never truncated under any priority
level. If the original goal text alone exceeds the \code{ACTIVE\_GOAL}
slot budget, \code{render()} raises \code{SlotOverflowError}. This is
intentional: a truncated goal is strictly worse than an error, because
the model would silently act on a fragment of the objective. The correct
fix is to increase the \code{active\_goal} allocation in
\code{SlotBudget} for goals that are genuinely long, or to compress the
goal text before constructing \code{GoalSummary}.

Second, the \code{critical} priority path attempts word-level truncation
of the last constraint rather than dropping it entirely. The reasoning
is that even a partial constraint --- ``Cost limit: \$2.00 per
[truncated]'' --- is more useful to the model than silence, because it
signals that a constraint exists even if its full text is unavailable.
The \code{[truncated]} suffix is a deliberate signal to the model, not a
hide-the-overflow artefact. Chapter~9's unit tests assert that a
\code{critical}-priority \code{GoalSummary} never silently drops its
first constraint, even at extreme slot pressure.

Third, \code{update\_step} returns \code{model\_copy(update=\{...\})}
rather than mutating \code{self}. \code{GoalSummary} instances live in
\code{AgentState}, which is a \code{TypedDict}. State in LangGraph is
replaced on write, not mutated in place; returning a new instance is the
correct pattern and avoids subtle bugs when the same \code{GoalSummary}
object is referenced from a checkpoint snapshot loaded by the recovery
node described in Chapter~3.

Chapter~7 extends this foundation into the full short-term memory layer:
\code{MessagesState}, \code{trim\_messages()}, and a summarisation node
that compresses accumulated \code{HISTORY} into a compact form before it
would otherwise saturate the slot. The context assembly pipeline built
here is the substrate that the memory layer writes into --- and the
\code{HISTORY} slot trim strategy built in \S\ref{sec:tokenizer} is the
first line of defence before summarisation ever fires.
