%% ============================================================
%%  Chapter 13 --- Semantic Caching
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%%  Dependencies: langchain-community >= 0.3.0,
%%                langchain-core >= 0.3.0,
%%                redis >= 5.0.0,
%%                prometheus-client >= 0.20.0,
%%                langfuse >= 2.35.0,
%%                pydantic-settings >= 2.3.0
%% ============================================================

\chapterquote{The cheapest LLM call is the one you never make.}{anon}

The production failure this chapter prevents is silent and cumulative. An
agent that calls an LLM for every incoming query pays the full inference
cost on each one, including queries it has answered correctly dozens of
times before. A customer-support assistant serving 50\,000 daily queries
with a 40\% repeat-question rate is spending \$0.02 per duplicate call
--- \$400 per day, \$146\,000 per year --- on computation whose result is
already known. The failure is not a crash; it is a margin leak that
compounds invisibly until someone examines the billing dashboard. The
second failure mode is subtler: when semantic caching is added without a
calibrated similarity threshold, the cache begins returning responses to
queries that are merely related, not identical, to the cached prompt. An
agent that answers a question about GPT-4o pricing with a cached response
from a question about GPT-3.5 pricing has not saved cost --- it has
amplified a stale answer at scale. Both failures share the same root
cause: caching was treated as an optimisation to add later rather than a
correctness constraint to design from the start.

The solution this chapter builds has four layers. The first is a
\emph{threshold layer}: a \code{RedisSemanticCache} configured with a
calibrated \code{score\_threshold} that controls how similar two queries
must be before the cached response is considered valid. The second is a
\emph{tier layer}: an exact Redis string cache checked before the
semantic cache, so that byte-identical queries always receive a
guaranteed-fresh response without incurring embedding overhead. The third
is an \emph{invalidation layer}: a tool-result hash scheme that prevents
tool-dependent LLM responses from being cached under prompt keys that
cannot reflect changes in the tool's output. The fourth is an
\emph{observability layer}: per-tier Prometheus counters and Langfuse
events that make hit rate a queryable signal rather than a guessed
property.

The sections must be read in order because each one closes a gap the
previous one left open. Section~\ref{sec:cache-threshold} establishes the
semantic cache and the threshold contract; without it, there is no cache
to query. Section~\ref{sec:cache-tiers} adds the exact cache tier and
defines the lookup order that makes the system correct; the two-tier
lookup function it produces becomes the call site that all later sections
instrument. Section~\ref{sec:cache-invalidation} handles the case the
first two sections cannot: tool-dependent responses whose staleness is
invisible to any similarity metric. Section~\ref{sec:cache-metrics}
closes the chapter by making the behaviour of all three prior sections
observable and tunable without log parsing.

This chapter introduces five terms defined in Table~\ref{tab:ch13-terms}.

\begin{table}[h]
\centering
\caption{Domain terms introduced in Chapter~13}
\label{tab:ch13-terms}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
\code{score\_threshold} &
  Cosine \emph{similarity} parameter on \code{RedisSemanticCache};
  higher values are stricter. The library default is \code{0.2}; the
  production minimum recommended here is \code{0.9}. Below \code{0.85}
  false hits become measurable; above \code{0.98} the cache degrades
  toward exact matching. \\
\code{CacheLookupResult} &
  \code{TypedDict} with \code{hit: bool}, \code{tier: str}
  (\code{"exact"} \textbar{} \code{"semantic"} \textbar{} \code{"miss"}),
  and \code{value: str \textbar{} None}; the unified return type
  across both cache tiers and all callers. \\
\code{CacheInvalidationEvent} &
  Pydantic model emitted when a tool result hash changes; carries
  \code{affected\_cache\_keys} and triggers immediate Redis key
  deletion via \code{client.delete()}. \\
\code{CacheMetricsEmitter} &
  Class that wraps \code{prometheus\_client.Counter} increments and
  \code{langfuse.event()} calls behind a single \code{emit()} method;
  decouples metric emission from cache lookup logic. \\
Two-tier cache &
  Exact Redis string cache (TTL 1\,h) checked before
  \code{RedisSemanticCache}; the \code{tier} field on every
  \code{CacheLookupResult} determines which Prometheus label is
  incremented. \\
\bottomrule
\end{tabularx}
\end{table}

The source layout for this chapter is:

\begin{verbatim}
src/
  cache/
    settings.py       # §13.4 -- CacheSettings (BaseSettings)
    semantic_cache.py # §13.1 -- build_semantic_cache()
    exact_cache.py    # §13.2 -- exact_cache_get/set,
                      #          two_tier_lookup, CacheLookupResult
    tool_cache.py     # §13.3 -- tool_result_cache_key,
                      #          check_and_invalidate,
                      #          CacheInvalidationEvent
    metrics.py        # §13.4 -- CacheMetricsEmitter,
                      #          CACHE_HITS, CACHE_MISSES
\end{verbatim}


%% ------------------------------------------------------------
\section{Score Threshold Is a Safety Parameter}
\label{sec:cache-threshold}
%% ------------------------------------------------------------

The design problem is that \code{RedisSemanticCache} ships with a
\code{score\_threshold} default of \code{0.2} --- a value low enough to
admit cache hits for queries that are merely topically related rather
than semantically equivalent. At \code{0.2}, a cosine similarity of just
20\% is enough to return a cached response: two prompts that share a
handful of domain-specific tokens but ask entirely different questions
will regularly collide in the cache. The naive approach --- deploying the
cache with default settings and trusting that the library chose a safe
default --- is wrong because the library chose a default optimised for
recall, not for correctness. The governing principle is that
\code{score\_threshold} must be understood as a correctness control and
set explicitly to a value whose safety can be reasoned about, not
inherited from a default whose implications have not been examined.

To reason about the threshold, it helps to state what cosine similarity
measures. Two prompt embeddings at cosine similarity 0.9 are within
approximately 26\textdegree{} of each other in the embedding space. At
that angle, paraphrases of the same question cluster tightly: ``What is
the refund policy?'' and ``How do I get a refund?'' are typically at
similarity 0.92--0.96 for a sentence-level encoder. At similarity 0.85,
related but distinct questions begin to cluster: ``What is the return
window?'' and ``What is the refund policy?'' may land at 0.87. Below
0.85, false hits --- where a cached response for one question is served
for a genuinely different question --- become measurable in A/B tests on
production traffic. This phenomenon is \emph{semantic hallucination
amplification}: the cache serves a plausible but wrong answer at cache
speed, with no indication that a cache hit occurred and no retry
mechanism to catch it. The production minimum recommended here is
\code{score\_threshold=0.9}. Above \code{0.98}, the threshold approaches
exact matching and the semantic hit rate collapses to near zero --- the
cache consumes Redis memory and embedding compute with no return.

\begin{notebox}
\code{RedisSemanticCache} is available from two packages.
\code{langchain\_community.cache.RedisSemanticCache} is the stable,
confirmed path (\texttt{langchain-community >= 0.3}). The newer
\texttt{langchain-redis} package (\texttt{>= 0.1}) re-exports a
\code{RedisSemanticCache} with an updated \code{distance\_threshold}
parameter (cosine \emph{distance} rather than similarity --- lower is
stricter). If your project uses \texttt{langchain-redis}, replace
\code{score\_threshold=0.9} with \code{distance\_threshold=0.1}
throughout --- the two values are equivalent (\(1 - 0.9 = 0.1\)).
All listings in this chapter use the \texttt{langchain-community} path
because its constructor signature is confirmed by the live API reference.
\code{LangCacheSemanticCache}, also in \texttt{langchain-redis}, is a
managed cloud alternative excluded here because it introduces an external
API dependency incompatible with the self-hosted deployment pattern this
book targets.
\end{notebox}

A second constraint governs the embedding model passed to
\code{RedisSemanticCache}: it must be identical to the model used by the
Qdrant vector store introduced in Chapter~10. The cache and the retrieval
pipeline operate in the same embedding space only if they use the same
encoder. If they differ, a query that the cache considers a near-miss may
be a strong retrieval hit in Qdrant's space, and vice versa. The
\code{score\_threshold} guarantee holds only within a single, consistent
embedding space. The enforcement mechanism is a shared \code{embeddings}
singleton injected from \code{CacheSettings}
(Section~\ref{sec:cache-metrics}) into both the cache factory and the
Qdrant client factory --- the same object reference, not two instances
of the same class.

\begin{warningbox}
LiteLLM-proxied completion calls do \emph{not} go through LangChain's
\code{set\_llm\_cache()} global. \code{set\_llm\_cache()} is a
side-effect on \code{langchain\_core.globals} that intercepts only
\code{ChatModel.invoke()} calls made through the LangChain abstraction.
For LiteLLM, caching is configured separately via
\code{litellm.cache = Cache(type="redis", host=..., port=...,} \\
\code{password=...)} before any \code{litellm.completion()} call. Both
must be configured. A system that sets only \code{set\_llm\_cache()} and
uses LiteLLM as its completion backend will have zero cache hits and no
error to explain why.
\end{warningbox}

With the threshold contract established, the factory function for the
semantic cache encodes the safety constraint directly. A misconfigured
\code{score\_threshold} supplied via environment variable causes an
immediate \code{AssertionError} during the application boot sequence ---
before any request is served --- rather than silently degrading
correctness for hours.

\begin{lstlisting}[language=Python,
  caption={Semantic cache factory with threshold assertion},
  label={lst:semantic-cache}]
# src/cache/semantic_cache.py
import litellm
from langchain_community.cache import RedisSemanticCache
from langchain_core.embeddings import Embeddings
from langchain_core.globals import set_llm_cache
from litellm.caching import Cache

def build_semantic_cache(
    redis_url: str,
    embeddings: Embeddings,
    score_threshold: float = 0.9,
) -> RedisSemanticCache:
    assert score_threshold >= 0.85, (
        f"score_threshold={score_threshold} risks semantic false hits; "
        "minimum safe value is 0.85."
    )
    cache = RedisSemanticCache(
        redis_url=redis_url,
        embedding=embeddings,
        score_threshold=score_threshold,
    )
    set_llm_cache(cache)
    litellm.cache = Cache(
        type="redis",
        host=redis_url.split("://")[1].split(":")[0],
        port=int(redis_url.split(":")[-1].split("/")[0]),
    )
    return cache
\end{lstlisting}

Three decisions in Listing~\ref{lst:semantic-cache} deserve explanation.
The \code{assert} fires at process startup, not at cache-query time,
because a wrong \code{score\_threshold} delivered via a misconfigured
\code{SEMANTIC\_SCORE\_THRESHOLD} environment variable would otherwise
produce plausible-looking responses for hours before anyone noticed the
elevated false-hit rate in Langfuse. The \code{set\_llm\_cache(cache)}
call registers the \code{RedisSemanticCache} instance globally for all
LangChain \code{ChatModel} calls; the separate \code{litellm.cache =
Cache(...)} registration targets LiteLLM-proxied calls --- the two
registrations cover different call paths on the same Redis backend and
are not redundant. The \code{redis\_url} parsing is intentionally naive:
in production, \code{CacheSettings} (Section~\ref{sec:cache-metrics})
provides \code{host} and \code{port} as separate fields so this
string-splitting is replaced by direct field access.

The semantic cache is now in place and correctly bounded. The problem it
cannot solve is lookup order: a semantic cache checked first will return
a near-match response even when the byte-identical prompt has a cached
response that is guaranteed correct, wasting embedding compute and
potentially serving an inferior answer. That ordering problem is what the
next section resolves.


%% ------------------------------------------------------------
\section{Lookup Order Is a Correctness Decision}
\label{sec:cache-tiers}
%% ------------------------------------------------------------

The design problem is that a semantic cache checked first will return a
near-match response even when the byte-identical prompt has a cached
response that is guaranteed correct. The near-match may be very close ---
similarity 0.97 --- but it is not the same prompt. An agent that answered
``What is the refund policy for order \#1042?'' should not have its
response served for ``What is the refund policy for order \#1043?'' even
if the two questions embed at similarity 0.97, because order numbers are
identity keys, not semantic content. The naive two-cache system that
checks the semantic cache first has no mechanism to detect this. The
governing principle is that the exact cache is checked first because an
exact key match is the only match that carries a \emph{correctness}
guarantee, and correctness takes precedence over throughput.

The exact cache uses a thin wrapper over \code{redis.Redis}: the key is
\code{"exact\_cache:" + hashlib.sha256(prompt.encode()).hexdigest()},
and the value is the JSON-serialised response stored with \code{SETEX}.
The SHA-256 key is deterministic across process restarts, collision-resistant
at any realistic corpus size, and --- critically --- does not include the
raw prompt text. Storing the raw prompt as a Redis key creates a memory
leak proportional to prompt length on every unique query, bounded only
by the TTL or by Redis eviction policy. The TTL for the exact cache tier
is 3\,600 seconds (1\,hour): short enough to ensure freshness for
high-frequency repeated queries, long enough to absorb traffic spikes
without every request hitting the LLM.

Serialisation uses \code{json.dumps} exclusively. Pickle is excluded
because it creates a deserialisation code-execution vulnerability if the
Redis instance is reachable from outside the private network. An attacker
with write access to Redis can craft a pickle payload that executes
arbitrary code when any process calls \code{redis.Redis.get()}. JSON has
no equivalent attack surface: a malformed JSON value raises
\code{json.JSONDecodeError}, not a shell.

\begin{lstlisting}[language=Python,
  caption={CacheLookupResult and exact cache read/write operations},
  label={lst:exact-cache}]
# src/cache/exact_cache.py
import hashlib, json
import redis as _redis
from typing_extensions import TypedDict

class CacheLookupResult(TypedDict):
    hit:   bool
    tier:  str          # "exact" | "semantic" | "miss"
    value: str | None

def _exact_key(prompt: str, prefix: str) -> str:
    digest = hashlib.sha256(prompt.encode()).hexdigest()
    return f"{prefix}:{digest}"

def exact_cache_get(
    client: _redis.Redis,
    prompt: str,
    prefix: str = "exact_cache",
) -> CacheLookupResult:
    raw = client.get(_exact_key(prompt, prefix))
    if raw is not None:
        return CacheLookupResult(
            hit=True, tier="exact", value=raw.decode()
        )
    return CacheLookupResult(hit=False, tier="miss", value=None)

def exact_cache_set(
    client: _redis.Redis,
    prompt: str,
    response: str,
    ttl: int = 3600,
    prefix: str = "exact_cache",
) -> None:
    client.setex(
        _exact_key(prompt, prefix), ttl, json.dumps(response)
    )
\end{lstlisting}

Two decisions in Listing~\ref{lst:exact-cache} deserve explanation.
\code{\_exact\_key()} is extracted as a private helper rather than
inlined in both \code{exact\_cache\_get} and \code{exact\_cache\_set}
because the key construction logic must be identical on read and write
--- any divergence between the two (a different prefix, a different
encoding) produces a permanent cache miss with no error to diagnose. A
shared function makes the identity guarantee structural. \code{raw.decode()}
converts the bytes returned by \code{redis.Redis.get()} to a Python
string because \code{redis-py} returns \code{bytes} by default. Using
\code{decode\_responses=True} on the client constructor is an
alternative, but it applies globally to all keys and can break binary
key consumers sharing the same client instance.

With the exact cache operations defined, the two-tier lookup function
enforces the ordering contract --- exact first, semantic second --- and
returns a \code{CacheLookupResult} whose \code{tier} field records which
layer was responsible for the hit or miss. This \code{tier} field is the
single piece of state that the observability layer in
Section~\ref{sec:cache-metrics} needs to route to the correct Prometheus
label.

\begin{lstlisting}[language=Python,
  caption={Two-tier lookup: exact then semantic, unified return type},
  label={lst:two-tier-lookup}]
# src/cache/exact_cache.py (continued)
from langchain_community.cache import RedisSemanticCache

def two_tier_lookup(
    client: _redis.Redis,
    semantic_cache: RedisSemanticCache,
    prompt: str,
    llm_string: str,
) -> CacheLookupResult:
    exact = exact_cache_get(client, prompt)
    if exact["hit"]:
        return exact

    generations = semantic_cache.lookup(prompt, llm_string)
    if generations:
        # Chat models return ChatGeneration; use .message.content
        first = generations[0][0] if generations[0] else None
        value = (
            first.message.content
            if hasattr(first, "message")
            else first.text
            if first is not None else None
        )
        return CacheLookupResult(
            hit=True, tier="semantic", value=value
        )

    return CacheLookupResult(hit=False, tier="miss", value=None)
\end{lstlisting}

Three decisions in Listing~\ref{lst:two-tier-lookup} deserve explanation.
The \code{llm\_string} parameter is required by LangChain's cache API:
it namespaces cache entries by model identifier so that a GPT-4o response
is never served for a GPT-4o-mini query with an identical prompt.
\code{semantic\_cache.lookup()} returns \code{list[list[Generation]] |
None}; the outer list corresponds to prompt variants, the inner to
generation candidates. The \code{hasattr(first, "message")} branch
handles the distinction between \code{ChatGeneration} (chat models, which
carry \code{.message.content}) and \code{Generation} (completion models,
which carry \code{.text}): both are valid return types depending on which
LangChain model class populated the cache entry, and accessing the wrong
attribute raises a silent \code{AttributeError} at runtime. The
\code{None} guard on \code{generations} is necessary because
\code{lookup()} returns \code{None} on a miss, not an empty list.

\begin{warningbox}
In async LangGraph nodes that call \code{exact\_cache\_get} or
\code{exact\_cache\_set} directly, use \code{redis.asyncio.Redis} and
\code{await client.get(key)}. The synchronous \code{redis.Redis} client
blocks the event loop when called inside \code{async def} node functions,
causing thread-pool starvation under concurrent graph execution.
\code{RedisSemanticCache} does not currently expose async lookup methods
in \texttt{langchain-community} --- wrap the \code{semantic\_cache.lookup()}
call with \code{asyncio.to\_thread()} in async contexts.
\end{warningbox}

The two-tier lookup resolves ordering and correctness for queries whose
relevant context is entirely within the prompt text. It cannot resolve
the case where the prompt is identical but the correct answer has changed
--- because the tool the agent called returned a different result. That
failure mode is what the next section addresses.


%% ------------------------------------------------------------
\section{Tool Outputs Must Not Be Cached by Prompt}
\label{sec:cache-invalidation}
%% ------------------------------------------------------------

The design problem is that tool-dependent LLM responses are stale from
the moment the underlying tool result changes, and no similarity metric
applied to the prompt can detect this. The naive approach caches the
agent's synthesised answer under the prompt key, as the first two cache
tiers do. For a query such as ``What is the current price of AAPL?'',
the prompt text is stable across time but the correct answer is not. An
exact cache hit at 14:00 for a query answered at 09:00 returns a price
that is five hours stale. A semantic cache hit is worse: it may return a
response originally generated for a different ticker that embeds close to
AAPL in financial-domain embedding space. The governing principle is that
the cache key for any response derived from a tool call must incorporate
the tool's identity and arguments, not the prompt text, and must be
invalidated \emph{immediately} when the tool's output changes.

The key construction rule is deterministic: the tool name and all
arguments are serialised to a single string, hashed with SHA-256, and
prefixed with \code{"tool\_cache:"}. The critical detail in argument
serialisation is that \code{json.dumps(tool\_args, sort\_keys=True)} must
be used --- not \code{sorted(tool\_args.items())} at the top level. The
built-in \code{sort\_keys=True} argument recurses into nested
dictionaries automatically. A top-level \code{sorted()} call sorts only
the outermost keys; a tool argument like
\code{\{"filters": \{"date": "2026-01-01", "category": "tech"\}\}}
produces a different hash depending on the insertion order of the inner
dict, which varies across Python versions and call sites. Cache key
instability on nested arguments means entries are never found on lookup,
turning the cache layer into a write-only data sink.

\begin{lstlisting}[language=Python,
  caption={Tool cache key construction and CacheInvalidationEvent model},
  label={lst:tool-cache-models}]
# src/cache/tool_cache.py
import hashlib, json
from datetime import datetime, timezone
from pydantic import BaseModel

class CacheInvalidationEvent(BaseModel):
    tool_name:            str
    tool_args:            dict
    old_result_hash:      str
    new_result_hash:      str
    invalidated_at:       datetime
    affected_cache_keys:  list[str]

def tool_result_cache_key(
    tool_name: str,
    tool_args: dict,
    prefix: str = "tool_cache",
) -> str:
    payload = tool_name + ":" + json.dumps(
        tool_args, sort_keys=True
    )
    digest = hashlib.sha256(payload.encode()).hexdigest()
    return f"{prefix}:{digest}"
\end{lstlisting}

Two decisions in Listing~\ref{lst:tool-cache-models} deserve explanation.
\code{CacheInvalidationEvent} carries both \code{old\_result\_hash} and
\code{new\_result\_hash} rather than just the new hash. A Langfuse event
that records only the new hash cannot answer ``how often does this tool's
result change?'' --- it needs both sides of the transition to measure
tool result volatility over time. Chapter~23 builds a dashboard panel
from these events. The \code{affected\_cache\_keys} field is a list
rather than a single string because a single tool call may have generated
multiple prompt-level cache entries (for example, a summary and a direct
answer both derived from the same \code{get\_stock\_price()} result); the
\code{check\_and\_invalidate} function populates this list by including
all keys whose hash has changed in a single transaction.

Invalidation uses immediate deletion via \code{client.delete()}, not TTL
expiry. TTL expiry sets a ceiling on how stale a cached response can be,
but between the last write and the TTL firing, every query served from
the cache receives the stale answer. For a tool whose output changes on
the order of seconds --- a price feed, a sensor reading, a queue depth
counter --- a one-hour TTL means up to one hour of wrong answers.
Immediate deletion on hash change closes the stale-serve window to the
latency of a single Redis \code{DEL} call, typically under one
millisecond on a co-located Redis instance.

\begin{lstlisting}[language=Python,
  caption={check\_and\_invalidate: hash comparison and immediate deletion},
  label={lst:check-invalidate}]
# src/cache/tool_cache.py (continued)
# hashlib, json, datetime, timezone imported above
import redis

_HASH_PREFIX = "tool_hash"

def check_and_invalidate(
    client: redis.Redis,
    tool_name: str,
    tool_args: dict,
    new_result: str,
) -> CacheInvalidationEvent | None:
    cache_key = tool_result_cache_key(tool_name, tool_args)
    hash_key  = tool_result_cache_key(
        tool_name, tool_args, prefix=_HASH_PREFIX
    )
    new_hash     = hashlib.sha256(new_result.encode()).hexdigest()
    old_hash_raw = client.get(hash_key)

    if old_hash_raw is None:
        client.set(hash_key, new_hash)
        return None                      # first call: no prior state

    old_hash = old_hash_raw.decode()
    if old_hash == new_hash:
        return None                      # result unchanged: cache valid

    client.delete(cache_key)
    client.set(hash_key, new_hash)
    return CacheInvalidationEvent(
        tool_name=tool_name,
        tool_args=tool_args,
        old_result_hash=old_hash,
        new_result_hash=new_hash,
        invalidated_at=datetime.now(tz=timezone.utc),
        affected_cache_keys=[cache_key],
    )
\end{lstlisting}

Three decisions in Listing~\ref{lst:check-invalidate} deserve explanation.
The hash of the tool result is stored under a separate \code{tool\_hash:}
key rather than embedded in the cached response value. This means the
hash check is a single \code{GET} on a short string key --- typically
under 64 bytes --- rather than a \code{GET} followed by deserialisation
of the full cached response. At high call rates the difference between a
64-byte \code{GET} and a multi-kilobyte JSON deserialisation per tool
invocation is measurable in Redis pipeline throughput. When
\code{old\_hash\_raw} is \code{None}, the function stores the new hash
and returns \code{None} because there is no prior cached state to
invalidate --- the first call for a given tool argument set is always a
cache population, not an invalidation. \code{datetime.now(tz=timezone.utc)}
is used rather than \code{datetime.utcnow()} because \code{utcnow()}
returns a naive datetime that carries no timezone information; a naive
UTC datetime is indistinguishable from a naive local datetime in
serialised form and produces silent bugs when timestamps are compared
across service boundaries.

The invalidation layer ensures that tool-dependent responses are never
served past the point where the tool's output changes. The remaining
problem is that none of this behaviour is visible: without structured
counters, there is no way to know whether the cache is providing value,
whether the threshold needs adjustment, or whether a specific tool is
invalidating its cache entries every five minutes. That visibility is
what the final section adds.


%% ------------------------------------------------------------
\section{Hit Rate Is a Metric, Not a Log Line}
\label{sec:cache-metrics}
%% ------------------------------------------------------------

The design problem is that a text log line saying
\code{"cache hit, tier=semantic"} is not aggregable without a log
parsing pipeline, a query language, and a retention policy. The engineer
who needs to know whether \code{score\_threshold} should be tightened
from \code{0.9} to \code{0.95} needs a \emph{rate} over a time window,
not a list of events. The naive approach --- logging cache outcomes as
structured JSON and deriving rates from log aggregation --- introduces a
dependency on the log pipeline's availability and query latency into what
should be a sub-second operational decision. The governing principle is
that cache behaviour is expressed as Prometheus counters, which are
queryable in constant time regardless of event volume, and as Langfuse
events that appear on the agent's trace timeline, giving both operational
and debugging views from a single emission point.

Two counters are used, not one. \code{cache\_hits\_total} and
\code{cache\_misses\_total} each carry a \code{cache\_tier} label with
values \code{"exact"}, \code{"semantic"}, and \code{"miss"}. Hit rate is
then expressed as a clean PromQL expression with no label-value ratio
tricks:

\[
  \frac{\texttt{rate(cache\_hits\_total[5m])}}
       {\texttt{rate(cache\_hits\_total[5m])} + \texttt{rate(cache\_misses\_total[5m])}}
\]

Adding a \code{by (cache\_tier)} clause breaks this down by tier. A
single counter with a \code{result=hit/miss} label requires a ratio
across two label values, which is valid PromQL but produces verbose alert
rule expressions that are harder to review and harder to test against a
local Prometheus instance during development.

\begin{notebox}
\code{prometheus\_client} does not initialise metrics with labels at
declaration time. A Prometheus scrape that arrives before the first cache
event will find no series at all for that label value, which causes alert
rules using \code{absent()} to fire spuriously on startup. The fix is to
call \code{.labels()} for every known label value once at module load
time, creating zero-valued series immediately so the alerting baseline is
established before any traffic arrives.
\end{notebox}

The \code{CacheMetricsEmitter} class wraps both the Prometheus increment
and the Langfuse event write behind a single \code{emit()} call. The
calling code has no direct import of either \code{prometheus\_client} or
\code{langfuse}: it receives an emitter by dependency injection and calls
\code{emit(result, prompt\_sha)}. This separation makes the emitter
testable in isolation by injecting a mock \code{Langfuse} client, and it
makes the calling code testable without a live Prometheus or Langfuse
endpoint. The \code{prompt\_sha} argument is the SHA-256 digest of the
prompt string, not the prompt itself: Langfuse event metadata must not
contain raw prompt text because it may include PII that should not appear
in the observability backend.

\begin{lstlisting}[language=Python,
  caption={Module-level counters with label pre-initialisation},
  label={lst:prom-counters}]
# src/cache/metrics.py
from prometheus_client import Counter
from langfuse import Langfuse
from src.cache.exact_cache import CacheLookupResult

CACHE_HITS = Counter(
    "cache_hits_total",
    "Cache hits by tier",
    ["cache_tier"],
)
CACHE_MISSES = Counter(
    "cache_misses_total",
    "Cache misses by tier",
    ["cache_tier"],
)

# Pre-initialise all label combinations so scrapes before the
# first cache event return zero-valued series, not absent series.
for _tier in ("exact", "semantic", "miss"):
    CACHE_HITS.labels(cache_tier=_tier)
    CACHE_MISSES.labels(cache_tier=_tier)
\end{lstlisting}

Two decisions in Listing~\ref{lst:prom-counters} deserve explanation.
Both \code{Counter} objects are declared at module level, not inside a
function or class \code{\_\_init\_\_}. \code{prometheus\_client} raises
\code{ValueError: Duplicated timeseries in CollectorRegistry} if a
\code{Counter} with the same name is registered twice in the same
process --- a mistake trivially made when the counter is instantiated
inside a factory function called more than once. Module-level declaration
makes the singleton property structural: Python's module import system
ensures the module body executes exactly once per process, regardless of
how many times the module is imported. The \code{for} loop iterates over
a fixed tuple of known tier values rather than a dynamic list, so the
pre-initialised series are always exactly the three values that the
\code{CacheLookupResult} \code{tier} field can take; adding a new tier
requires a matching entry here, and omitting it produces an absent-series
alert on the new tier's first occurrence.

\begin{lstlisting}[language=Python,
  caption={CacheMetricsEmitter: dual Prometheus and Langfuse emission},
  label={lst:cache-emitter}]
# src/cache/metrics.py (continued)

class CacheMetricsEmitter:
    def __init__(self, langfuse: Langfuse) -> None:
        self._langfuse = langfuse

    def emit(
        self,
        result: CacheLookupResult,
        prompt_sha: str,
    ) -> None:
        tier = result["tier"]
        if result["hit"]:
            CACHE_HITS.labels(cache_tier=tier).inc()
        else:
            CACHE_MISSES.labels(cache_tier=tier).inc()

        self._langfuse.event(
            name="cache_lookup",
            metadata={
                "hit":        result["hit"],
                "tier":       tier,
                "prompt_sha": prompt_sha,
            },
        )
\end{lstlisting}

Two decisions in Listing~\ref{lst:cache-emitter} deserve explanation.
The Prometheus increment uses \code{tier} from the \code{CacheLookupResult}
even for misses: the \code{tier} field is set to \code{"miss"} by
\code{two\_tier\_lookup()} when both cache layers return nothing, so the
label is always populated. Using the field value rather than a
hard-coded \code{"miss"} string means that a future cache tier --- for
example, a local in-process LRU cache added before the exact Redis check
--- is automatically instrumented by populating the \code{tier} field
correctly, with no change to \code{CacheMetricsEmitter}. The Langfuse
\code{event()} call uses \code{prompt\_sha} in metadata rather than the
raw prompt text; a SHA allows correlation of cache events to specific
queries in debugging sessions where the raw prompt is already available
from application logs, without exposing PII in the observability
platform.

The \code{CacheSettings} model and the \code{build\_all\_caches()} factory
close the configuration surface. Chapter~12 introduced \code{BaseSettings}
as a pattern that converts missing credentials into startup failures
rather than mid-request errors; the same discipline applies here.

\begin{lstlisting}[language=Python,
  caption={CacheSettings and build\_all\_caches factory},
  label={lst:cache-settings}]
# src/cache/settings.py
import redis
from langchain_community.cache import RedisSemanticCache
from langchain_core.embeddings import Embeddings
from langfuse import Langfuse
from pydantic_settings import BaseSettings
from src.cache.semantic_cache import build_semantic_cache
from src.cache.metrics import CacheMetricsEmitter

class CacheSettings(BaseSettings):
    redis_host:                  str        = "localhost"
    redis_port:                  int        = 6379
    redis_password:              str | None = None
    semantic_score_threshold:    float      = 0.9
    exact_ttl_seconds:           int        = 3600
    semantic_ttl_seconds:        int        = 86400
    prometheus_push_gateway_url: str | None = None

    model_config = {"env_file": ".env", "env_file_encoding": "utf-8"}

    @property
    def redis_url(self) -> str:
        if self.redis_password:
            return (f"redis://:{self.redis_password}"
                    f"@{self.redis_host}:{self.redis_port}")
        return f"redis://{self.redis_host}:{self.redis_port}"

def build_all_caches(
    settings: CacheSettings,
    embeddings: Embeddings,
    langfuse: Langfuse,
) -> tuple[redis.Redis, RedisSemanticCache, CacheMetricsEmitter]:
    client  = redis.Redis(host=settings.redis_host,
                          port=settings.redis_port,
                          password=settings.redis_password)
    sem     = build_semantic_cache(
                  settings.redis_url, embeddings,
                  settings.semantic_score_threshold)
    emitter = CacheMetricsEmitter(langfuse)
    return client, sem, emitter
\end{lstlisting}

Three decisions in Listing~\ref{lst:cache-settings} deserve explanation.
\code{CacheSettings} exposes \code{redis\_host}, \code{redis\_port}, and
\code{redis\_password} as separate fields rather than a single
\code{redis\_url} string, eliminating the URL-parsing fragility shown in
Listing~\ref{lst:semantic-cache}. The \code{redis\_url} property
assembles the URL from validated components and is the only place where
URL construction logic lives. \code{build\_all\_caches()} returns a typed
tuple of three objects rather than a single wrapper class: the
\code{redis.Redis} client is consumed by \code{exact\_cache\_get} and
\code{check\_and\_invalidate}; the \code{RedisSemanticCache} is consumed
by \code{two\_tier\_lookup}; the \code{CacheMetricsEmitter} is consumed
by the LangGraph node that calls \code{two\_tier\_lookup} and then emits.
A wrapper class holding all three as attributes would couple those
three call sites to a shared type none of them individually needs.
\code{semantic\_ttl\_seconds} is declared in \code{CacheSettings} even
though it is not currently passed to the \code{RedisSemanticCache}
constructor --- TTL at the \code{redisvl} index level requires index
schema configuration applied at deployment time, outside the scope of
this chapter. The field is declared now so the environment variable is
documented and the value is available when the index schema is configured.

\begin{tipbox}
Set \code{PROMETHEUS\_PUSH\_GATEWAY\_URL} only for short-lived worker
processes --- Celery tasks, one-shot warm-up scripts --- that exit before
Prometheus can scrape them. Long-running FastAPI services should mount
\code{prometheus\_client.make\_asgi\_app()} at \code{/metrics} and allow
Prometheus to scrape passively. The push gateway adds a staleness window
equal to the push interval; passive scraping avoids this window entirely.
\end{tipbox}

\begin{notebox}
Langfuse Python SDK v3 (GA June 2026) is OTEL-based and replaces the
\code{langfuse\_client.event()} call shown here with OTEL span attribute
writes. The \code{CacheMetricsEmitter} interface is unchanged --- only
the \code{Langfuse} client injected into it is swapped for the v3
client. Chapter~23 performs the full migration and builds the cache hit
rate dashboard panel from the \code{cache\_lookup} events emitted here.
\end{notebox}

Chapter~14 opens with the problem the caching layer does not solve:
the agent still calls the same \emph{tools} repeatedly when the tool's
output has not changed, paying the full tool latency and rate-limit cost
on every invocation. The tool memoisation and rate-limit coordination
patterns that address this are the opening problem of Chapter~14.
