%% ============================================================
%%  Chapter 3 — State Schema, Checkpointing & Recovery
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%% ============================================================

\chapterquote{The shortest path to a production incident is a stateful system
that does not know what state it is in.}{Attributed to Werner Vogels, adapted}

Chapter~1 established \code{AgentState} as the data contract every node reads
and writes. Chapter~2 built the termination and guard layer on top of it. Both
chapters assumed that state was always freshly initialised---no persistence, no
recovery, no versioning. That assumption collapses the moment a long-horizon task
is interrupted, a deployment is rolled forward with a schema change, or two
concurrent agent runs share the same checkpointer pool.

This chapter closes that gap. It defines three things that must be true for any
production agent whose state outlives a single process invocation: the state
schema must carry enough structure to survive serialisation and reload without
data loss; the checkpointer must match the durability and concurrency requirements
of the deployment environment; and the graph must be able to resume from a saved
checkpoint, validate its own consistency, and re-anchor the original goal before
continuing---without accepting a new goal from the caller.

The running example remains the earnings-report agent. In this chapter it
acquires a PostgreSQL-backed checkpointer, a checkpoint frequency policy, a
\code{RecoveryNode} that re-anchors the goal on resume, and a schema migration
system that ensures checkpoints written before a schema change can still be
loaded after it.

%% ------------------------------------------------------------
\section{BDI Fields in \texttt{AgentState}: The Mandatory Named Contract}
\label{sec:bdi-fields}
%% ------------------------------------------------------------

Chapter~1's \code{AgentState} carried \code{goal}, \code{beliefs}, and
\code{intentions} as typed fields, but the rationale was deferred. Now that the
checkpointer and recovery system depend on them, the design choices have concrete
consequences.

The first choice is \emph{named fields versus a generic context dict}. A generic
\code{context: dict} cannot carry reducers---LangGraph's merge model requires
each field's reduction semantics to be declared at the schema level. A generic
dict cannot be inspected by guard nodes without key-name conventions that drift
and diverge across contributors. And a generic dict serialises to JSON without
type information, making schema migration detection impossible.

Named BDI fields solve all three problems at once. \code{goal} is the desire: the
objective the agent pursues. \code{beliefs} is the belief store: an
append-only log of every observation the agent has made since the task began.
\code{intentions} is the intention list: the committed next steps derived from the
current reasoning.

The second choice is \emph{immutability of \texttt{goal}}. No node should ever
return \code{goal} in its update dict after the initial state is constructed.
Overwriting \code{goal} mid-run is the definition of goal drift---the same
phenomenon Chapter~2's divergence detector exists to catch. \code{TypedDict}
cannot enforce write-once semantics at runtime, so this is a team convention
backed by a lint rule: any \code{return \{"goal": ...\}} in a node function
outside \code{make\_initial\_state} is a reviewer error.

The third choice is \emph{append-only reducers on all list fields}. When a node
returns \code{\{"beliefs": ["new observation"]\}}, LangGraph appends to the
existing list rather than replacing it---because \code{beliefs} is annotated with
\code{Annotated[list[str], add]}. Without that annotation, the return value
replaces the entire \code{beliefs} list. That silent replacement is the most
common data-loss bug in early LangGraph state designs, and it is invisible in
unit tests that start from an empty state.

The updated schema also adds two fields required by Chapter~2's goal divergence
detector---\code{goal\_embedding} and \code{goal\_divergence\_checked\_at}---and
a \code{schema\_version} integer that the migration system in
\S\ref{sec:migration} depends on. All fields must be JSON-serialisable: no Pydantic
object instances, no dataclasses, no \code{datetime} objects. The checkpointer
serialises the full state dict; any non-primitive type will fail silently or
raise a \code{TypeError} at checkpoint time.

\begin{lstlisting}[language=Python,
  caption={\code{AgentState} v2: BDI fields, divergence fields, and schema version},
  label={lst:agentstate-v2}]
from __future__ import annotations
from operator import add
from typing import Annotated, Literal
from typing_extensions import TypedDict
from langgraph.graph.message import add_messages

class AgentState(TypedDict):
    # --- BDI core (mandatory; goal is write-once after entry) ---
    goal:             str
    beliefs:          Annotated[list[str], add]
    intentions:       Annotated[list[str], add]
    # --- resource tracking ---
    token_budget:     int
    cycle_count:      int
    cycle_traces:     Annotated[list[dict], add]
    # --- message bus (LangGraph dedup reducer) ---
    messages:         Annotated[list[dict], add_messages]
    # --- control signals ---
    last_action:      dict | None
    reflect_decision: str | None
    final_answer:     str | None
    # --- goal divergence (Ch.2 ss2.7) ---
    goal_embedding:             list[float] | None
    goal_divergence_checked_at: int | None
    # --- policy (stored as plain dict for checkpointer compat) ---
    _policy:          dict | None
    # --- recovery flag (one-shot; cleared after first post-recovery cycle) ---
    _recovery_cycle:  bool
    # --- migration key ---
    schema_version:   int
\end{lstlisting}

\code{make\_initial\_state} is the only place that sets \code{goal} and
\code{schema\_version}. It also converts the \code{TerminationPolicy} Pydantic
object to a plain dict before storing it, satisfying the JSON-serialisability
requirement.

\begin{lstlisting}[language=Python,
  caption={\code{make\_initial\_state}: single entry point for state construction},
  label={lst:make-initial-state}]
from src.agent.termination import TerminationPolicy

CURRENT_SCHEMA_VERSION = 2

def make_initial_state(
    inp: AgentStateInput,
    policy: TerminationPolicy,
) -> AgentState:
    return AgentState(
        goal=inp.goal,
        beliefs=[],
        intentions=[],
        token_budget=inp.token_budget,
        cycle_count=0,
        cycle_traces=[],
        messages=[],
        last_action=None,
        reflect_decision=None,
        final_answer=None,
        goal_embedding=None,
        goal_divergence_checked_at=None,
        _policy=policy.model_dump(),
        _recovery_cycle=False,
        schema_version=CURRENT_SCHEMA_VERSION,
    )
\end{lstlisting}

\begin{warningbox}
\code{\_policy} is stored in state as a plain \code{dict}---not as a
\code{TerminationPolicy} instance---because the checkpointer cannot serialise
Pydantic models. When any node needs the policy object, it reconstructs it with
\lstinline|TerminationPolicy.model_validate(state["_policy"])|. This is cheap and
correct. Storing the live Pydantic object instead is a common mistake that
produces silent serialisation failures on the first checkpoint write.
\end{warningbox}

%% ------------------------------------------------------------
\section{\texttt{SqliteSaver} (Dev) vs.\ \texttt{AsyncPostgresSaver} (Production)}
\label{sec:checkpointers}
%% ------------------------------------------------------------

\code{SqliteSaver} and \code{AsyncPostgresSaver} implement the same LangGraph
\code{BaseCheckpointSaver} interface, but they differ in every property that
matters in production: concurrency, durability, and connection semantics.

\code{SqliteSaver} serialises all writes through a single file lock. Under
concurrent agent runs it produces \code{OperationalError: database is locked}
errors. It has no connection pool, no replication, and no HA failover path.
None of these limitations surface in single-run tests or development notebooks,
which is exactly why they are dangerous---by the time they appear, the system is
already deployed.

\begin{warningbox}
\code{SqliteSaver} is a development and testing tool. It must never be used in a
staging or production environment that handles concurrent agent runs or requires
durability across process restarts. The rule is absolute: if the process that
wrote the checkpoint can exit before the run completes, \code{AsyncPostgresSaver}
is the minimum viable production choice.
\end{warningbox}

\code{AsyncPostgresSaver} requires an \code{asyncpg} connection pool. A pool,
not a single connection: a single connection blocks while a checkpoint write is
in flight, creating head-of-line contention across concurrent graph invocations.
The pool distributes writes across multiple server connections, bounded by the
Postgres \code{max\_connections} setting.

The pool \code{min\_size} governs warm connections. With \code{min\_size=2},
two connections are established at startup and held ready; the first checkpoint
write after cold start does not pay the connection handshake cost. The
\code{max\_size} ceiling must be calibrated against the Postgres server's
\code{max\_connections} and the number of application instances: if four
application pods each hold a pool with \code{max\_size=10}, the database must
tolerate at least 40 connections plus headroom for administrative sessions.

\code{await saver.setup()} creates the LangGraph checkpoint tables if they do
not exist. Omitting this call does not produce an error at startup---it produces a
\code{relation does not exist} error on the first checkpoint write during a
production run. The setup call is idempotent and costs one query; there is no
reason to skip it.

\begin{lstlisting}[language=Python,
  caption={Checkpointer factories: dev-only \code{SqliteSaver} and production \code{AsyncPostgresSaver}},
  label={lst:checkpointer}]
# src/agent/checkpointer.py
import asyncpg
from langgraph.checkpoint.sqlite import SqliteSaver
from langgraph.checkpoint.postgres.aio import AsyncPostgresSaver

# DEVELOPMENT AND TESTS ONLY
def make_dev_checkpointer() -> SqliteSaver:
    """In-memory SQLite. Never use outside tests."""
    return SqliteSaver.from_conn_string(":memory:")

# PRODUCTION
async def make_prod_checkpointer(
    dsn: str,
    min_size: int = 2,
    max_size: int = 10,
) -> AsyncPostgresSaver:
    pool = await asyncpg.create_pool(
        dsn,
        min_size=min_size,
        max_size=max_size,
    )
    saver = AsyncPostgresSaver.from_pool(pool)
    await saver.setup()  # idempotent; creates tables if absent
    return saver

async def close_prod_checkpointer(
    saver: AsyncPostgresSaver,
) -> None:
    """Call from the application shutdown hook."""
    await saver.conn.close()
\end{lstlisting}

The checkpointer factory is initialised once in the application lifespan and
injected into \code{build\_graph()}. Chapter~4 shows how to wire
\code{make\_prod\_checkpointer} into a FastAPI \code{@asynccontextmanager}
lifespan so that the pool is established before the first request and gracefully
closed on shutdown.

%% ------------------------------------------------------------
\section{Checkpoint Strategy: Every-Cycle vs.\ Milestone}
\label{sec:checkpoint-strategy}
%% ------------------------------------------------------------

When a checkpointer is attached, LangGraph writes state after every node
execution by default. For a four-node graph (Perceive, Reason, Act, Reflect)
running ten cycles, that is 40 checkpoint writes per run. Each write is a
synchronous-in-the-context-of-the-node round trip to the database. At high cycle
counts or with expensive serialisation (large \code{beliefs} lists, large message
histories), this overhead is not negligible.

The alternative is milestone checkpointing: write only after specific high-value
nodes, and skip checkpoint writes at intermediate nodes within a cycle. The cost
reduction is significant---two or three writes per run regardless of cycle count.
The cost is coarser recovery granularity: if the process is interrupted mid-cycle
between two un-checkpointed nodes, the agent resumes from the last milestone, not
the last node, and re-executes the intervening work.

Table~\ref{tab:checkpoint-policy} frames the decision.

\begin{table}[h]
\centering
\caption{Checkpoint strategy trade-offs}
\label{tab:checkpoint-policy}
\begin{tabular}{lll}
\toprule
\textbf{Factor} & \textbf{Every-cycle} & \textbf{Milestone} \\
\midrule
Recovery fidelity & Node-level (finest) & Cycle-level (coarser) \\
Write volume & $O(\text{nodes} \times \text{cycles})$ & $O(\text{milestones})$ \\
Latency per cycle & +1--4 DB writes & +0--1 DB writes \\
Re-work on recovery & Minimal & At most one partial cycle \\
Recommended for & Long-horizon; expensive tools & Short multi-step; cheap tools \\
\bottomrule
\end{tabular}
\end{table}

For the earnings-report agent---three to five cycles, two tool calls per run,
modest state size---milestone checkpointing is the correct default. For a
long-horizon research agent that may run 50 cycles and invoke expensive external
APIs, every-cycle checkpointing is worth its write overhead because recovering
from the last node is far cheaper than re-running three cycles of expensive tool
calls.

\begin{lstlisting}[language=Python,
  caption={\code{CheckpointPolicy} and milestone-aware \code{build\_graph()}},
  label={lst:checkpoint-policy}]
from enum import Enum

class CheckpointPolicy(str, Enum):
    EVERY_CYCLE = "every_cycle"
    MILESTONE   = "milestone"

MILESTONE_NODES: frozenset[str] = frozenset(
    {"reflect_node", "action_node"}
)

def build_graph(
    checkpointer=None,
    checkpoint_policy: CheckpointPolicy = CheckpointPolicy.EVERY_CYCLE,
):
    from src.agent.graph import _wire_edges  # shared wiring helper
    b = StateGraph(AgentState)
    for name, fn in _ALL_NODES:
        if (checkpoint_policy == CheckpointPolicy.MILESTONE
                and name not in MILESTONE_NODES):
            # Wrap non-milestone nodes to skip checkpointing
            b.add_node(name, fn, checkpoint=False)
        else:
            b.add_node(name, fn)
    _wire_edges(b)
    return b.compile(
        checkpointer=checkpointer,
        recursion_limit=25,
    )
\end{lstlisting}

\begin{notebox}
The \code{checkpoint=False} keyword on \code{add\_node} is supported in
LangGraph~$\geq$~0.2.0. If your version does not yet expose this parameter,
the practical alternative is to omit the checkpointer for non-milestone nodes by
building two sub-graphs---one with the checkpointer and one without---and
chaining them. Chapter~9 revisits this when checkpoint write latency appears in
the cost attribution dashboard.
\end{notebox}

%% ------------------------------------------------------------
\section{State Reducers: \texttt{add\_messages} and Custom Append-Only Reducers}
\label{sec:reducers}
%% ------------------------------------------------------------

A reducer is the merge function LangGraph applies when a node returns a field
value. Without a reducer, the return value replaces the existing field value.
With a reducer, the existing value and the returned value are combined according
to the reducer's logic.

This distinction is invisible until it silently destroys data. Consider a node
that successfully updates \code{beliefs} during a cycle and returns
\code{\{"beliefs": ["Q3 revenue retrieved"]\}}. Without the \code{add} reducer,
every prior observation is gone---the list now contains a single entry. With the
\code{add} reducer, the new observation is appended to the accumulating record.
The difference is not apparent in tests that start from an empty \code{beliefs}
list; it only surfaces when the list has accumulated three or four entries from
prior cycles and a node wipes them.

\code{add\_messages} is LangGraph's built-in reducer for the \code{messages}
field. It does more than append: it deduplicates by message \code{id} and
processes \code{RemoveMessage} objects, which allow a later node to delete a
specific message from the history. This behaviour is correct for the message bus
but wrong for \code{beliefs}, \code{intentions}, and \code{cycle\_traces}:
beliefs must accumulate every observation, including re-observations; intentions
must retain their history; and \code{cycle\_traces} must never be modified after
a trace is written. For these fields, \code{operator.add} on lists is the
correct reducer---it is simply list concatenation.

\begin{lstlisting}[language=Python,
  caption={Reducers side-by-side: \code{add} for beliefs, \code{add\_messages} for messages},
  label={lst:reducers}]
from operator import add
from langgraph.graph.message import add_messages
from typing import Annotated

# --- operator.add: pure list concatenation, no dedup, no removals ---
beliefs:      Annotated[list[str], add]   # every observation kept
intentions:   Annotated[list[str], add]   # every committed step kept
cycle_traces: Annotated[list[dict], add]  # every cycle record kept

# --- add_messages: dedup by id + supports RemoveMessage ---
messages: Annotated[list[dict], add_messages]  # message bus only

# --- strict append reducer: raises if caller attempts to shrink the list ---
def append_only(existing: list, incoming: list) -> list:
    """Use for fields that must never lose history."""
    if not incoming:
        return existing
    if len(incoming) < 0:  # guard type: incoming is always a list
        raise ValueError("append_only received non-list incoming")
    return existing + incoming
\end{lstlisting}

The \code{append\_only} variant is useful when a field's immutability must be
detectable at runtime---for example, in a custom field that records audit events.
The base \code{add} reducer is sufficient for all standard fields in
\code{AgentState}.

\begin{warningbox}
Do not use \code{add\_messages} as a generic append reducer. If a node returns a
message with the same \code{id} as one already in the list, \code{add\_messages}
will silently replace it. For fields where duplicate entries are semantically
correct---such as a belief that the agent re-observed---use \code{operator.add}.
\end{warningbox}

%% ------------------------------------------------------------
\section{\texttt{RecoveryNode}: Resume, Validate, Re-anchor}
\label{sec:recovery-node}
%% ------------------------------------------------------------

When an agent run is interrupted and resumed from a checkpoint, LangGraph restores
state automatically. The restored state is the JSON-deserialised dict that the
checkpointer wrote at the last saved node. That is \emph{mechanically} correct but
not \emph{semantically} sufficient for re-entry into the loop.

Three things can be wrong with a mechanically restored state. First, the
\code{reflect\_decision} field may hold a stale \code{TERMINATE} or
\code{ESCALATE} value from the cycle that was interrupted. The routing function
\code{route\_from\_reflect} (Chapter~1 \S\ref{sec:graph}) routes both values to
\code{END}---so a restored state with a stale \code{TERMINATE} would immediately
exit the loop before executing a single cycle. Second, the \code{schema\_version}
may be lower than \code{CURRENT\_SCHEMA\_VERSION} if the schema was migrated
between the checkpoint write and the resume---in which case the migration step
in \S\ref{sec:migration} must run first. Third, the goal re-injection into
\code{beliefs} must be explicit: after several cycles the goal may have drifted
below the relevance horizon in the \code{beliefs} window, and the first
post-recovery \code{reason\_node} call will not see it unless it is re-injected.

The third point is the one that matters most, and it introduces a design
constraint: \emph{the goal on recovery must come from the checkpoint, not from
the caller}. If the resuming caller supplies a new goal---even a slightly
reworded one---the agent continues a different objective using the accumulated
beliefs, intentions, and cycle traces from the original task. The result is
incoherent state that can produce plausible-looking but incorrect output.

\begin{lstlisting}[language=Python,
  caption={\code{validate\_restored\_state}: consistency checks before resuming},
  label={lst:validate-recovery}]
# src/agent/nodes/recovery.py
from dataclasses import dataclass
from src.agent.state import CURRENT_SCHEMA_VERSION

@dataclass
class RecoveryCheckResult:
    valid:  bool
    errors: list[str]

def validate_restored_state(
    state: AgentState,
) -> RecoveryCheckResult:
    errors: list[str] = []
    if not state.get("goal", "").strip():
        errors.append("goal is empty")
    if state.get("token_budget", -1) < 0:
        errors.append("token_budget is negative")
    if state.get("cycle_count", -1) < 0:
        errors.append("cycle_count is negative")
    if state.get("schema_version", 0) != CURRENT_SCHEMA_VERSION:
        errors.append(
            f"schema_version={state.get('schema_version')} "
            f"!= CURRENT={CURRENT_SCHEMA_VERSION}; "
            "run migrate_state() before resuming"
        )
    return RecoveryCheckResult(
        valid=len(errors) == 0, errors=errors
    )
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={\code{recovery\_node}: clear stale decisions, re-anchor goal, set recovery flag},
  label={lst:recovery-node}]
def recovery_node(state: AgentState) -> dict:
    check = validate_restored_state(state)
    if not check.valid:
        return {
            "reflect_decision": "ESCALATE",
            "final_answer": (
                f"Recovery aborted: {'; '.join(check.errors)}"
            ),
        }
    goal = state["goal"]
    return {
        # Re-anchor the original goal as the highest-priority belief
        "beliefs":          [f"[recovery] re-anchoring goal: {goal}"],
        # Clear any stale routing decision from the interrupted cycle
        "reflect_decision": None,
        # Signal to perceive_node to log the recovery event
        "_recovery_cycle":  True,
    }
\end{lstlisting}

The \code{recovery\_node} is wired as a conditional entry point: the graph runner
passes a \code{\{"configurable": \{"resume": True\}\}} flag when invoking from a
checkpoint, and the entry-point router directs the run to \code{recovery\_node}
rather than \code{perceive\_node}. After \code{recovery\_node} completes, the
conditional edge routes to \code{perceive\_node} for the first post-recovery
cycle.

\begin{lstlisting}[language=Python,
  caption={Recovery entry-point router and updated \code{build\_graph()}},
  label={lst:recovery-wiring}]
def route_entry_point(state: AgentState,
                      config: dict) -> str:
    if (config or {}).get("configurable", {}).get("resume", False):
        return "recovery_node"
    return "perceive_node"

def build_graph(checkpointer=None,
                checkpoint_policy=CheckpointPolicy.EVERY_CYCLE):
    b = StateGraph(AgentState)
    b.add_node("recovery_node",  recovery_node)
    b.add_node("perceive_node",  perceive_node)
    b.add_node("reason_node",    reason_node)
    b.add_node("action_node",    action_node)
    b.add_node("reflect_node",   reflect_node)
    # Conditional entry: fresh run or resume
    b.set_conditional_entry_point(
        route_entry_point,
        {"recovery_node": "recovery_node",
         "perceive_node": "perceive_node"},
    )
    b.add_edge("recovery_node", "perceive_node")
    b.add_edge("perceive_node", "reason_node")
    b.add_conditional_edges(
        "reason_node", route_from_reason,
        {"action_node": "action_node",
         "perceive_node": "perceive_node", "end": END})
    b.add_edge("action_node", "reflect_node")
    b.add_conditional_edges(
        "reflect_node", route_from_reflect,
        {"perceive_node": "perceive_node",
         "action_node": "action_node", "end": END})
    return b.compile(
        checkpointer=checkpointer,
        recursion_limit=25,
    )
\end{lstlisting}

%% ------------------------------------------------------------
\section{State Schema Migration: Versioned \texttt{TypedDict}}
\label{sec:migration}
%% ------------------------------------------------------------

Schema migration for agent state is not optional. It is the predictable
consequence of a system that evolves. Every time a new field is added to
\code{AgentState}---and this book adds several across the coming chapters---any
checkpoint written before that addition will be missing the field. The first
node that reads the new field from a restored checkpoint raises a \code{KeyError}.
In a long-running production agent, that \code{KeyError} surfaces inside a live
run, not at startup.

The \code{schema\_version} integer in \code{AgentState} is the migration key.
It starts at \code{1} (Chapter~1's original schema). Chapter~3 introduces two new
fields---\code{goal\_embedding} and \code{goal\_divergence\_checked\_at}---and
increments it to \code{2}. The migration system applies all version-to-version
transformations in sequence until the loaded dict matches
\code{CURRENT\_SCHEMA\_VERSION}.

The migration output is a plain \code{dict}, not a Pydantic object. The migrated
dict goes directly back into the graph's state machinery, which operates on plain
dicts. Pydantic objects are only used at the external entry boundary (Chapter~1
\S\ref{sec:agentstate}).

The backward-compatibility rules are simple. New optional fields are added with
\code{setdefault} and a safe default (\code{None} for optionals,
\code{0} for counters). Removed fields are deleted from the dict. Renamed fields
require a copy-and-delete step. These three operations cover the entire migration
surface for all changes planned in this book.

\begin{lstlisting}[language=Python,
  caption={Version constants, v1\;$\to$\;v2 migration, and dispatcher},
  label={lst:migration-schema}]
# src/agent/state.py
from typing import Callable

CURRENT_SCHEMA_VERSION = 2

def _migrate_v1_to_v2(state: dict) -> dict:
    """Add Ch.2 goal divergence fields (default None)."""
    state.setdefault("goal_embedding",             None)
    state.setdefault("goal_divergence_checked_at", None)
    state.setdefault("_recovery_cycle",            False)
    state["schema_version"] = 2
    return state

_MIGRATIONS: dict[int, Callable[[dict], dict]] = {
    1: _migrate_v1_to_v2,
    # future: 2: _migrate_v2_to_v3, etc.
}

def migrate_state(raw: dict) -> AgentState:
    """Upgrade raw checkpoint dict to CURRENT_SCHEMA_VERSION in-place."""
    v = raw.get("schema_version", 1)
    while v < CURRENT_SCHEMA_VERSION:
        if v not in _MIGRATIONS:
            raise ValueError(
                f"No migration registered for schema_version={v}"
            )
        raw = _MIGRATIONS[v](raw)
        v   = raw["schema_version"]
    return raw  # type: ignore[return-value]
\end{lstlisting}

The runner calls \code{migrate\_state} immediately after loading a checkpoint,
before passing the state to any node. This isolates migration to a single call
site: changing the migration logic does not require touching the graph, the nodes,
or the checkpointer.

\begin{lstlisting}[language=Python,
  caption={\code{resume\_agent}: load checkpoint, migrate, invoke},
  label={lst:resume-agent}]
# src/agent/runner.py
from langgraph.errors import GraphRecursionError
from src.agent.state import migrate_state

async def resume_agent(
    graph,
    config: dict,
) -> AgentRunResult:
    snapshot  = await graph.aget_state(config)
    raw_state = dict(snapshot.values)
    state     = migrate_state(raw_state)
    try:
        result = await graph.ainvoke(
            state,
            config={**config,
                    "configurable": {
                        **config.get("configurable", {}),
                        "resume": True,
                    }},
        )
        return AgentRunResult(state=result, stopped_by="completion")
    except GraphRecursionError:
        return AgentRunResult(
            state=None, stopped_by="recursion_limit"
        )
\end{lstlisting}

The \code{resume\_agent} function is the production resume path. The \code{resume:
True} flag in \code{configurable} triggers \code{route\_entry\_point} to direct
the first step to \code{recovery\_node} rather than \code{perceive\_node}, which
then validates the state and re-anchors the goal before the agent continues.

\begin{tipbox}
Add a \code{migration\_applied} boolean to the \code{AgentRunResult} return value.
When \code{True}, it indicates that schema migration ran during this resume---a
useful signal for alerting when old checkpoints accumulate, which may indicate a
deployment that is not completing its runs before the next schema change is
shipped.
\end{tipbox}

%% ------------------------------------------------------------
\section{Test Suite: Checkpointer and Recovery Tests}
\label{sec:ch3-tests}
%% ------------------------------------------------------------

The state management layer has three independently testable units: the reducers,
the \code{recovery\_node}, and the \code{migrate\_state} function. None of them
require a compiled graph or a live database connection.

Reducer tests assert the merge semantics directly by calling the reducer function
with two lists and asserting the result. Recovery tests construct a minimal
\code{AgentState} dict with a known inconsistency and assert that
\code{validate\_restored\_state} catches it. Migration tests construct a v1 dict
(missing the new fields) and assert that \code{migrate\_state} produces a v2
dict with the correct defaults.

\begin{lstlisting}[language=Python,
  caption={State management unit tests: reducers, recovery, and migration},
  label={lst:ch3-tests}]
# tests/test_state.py
from operator import add
from src.agent.state import migrate_state, CURRENT_SCHEMA_VERSION
from src.agent.nodes.recovery import (
    validate_restored_state, recovery_node
)

# --- Reducer tests ---
def test_add_reducer_appends_not_replaces():
    existing = ["belief-1", "belief-2"]
    incoming = ["belief-3"]
    assert add(existing, incoming) == ["belief-1", "belief-2", "belief-3"]

def test_add_reducer_empty_incoming_preserves_existing():
    existing = ["belief-1"]
    assert add(existing, []) == ["belief-1"]

# --- Recovery validation tests ---
def test_recovery_fails_on_empty_goal():
    state = _make_v2_state(goal="")
    result = validate_restored_state(state)
    assert not result.valid
    assert any("goal" in e for e in result.errors)

def test_recovery_fails_on_schema_mismatch():
    state = _make_v2_state(schema_version=1)
    result = validate_restored_state(state)
    assert not result.valid
    assert any("schema_version" in e for e in result.errors)

def test_recovery_node_reanchors_goal():
    state = _make_v2_state(goal="retrieve Q3 earnings")
    updates = recovery_node(state)
    assert updates["reflect_decision"] is None
    assert updates["_recovery_cycle"] is True
    assert "retrieve Q3 earnings" in updates["beliefs"][0]

def test_recovery_node_escalates_on_invalid_state():
    state = _make_v2_state(goal="")  # invalid: empty goal
    updates = recovery_node(state)
    assert updates["reflect_decision"] == "ESCALATE"

# --- Migration tests ---
def test_migrate_v1_to_v2_adds_missing_fields():
    v1_state = {
        "goal": "test", "beliefs": [], "intentions": [],
        "token_budget": 1000, "cycle_count": 0,
        "cycle_traces": [], "messages": [],
        "last_action": None, "reflect_decision": None,
        "final_answer": None, "_policy": None,
        "schema_version": 1,
    }
    v2 = migrate_state(v1_state)
    assert v2["schema_version"] == CURRENT_SCHEMA_VERSION
    assert v2["goal_embedding"] is None
    assert v2["goal_divergence_checked_at"] is None
    assert v2["_recovery_cycle"] is False

def test_migrate_already_current_is_noop():
    v2_state = _make_v2_state()
    assert migrate_state(dict(v2_state))["schema_version"] == 2

# --- Fixture ---
def _make_v2_state(**overrides) -> dict:
    base = {
        "goal": "test goal", "beliefs": [], "intentions": [],
        "token_budget": 1000, "cycle_count": 0,
        "cycle_traces": [], "messages": [],
        "last_action": None, "reflect_decision": None,
        "final_answer": None, "_policy": None,
        "goal_embedding": None,
        "goal_divergence_checked_at": None,
        "_recovery_cycle": False,
        "schema_version": 2,
    }
    base.update(overrides)
    return base
\end{lstlisting}

Chapter~28 extends this suite with an integration test that compiles the full
graph with \code{make\_dev\_checkpointer()}, runs two cycles, serialises the
checkpoint to a JSON string representing a v1 schema, and then calls
\code{resume\_agent} to verify that the migration, recovery, and re-entry path
all complete without error and produce a valid \code{FinalAnswer}.

\begin{takeawaybox}
State schema design and checkpointing are not infrastructure concerns layered on
top of the agent---they are structural constraints that determine what the agent
can survive. \code{AgentState} must use named BDI fields with explicit reducers
because the merge semantics are the contract every node relies on. The
\code{goal} field is write-once: \code{TypedDict} cannot enforce this, so the
team convention must. \code{SqliteSaver} is a testing tool; \code{AsyncPostgresSaver}
with a properly sized connection pool is the minimum viable production
checkpointer. Checkpoint frequency is a cost/latency trade-off decided by task
type, not set-and-forgotten. \code{RecoveryNode} must re-anchor the goal from the
checkpoint, never from the caller, because accepting a new goal on recovery is
itself a goal-drift vector. \code{migrate\_state} is called by the runner before
every resume: it ensures that a checkpoint written by any prior schema version
can be loaded cleanly, and that the \code{KeyError} from a missing field is
caught at startup, not inside a live production run.
\end{takeawaybox}
