%% ============================================================
%%  CHAPTER 25 — Continuous Quality Without Guessing
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%% ============================================================


An agent that passes all unit tests in CI can still silently degrade in
production the moment a new LLM version ships, the retriever's index is
rebuilt with different chunking, or the user population drifts toward a
subdomain the golden set never represented.  The failure mode is invisible:
no exception is raised, no latency spike appears in the dashboards built in
Chapter~\ref{chap:observability}, and the agent continues returning
responses --- responses that are subtly less faithful, less relevant, or
less grounded than they were last week.  Without a measurement loop that
runs both continuously on live traffic \emph{and} in deterministic CI runs,
quality degradation is discovered only through user complaints.

The solution has four layers.  The first is a \emph{golden dataset} --- a
version-controlled collection of (input, expected output, metadata) triples
stored in Langfuse that acts as the single source of truth for all
evaluation.  The second is \emph{online evaluation}: lightweight
LLM-as-judge scoring attached fire-and-forget to every production trace,
providing continuous signal that catches distribution shift without blocking
the user's response.  The third is \emph{offline evaluation}: RAGAS and
DeepEval metrics computed in CI against the golden dataset, providing
statistically meaningful pass/fail gates on every pull request.  The fourth
is a \emph{CI harness} that wires all three layers together, pins the
dataset to a version-stamped snapshot so parallel PRs are comparable, and
enforces machine-readable success criteria that prevent metric inflation
from creeping thresholds.

The sections are ordered by dependency.  \S\ref{sec:golden-dataset}
establishes the dataset structure because every later section reads from it.
\S\ref{sec:online-eval} builds online judging because it produces additional
trace scores that enrich the dataset items promoted from production.
\S\ref{sec:ragas-metrics} and \S\ref{sec:deepeval-metrics} build the offline
suites that consume the same dataset.  \S\ref{sec:success-criteria} defines
the immutable threshold contract before \S\ref{sec:ci-harness} assembles the
harness that enforces it, so the harness never hard-codes a number.

\begin{table}[ht]
\centering
\caption{Domain terms introduced in Chapter~\ref{chap:eval}}
\label{tab:eval-terms}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{P{0.24\linewidth} P{0.70\linewidth}}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
Golden dataset
  & A version-controlled collection of labelled (input, expected\,output)
    pairs used as the reference population for all offline evaluation. \\
Online evaluation
  & Metric computation performed on live production traces, asynchronously,
    without blocking the user's response path. \\
Offline evaluation
  & Metric computation performed in CI against a pinned golden dataset
    snapshot; provides deterministic pass/fail gates. \\
LLM-as-judge
  & A separate LLM invocation --- distinct from the agent's model --- that
    scores a response against a rubric, returning a numeric value and
    a natural-language rationale. \\
Faithfulness
  & The fraction of factual claims in a generated answer that are directly
    supported by the retrieved contexts; measures hallucination risk. \\
Answer Relevancy
  & A metric that reverse-generates hypothetical questions from an answer
    and measures their embedding cosine similarity to the original query;
    measures off-topic or evasive responses. \\
Context Precision
  & The fraction of retrieved chunks that are actually useful for producing
    the reference answer; measures retriever over-fetching. \\
Context Recall
  & The fraction of reference-answer statements that are attributable to
    at least one retrieved chunk; measures retriever under-fetching. \\
Success criteria
  & A machine-readable, version-controlled mapping from metric name to
    minimum passing threshold; the single authoritative source CI reads. \\
Score config
  & A Langfuse server-side object that registers a metric name, data type,
    and rubric description so scores appear labelled in the UI. \\
\bottomrule
\end{tabular}
\end{table}

\begin{sourcelayout}
\dirtree{%
  .1 src/.
  .2 eval/.
  .3 dataset\_manager.py \DTcomment{\S\ref{sec:golden-dataset}}.
  .3 online\_judge.py \DTcomment{\S\ref{sec:online-eval}}.
  .3 ragas\_suite.py \DTcomment{\S\ref{sec:ragas-metrics}}.
  .3 deepeval\_suite.py \DTcomment{\S\ref{sec:deepeval-metrics}}.
  .3 success\_criteria.py \DTcomment{\S\ref{sec:success-criteria}}.
  .3 run\_eval.py \DTcomment{\S\ref{sec:ci-harness}}.
  .2 tests/.
  .3 test\_eval\_ci.py \DTcomment{\S\ref{sec:ci-harness}}.
  .1 .github/workflows/.
  .2 eval.yml \DTcomment{\S\ref{sec:ci-harness}}.
}
\end{sourcelayout}


%%------------------------------------------------------------
\section{The Dataset Is the Single Source of Truth}
\label{sec:golden-dataset}
%%------------------------------------------------------------

The design problem is reproducibility: if two engineers run the evaluation
suite on the same commit and get different scores, the metric is meaningless
as a gate.  The naive approach of storing golden cases in a JSON file
committed to the repository makes this worse, not better --- the file drifts
independently of the live traces it was supposed to represent, there is no
audit trail of who added or removed items, and a PR that renames a field
silently breaks every future run.  The principle the solution follows is
that the golden dataset must be a \emph{server-side, versioned artefact}
whose items carry provenance metadata and whose state at any point in time
is reachable by a deterministic identifier that CI can pin.

Langfuse datasets fulfil this contract.  An \emph{item} is a triple
composed of an \code{input} dict, an \code{expected\_output} dict, and a
\code{metadata} dict.  The server validates item shape against
\code{input\_schema} and \code{expected\_output\_schema} JSON Schema objects
supplied when the dataset is created, rejecting malformed uploads at
write-time rather than at eval-time.  Every mutation --- \code{upsert\_item},
archive, delete --- increments an implicit version tracked by the server's
\code{updatedAt} timestamp; CI pins to a specific snapshot by passing that
ISO-8601 string to \code{get\_dataset} as the \code{version} argument.  The
default call returns the latest state, so local development always sees
current data while CI sees a frozen snapshot.

Items arrive from two sources whose provenance must be distinguished.
\emph{Production-promoted} items are created when a human operator labels a
live trace as correct in the Langfuse UI: the trace's input and the
operator's confirmed output become an item with
\code{metadata["source"] = "production"}.  \emph{Synthetic} items are
generated by an LLM to cover adversarial cases, rare domains, or edge
inputs the production population never includes; they carry
\code{metadata["source"] = "synthetic"}.  The distinction matters for
selective runs: a deployment that only touches the security guardrails layer
can run evaluation against \code{domain="security"} items only, cutting
CI cost by an order of magnitude while preserving signal relevance.

\begin{warningbox}
Langfuse SDK v4 uses \code{get\_client()} --- a singleton backed by the
OpenTelemetry foundation --- instead of the \code{Langfuse()} constructor
used in SDK v2 and v3.  Calling \code{Langfuse()} in v4 raises a
deprecation warning and will fail in a future minor.  All code in this
chapter uses \code{get\_client()} exclusively.
\end{warningbox}

The \code{ensure\_dataset} helper is intentionally idempotent.  Calling it
twice on a dataset that already exists must not raise an error.
The Langfuse server returns the existing dataset unchanged when
\code{create\_dataset} is called with a name that already exists, so the
function is safe to invoke on every CI startup without a prior existence
check.

\begin{lstlisting}[language=Python, caption={Dataset manager: schema
    enforcement\, idempotent creation\, and version-pinned fetch},
    label={lst:dataset-manager}]
# src/eval/dataset_manager.py
from __future__ import annotations
from langfuse import get_client

DATASET_NAME = "evaluation/agent-golden-v1"

GOLDEN_INPUT_SCHEMA: dict = {
    "type": "object",
    "required": ["query", "retrieved_contexts"],
    "properties": {
        "query":              {"type": "string"},
        "retrieved_contexts": {
            "type": "array", "items": {"type": "string"}
        },
    },
}
GOLDEN_OUTPUT_SCHEMA: dict = {
    "type": "object",
    "required": ["answer"],
    "properties": {"answer": {"type": "string"}},
}

def ensure_dataset() -> None:
    get_client().create_dataset(
        name=DATASET_NAME,
        input_schema=GOLDEN_INPUT_SCHEMA,
        expected_output_schema=GOLDEN_OUTPUT_SCHEMA,
    )

def upsert_golden_item(
    item_id: str, query: str, contexts: list[str],
    expected_answer: str, metadata: dict,
) -> None:
    get_client().get_dataset(DATASET_NAME).upsert_item(
        id=item_id,
        input={"query": query, "retrieved_contexts": contexts},
        expected_output={"answer": expected_answer},
        metadata=metadata,
    )

def fetch_dataset_at_version(version_ts: str) -> list[dict]:
    ds = get_client().get_dataset(DATASET_NAME, version=version_ts)
    return [
        {"id": it.id, "input": it.input,
         "expected_output": it.expected_output,
         "metadata": it.metadata}
        for it in ds.items
    ]
\end{lstlisting}

Three decisions in Listing~\ref{lst:dataset-manager} deserve explanation.
First, \code{DATASET\_NAME} uses the \texttt{folder/name} format
(\code{evaluation/agent-golden-v1}) rather than a flat name; Langfuse
displays the folder prefix as a namespace in the UI, preventing name
collisions when multiple projects share a server.  Second, the
\code{version\_ts} parameter to \code{fetch\_dataset\_at\_version} is an
ISO-8601 string --- it is the value stored in the CI workflow's
\code{EVAL\_DATASET\_VERSION} secret, not a sequential integer, which means
rolling back to any historical state requires only updating that secret
rather than rewriting history.  Third, \code{upsert\_item} is used instead
of a create/update conditional because \code{upsert\_item} is idempotent on
\code{id}: a re-promoted trace does not produce a duplicate item but instead
updates the existing one in place.

The dataset structure and versioning contract are now established.  The next
problem is measuring quality on live traces without touching the user's
response path.

%%------------------------------------------------------------
\section{Judgment Must Not Block the User}
\label{sec:online-eval}
%%------------------------------------------------------------

The design problem is latency asymmetry: an LLM-as-judge call takes
100--800\,ms, which is a large fraction of the agent's total response time
budget.  Inserting the judge into the graph's critical path --- as a
post-processing node that runs before the response is returned --- inflates
p95 latency by the full judge call duration for every user request.  The
naive approach, calling the judge synchronously, is unjustifiable because
the judge score is not needed to produce the response; it is observability
data.  The principle is that observability must be fire-and-forget: enqueue
the work as a non-blocking \code{asyncio} task and return the user's answer
immediately.

Score attachment in Langfuse SDK v4 follows a single path:
\code{get\_client().create\_score(trace\_id=..., name=..., value=...,
data\_type=..., comment=...)}.  The \code{data\_type} argument is a plain
string from the set \code{"NUMERIC"}, \code{"CATEGORICAL"},
\code{"BOOLEAN"}, and \code{"TEXT"} --- it is not an imported enum.
The score object's target is determined by exactly one of the four
mutually-exclusive identifier fields: \code{trace\_id} attaches to the
trace root, \code{observation\_id} narrows to a specific span,
\code{session\_id} attaches to a session, and \code{dataset\_run\_id}
attaches to a dataset run item.  Passing more than one raises a validation
error on the server.  For online faithfulness and relevancy the trace root
is the correct granularity because both metrics characterise the full
response, not any individual retrieval or generation step.

Before scores can be created they must be registered as \emph{score
configs} on the Langfuse server.  A score config records the metric's name,
data type, and description so the UI can display labelled scores with
their definitions inline.  Score configs are \emph{immutable once created}:
the server rejects a second \code{create} call with the same name.  The
\code{ensure\_score\_configs} function must therefore guard against
duplicate-creation errors rather than calling \code{create} unconditionally.
In SDK v4 the correct API path is
\code{get\_client().api.score\_configs.create(request=...)} using the
\code{CreateScoreConfigRequest} model; there is no top-level convenience
method.

\begin{notebox}
The judge model must be \emph{different} from the agent's own model.
Using the same model as both author and critic produces optimistic
self-grading: the judge internalises the same biases that caused the
response and rates it higher than an independent evaluator would.
A fast, cheap model in a different model family --- \code{claude-haiku-4.5}
when the agent uses \code{gemini-3.1-pro}, or vice versa --- provides
genuine independence while keeping judge latency low.
\end{notebox}

The judge model is injected through \code{RunnableConfig.configurable}
rather than hard-coded.  Chapter~\ref{chap:config} introduced
\code{RunnableConfig.configurable} as the mechanism for per-invocation
parameter overrides without modifying source; reading the judge model from
it means the test suite can swap to a deterministic mock without altering
the production code path.

\begin{lstlisting}[language=Python, caption={Online judge: fire-and-forget
    scoring attached to every production trace},
    label={lst:online-judge}]
# src/eval/online_judge.py
from __future__ import annotations
import asyncio, json
from langfuse import get_client
from langfuse.api import CreateScoreConfigRequest
import litellm

FAITHFULNESS_RUBRIC = (
    "Score 0.0-1.0. 1.0 = every factual claim in the answer is "
    "directly supported by at least one retrieved context. "
    "0.0 = answer introduces facts absent from all contexts."
)
RELEVANCY_RUBRIC = (
    "Score 0.0-1.0. 1.0 = answer directly and completely addresses "
    "the question. Penalise irrelevant content or evasion."
)
_SCORE_CONFIGS = {
    "faithfulness_online":     FAITHFULNESS_RUBRIC,
    "answer_relevancy_online": RELEVANCY_RUBRIC,
}

def ensure_score_configs() -> None:
    """Idempotent: silently skips configs that already exist."""
    lf = get_client()
    existing = {sc.name for sc in lf.api.score_configs.list().data}
    for name, rubric in _SCORE_CONFIGS.items():
        if name not in existing:
            lf.api.score_configs.create(
                request=CreateScoreConfigRequest(
                    name=name,
                    data_type="NUMERIC",
                    description=rubric,
                )
            )

async def _call_judge(
    query: str, answer: str,
    contexts: list[str], judge_model: str,
) -> dict:
    prompt = (
        f"Query: {query}\nContexts: {contexts}\nAnswer: {answer}\n"
        'Return JSON with keys "faithfulness" (0-1), '
        '"answer_relevancy" (0-1), '
        '"faithfulness_reason" (string), '
        '"relevancy_reason" (string).'
    )
    resp = await litellm.acompletion(
        model=judge_model,
        messages=[{"role": "user", "content": prompt}],
        response_format={"type": "json_object"},
    )
    return json.loads(resp.choices[0].message.content)

async def run_online_judge(
    trace_id: str, query: str, answer: str,
    contexts: list[str], judge_model: str,
) -> None:
    scores = await _call_judge(query, answer, contexts, judge_model)
    lf = get_client()
    lf.create_score(
        trace_id=trace_id,
        name="faithfulness_online",
        value=scores["faithfulness"],
        data_type="NUMERIC",
        comment=scores.get("faithfulness_reason"),
    )
    lf.create_score(
        trace_id=trace_id,
        name="answer_relevancy_online",
        value=scores["answer_relevancy"],
        data_type="NUMERIC",
        comment=scores.get("relevancy_reason"),
    )

def attach_online_judge(
    trace_id: str, query: str, answer: str,
    contexts: list[str], judge_model: str,
) -> None:
    """Schedule judging as a background task; must be called
    from within a running asyncio event loop (ainvoke context)."""
    asyncio.get_event_loop().create_task(
        run_online_judge(
            trace_id, query, answer, contexts, judge_model)
    )
\end{lstlisting}

Four decisions in Listing~\ref{lst:online-judge} deserve explanation.
First, \code{ensure\_score\_configs} queries \code{lf.api.score\_configs.list()}
before creating; score configs are immutable once registered, so a second
unconditional \code{create} call for the same name would raise a server-side
validation error on every CI run after the first.  The existence check makes
the function safe to call on startup across any number of deployments.
Second, the JSON prompt uses explicit key names in the instruction string
rather than embedding JSON template literals; this avoids double-escaping
issues that arise in f-strings that embed brace-delimited JSON.  Third, both
\code{create\_score} calls use \code{trace\_id} as the sole target
identifier; the Langfuse API treats the four target fields
(\code{trace\_id}, \code{observation\_id}, \code{session\_id},
\code{dataset\_run\_id}) as mutually exclusive, and supplying more than one
produces a server validation error.  Fourth, \code{attach\_online\_judge}
calls \code{asyncio.get\_event\_loop().create\_task} rather than
\code{asyncio.create\_task}: the latter is the preferred form inside a
coroutine, but since \code{attach\_online\_judge} is a synchronous function
called from a LangGraph node, the explicit loop reference avoids the
\code{RuntimeError} raised when \code{create\_task} is invoked from a
synchronous frame.

\begin{warningbox}
\code{attach\_online\_judge} relies on a running asyncio event loop already
existing in the calling thread.  This is guaranteed when the LangGraph graph
is invoked via \code{ainvoke} or \code{astream}, but not when it is invoked
via the synchronous \code{invoke}.  If synchronous invocation is required,
fire judging from a dedicated background thread using
\code{asyncio.run\_coroutine\_threadsafe}.
\end{warningbox}

Online scores now accumulate on every production trace.  The next problem
is computing the statistically richer offline metrics --- faithfulness,
relevancy, context precision, and context recall --- in a way that is
reproducible, configurable, and aware of its own timeout boundaries.

%%------------------------------------------------------------
\section{Retrieval Quality Has Four Distinct Failure Modes}
\label{sec:ragas-metrics}
%%------------------------------------------------------------

The design problem is that ``quality'' is not a single dimension.  A
retriever that over-fetches --- returning fifty chunks when three are
relevant --- scores high on recall but wastes tokens and degrades
faithfulness because the model must ignore irrelevant material.  A
generator that hedges excessively scores high on faithfulness but low on
answer relevancy because it refuses to commit to a direct answer.  The
naive approach of tracking a single aggregate quality score conflates these
independent failure modes and makes it impossible to diagnose which layer
--- retriever or generator --- regressed.  The principle is that each
metric must measure one failure mode in one layer, with no overlap.

RAGAS v0.2 provides four metrics that partition the failure space cleanly.
\emph{Faithfulness} measures whether the generator invents facts not present
in the retrieved contexts; it requires \code{user\_input}, \code{response},
and \code{retrieved\_contexts} but not a reference answer, because it checks
the response against the contexts, not against ground truth.
\emph{AnswerRelevancy} measures whether the response addresses the question;
it reverse-generates hypothetical questions from the answer, measures their
embedding similarity to the original query, and requires only
\code{user\_input} and \code{response}.  \emph{ContextPrecision} measures
what fraction of retrieved chunks were actually needed; it requires a
reference answer because it must determine which chunks are useful for
producing the ground truth.  \emph{ContextRecall} measures whether every
claim in the reference answer can be attributed to at least one retrieved
chunk; it requires \code{retrieved\_contexts} and \code{reference}.

\begin{notebox}
RAGAS \code{SingleTurnSample} field names are \code{user\_input},
\code{response}, \code{retrieved\_contexts}, and \code{reference} ---
not \code{query}, \code{answer}, \code{contexts}, or \code{ground\_truth}.
Using the wrong names produces silent \code{None} fields which RAGAS treats
as missing data, skipping the metric entirely without raising an exception.
Map all incoming data explicitly at the boundary.
\end{notebox}

RAGAS uses its own LLM client for judge calls by default.  Allowing this
means the judge bypasses the LiteLLM retry-and-fallback layer established
in Chapter~\ref{chap:litellm}, opening a second unmonitored token spend path
and a different error-handling surface.  Injecting a
\code{LangchainLLMWrapper} around the same \code{ChatLiteLLM} instance used
by the agent closes both gaps: all RAGAS judge calls flow through the same
provider, inherit the same retry policy, and appear in the same Langfuse
trace tree.

The \code{RunConfig} timeout deserves explicit attention.  RAGAS's default
timeout is thirty seconds.  A judge call to a provider under load --- common
in CI environments running parallel matrix jobs --- can easily exceed that
limit, causing RAGAS to silently return \code{nan} for the affected metric
rather than raising an exception.  In RAGAS v0.2 the \code{RunConfig}
is passed \emph{per metric} at construction time, not to the top-level
\code{evaluate()} call; setting it once on a shared constant and passing
it to every metric constructor is the correct pattern.

\begin{lstlisting}[language=Python, caption={RAGAS suite: four metrics\,
    per-metric RunConfig\, and nan-filtered results},
    label={lst:ragas-suite}]
# src/eval/ragas_suite.py
from __future__ import annotations
from ragas import EvaluationDataset, evaluate
from ragas.run_config import RunConfig
from ragas.dataset_schema import SingleTurnSample
from ragas.llms import LangchainLLMWrapper
from ragas.embeddings import LangchainEmbeddingsWrapper
from ragas.metrics import (
    Faithfulness, AnswerRelevancy,
    ContextPrecision, ContextRecall,
)

DEFAULT_RUN_CONFIG = RunConfig(timeout=120, max_retries=3, max_wait=60)

def build_ragas_dataset(items: list[dict]) -> EvaluationDataset:
    samples = [
        SingleTurnSample(
            user_input=it["input"]["query"],
            response=it["run_output"]["answer"],
            retrieved_contexts=it["run_output"].get(
                "retrieved_contexts", []),
            reference=it["expected_output"].get("answer"),
        )
        for it in items
    ]
    return EvaluationDataset(samples=samples)

def run_ragas_suite(
    dataset: EvaluationDataset,
    llm: LangchainLLMWrapper,
    embeddings: LangchainEmbeddingsWrapper,
    run_config: RunConfig = DEFAULT_RUN_CONFIG,
) -> dict[str, float]:
    metrics = [
        Faithfulness(llm=llm, run_config=run_config),
        AnswerRelevancy(
            llm=llm, embeddings=embeddings, run_config=run_config),
        ContextPrecision(llm=llm, run_config=run_config),
        ContextRecall(llm=llm, run_config=run_config),
    ]
    result = evaluate(
        dataset=dataset,
        metrics=metrics,
        llm=llm,
        embeddings=embeddings,
    )
    return {
        k: float(v)
        for k, v in result.items()
        if v is not None and str(v) != "nan"
    }
\end{lstlisting}

Three decisions in Listing~\ref{lst:ragas-suite} deserve explanation.
First, \code{RunConfig} is passed to each metric's constructor rather than
to \code{evaluate()}; in RAGAS v0.2, \code{evaluate()} does not accept a
\code{run\_config} keyword argument --- the timeout and retry policy are
properties of each metric, not of the evaluation loop.  Passing it to
\code{evaluate()} would be silently ignored, producing the same thirty-second
timeout failures the configuration was meant to prevent.  Second, the dict
comprehension filters out \code{None} and \code{"nan"} values explicitly;
a caller that receives \code{nan} and compares it against a float threshold
will silently pass the gate, which is a worse failure mode than surfacing the
missing metric as an absent key.  Third, the \code{llm} and \code{embeddings}
arguments are passed \emph{both} at metric constructor level and at
\code{evaluate()} level; some RAGAS metric implementations read the LLM from
the metric object while others fall back to the evaluate-level argument, and
supplying both ensures consistent coverage regardless of which internal path
a given metric version takes.

RAGAS measures retrieval and generation quality in aggregate across the
dataset.  The next section adds a complementary framework that measures
quality at the level of individual test cases, enabling per-item failure
reports rather than dataset-level mean scores.

%%------------------------------------------------------------
\section{Item-Level Failures Reveal What Means Cannot}
\label{sec:deepeval-metrics}
%%------------------------------------------------------------

The design problem is that dataset-level mean scores obscure the
distribution of failures.  A faithfulness mean of 0.83 across fifty items
is consistent with forty-eight perfect items and two complete hallucinations,
which is a very different risk profile from fifty items each scoring 0.83.
The naive approach of accepting a passing mean score as evidence of quality
misses the tail: the two hallucinating items may correspond to the agent's
most common query pattern.  The principle is that every evaluation framework
must be capable of surfacing which specific items failed and why, so
engineers can trace failures back to the golden dataset item, the trace, and
the retriever invocation that produced them.

DeepEval operates at the test-case level: each \code{LLMTestCase} maps to a
single golden item, and each metric's \code{measure()} call assigns a score
and a \code{reason} string to that specific case.  This means a failing CI
run produces a per-item report --- item ID, metric name, score, reason ---
that is actionable without additional querying.  The \code{ItemResult}
dataclass defined below carries all four fields so downstream code can branch
on \code{.passed} without string key lookups.

\begin{notebox}
DeepEval's built-in metrics (\code{FaithfulnessMetric},
\code{AnswerRelevancyMetric}) and \code{GEval} accept a \code{model}
argument that must be either an OpenAI-compatible model name string or a
subclass of \code{DeepEvalBaseLLM}.  Passing a bare non-OpenAI model string
such as \code{"claude-haiku-4.5"} raises a \code{ValueError} at runtime
because DeepEval cannot resolve it to a provider client.  The clean solution
is to implement a thin \code{DeepEvalBaseLLM} subclass whose
\code{generate} and \code{a\_generate} methods delegate to
\code{litellm.completion} and \code{litellm.acompletion} respectively, then
pass an instance of that class as the \code{model} argument.
\end{notebox}

\code{GEval} is DeepEval's configurable rubric-based metric.  It evaluates
a test case against a natural-language criteria string, producing a 0--1
score and a reason.  The \code{evaluation\_params} argument constrains the
judge's attention to only the specified \code{LLMTestCaseParams} fields,
preventing the judge from reading \code{expected\_output} as a shortcut
and inflating scores.  This makes \code{GEval} the right choice for
domain-specific quality dimensions that the built-in metrics do not cover.

\begin{tipbox}
For non-OpenAI models, wrap your LiteLLM call in a \code{DeepEvalBaseLLM}
subclass and pass the instance directly:
\texttt{FaithfulnessMetric(threshold=0.85, model=LiteLLMJudge())}.
Routing through LiteLLM means all provider failover, retry, and spend
tracking configured in Chapter~\ref{chap:litellm} applies automatically
to DeepEval judge calls as well.
\end{tipbox}

\begin{lstlisting}[language=Python, caption={DeepEval suite: per-item
    scoring with built-in and rubric-based metrics},
    label={lst:deepeval-suite}]
# src/eval/deepeval_suite.py
from __future__ import annotations
from dataclasses import dataclass
from deepeval.test_case import LLMTestCase, LLMTestCaseParams
from deepeval.metrics import (
    FaithfulnessMetric, AnswerRelevancyMetric, GEval,
)

@dataclass
class ItemResult:
    item_id: str
    metric:  str
    score:   float
    reason:  str
    passed:  bool

def build_test_case(item: dict) -> LLMTestCase:
    return LLMTestCase(
        input=item["input"]["query"],
        actual_output=item["run_output"]["answer"],
        expected_output=item["expected_output"].get("answer", ""),
        retrieval_context=item["run_output"].get(
            "retrieved_contexts", []),
    )

def run_deepeval_suite(
    items: list[dict],
    judge_model: object,          # DeepEvalBaseLLM instance
    faithfulness_threshold: float,
    relevancy_threshold:    float,
) -> list[ItemResult]:
    metrics = [
        FaithfulnessMetric(
            threshold=faithfulness_threshold, model=judge_model),
        AnswerRelevancyMetric(
            threshold=relevancy_threshold,    model=judge_model),
        GEval(
            name="citation_coverage",
            criteria=(
                "Does the answer cite or reference the specific "
                "source material used from the retrieved contexts?"
            ),
            evaluation_params=[
                LLMTestCaseParams.INPUT,
                LLMTestCaseParams.ACTUAL_OUTPUT,
                LLMTestCaseParams.RETRIEVAL_CONTEXT,
            ],
            threshold=0.70,
            model=judge_model,
        ),
    ]
    results: list[ItemResult] = []
    for item in items:
        tc = build_test_case(item)
        for metric in metrics:
            metric.measure(tc)
            results.append(ItemResult(
                item_id=item["id"],
                metric=metric.name,
                score=float(metric.score),
                reason=metric.reason or "",
                passed=metric.is_successful(),
            ))
    return results
\end{lstlisting}

Three decisions in Listing~\ref{lst:deepeval-suite} deserve explanation.
First, thresholds are passed in as arguments rather than embedded in the
metric constructors, because the \code{SuccessCriteria} dataclass in
\S\ref{sec:success-criteria} is the single owner of all threshold values;
duplicating them here would create the drift the success-criteria design is
specifically built to prevent.  Second, \code{judge\_model} is typed as
\code{object} rather than \code{str}; this signals to every caller that a
\code{DeepEvalBaseLLM} instance is expected, not a bare model-name string,
and prevents the \code{ValueError} that would otherwise surface only at
runtime on non-OpenAI providers.  Third, \code{metric.reason or ""} guards
against a \code{None} reason field: some metric implementations return
\code{None} when the judge call succeeds but produces no rationale text,
and a \code{None} reason stored in \code{ItemResult} causes downstream
string formatting in the CI report to raise a \code{TypeError}.

Item-level results are now available.  Before the harness can aggregate
them into a pass/fail decision, it needs a machine-readable definition of
what ``passing'' means --- one that is version-controlled, immutable after
declaration, and readable by CI without parsing prose.

%%------------------------------------------------------------
\section{Thresholds Must Be Declared, Not Inferred}
\label{sec:success-criteria}
%%------------------------------------------------------------

The design problem is threshold drift: without an explicit, version-controlled
definition of what constitutes a passing score, thresholds tend to migrate
toward whatever the system currently achieves.  The engineer running CI
notices a score of 0.79 against a 0.80 threshold, adjusts the threshold to
0.78 ``temporarily'', commits, and forgets.  Six months later the thresholds
have been lowered four times and no longer reflect product quality
requirements.  The naive approach of embedding thresholds as magic numbers
inside test files or environment variables makes them invisible in code
review.  The principle is that all thresholds live in a single, reviewed,
immutable module that is the only file CI reads for pass/fail decisions.

The implementation uses a \code{@dataclass(frozen=True)}: frozen prevents
post-construction mutation, and \code{\_\_post\_init\_\_} validates that every
threshold is in the valid range at construction time.  Freezing means that
any attempt to update a threshold after the object is created raises a
\code{FrozenInstanceError} immediately, making accidental runtime drift
impossible.

\begin{lstlisting}[language=Python, caption={Success criteria: a frozen\,
    validated threshold contract},
    label={lst:success-criteria}]
# src/eval/success_criteria.py
from __future__ import annotations
from dataclasses import dataclass, fields

@dataclass(frozen=True)
class SuccessCriteria:
    # RAGAS metrics (dataset-level means)
    faithfulness_min:       float = 0.85
    answer_relevancy_min:   float = 0.80
    context_precision_min:  float = 0.75
    context_recall_min:     float = 0.80
    # DeepEval per-item pass rate (fraction of items that pass)
    deepeval_pass_rate_min: float = 0.90
    # Online judge — alert threshold only, not a CI gate
    online_faithfulness_alert: float = 0.70

    def __post_init__(self) -> None:
        for f in fields(self):
            v = getattr(self, f.name)
            if not (0.0 <= v <= 1.0):
                raise ValueError(
                    f"Threshold {f.name}={v} must be in [0, 1]"
                )

    def ragas_passes(self, scores: dict[str, float]) -> bool:
        checks = {
            "faithfulness":      self.faithfulness_min,
            "answer_relevancy":  self.answer_relevancy_min,
            "context_precision": self.context_precision_min,
            "context_recall":    self.context_recall_min,
        }
        return all(
            scores.get(k, 0.0) >= v for k, v in checks.items()
        )

    def deepeval_passes(self, results: list) -> bool:
        if not results:
            return False
        n_passed = sum(1 for r in results if r.passed)
        return (n_passed / len(results)) >= self.deepeval_pass_rate_min

THRESHOLDS = SuccessCriteria()
\end{lstlisting}

Three decisions in Listing~\ref{lst:success-criteria} deserve explanation.
First, \code{fields(self)} from the \code{dataclasses} standard library is
used instead of \code{self.\_\_dataclass\_fields\_.items()}; the latter
returns \code{(name, Field)} pairs where the \code{Field} object is not the
current value, making it easy to accidentally validate the descriptor rather
than the threshold.  \code{fields(self)} returns \code{Field} objects whose
\code{.name} attribute is then passed to \code{getattr}, keeping the intent
unambiguous.  Second, \code{deepeval\_pass\_rate\_min} is a pass-rate
fraction rather than a per-metric threshold, because DeepEval operates at
item level: what matters for a CI gate is that no more than a specified
fraction of items fail across all metrics, not that every metric achieves a
certain mean.  Third, \code{THRESHOLDS = SuccessCriteria()} creates a
module-level singleton that every other module imports directly; this ensures
that all consumers see the same values and that a change to any threshold
triggers a visible diff on \code{src/eval/success\_criteria.py} in every pull
request that touches it.

With the threshold contract defined, the harness can now assemble the four
layers into a single reproducible CI run.

%%------------------------------------------------------------
\section{A Harness Is Not a Test File}
\label{sec:ci-harness}
%%------------------------------------------------------------

The design problem is that evaluation is not unit testing.  A unit test is
deterministic, sub-second, and tests a function's contract.  An evaluation
run is probabilistic, multi-minute, and measures a statistical distribution
of outputs across a population of inputs.  Treating evaluation as a pytest
test file --- one \code{assert} per item, failing the first assertion ---
produces a misleading CI report: a single failing item halts the run,
preventing the collection of data from the remaining items, and the engineer
cannot tell whether the run would have achieved 95\,\% pass rate or 20\,\%.
The principle is that the harness must \emph{collect all results before
making a pass/fail decision}, which requires a separate runner layer that
aggregates before asserting.

The harness has three responsibilities: pin the dataset to a version-stamped
snapshot, invoke the agent on every item concurrently (bounded by a semaphore
to avoid exhausting provider rate limits), and compute RAGAS and DeepEval
scores on the collected outputs before consulting \code{THRESHOLDS} for the
final decision.  The pytest integration is thin: a single test function calls
the harness and asserts on its boolean return value, so the full item-level
report is available in the CI log regardless of whether the test passes or
fails.

The runner function accepts the LLM, embeddings, and judge model as
explicit parameters rather than constructing them internally.  This keeps
the runner free of provider configuration so the test entry point can swap
providers without modifying source, and so the runner can be exercised
in isolation with deterministic stubs.

\begin{notebox}
The \code{run\_all\_items} function does not construct the \code{llm},
\code{embeddings}, or \code{judge\_model} objects.  These are built at the
test entry point (Listing~\ref{lst:test-eval-ci}) using \code{ChatLiteLLM}
and \code{LangchainLLMWrapper}, then injected here.  This separation means
the runner is testable in isolation: unit tests can pass
\code{unittest.mock.AsyncMock} for the agent and stub objects for the
LLM wrappers without touching any network.
\end{notebox}

\begin{lstlisting}[language=Python, caption={Evaluation runner: concurrent
    agent invocation with bounded parallelism},
    label={lst:eval-runner}]
# src/eval/run_eval.py
from __future__ import annotations
import asyncio
from ragas.llms import LangchainLLMWrapper
from ragas.embeddings import LangchainEmbeddingsWrapper
from src.eval.dataset_manager import fetch_dataset_at_version
from src.eval.ragas_suite import build_ragas_dataset, run_ragas_suite
from src.eval.deepeval_suite import run_deepeval_suite
from src.eval.success_criteria import THRESHOLDS

MAX_CONCURRENT = 8

async def _run_item(
    agent, item: dict, sem: asyncio.Semaphore,
) -> dict:
    async with sem:
        result = await agent.ainvoke(item["input"])
        return {**item, "run_output": result}

async def run_all_items(
    agent,
    version_ts: str,
    llm: LangchainLLMWrapper,
    embeddings: LangchainEmbeddingsWrapper,
    judge_model: object,
) -> tuple[bool, dict]:
    items = fetch_dataset_at_version(version_ts)
    sem   = asyncio.Semaphore(MAX_CONCURRENT)
    tasks = [_run_item(agent, it, sem) for it in items]
    run_results = await asyncio.gather(*tasks, return_exceptions=True)

    ok_items = [r for r in run_results
                if not isinstance(r, BaseException)]
    n_failed  = len(run_results) - len(ok_items)
    if n_failed:
        print(f"[eval] {n_failed} items raised exceptions; skipped.")

    ragas_scores = run_ragas_suite(
        build_ragas_dataset(ok_items), llm, embeddings)

    deepeval_res = run_deepeval_suite(
        ok_items,
        judge_model=judge_model,
        faithfulness_threshold=THRESHOLDS.faithfulness_min,
        relevancy_threshold=THRESHOLDS.answer_relevancy_min,
    )

    passed = (
        THRESHOLDS.ragas_passes(ragas_scores)
        and THRESHOLDS.deepeval_passes(deepeval_res)
    )
    report = {
        "ragas":         ragas_scores,
        "deepeval":      deepeval_res,
        "passed":        passed,
        "skipped_items": n_failed,
    }
    return passed, report
\end{lstlisting}

Three decisions in Listing~\ref{lst:eval-runner} deserve explanation.
First, \code{asyncio.gather(..., return\_exceptions=True)} returns
\code{BaseException} objects mixed into the result list rather than raising
on the first failure; the subsequent \code{isinstance} filter makes the
mixed-type contract explicit.  An item that raises --- because the agent
returned an unexpected schema or the provider timed out --- is logged and
skipped rather than aborting the run, because the purpose of the harness is
to collect data, and a partially complete run is more useful than no run.
Second, \code{MAX\_CONCURRENT = 8} is a module-level constant rather than
an environment variable because it protects against provider rate limits and
is a property of the evaluation workload, not the deployment environment;
changing it requires a code review, not a secret rotation.  Third,
\code{llm}, \code{embeddings}, and \code{judge\_model} are explicit
parameters so that the test entry point below owns the construction logic
and the runner stays portable across providers.

\begin{lstlisting}[language=Python, caption={pytest entry point: dependency
    construction and collect-before-assert},
    label={lst:test-eval-ci}]
# src/tests/test_eval_ci.py
import os
import pytest
from langchain_community.chat_models import ChatLiteLLM
from langchain_community.embeddings import HuggingFaceEmbeddings
from ragas.llms import LangchainLLMWrapper
from ragas.embeddings import LangchainEmbeddingsWrapper
from src.eval.run_eval import run_all_items
from src.agent.graph import build_agent
from src.eval.deepeval_suite import LiteLLMJudge  # DeepEvalBaseLLM wrapper

VERSION_TS  = os.environ["EVAL_DATASET_VERSION"]
JUDGE_MODEL = os.environ.get("EVAL_JUDGE_MODEL", "claude-haiku-4.5")

@pytest.mark.asyncio
async def test_eval_suite_passes() -> None:
    agent      = build_agent()
    llm        = LangchainLLMWrapper(
                     ChatLiteLLM(model=JUDGE_MODEL))
    embeddings = LangchainEmbeddingsWrapper(
                     HuggingFaceEmbeddings(
                         model_name="BAAI/bge-small-en-v1.5"))
    judge      = LiteLLMJudge(model=JUDGE_MODEL)

    passed, report = await run_all_items(
        agent, VERSION_TS, llm, embeddings, judge)

    print(report)   # always visible in pytest -s and CI captured output
    n_passed = sum(r.passed for r in report["deepeval"])
    n_total  = len(report["deepeval"])
    assert passed, (
        f"Evaluation suite FAILED\n"
        f"  RAGAS scores      : {report['ragas']}\n"
        f"  DeepEval pass rate: {n_passed}/{n_total}\n"
        f"  Skipped items     : {report['skipped_items']}"
    )
\end{lstlisting}

Three decisions in Listing~\ref{lst:test-eval-ci} deserve explanation.
First, \code{print(report)} runs unconditionally before the \code{assert}:
pytest captures stdout and includes it in the failure report, so the full
metric breakdown appears in the CI log even when the assertion fails --- the
engineer does not need to re-run the suite to see which metrics missed their
thresholds.  Second, \code{JUDGE\_MODEL} falls back to
\code{"claude-haiku-4.5"} from an environment variable, making it trivially
overridable in CI without modifying source; the default is a fast, cheap
model from a different family than the typical agent model, preserving judge
independence.  Third, the \code{LiteLLMJudge} wrapper imported from
\code{deepeval\_suite} is the \code{DeepEvalBaseLLM} subclass discussed in
\S\ref{sec:deepeval-metrics}; importing it here rather than constructing a
raw model string is the type-safe pattern that surfaces provider resolution
errors at import time rather than inside a long-running eval loop.

\begin{lstlisting}[language=yaml, caption={GitHub Actions workflow: pinned
    dataset version and compound cache key},
    label={lst:eval-workflow}]
# .github/workflows/eval.yml
name: Evaluation Suite
on:
  pull_request:
    branches: [main]
  schedule:
    - cron: "0 3 * * *"    # nightly against latest dataset state

jobs:
  eval:
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4

      - uses: actions/setup-python@v5
        with:
          python-version: "3.12"

      - name: Restore RAGAS metric cache
        uses: actions/cache@v4
        with:
          path: .ragas-cache
          key: >-
            ragas-${{ hashFiles('src/eval/success_criteria.py') }}-
            ${{ secrets.EVAL_DATASET_VERSION }}

      - name: Install dependencies
        run: pip install -e ".[eval]"

      - name: Run evaluation suite
        env:
          EVAL_DATASET_VERSION: ${{ secrets.EVAL_DATASET_VERSION }}
          EVAL_JUDGE_MODEL:     ${{ secrets.EVAL_JUDGE_MODEL }}
          LANGFUSE_PUBLIC_KEY:  ${{ secrets.LANGFUSE_PUBLIC_KEY }}
          LANGFUSE_SECRET_KEY:  ${{ secrets.LANGFUSE_SECRET_KEY }}
          LANGFUSE_BASE_URL:    ${{ secrets.LANGFUSE_BASE_URL }}
          OPENAI_API_KEY:       ${{ secrets.OPENAI_API_KEY }}
          ANTHROPIC_API_KEY:    ${{ secrets.ANTHROPIC_API_KEY }}
        run: pytest src/tests/test_eval_ci.py -v --tb=short
\end{lstlisting}

Three decisions in Listing~\ref{lst:eval-workflow} deserve explanation.
First, the cache key uses the YAML \code{>-} block scalar (folded,
strip-trailing-newline) rather than a backslash line continuation; a
backslash is a shell continuation character, not a YAML one, and would be
treated as a literal character in the key string, making cache hits
impossible.  Second, the cache key combines a hash of
\code{success\_criteria.py} with the dataset version secret: changing either
the thresholds or the pinned dataset snapshot invalidates the cache and
forces a fresh metric computation, preventing a scenario where stale cached
results from a higher-threshold run are accepted under a lowered threshold.
Third, \code{EVAL\_JUDGE\_MODEL} is exposed as a secret rather than
hard-coded so the team can upgrade the judge model independently of the
code --- useful when a new model family becomes available and the team wants
to compare judge scores across a model transition without a code diff.

\begin{tipbox}
Upload the evaluation report JSON as a GitHub Actions artefact using
\texttt{actions/upload-artifact} after the pytest step.  GitHub retains
artefacts for 90 days by default; storing the per-item DeepEval results and
RAGAS mean scores in a structured JSON file lets the team track metric
trends over time and run anomaly detection across the artefact history
without re-running the full suite.
\end{tipbox}

The evaluation infrastructure built in this chapter closes the quality loop:
golden items are curated in Langfuse, online judges fire silently on every
production trace, offline RAGAS and DeepEval suites gate every pull request,
and the success-criteria contract prevents threshold drift.  The remaining
open question is what happens when an evaluation run reveals a failure that
requires changing the agent's retrieval strategy --- a change that affects
the entire system simultaneously rather than a single node.
Chapter~\ref{chap:advanced-rag} examines how to restructure the retrieval
graph so that strategy changes are localised to a swappable subgraph rather
than scattered across every node that touches a retriever.
