%% ============================================================
%%  Chapter 12 --- Knowledge Graph Retrieval
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%%  Dependencies: langchain-neo4j >= 0.4.0,
%%                langchain-community >= 0.3.0,
%%                litellm >= 1.40.0, instructor >= 1.3.0,
%%                mem0ai >= 0.1.35, pydantic-settings >= 2.3.0
%% ============================================================

\chapterquote{A graph does not store facts. It stores the relationships that make facts answerable.}{anon}

The production failure this chapter prevents is invisible until it becomes
expensive. A customer-support agent with a dense vector memory can retrieve
every conversation in which a user mentioned a particular tool. What it
cannot answer is: \emph{which tools has this user used across all workflows
that also involved a specific model?} That question requires traversing a
path through a graph. The dense retriever returns passages that are
semantically similar to the query string, but semantic similarity is not
the same as relational reachability. The relevant memories may contain no
overlapping vocabulary with the query at all --- they are linked by structure,
not by words. Teams discover this failure mode when an agent that passed all
similarity-based retrieval tests begins producing answers that contradict
known facts about the user's history, because the facts were relational and
the retriever was not.

The solution this chapter builds has four layers. The first is a
\emph{connection layer}: a settings-driven \code{Neo4jGraph} client that
serves as the shared interface to the graph database, together with a
\code{GraphCypherQAChain} that translates natural-language questions into
validated Cypher queries. The second is an \emph{extraction layer}: an
LLM-driven pipeline that reads raw memory text, produces validated
\code{(subject, predicate, object)} triples, and writes them to Neo4j
using deterministic merge semantics. The third is a \emph{query layer}: a
multi-hop Cypher execution path for relational questions, with a mandatory
validation step that prevents LLM-hallucinated schema references from
reaching the database. The fourth is a \emph{routing layer}: a LangGraph
classifier node that decides per query whether to call the graph path, the
vector path from Chapter~11, or both --- merging the results through the
same \code{reciprocal\_rank\_fusion()} function already built. A fifth section
closes the chapter by establishing the write-ordering contract between
Neo4j and mem0 that keeps the two stores consistent without a distributed
transaction.

The sections must be read in order because each one hands a resolved
problem to the next. Section~\ref{sec:kg-connection} establishes the
connection and chain singletons that all other sections depend on.
Section~\ref{sec:kg-extraction} populates the graph; without data, the
query layer is untestable. Section~\ref{sec:kg-multihop} shows what the
graph enables that vectors cannot. Section~\ref{sec:kg-routing} wraps both
retrieval paths inside a single LangGraph dispatcher that makes the choice
transparent and auditable. Section~\ref{sec:kg-consistency} closes the
write path by ensuring that every memory written to Neo4j is immediately
cross-referenced in mem0.

This chapter introduces several new domain terms defined in
Table~\ref{tab:ch12-terms}.

\begin{table}[h]
\centering
\caption{Domain terms introduced in Chapter~12}
\label{tab:ch12-terms}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
\code{Triple} & A Pydantic model representing one \code{(subject, predicate, object\_)} fact extracted from memory text; maps to one Neo4j \code{Relationship} with two \code{Node} endpoints. \\
\code{GraphDocument} & A \code{langchain\_community} wrapper that bundles a list of \code{Node} and \code{Relationship} objects with a source \code{Document}; the unit of write for \code{add\_graph\_documents()}. \\
\code{CypherQueryCorrector} & A \code{langchain\_neo4j} post-processor that validates LLM-generated Cypher against the live graph schema before execution; activated automatically by \code{validate\_cypher=True}. \\
\code{RoutingDecision} & A three-valued \code{str} enum (\code{GRAPH}, \code{VECTOR}, \code{HYBRID}) that a LangGraph classifier node writes to state; read by the conditional edge to dispatch to the correct retrieval branch. \\
\code{ConsistencyResult} & A Pydantic model recording whether the Neo4j write and the mem0 additive tag succeeded; used for Langfuse alerting and retry scheduling. \\
\bottomrule
\end{tabularx}
\end{table}

The source layout for this chapter is:

\begin{verbatim}
src/
  graph/
    settings.py        # §12.1
    connection.py      # §12.1
    models.py          # §12.2
    extractor.py       # §12.2
    query.py           # §12.3
    routing.py         # §12.4
    kg_graph.py        # §12.4
    consistency.py     # §12.5
prompts/
  entity_extraction_v1.jinja2   # §12.2
  cypher_generation_v1.py       # §12.3
\end{verbatim}


%% ------------------------------------------------------------
\section{One Connection, Two Chains, Zero Hardcodes}
\label{sec:kg-connection}
%% ------------------------------------------------------------

The design problem is that a Neo4j connection requires a URL, a username,
and a password, and in every codebase where these values have been written
directly into a function call, at least one of them has eventually appeared
in a git commit. The naive approach --- passing strings from a
\code{settings.py} file that reads from a \code{.env} --- is not wrong, but
it leaves the secret type invisible: a plain \code{str} field for a password
is a loading dock left permanently open. The governing principle is that
credentials live in typed settings objects, passwords are \code{SecretStr}
so that their values are never included in log output or \code{repr()}, and
the connection object is constructed once and shared by reference.

\code{pydantic-settings} --- a library that maps environment variables to
Pydantic models with automatic type coercion --- is the mechanism. A
\code{BaseSettings} subclass declares each required variable as a typed
field; the framework reads the corresponding environment variable at
construction time and raises a \code{ValidationError} immediately if any
required variable is absent. This means the process fails at startup, not
mid-request, which is the correct failure mode for missing credentials.

\begin{lstlisting}[language=Python,
  caption={Neo4j settings model with typed secret field},
  label={lst:neo4j-settings}]
# src/graph/settings.py
from pydantic import SecretStr
from pydantic_settings import BaseSettings

class Neo4jSettings(BaseSettings):
    neo4j_url: str           # bolt://host:7687
    neo4j_user: str
    neo4j_password: SecretStr
    neo4j_database: str = "neo4j"

    model_config = {
        "env_file": ".env",
        "env_file_encoding": "utf-8",
    }
\end{lstlisting}

Two decisions in Listing~\ref{lst:neo4j-settings} deserve explanation.
\code{SecretStr} is not cosmetic: Pydantic renders its value as
\code{**********} in any \code{str()} or \code{repr()} call, and
\code{get\_secret\_value()} is the only method that returns the raw string.
This prevents the password from appearing in tracebacks, log lines, or any
serialisation that Langfuse (Chapter~23) may attach to a span.
The \code{neo4j\_database} field defaults to \code{"neo4j"} because the
default database name in both Community and Enterprise editions is
\code{"neo4j"}; teams that use named databases for tenant isolation can
override it per environment without changing code.

With the settings model defined, the connection and chain singletons can
be built. \code{Neo4jGraph} is the low-level interface: it wraps a Neo4j
Bolt driver and exposes \code{query()}, \code{refresh\_schema()}, and
\code{add\_graph\_documents()}. Two constructor parameters that affect
production behaviour need explicit values. \code{sanitize=True} strips
list-valued properties with more than 128 elements from query results ---
without it, a node that accumulated embedding-like float arrays in its
payload would flood the context window of the Cypher QA chain.
\code{enhanced\_schema=False} is the correct production default: when
\code{True}, the schema refresh scans the entire database for example
values, which is a full-table scan at the scale of an agent memory graph
and will time out or degrade throughput during peak load.

\begin{warningbox}
\code{Neo4jGraph} implements \code{\_\_enter\_\_} and \code{\_\_exit\_\_},
and the official documentation carries an explicit \textbf{Danger} notice
against relying on \code{\_\_del\_\_} for resource cleanup. Always open a
\code{Neo4jGraph} inside a \code{with} block or call \code{graph.close()}
in a \code{finally} clause. A connection that is not explicitly closed will
hold a Bolt session open until the process exits, which can exhaust the
connection pool under concurrent load.
\end{warningbox}

\code{GraphCypherQAChain} --- a LangChain chain that translates a
natural-language question into a Cypher query, executes it, and
post-processes the raw results into a natural-language answer --- is the
high-level interface. It can be constructed with a single \code{llm=}
argument, but in production the Cypher generation and the answer
post-processing are different tasks with different accuracy requirements.
Cypher generation requires a model that reliably follows schema constraints
and produces syntactically valid query strings; answer post-processing is
closer to summarisation and tolerates a cheaper, faster model. The
\code{from\_llm()} factory accepts separate \code{cypher\_llm=} and
\code{qa\_llm=} arguments for this split.

\begin{notebox}
\code{allow\_dangerous\_requests=True} is a class attribute on
\code{GraphCypherQAChain}, forwarded to \code{\_\_init\_\_} via
\code{**kwargs} from \code{from\_llm()}. The chain raises a
\code{ValueError} at construction time if this attribute is not set to
\code{True}. This is a deliberate security opt-in: the chain executes
LLM-generated Cypher against a live database, which is a write-capable
interface. The flag exists so that the decision to accept that risk is
explicit in every call site.
\end{notebox}

\begin{lstlisting}[language=Python,
  caption={Neo4jGraph and GraphCypherQAChain factory functions},
  label={lst:neo4j-connection}]
# src/graph/connection.py
from langchain_neo4j import Neo4jGraph, GraphCypherQAChain
from langchain_community.chat_models import ChatLiteLLM
from src.graph.settings import Neo4jSettings

def build_neo4j_graph(settings: Neo4jSettings) -> Neo4jGraph:
    return Neo4jGraph(
        url=settings.neo4j_url,
        username=settings.neo4j_user,
        password=settings.neo4j_password.get_secret_value(),
        database=settings.neo4j_database,
        sanitize=True,
        refresh_schema=True,
        enhanced_schema=False,
    )

def build_cypher_qa_chain(
    graph: Neo4jGraph,
    cypher_llm_model: str = "gpt-4o",
    qa_llm_model: str = "gpt-4o-mini",
) -> GraphCypherQAChain:
    cypher_llm = ChatLiteLLM(model=cypher_llm_model)
    qa_llm     = ChatLiteLLM(model=qa_llm_model)
    return GraphCypherQAChain.from_llm(
        cypher_llm=cypher_llm,
        qa_llm=qa_llm,
        graph=graph,
        validate_cypher=True,
        return_intermediate_steps=True,
        use_function_response=False,
        allow_dangerous_requests=True,
    )
\end{lstlisting}

Three decisions in Listing~\ref{lst:neo4j-connection} deserve explanation.
\code{return\_intermediate\_steps=True} is not optional for a production
system: it causes \code{chain.invoke()} to return the generated Cypher
string and the raw database results alongside the final answer, which
Chapter~23 attaches as span metadata in Langfuse. Without it, a wrong
answer is undebuggable because the Cypher that produced it is invisible.
\code{use\_function\_response=False} is set explicitly because not all
models proxied through LiteLLM support the function-calling wrapping that
this flag activates; leaving it at its default avoids a
provider-specific capability assumption. The \code{ChatLiteLLM} wrapper is
used rather than a provider-specific chat model to stay consistent with the
LiteLLM abstraction established in Chapter~5: swapping models requires a
string change, not a dependency swap.

\begin{notebox}
\code{langchain-neo4j} also exports \code{Neo4jSaver} (synchronous) and
\code{AsyncNeo4jSaver} (asynchronous), both implementing
\code{BaseCheckpointSaver} --- the LangGraph checkpoint saver interface
introduced in Chapter~3. If Neo4j is already in the infrastructure stack,
the same instance can serve as a LangGraph checkpoint backend. The
\code{Neo4jSaver} is opened as a context manager:
\code{with Neo4jSaver.from\_conn\_string(uri, user, password) as saver:},
and \code{saver.setup()} must be called once inside that context to create
the required indexes. The library also exports
\code{Neo4jChatMessageHistory}, a \code{BaseChatMessageHistory}
implementation that stores conversation turns as graph nodes, which
Chapter~7 covers as an alternative to in-memory message history.
\end{notebox}

The connection and chain are now singletons that every subsequent section
consumes. The next problem is filling the graph with structured knowledge
extracted from raw memory text.


%% ------------------------------------------------------------
\section{Structure Is Extracted, Not Stored}
\label{sec:kg-extraction}
%% ------------------------------------------------------------

The design problem is that memories arrive as unstructured text. A memory
entry such as ``The user completed a multi-step data pipeline using GPT-4o
and LangGraph, triggered by a Celery beat task'' contains at least three
facts --- user used GPT-4o, user used LangGraph, pipeline triggered by
Celery --- but a vector store represents this as a single embedding. A
knowledge graph can represent each fact as a separate, traversable edge.
The naive approach is to write the text directly to Neo4j as a node
property, which stores the string but not the structure: queries that ask
``what tools did this user use?'' still require full-text search, not graph
traversal. The governing principle is that structure must be extracted
before the write, validated before it reaches the database, and stored in
a form that makes relational queries possible.

The extraction pipeline has three responsibilities: call the LLM to
produce a list of \code{(subject, predicate, object)} triples, validate
the list with a Pydantic model that rejects empty output, and convert the
validated triples into \code{GraphDocument} objects that
\code{add\_graph\_documents()} can consume. Each responsibility is
isolated in its own function so that each can be tested and replaced
independently.

\begin{lstlisting}[language=Python,
  caption={Triple and ExtractionResult Pydantic models},
  label={lst:triple-models}]
# src/graph/models.py
from pydantic import BaseModel, model_validator

class Triple(BaseModel):
    subject:      str
    predicate:    str
    object_:      str           # not "object" — Python builtin
    subject_type: str = "Entity"
    object_type:  str = "Entity"

class ExtractionResult(BaseModel):
    triples: list[Triple]

    @model_validator(mode="after")
    def must_have_triples(self) -> "ExtractionResult":
        if not self.triples:
            raise ValueError(
                "LLM extraction returned zero triples; "
                "refusing to write empty graph document"
            )
        return self
\end{lstlisting}

Two decisions in Listing~\ref{lst:triple-models} deserve explanation.
The field is named \code{object\_} rather than \code{object} because
\code{object} is a Python builtin; shadowing it inside a class body causes
no runtime error but produces confusing behaviour when the class is
introspected or serialised. The \code{subject\_type} and \code{object\_type}
fields default to \code{"Entity"}: this default activates the
\code{\_\_Entity\_\_} secondary index that
\code{add\_graph\_documents(baseEntityLabel=True)} creates, which
accelerates MERGE operations on frequently-written node types. Teams whose
memory corpus has a known entity taxonomy can override these fields with
domain-specific labels (e.g., \code{"User"}, \code{"Tool"},
\code{"Workflow"}) by prompting the LLM to populate them.

The extraction function calls the same \code{instructor}-patched LiteLLM
client established in Chapter~11 \S11.6. The conversion function builds
\code{GraphDocument} objects from the validated triples. One detail in the
conversion is non-obvious: \code{add\_graph\_documents()} merges source
documents on the \code{id} field in \code{Document.metadata} when it is
present, and falls back to an MD5 hash of \code{page\_content} when it is
not. Using the MD5 fallback means that two memory entries with different
IDs but identical text would be merged into a single source node,
destroying provenance. Passing the \code{memory\_id} explicitly as
\code{metadata=\{"id": memory\_id\}} prevents this.

\begin{lstlisting}[language=Python,
  caption={Triple extraction and GraphDocument conversion},
  label={lst:extractor}]
# src/graph/extractor.py
import instructor, litellm
from langchain_core.documents import Document
from langchain_community.graphs.graph_document import (
    GraphDocument, Node, Relationship,
)
from src.graph.models import Triple, ExtractionResult

_client = instructor.from_litellm(litellm.completion)

def extract_triples(
    text: str, model: str = "gpt-4o"
) -> ExtractionResult:
    return _client.chat.completions.create(
        model=model,
        messages=[{"role": "user", "content": text}],
        response_model=ExtractionResult,
    )

def triples_to_graph_documents(
    result: ExtractionResult,
    source_text: str,
    memory_id: str,
) -> list[GraphDocument]:
    source = Document(
        page_content=source_text,
        metadata={"id": memory_id},
    )
    docs = []
    for t in result.triples:
        subj = Node(id=t.subject,  type=t.subject_type)
        obj  = Node(id=t.object_,  type=t.object_type)
        rel  = Relationship(source=subj, target=obj, type=t.predicate)
        docs.append(GraphDocument(
            nodes=[subj, obj], relationships=[rel], source=source,
        ))
    return docs
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={write\_triples\_to\_neo4j with deferred schema refresh},
  label={lst:write-triples}]
# src/graph/extractor.py (continued)
from langchain_neo4j import Neo4jGraph

def write_triples_to_neo4j(
    graph: Neo4jGraph,
    result: ExtractionResult,
    source_text: str,
    memory_id: str,
) -> None:
    graph_docs = triples_to_graph_documents(
        result, source_text, memory_id
    )
    graph.add_graph_documents(
        graph_docs,
        include_source=True,
        baseEntityLabel=True,
    )
    graph.refresh_schema()
\end{lstlisting}

Three decisions across Listings~\ref{lst:extractor}
and~\ref{lst:write-triples} deserve explanation. Each \code{Triple}
produces a separate \code{GraphDocument} with its own source reference
rather than packing all triples into one document. This is deliberate:
\code{add\_graph\_documents()} runs one MERGE transaction per document,
and batching multiple relationships into a single document with a shared
source node is correct only when all relationships have identical
provenance. Since each triple is an independently extractable fact, its
provenance is the source document, not the other triples.
\code{include\_source=True} writes the \code{Document} as a
\code{:Document} node in the graph and links it to each entity via a
\code{MENTIONS} relationship, enabling future queries such as ``which
memory entries mention GPT-4o?'' --- a provenance query that is free to
express in Cypher but impossible to express as a vector similarity search.
\code{graph.refresh\_schema()} is called once after the batch write, not
inside the loop: schema refresh is a full database scan that reads all
node labels and relationship types, so calling it per triple would make
write throughput proportional to schema scan time rather than write time.

The graph now contains structured knowledge. The next problem is querying
it for multi-hop relational questions that vectors cannot answer.


%% ------------------------------------------------------------
\section{Relational Questions Require Graph Traversal}
\label{sec:kg-multihop}
%% ------------------------------------------------------------

The design problem is that LLMs generate syntactically valid Cypher that
references node labels, relationship types, and property names that do not
exist in the live schema. This is not a fringe case: it is the expected
behaviour of a model that has seen Cypher in training data from many
different graph schemas. A hallucinated label such as \code{:Workflow}
when the actual label is \code{:Process} produces an empty result
silently, because Cypher's \code{MATCH} returns zero rows for a
non-existent label rather than raising an error. The agent interprets the
empty result as ``no relevant history found'' and proceeds to answer
incorrectly. Without schema validation, this failure is invisible and
reproducible on every query that uses the hallucinated label.

\code{validate\_cypher=True} on \code{GraphCypherQAChain} activates the
\emph{CypherQueryCorrector} --- a post-processor that parses the generated
Cypher, checks every label and relationship type against
\code{Neo4jGraph.get\_structured\_schema}, and either corrects minor
discrepancies or raises a \code{CypherQueryCorrectorException} that the
chain catches and retries. It is instantiated automatically from the graph
schema when the chain is constructed; no manual setup is required. The
corrector cannot fix every hallucination --- a label that bears no
resemblance to any schema label cannot be corrected --- but it eliminates
the most common class of failure: capitalisation differences and
near-matches that the LLM produces when a schema label is in its context
but not exactly recalled.

\begin{warningbox}
The second class of Cypher safety risk is data exfiltration: an LLM that
generates \code{MATCH (n) RETURN n} with no \code{WHERE} clause will
return every node in the database. The \code{validate\_cypher} corrector
does not prevent this. The defence is to inject a mandatory tenant
\code{WHERE} clause into every generated query via a custom
\code{cypher\_prompt} that the LLM cannot override. The tenant filter is
not a best-effort guideline in the prompt --- it is structural scaffolding
that the template imposes on every query regardless of what the LLM
generates inside it.
\end{warningbox}

The \code{cypher\_prompt} is a LangChain \code{PromptTemplate} --- not a
Jinja2 template --- because \code{GraphCypherQAChain.from\_llm()} accepts
a \code{PromptTemplate} directly for its \code{cypher\_prompt=} parameter.
The template variables \code{\{schema\}} and \code{\{question\}} are
consumed by the chain; additional literal text in the template is passed
to the LLM verbatim. The tenant filter is expressed as a structured rule
immediately before the output scaffold, making it part of the Cypher the
LLM is asked to complete rather than a separate instruction it can reason
around.

\begin{lstlisting}[language=Python,
  caption={Tenant-filtered Cypher generation prompt template},
  label={lst:cypher-prompt}]
# prompts/cypher_generation_v1.py
from langchain_core.prompts import PromptTemplate

_TEMPLATE = """\
You are a Cypher expert. Use only the schema below.
Do not reference labels or properties not in the schema.

Schema:
{schema}

Rules:
- Every MATCH must include a WHERE clause that filters
  on the user_id property of a node already in scope.
- Return only fields needed to answer the question.
- Do not use LIMIT unless the question asks for a count.

Question: {question}
Cypher:"""

def build_tenant_cypher_prompt(user_id: str) -> PromptTemplate:
    # Inject user_id as structural scaffolding before the
    # PromptTemplate is constructed so it cannot be omitted
    # at invocation time.
    filled = _TEMPLATE.replace(
        "on the user_id property",
        f"on the user_id = '{user_id}' property",
    )
    return PromptTemplate(
        input_variables=["schema", "question"],
        template=filled,
    )
\end{lstlisting}

Two decisions in Listing~\ref{lst:cypher-prompt} deserve explanation.
The \code{user\_id} is substituted into the template string before the
\code{PromptTemplate} is constructed rather than being passed as an input
variable. If \code{user\_id} were a template variable, the chain would
need to supply it at invocation time, creating a call-site dependency that
is easy to forget and produces a silent empty filter when the value is
missing. Substituting it at construction time makes the tenant filter a
structural property of the chain instance, not a runtime parameter.
The resulting \code{filled} string still contains \code{\{schema\}} and
\code{\{question\}} as live template variables, which the chain fills
automatically on every \code{invoke()} call.

With the prompt in place, the query execution function surfaces
intermediate steps for observability.

\begin{lstlisting}[language=Python,
  caption={run\_graph\_query with intermediate step extraction},
  label={lst:run-graph-query}]
# src/graph/query.py
from langchain_neo4j import GraphCypherQAChain, Neo4jGraph

def run_graph_query(
    chain: GraphCypherQAChain,
    question: str,
) -> dict:
    result = chain.invoke({"query": question})
    steps  = result.get("intermediate_steps", [])
    return {
        "answer":     result["result"],
        "cypher":     steps[0].get("query", "")     if steps else "",
        "db_results": steps[1].get("context", [])   if len(steps) > 1 else [],
    }

def direct_cypher_query(
    graph: Neo4jGraph,
    cypher: str,
    params: dict,
    session_params: dict | None = None,
) -> list[dict]:
    return graph.query(
        cypher,
        params=params,
        session_params=session_params or {},
    )
\end{lstlisting}

Three decisions in Listing~\ref{lst:run-graph-query} deserve explanation.
The \code{intermediate\_steps} list is accessed with defensive index
guards rather than direct subscript, because not all error-recovery paths
in \code{GraphCypherQAChain} populate both steps: a corrector failure may
short-circuit execution after the Cypher is generated but before the
database call. The step dicts are accessed via \code{.get("query")} and
\code{.get("context")} rather than positional subscript because
\code{langchain-neo4j} v0.4 stores steps as named dicts with these keys,
not as bare strings or lists. \code{direct\_cypher\_query()} exists as a
separate function for write operations and schema management calls that
should not pass through the LLM-mediated chain; the \code{session\_params}
argument accepts Neo4j session-level configuration such as database routing
mode for teams running Enterprise Edition routing policies.

The graph is now queryable for relational questions. The remaining problem
is deciding, per incoming query, whether the graph path, the vector path,
or both should be invoked.


%% ------------------------------------------------------------
\section{The Retrieval Decision Is a First-Class Node}
\label{sec:kg-routing}
%% ------------------------------------------------------------

The design problem is that the choice between graph traversal and vector
search is made implicitly in most systems --- either hardcoded in the
agent prompt, or left to the LLM to reason about in the same turn that
generates the answer. Both approaches produce unauditable routing
decisions: there is no record of why a particular retrieval path was
chosen, and a wrong choice is invisible until the answer is wrong. The
governing principle is that routing is a first-class node in the
LangGraph state machine, its decision is written to state as a typed
value, and the conditional edge that reads it makes the dispatch
explicit and testable.

The routing decision has three possible values, not two.
\emph{GRAPH} covers queries that name specific entities and ask about
the relationships between them --- counting, ordering, path traversal, or
any question that requires following edges rather than finding similar
text. \emph{VECTOR} covers semantically open queries where the user
describes a memory rather than naming an entity; similarity search over
dense embeddings is the correct tool. \emph{HYBRID} covers queries that
do both --- the user names an entity but also needs semantic context about
it that the graph does not carry as a property. A two-value enum forces
the caller to pick one and silently discards the other signal; three
values make the hybrid case explicit.

\begin{lstlisting}[language=Python,
  caption={RoutingDecision enum, RoutingResult model, and KGState},
  label={lst:routing-types}]
# src/graph/routing.py
from enum import Enum
from pydantic import BaseModel
from typing_extensions import TypedDict
from src.retrieval.types import RetrievedPassage

class RoutingDecision(str, Enum):
    GRAPH  = "graph"
    VECTOR = "vector"
    HYBRID = "hybrid"

class RoutingResult(BaseModel):
    decision: RoutingDecision
    reason: str

class KGState(TypedDict):
    query:             str
    user_id:           str
    routing_decision:  RoutingDecision | None
    graph_results:     list[dict]
    vector_passages:   list[RetrievedPassage]
    final_context:     str
\end{lstlisting}

Two decisions in Listing~\ref{lst:routing-types} deserve explanation.
\code{RoutingDecision(str, Enum)} inherits from both \code{str} and
\code{Enum}: the \code{str} mixin ensures that
\code{routing\_decision.value} returns a plain Python string rather than
an enum object, which is required because \code{add\_conditional\_edges()}
path map keys must be plain strings --- passing an enum object as a
dictionary key would silently miss all path map lookups at runtime.
The \code{reason} field on \code{RoutingResult} is not used
programmatically; it is logged to Langfuse (Chapter~23) as span metadata
so that routing decisions can be audited and their quality assessed
independently of the final answer quality.

The \code{HYBRID} branch requires converting \code{graph\_results} ---
a \code{list[dict]} returned by \code{run\_graph\_query()} --- into
\code{list[RetrievedPassage]} so that the same
\code{reciprocal\_rank\_fusion()} function from Chapter~11 \S11.4 can
merge them with vector results. This conversion is the only place where
the graph retrieval pipeline touches the vector retrieval pipeline's type
system, and it is intentionally the thinnest possible bridge: one list
comprehension that assigns rank by position and sets \code{source="graph"}.

\begin{lstlisting}[language=Python,
  caption={Routing classifier node, route function, and hybrid merge},
  label={lst:routing-nodes}]
# src/graph/kg_graph.py
import instructor, litellm
from langgraph.graph import StateGraph
from src.graph.routing import RoutingDecision, RoutingResult, KGState
from src.retrieval.fusion import reciprocal_rank_fusion
from src.retrieval.types import RetrievedPassage

_client = instructor.from_litellm(litellm.completion)
CLASSIFIER_MODEL = "gpt-4o-mini"

def routing_classifier_node(state: KGState) -> dict:
    result = _client.chat.completions.create(
        model=CLASSIFIER_MODEL,
        messages=[{
            "role": "user",
            "content": (
                "Classify this query: graph traversal (GRAPH), "
                "semantic similarity (VECTOR), or both (HYBRID).\n"
                f"Query: {state['query']}"
            ),
        }],
        response_model=RoutingResult,
    )
    return {"routing_decision": result.decision}

def route_query(state: KGState) -> str:
    return state["routing_decision"].value

def hybrid_merge_node(state: KGState) -> dict:
    graph_passages = [
        RetrievedPassage(
            id=str(i), text=str(r), score=1.0,
            rank=i + 1, source="graph", meta=r,
        )
        for i, r in enumerate(state["graph_results"])
    ]
    fused = reciprocal_rank_fusion(
        [graph_passages, state["vector_passages"]]
    )
    return {"final_context": "\n".join(p["text"] for p in fused)}
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={Stub retrieval nodes and KG StateGraph assembly},
  label={lst:kg-graph}]
# src/graph/kg_graph.py (continued)

def graph_retrieval_node(state: KGState) -> dict:
    # Inject GraphCypherQAChain via context_schema (Ch. 12 §12.3).
    raise NotImplementedError("inject chain via context_schema")

def vector_retrieval_node(state: KGState) -> dict:
    # Inject QdrantVectorStore + Ranker via context_schema (Ch. 11).
    raise NotImplementedError("inject store via context_schema")

def build_kg_graph() -> StateGraph:
    builder = StateGraph(state_schema=KGState)
    builder.add_node("routing_classifier", routing_classifier_node)
    builder.add_node("graph_retrieval",    graph_retrieval_node)
    builder.add_node("vector_retrieval",   vector_retrieval_node)
    builder.add_node("hybrid_merge",       hybrid_merge_node)
    builder.set_entry_point("routing_classifier")
    builder.add_conditional_edges(
        "routing_classifier",
        route_query,
        {
            "graph":  "graph_retrieval",
            "vector": "vector_retrieval",
            "hybrid": "hybrid_merge",
        },
    )
    builder.add_edge("graph_retrieval",  "__end__")
    builder.add_edge("vector_retrieval", "__end__")
    builder.add_edge("hybrid_merge",     "__end__")
    return builder.compile()
\end{lstlisting}

Three decisions across Listings~\ref{lst:routing-nodes}
and~\ref{lst:kg-graph} deserve explanation.
\code{StateGraph(state\_schema=KGState)} uses the \code{state\_schema=}
parameter, not the deprecated \code{config\_schema=} removed in
LangGraph v0.6.0. Runtime singletons such as the
\code{GraphCypherQAChain} and the \code{QdrantVectorStore} are injected
via \code{context\_schema=} when immutable per-run context is needed, which
keeps the node signatures free of infrastructure dependencies. The
\code{route\_query()} function returns
\code{state["routing\_decision"].value} --- the plain string value ---
because \code{add\_conditional\_edges()} path map keys must be plain
strings; returning the enum object directly would silently fail all path
map lookups. In \code{hybrid\_merge\_node()}, graph results are converted
to \code{RetrievedPassage} with \code{source="graph"} and rank assigned
by position, then merged via the existing \code{reciprocal\_rank\_fusion()}
with its default \code{k=60}: no new merging logic is introduced, and
Chapter~11's variadic \code{ranked\_lists} argument is exactly the
extension point that makes this zero-change integration possible.

The routing layer makes the retrieval decision transparent and
dispatchable. The final problem is ensuring that every memory written to
Neo4j is immediately cross-referenced in mem0, so that the two stores do
not drift apart.


%% ------------------------------------------------------------
\section{Two Stores Converge, Never Synchronise}
\label{sec:kg-consistency}
%% ------------------------------------------------------------

The design problem is that the system now writes to two stores on the same
event: \code{write\_triples\_to\_neo4j()} writes structured facts to Neo4j,
and mem0 holds the original memory blob. If the two writes happen
independently, a failure in either one produces an inconsistency that is
silent at the call site. The naive approach is to treat them as equivalent
and execute them in an arbitrary order, accepting that either can fail
first. The governing principle is that the stores are not equivalent ---
Neo4j holds the canonical structured representation, and the mem0 metadata
tag is a derived cross-reference --- and this asymmetry determines a
non-arbitrary write order.

Neo4j must be written first. The reasoning is about failure modes, not
performance. If the Neo4j write fails, the mem0 additive tag is never
created, and no cross-reference points to a non-existent entity; the
system remains consistent. If Neo4j succeeds and the mem0 tag write
subsequently fails, the entity exists in the graph and is queryable via
Cypher, the mem0 memory blob is retrievable via vector search, and only
the cross-reference link is missing --- a gap that can be retried
independently using the tenacity patterns from Chapter~27. If mem0 were
written first and the Neo4j write then failed, the cross-reference would
point to an entity that does not exist in the graph, producing a silent
query failure with no self-healing path.

\begin{notebox}
The mem0 v3 API is \emph{additive-only}. There is no
\code{MemoryClient.update(metadata=...)} method that patches the metadata
of an existing memory in place. The correct pattern for attaching a
\code{graph\_entities} tag to a memory is to call \code{client.add()}
with \code{infer=False} and a structured message. This creates a new
additive memory entry that carries the cross-reference; the original
memory is not mutated. The \code{infer=False} flag bypasses the mem0
extraction pipeline, which would otherwise attempt to derive additional
memories from the message content and could produce spurious entries.
\end{notebox}

The \code{ConsistencyResult} model records both write outcomes and the
\code{event\_id} returned by the mem0 \code{add()} call. The mem0 v3
\code{add()} endpoint is asynchronous at the platform level: it returns
\code{\{"status": "PENDING", "event\_id": "..."\}} immediately, and the
tag does not appear in \code{client.search()} results until the background
job completes. The \code{event\_id} is stored so that the Langfuse alert
(Chapter~23) and the retry scheduler (Chapter~27) have a stable identifier
to reference when polling for completion or re-submitting a failed tag.

\begin{lstlisting}[language=Python,
  caption={ConsistencyResult model and sync\_graph\_to\_mem0()},
  label={lst:consistency}]
# src/graph/consistency.py
from pydantic import BaseModel
from mem0 import MemoryClient
from langchain_neo4j import Neo4jGraph
from src.graph.models import ExtractionResult
from src.graph.extractor import write_triples_to_neo4j

class ConsistencyResult(BaseModel):
    neo4j_written:   bool
    mem0_add_queued: bool
    mem0_event_id:   str | None
    memory_id:       str
    entity_count:    int

def sync_graph_to_mem0(
    graph: Neo4jGraph,
    mem0_client: MemoryClient,
    memory_id: str,
    text: str,
    result: ExtractionResult,
    user_id: str,
) -> ConsistencyResult:
    write_triples_to_neo4j(graph, result, text, memory_id)
    entity_names = [t.subject for t in result.triples]
    resp = mem0_client.add(
        messages=[{
            "role": "system",
            "content": f"Linked graph entities: {entity_names}",
        }],
        user_id=user_id,
        metadata={
            "graph_entities":    entity_names,
            "linked_memory_id":  memory_id,
        },
        infer=False,
    )
    event_id = (
        resp.get("event_id") if isinstance(resp, dict) else None
    )
    return ConsistencyResult(
        neo4j_written=True,
        mem0_add_queued=event_id is not None,
        mem0_event_id=event_id,
        memory_id=memory_id,
        entity_count=len(entity_names),
    )
\end{lstlisting}

Three decisions in Listing~\ref{lst:consistency} deserve explanation.
\code{neo4j\_written=True} is set on the success path only because
\code{write\_triples\_to\_neo4j()} raises on failure --- the
\code{ConsistencyResult} is never constructed when the Neo4j write fails.
Chapter~27 wraps \code{sync\_graph\_to\_mem0()} with a tenacity retry
decorator that catches those exceptions and handles the failure path
without this function needing to encode retry logic. The
\code{entity\_names} list uses only the \code{subject} field of each
triple, not \code{object\_}: object-position values are often generic
strings (\code{"true"}, \code{"None"}, \code{"2026-01-01"}) that pollute
the cross-reference tag with values that carry no graph traversal utility.
The \code{isinstance(resp, dict)} guard handles the case where the mem0
client version returns a response object rather than a raw dict; the
\code{event\_id} extraction degrades to \code{None} rather than raising an
\code{AttributeError} that would incorrectly mark the Neo4j write as
failed.

Chapter~13 builds the semantic caching layer that sits in front of the
entire retrieval system --- both the vector pipeline from Chapter~11 and
the graph pipeline from this chapter --- and prevents redundant LLM and
database calls for queries that have been answered recently.
