%% ============================================================
%%  Chapter 2 -- Termination, Resource Limits & Loop Failure Guards
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%% ============================================================

\chapterquote{A system that cannot stop is not autonomous. It is runaway.}
{Attributed to Edsger W.\ Dijkstra, adapted}

Chapter~1 built the smallest complete agent loop you can put into production: four
typed nodes, conditional routing, a checkpointer, and Langfuse spans on every
cycle. It left one question open: what happens when the loop does not stop?

The answer is expensive and embarrassing. A loop without explicit termination
guards can exhaust a token budget, run up a four-figure API bill overnight, or
produce thousands of log entries documenting a system that was effectively frozen
but still billing. None of these failures produce a loud error. They produce
silent accumulation until a threshold is crossed that was never set.

This chapter installs four independent layers of protection. The first is a
\code{TerminationPolicy} Pydantic model that consolidates all resource axes into
a single validated, serialisable config object. The second is LangGraph's native
\code{recursion\_limit}, a structural ceiling that fires regardless of
application logic. The third is an application-level cycle counter checked by a
routing function at the start of every cycle. The fourth is a suite of semantic
detectors---infinite loop, stuck state, and goal divergence---each with its own
dedicated recovery branch.

Two-layer guard design is the central principle. Neither the LangGraph ceiling
nor the application counter is sufficient on its own. Understanding why both are
required, and why they must be independent, is the core engineering lesson of
this chapter.

%% ------------------------------------------------------------
\section{\texttt{TerminationPolicy}: The Four-Axis Config Object}
\label{sec:termination-policy}
%% ------------------------------------------------------------

Termination limits almost always start as constants in the code. A
\code{MAX\_CYCLES = 10} at the top of \code{agent.py}, a
\code{TOKEN\_BUDGET = 8000} two files away, a wall-clock check that someone
added after an incident and stored in a config YAML under a different key name.
Over time these constants diverge. Different task types need different limits.
A deployment rolls out with the wrong value because there was no single place
to update. A checkpoint is written with the constants at one set of values;
the agent is resumed after the constants have changed.

A single Pydantic \code{BaseModel} solves all of these at once.

The choice of \code{BaseModel} over \code{@dataclass} is deliberate and
consequential. Pydantic provides three things a dataclass does not: free JSON
serialisation via \code{model\_dump()}, which means the policy can be stored
inside \code{AgentState} and survive a checkpoint round-trip; field validators
that enforce invariants at construction time, not at the first use of a value
in a guard condition; and \code{model\_config = \{"frozen": True\}}, which makes
the object immutable after construction so no node can silently relax a limit
during a run.

The four axes cover every dimension of runaway cost. \code{max\_cycles} bounds
the number of complete Perceive-Reason-Act-Reflect iterations. \code{token\_ceiling}
bounds cumulative token consumption across all model calls in the run.
\code{wall\_clock\_timeout} bounds elapsed real time in seconds---useful for
user-facing tasks where latency is a product requirement, not just a cost concern.
\code{cost\_cap} bounds total spend in USD, not in tokens. The distinction between
\code{token\_ceiling} and \code{cost\_cap} matters: a run that calls a cheap model
twenty times may exhaust the token ceiling while staying well under the cost cap;
a run that makes two calls to a premium model may breach the cost cap after a
fraction of the token ceiling. Both axes are needed.

\code{wall\_clock\_timeout} and \code{cost\_cap} are typed \code{float | None}.
A policy with only \code{max\_cycles} and \code{token\_ceiling} set is valid;
the orchestrator simply skips the checks for \code{None} axes. The alternative---
omitting those fields entirely---creates a different, more dangerous situation:
the code never checks for a wall-clock limit at all, so no future operator can
enable it without a code change.

\begin{lstlisting}[language=Python,
  caption={\code{TerminationPolicy}: validated, immutable, serialisable},
  label={lst:terminationpolicy}]
# src/agent/termination.py
from pydantic import BaseModel, field_validator

class TerminationPolicy(BaseModel):
    max_cycles:         int
    token_ceiling:      int
    wall_clock_timeout: float | None = None  # seconds
    cost_cap:           float | None = None  # USD
    model_config = {"frozen": True}

    @field_validator("max_cycles")
    @classmethod
    def max_cycles_positive(cls, v: int) -> int:
        if v <= 0:
            raise ValueError(f"max_cycles must be > 0, got {v}")
        return v

    @field_validator("token_ceiling")
    @classmethod
    def ceiling_positive(cls, v: int) -> int:
        if v <= 0:
            raise ValueError(
                f"token_ceiling must be > 0, got {v}"
            )
        return v
\end{lstlisting}

\begin{warningbox}
Do not store a live \code{TerminationPolicy} instance in \code{AgentState}.
The checkpointer serialises state as JSON; Pydantic objects are not
JSON-serialisable by default. Store \code{policy.model\_dump()} and reconstruct
with \code{TerminationPolicy.model\_validate(state["_policy"])} when a node needs
the object. Chapter~3 shows this pattern in \code{make\_initial\_state}.
\end{warningbox}

%% ------------------------------------------------------------
\section{LangGraph Native Recursion Guard: \texttt{recursion\_limit}
         and \texttt{GraphRecursionError}}
\label{sec:recursion-guard}
%% ------------------------------------------------------------

\code{recursion\_limit} is a structural property of the compiled graph. It is set
in \code{b.compile(recursion\_limit=N)}, not in \code{config["configurable"]}.
This distinction is not a naming convention---it reflects what the parameter
actually controls. A \code{config["configurable"]} value can be changed per run
because it is a per-run parameter passed through the runtime. \code{recursion\_limit}
is baked into the compiled graph object; changing it requires recompiling.

LangGraph counts \emph{node executions}, not cycles. A four-node graph running
ten cycles executes 40 node steps. Setting \code{recursion\_limit = max\_cycles}
would trigger after the first ten node steps---after two and a half cycles, not
ten. The correct formula is \code{recursion\_limit = max\_cycles * 4} for a
four-node graph, with headroom added for recovery and guard nodes.

\code{recursion\_limit} is the last line of defence. It fires when every other
limit has failed to stop the loop---when the application-level cycle counter
was not incremented due to a node bug, when the token budget check was bypassed
by an exception handler, when a routing bug caused the graph to cycle through
a two-node sub-path rather than the full four-node path. It is structural: no
amount of application code can suppress it once the node-step count is reached.

An uncaught \code{GraphRecursionError} in production means the caller receives
a Python exception instead of a structured \code{AgentRunResult}. The calling
service crashes, logs a stack trace, and has no structured record of what the
agent accomplished before it was stopped. Catching it explicitly and converting
it to a \code{stopped\_by="recursion\_limit"} result is not optional.

\begin{warningbox}
Never raise \code{recursion\_limit} as a workaround for missing termination
logic. The correct response to a loop that reaches the recursion limit is to
add proper termination guards---not to increase the number. A higher limit is
a more expensive infinite loop.
\end{warningbox}

\begin{lstlisting}[language=Python,
  caption={\code{AgentRunResult} and \code{run\_agent} with explicit
           \code{GraphRecursionError} handling},
  label={lst:runner}]
# src/agent/runner.py
from dataclasses import dataclass
from langgraph.errors import GraphRecursionError

@dataclass
class AgentRunResult:
    state:      dict | None
    stopped_by: str   # "completion" | "recursion_limit"

def run_agent(
    graph,
    initial_state: dict,
    config: dict,
) -> AgentRunResult:
    try:
        result = graph.invoke(initial_state, config=config)
        return AgentRunResult(state=result,
                              stopped_by="completion")
    except GraphRecursionError:
        return AgentRunResult(state=None,
                              stopped_by="recursion_limit")
\end{lstlisting}

The \code{stopped\_by} string becomes the \code{termination\_status} field in the
final \code{CycleTrace} record (Chapter~1 \S\ref{sec:cycle-trace}). A result with
\code{stopped\_by="recursion\_limit"} is incomplete even if the agent produced
partial output. Logging it as a success because a \code{final\_answer} field
happens to be present is a monitoring error.

%% ------------------------------------------------------------
\section{Application-Level Max-Cycle Policy: The Second Guard Layer}
\label{sec:app-cycle-guard}
%% ------------------------------------------------------------

\code{recursion\_limit} alone is not sufficient. The failure mode it does not
cover is the one that occurs when \code{cycle\_count} is not incremented
correctly. Consider a node bug that returns an empty dict instead of
\code{\{"cycle\_count": state["cycle\_count"] + 1\}}. From LangGraph's perspective,
the graph is still executing---each node execution counts toward the recursion
limit---but from the application's perspective, the cycle counter never advances.
The application-level guard never fires. The graph hits \code{recursion\_limit}
at \code{max\_cycles * 4} node steps, which may be far more than the intended
number of semantic cycles.

Application-level alone is also not sufficient. It depends entirely on the
\code{cycle\_count} increment being correct in every code path. The
\code{recursion\_limit} is structural: it counts node executions in the LangGraph
runtime, independently of any value in \code{AgentState}. Application code
cannot suppress it, reset it, or bypass it. These two independence properties
are precisely why both layers are necessary.

The application guard lives in \code{pre\_cycle\_gate}, a routing function inserted
at the loop entry point. It runs before \code{perceive\_node} on every cycle. If
either the cycle limit or the token floor is breached, it routes to a dedicated
recovery node rather than continuing. The gate checks both axes so that there is
one decision point at loop entry, not two separate routing functions that could
fire at different times.

\code{cycle\_count} is incremented inside \code{perceive\_node}, the first node
of each cycle. It must not be incremented inside \code{reflect\_node}: if a cycle
fails mid-execution and is retried, the count must reflect cycles \emph{entered},
not cycles \emph{completed}. An incomplete cycle that consumed tokens still
counts against the limit.

\begin{lstlisting}[language=Python,
  caption={\code{pre\_cycle\_gate}: the two-check loop entry guard},
  label={lst:precyclegate}]
# src/agent/graph.py
from src.agent.termination import TerminationPolicy

_MINIMUM_SAFE_BUDGET = 500   # from src/agent/nodes/reason.py

def pre_cycle_gate(state: AgentState) -> str:
    policy = TerminationPolicy.model_validate(
        state.get("_policy") or {}
    )
    if state["cycle_count"] >= policy.max_cycles:
        return "cycle_limit_exceeded"
    if state["token_budget"] < _MINIMUM_SAFE_BUDGET:
        return "budget_exhausted"
    return "perceive_node"

def build_graph(policy: TerminationPolicy,
                checkpointer=None):
    b = StateGraph(AgentState)
    b.add_node("perceive_node",        perceive_node)
    b.add_node("reason_node",          reason_node)
    b.add_node("action_node",          action_node)
    b.add_node("reflect_node",         reflect_node)
    b.add_node("cycle_limit_exceeded", cycle_limit_node)
    b.add_node("budget_exhausted",     budget_exhausted_node)
    b.set_conditional_entry_point(
        pre_cycle_gate,
        {"perceive_node":        "perceive_node",
         "cycle_limit_exceeded": "cycle_limit_exceeded",
         "budget_exhausted":     "budget_exhausted"},
    )
    # ... remaining edges omitted for brevity (see ss2.8)
    return b.compile(
        checkpointer=checkpointer,
        recursion_limit=policy.max_cycles * 4,
    )
\end{lstlisting}

\begin{notebox}
\code{recursion\_limit} is derived from \code{policy.max\_cycles} at compile time,
not from a hard-coded constant. This means the structural ceiling and the
application ceiling are always proportional---a policy change updates both
without requiring a separate edit to \code{compile()}.
\end{notebox}

%% ------------------------------------------------------------
\section{Cumulative Token Budget: \texttt{token\_budget} Graceful Stop}
\label{sec:token-budget}
%% ------------------------------------------------------------

Chapter~1's \code{reason\_node} raises \code{BudgetExhaustedError} when
\code{token\_budget} falls below \code{\_MINIMUM\_SAFE\_BUDGET}. That behaviour
is correct inside a unit test where the node is called in isolation. In
production, an exception raised inside a node propagates uncaught through the
graph unless explicitly handled by an error edge, producing the same unstructured
failure outcome as an uncaught \code{GraphRecursionError}.

The production pattern is different: the budget check moves to \code{pre\_cycle\_gate},
before the node is ever invoked. When the gate detects a sub-threshold budget, it
routes to \code{budget\_exhausted\_node}, which writes a structured
\code{termination\_status} and routes to \code{END}. No exception is raised; the
caller receives a clean \code{AgentRunResult}.

\code{token\_budget} is a floor, not a count-up counter. It starts at
\code{TerminationPolicy.token\_ceiling} and is decremented by \code{reason\_node}
each cycle. It must never go negative: \code{max(0, remaining - used)} is the
correct update expression. The floor semantics matter because the budget is also
read by the \code{pre\_cycle\_gate} as a routing signal---a negative value would
pass the \code{< \_MINIMUM\_SAFE\_BUDGET} check correctly, but any arithmetic on
it in downstream nodes would produce nonsensical results.

The \code{pre\_cycle\_gate} intentionally combines both checks---cycle count and
token budget---into one routing function. Two separate routing functions at the
loop entry would require two conditional entry points, which LangGraph does not
natively support, and would also create an ordering dependency: which check fires
first when both limits are breached simultaneously? A single gate with an explicit
ordering (cycle count checked before token budget) is unambiguous.

%% ------------------------------------------------------------
\section{Infinite Loop Detector: \texttt{goal\_delta} and Repeated Actions}
\label{sec:infinite-loop}
%% ------------------------------------------------------------

The cycle counter and token budget guard against \emph{resource overrun}.
They do not guard against \emph{semantic stagnation}: a loop that executes
correctly, calls tools successfully, decrements the budget, increments the cycle
counter, and makes no real progress toward the goal. This is the infinite loop
failure mode, and it is silent precisely because the instrumentation metrics
all look healthy.

The signal is \code{goal\_delta}, introduced in Chapter~1 \S\ref{sec:reflect}.
\code{ReflectNode} emits \code{goal\_delta: Literal["ADVANCED", "NO\_CHANGE",
"REGRESSED"]} in every \code{CycleTrace}. A single \code{NO\_CHANGE} is not
alarm-worthy---a cycle that retrieved context without yet acting on it may
legitimately report no goal advancement. Three consecutive \code{NO\_CHANGE}
values, with the same tool being called each time, is the definition of an
infinite loop.

The detector is a pure function of \code{AgentState}. It reads
\code{state["cycle\_traces"]}, inspects the last \code{window} entries, and
returns a typed signal object. No model call, no I/O, no side effects. This
makes it testable with synthetic trace sequences (\S\ref{sec:ch2-tests}) and
cheap to run on every cycle.

\begin{lstlisting}[language=Python,
  caption={\code{InfiniteLoopSignal}, \code{detect\_infinite\_loop}, and
           \code{detect\_repeated\_action}},
  label={lst:infiniteloop}]
# src/agent/guards.py
from dataclasses import dataclass
from src.agent.state import AgentState

@dataclass
class InfiniteLoopSignal:
    triggered:             bool
    consecutive_no_change: int
    repeated_action:       str | None  # tool name, or None

def detect_repeated_action(
    state: AgentState,
    window: int = 3,
) -> str | None:
    traces = state.get("cycle_traces", [])
    if len(traces) < window:
        return None
    recent = [t.get("action_invoked") for t in traces[-window:]]
    if all(a is not None and a == recent[0] for a in recent):
        return recent[0]
    return None

def detect_infinite_loop(
    state: AgentState,
    window: int = 3,
) -> InfiniteLoopSignal:
    traces = state.get("cycle_traces", [])
    if len(traces) < window:
        return InfiniteLoopSignal(
            triggered=False,
            consecutive_no_change=0,
            repeated_action=None,
        )
    recent  = traces[-window:]
    no_chg  = sum(1 for t in recent
                  if t.get("goal_delta") == "NO_CHANGE")
    rep_act = detect_repeated_action(state, window)
    triggered = (no_chg >= window)
    return InfiniteLoopSignal(
        triggered=triggered,
        consecutive_no_change=no_chg,
        repeated_action=rep_act,
    )
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={\code{loop\_guard\_node}: calls the detector and routes on trigger},
  label={lst:loop-guard-node}]
def loop_guard_node(state: AgentState) -> dict:
    signal = detect_infinite_loop(state)
    if signal.triggered:
        return {
            "reflect_decision": "ESCALATE",
            "beliefs": [
                f"[loop-guard] infinite loop detected: "
                f"{signal.consecutive_no_change} consecutive "
                f"NO_CHANGE cycles, "
                f"repeated_action={signal.repeated_action}"
            ],
        }
    return {}
\end{lstlisting}

%% ------------------------------------------------------------
\section{Stuck-State Detector: No Valid Output for N Cycles}
\label{sec:stuck-state}
%% ------------------------------------------------------------

A stuck state is distinct from an infinite loop. An infinite loop \emph{makes
progress in form}---it calls tools, tools return results, \code{CycleTrace} records
are written---while making no progress toward the goal. A stuck state does not
even produce a valid next action: the model returns \code{ReasoningTrace} cycle
after cycle without committing to either a tool call or a final answer.

Chapter~1 \S\ref{sec:reason} established that a single \code{ReasoningTrace} is
legitimate. The model may need to externalise an intermediate deliberation step
before it is ready to commit to an action. What is not legitimate is \emph{N
consecutive} \code{ReasoningTrace} outputs with no intervening \code{TOOL\_CALL}
or \code{FINAL\_ANSWER}. That pattern means the model is circling without
committing---a symptom of an under-specified goal, an ambiguous tool set, or a
prompt that encourages over-deliberation.

The detection signal is \code{CycleTrace.reason\_output\_type}. The stuck-state
detector reads the last \code{window} traces and counts entries where
\code{reason\_output\_type} is neither \code{"TOOL\_CALL"} nor
\code{"FINAL\_ANSWER"}. When the count equals \code{window}, the detector
triggers.

\begin{lstlisting}[language=Python,
  caption={\code{StuckStateSignal} and \code{detect\_stuck\_state}},
  label={lst:stuckstate}]
@dataclass
class StuckStateSignal:
    triggered:               bool
    consecutive_non_actions: int
    last_output_types:       list[str]

def detect_stuck_state(
    state: AgentState,
    window: int = 3,
) -> StuckStateSignal:
    traces = state.get("cycle_traces", [])
    if len(traces) < window:
        return StuckStateSignal(
            triggered=False,
            consecutive_non_actions=0,
            last_output_types=[],
        )
    recent = traces[-window:]
    types  = [t.get("reason_output_type", "") for t in recent]
    non_actions = sum(
        1 for ot in types
        if ot not in ("TOOL_CALL", "FINAL_ANSWER")
    )
    return StuckStateSignal(
        triggered=(non_actions >= window),
        consecutive_non_actions=non_actions,
        last_output_types=types,
    )

def stuck_state_guard_node(state: AgentState) -> dict:
    signal = detect_stuck_state(state)
    if signal.triggered:
        return {
            "reflect_decision": "ESCALATE",
            "beliefs": [
                f"[stuck-guard] {signal.consecutive_non_actions} "
                f"consecutive non-action outputs: "
                f"{signal.last_output_types}"
            ],
        }
    return {}
\end{lstlisting}

%% ------------------------------------------------------------
\section{Goal Divergence Detector: Cosine Distance and Rollback}
\label{sec:goal-divergence}
%% ------------------------------------------------------------

Goal divergence is the most dangerous failure mode covered in this chapter.
An infinite loop and a stuck state are both visible in the metrics: cycle count
advances without a \code{FINAL\_ANSWER}, or \code{reason\_output\_type} shows
a long run of \code{REASONING\_TRACE}. Goal divergence is invisible in the same
metrics: the loop is executing, tools are being called, \code{goal\_delta} may
even show \code{ADVANCED} occasionally---but the agent is advancing toward the
wrong objective.

This happens because the \code{goal} field in \code{AgentState} is immutable
by convention, not by runtime enforcement. As the \code{beliefs} list grows and
the model's sliding context window fills with tool results and prior reasoning,
the original goal text recedes. The model continues to reason about the most
recent beliefs, which may have accumulated around a subtask rather than the
primary objective. The agent is solving a problem---just not the right one.

Embedding-based cosine distance gives a quantitative measure of this drift.
The original goal is embedded once at graph entry and stored in
\code{AgentState.goal\_embedding}. At each detection check, the current goal
summary---reconstructed from the most recent beliefs or explicitly tracked---is
embedded and compared. When the cosine \emph{distance} exceeds
\code{DIVERGENCE\_THRESHOLD = 0.25}, the detector triggers.

Two cost concerns must be addressed. First, embedding is not free: a single
embedding call costs tokens and adds latency. The solution is to cache the
original embedding in \code{AgentState} (one call at entry, never again) and
to check divergence only every \code{N\_DIVERGENCE\_CHECK\_INTERVAL} cycles
(default 5), not every cycle. Second, the check-interval must be calibrated
against the expected run length: checking every 5 cycles in a 6-cycle run means
the detector fires only once, which is sufficient.

\begin{lstlisting}[language=Python,
  caption={\code{cosine\_distance}, \code{embed\_text},
           and \code{detect\_goal\_divergence}},
  label={lst:goaldivergence-fns}]
import math
from litellm import embedding as litellm_embedding

_EMBED_MODEL              = "text-embedding-3-small"
DIVERGENCE_THRESHOLD      = 0.25
N_DIVERGENCE_CHECK_INTERVAL = 5

def embed_text(text: str) -> list[float]:
    resp = litellm_embedding(model=_EMBED_MODEL, input=[text])
    return resp.data[0]["embedding"]

def cosine_distance(
    a: list[float], b: list[float]
) -> float:
    dot   = sum(x * y for x, y in zip(a, b))
    mag_a = math.sqrt(sum(x * x for x in a))
    mag_b = math.sqrt(sum(y * y for y in b))
    if mag_a == 0 or mag_b == 0:
        return 1.0
    return 1.0 - dot / (mag_a * mag_b)

@dataclass
class GoalDivergenceSignal:
    triggered:           bool
    cosine_distance:     float
    rollback_checkpoint: str | None

def detect_goal_divergence(
    state: AgentState,
    threshold: float = DIVERGENCE_THRESHOLD,
) -> GoalDivergenceSignal:
    orig_emb = state.get("goal_embedding")
    if orig_emb is None:
        return GoalDivergenceSignal(
            triggered=False, cosine_distance=0.0,
            rollback_checkpoint=None)
    current_summary = " ".join(
        state.get("beliefs", [])[-3:]
    ) or state["goal"]
    curr_emb = embed_text(current_summary)
    dist     = cosine_distance(orig_emb, curr_emb)
    return GoalDivergenceSignal(
        triggered=(dist > threshold),
        cosine_distance=dist,
        rollback_checkpoint=None,  # set by guard node
    )
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={\code{goal\_divergence\_guard\_node}: every-N-cycles check with rollback},
  label={lst:goaldivergence-node}]
def goal_divergence_guard_node(
    state: AgentState,
    config: dict,
) -> dict:
    cycle = state.get("cycle_count", 0)
    last  = state.get("goal_divergence_checked_at") or 0
    if cycle - last < N_DIVERGENCE_CHECK_INTERVAL:
        return {}  # not yet due for a check

    signal = detect_goal_divergence(state)
    updates: dict = {
        "goal_divergence_checked_at": cycle
    }
    if signal.triggered:
        updates["reflect_decision"] = "ESCALATE"
        updates["beliefs"] = [
            f"[divergence-guard] cosine distance "
            f"{signal.cosine_distance:.3f} exceeds threshold "
            f"{DIVERGENCE_THRESHOLD}; initiating rollback"
        ]
    return updates
\end{lstlisting}

\begin{tipbox}
Chapter~6 covers embedding model selection in detail. For the divergence
detector, latency and cost matter more than quality: the goal embedding is a
short text compared to another short text. \code{text-embedding-3-small} is
an appropriate default. Chapter~6 also shows how to cache the model's tokeniser
for faster batch embedding when multiple agents are running concurrently.
\end{tipbox}

%% ------------------------------------------------------------
\section{Per-Failure-Mode LangGraph Recovery Branches}
\label{sec:recovery-branches}
%% ------------------------------------------------------------

Three detectors produce three different signals, and those signals require three
different recovery strategies. Routing all three to a single
\code{failure\_node} is the obvious shortcut, and it is the wrong one: the
recovery from an infinite loop (increment a retry counter, check if the retry
ceiling is reached, escalate if so) is completely different from the recovery
from a stuck state (escalate to a human handoff immediately) and from the
recovery from goal divergence (roll back to the last aligned checkpoint, re-anchor
the goal in beliefs, re-enter the perceive cycle).

Merging these paths into a generic handler means that the most expensive
recovery action---checkpoint rollback with goal re-anchoring---runs for a loop
that could have been resolved with a simple retry increment. It also means that
the \code{termination\_status} field in the \code{CycleTrace} becomes meaningless:
``failure'' instead of ``infinite loop after N cycles'' or ``goal drift above
threshold at cycle K''.

Each guard node returns a routing signal via \code{reflect\_decision}. The
routing function after the guard layer reads that value and directs the run to
the appropriate recovery node. Recovery nodes write a structured
\code{termination\_status} and route to \code{END}---except
\code{goal\_divergence\_recovery\_node}, which routes back to \code{perceive\_node}
after re-anchoring the goal.

\begin{lstlisting}[language=Python,
  caption={Three recovery nodes with distinct strategies},
  label={lst:recovery-nodes}]
# src/agent/guards.py  (continued)

def infinite_loop_recovery_node(
    state: AgentState,
) -> dict:
    policy = TerminationPolicy.model_validate(
        state.get("_policy") or {}
    )
    retries = state.get("_retry_count", 0) + 1
    if retries >= getattr(policy, "max_retries", 3):
        return {
            "reflect_decision": "ESCALATE",
            "final_answer": (
                f"Infinite loop detected after {retries} retries; "
                "escalating."
            ),
            "_retry_count": retries,
        }
    return {
        "reflect_decision": "RETRY",
        "_retry_count":     retries,
    }

def stuck_state_recovery_node(
    state: AgentState,
) -> dict:
    return {
        "reflect_decision": "ESCALATE",
        "final_answer": (
            "Agent is stuck: no valid action produced after "
            "3 consecutive cycles. Escalating for human review."
        ),
    }

def goal_divergence_recovery_node(
    state: AgentState,
    config: dict,
) -> dict:
    goal = state["goal"]
    return {
        "beliefs":          [f"[divergence-recovery] goal "
                             f"re-anchored: {goal}"],
        "reflect_decision": None,  # clear stale decision
    }
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={\code{build\_graph} with guard nodes, recovery nodes, and wiring},
  label={lst:graph-ch2}]
def build_graph(
    policy: TerminationPolicy,
    checkpointer=None,
):
    b = StateGraph(AgentState)
    # -- core nodes --
    for name, fn in [
        ("perceive_node",  perceive_node),
        ("reason_node",    reason_node),
        ("action_node",    action_node),
        ("reflect_node",   reflect_node),
    ]:
        b.add_node(name, fn)
    # -- resource limit nodes --
    b.add_node("cycle_limit_exceeded",  cycle_limit_node)
    b.add_node("budget_exhausted",      budget_exhausted_node)
    # -- guard nodes --
    b.add_node("loop_guard_node",       loop_guard_node)
    b.add_node("stuck_state_guard_node",stuck_state_guard_node)
    b.add_node("goal_divergence_guard", goal_divergence_guard_node)
    # -- recovery nodes --
    b.add_node("infinite_loop_recovery",   infinite_loop_recovery_node)
    b.add_node("stuck_state_recovery",     stuck_state_recovery_node)
    b.add_node("goal_divergence_recovery", goal_divergence_recovery_node)
    # -- entry --
    b.set_conditional_entry_point(
        pre_cycle_gate,
        {"perceive_node":        "perceive_node",
         "cycle_limit_exceeded": "cycle_limit_exceeded",
         "budget_exhausted":     "budget_exhausted"},
    )
    # -- core loop --
    b.add_edge("perceive_node", "reason_node")
    b.add_conditional_edges(
        "reason_node", route_from_reason,
        {"action_node":   "action_node",
         "perceive_node": "perceive_node", "end": END})
    b.add_edge("action_node",  "reflect_node")
    b.add_edge("reflect_node", "loop_guard_node")
    b.add_edge("loop_guard_node", "stuck_state_guard_node")
    b.add_edge("stuck_state_guard_node", "goal_divergence_guard")
    # -- guard routing --
    b.add_conditional_edges(
        "goal_divergence_guard", _route_after_guards,
        {"infinite_loop_recovery":   "infinite_loop_recovery",
         "stuck_state_recovery":     "stuck_state_recovery",
         "goal_divergence_recovery": "goal_divergence_recovery",
         "continue":                 "perceive_node",
         "end":                      END})
    # -- recovery exits --
    b.add_conditional_edges(
        "infinite_loop_recovery", _route_after_recovery,
        {"perceive_node": "perceive_node", "end": END})
    b.add_edge("stuck_state_recovery",     END)
    b.add_edge("goal_divergence_recovery", "perceive_node")
    b.add_edge("cycle_limit_exceeded",     END)
    b.add_edge("budget_exhausted",         END)
    return b.compile(
        checkpointer=checkpointer,
        recursion_limit=policy.max_cycles * 4,
    )
\end{lstlisting}

%% ------------------------------------------------------------
\section{Test Suite: Guard Triggering Tests}
\label{sec:ch2-tests}
%% ------------------------------------------------------------

Every detector in this chapter is a pure function of \code{AgentState}. No
compiled graph is required to test them. Tests construct synthetic
\code{cycle\_traces} lists using a \code{make\_cycle\_trace} helper and pass
them directly to the detector functions. This makes the test cases fast,
deterministic, and independent of LangGraph's execution engine.

The \code{GraphRecursionError} path requires one integration test: a minimal
graph compiled with \code{recursion\_limit=2} and a node that always returns
\code{CONTINUE}, invoked via \code{run\_agent} to confirm the error is caught
and converted to a structured result.

\begin{lstlisting}[language=Python,
  caption={Guard test fixture and detector unit tests},
  label={lst:guard-tests}]
# tests/test_guards.py
from src.agent.guards import (
    detect_infinite_loop, detect_stuck_state,
    detect_goal_divergence,
)
from unittest.mock import patch

def make_cycle_trace(
    goal_delta:         str       = "NO_CHANGE",
    reason_output_type: str       = "REASONING_TRACE",
    action_invoked:     str|None  = None,
) -> dict:
    return {
        "goal_delta":          goal_delta,
        "reason_output_type":  reason_output_type,
        "action_invoked":      action_invoked,
    }

def _state_with_traces(traces: list[dict]) -> dict:
    return {"cycle_traces": traces, "goal": "test goal",
            "beliefs": [], "goal_embedding": None,
            "goal_divergence_checked_at": None,
            "cycle_count": len(traces)}

def test_infinite_loop_triggers_after_3_no_change():
    traces = [make_cycle_trace(goal_delta="NO_CHANGE",
                               action_invoked="fetch_report")] * 3
    sig = detect_infinite_loop(_state_with_traces(traces))
    assert sig.triggered
    assert sig.consecutive_no_change == 3
    assert sig.repeated_action == "fetch_report"

def test_infinite_loop_no_trigger_below_window():
    traces = [make_cycle_trace(goal_delta="NO_CHANGE")] * 2
    assert not detect_infinite_loop(
        _state_with_traces(traces)).triggered

def test_stuck_state_triggers_after_3_reasoning_traces():
    traces = [make_cycle_trace(
        reason_output_type="REASONING_TRACE")] * 3
    sig = detect_stuck_state(_state_with_traces(traces))
    assert sig.triggered
    assert sig.consecutive_non_actions == 3

def test_stuck_state_no_trigger_with_tool_call():
    traces = [
        make_cycle_trace(reason_output_type="REASONING_TRACE"),
        make_cycle_trace(reason_output_type="TOOL_CALL"),
        make_cycle_trace(reason_output_type="REASONING_TRACE"),
    ]
    assert not detect_stuck_state(
        _state_with_traces(traces)).triggered

def test_goal_divergence_triggers_above_threshold():
    orig_emb = [1.0, 0.0]
    with patch("src.agent.guards.embed_text",
               return_value=[0.0, 1.0]):
        state = _state_with_traces([])
        state["goal_embedding"] = orig_emb
        state["beliefs"]        = ["drifted context"]
        sig = detect_goal_divergence(state, threshold=0.1)
    assert sig.triggered
    assert sig.cosine_distance > 0.1

def test_goal_divergence_no_trigger_below_threshold():
    orig_emb = [1.0, 0.0]
    with patch("src.agent.guards.embed_text",
               return_value=[0.99, 0.14]):
        state = _state_with_traces([])
        state["goal_embedding"] = orig_emb
        state["beliefs"]        = ["still on track"]
        sig = detect_goal_divergence(state, threshold=0.25)
    assert not sig.triggered
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={\code{test\_graphrecursion\_caught\_by\_runner}: integration test},
  label={lst:runner-test}]
# tests/test_runner.py
from langgraph.graph import StateGraph, END
from src.agent.runner import run_agent, AgentRunResult
from src.agent.termination import TerminationPolicy
from src.agent.state import AgentState

def test_graphrecursion_caught_by_runner():
    """A graph that always loops must return stopped_by=recursion_limit."""
    b = StateGraph(AgentState)

    def looping_node(state):
        return {"cycle_count": state["cycle_count"] + 1}

    b.add_node("looping_node", looping_node)
    b.set_entry_point("looping_node")
    b.add_edge("looping_node", "looping_node")
    graph = b.compile(recursion_limit=2)

    policy  = TerminationPolicy(max_cycles=10,
                                 token_ceiling=8000)
    initial = {"cycle_count": 0, "token_budget": 8000,
               "goal": "test", "beliefs": [],
               "intentions": [], "messages": [],
               "cycle_traces": [], "last_action": None,
               "reflect_decision": None,
               "final_answer": None, "_policy": None,
               "goal_embedding": None,
               "goal_divergence_checked_at": None,
               "_recovery_cycle": False,
               "schema_version": 2}
    result = run_agent(graph, initial, config={})
    assert result.stopped_by == "recursion_limit"
    assert result.state is None
\end{lstlisting}

Chapter~28 extends this suite to full integration tests: a compiled production
graph with \code{SqliteSaver} checkpointing, synthetic multi-cycle state
sequences that exercise every detector and every recovery branch, and assertions
on the complete \code{CycleTrace} record emitted by each run.

\begin{takeawaybox}
Termination is a first-class design problem, not an afterthought. Every
production agent needs four independent axes of resource protection:
\code{max\_cycles}, \code{token\_ceiling}, \code{wall\_clock\_timeout}, and
\code{cost\_cap}---collected in a single \code{TerminationPolicy} object that is
validated at construction and immutable at runtime. Two guard layers are required
because neither alone is sufficient: \code{recursion\_limit} is structural and
cannot be bypassed, but knows nothing about semantic progress;
the application cycle counter knows about semantic progress, but can be defeated
by a node bug that skips the increment. Three semantic detectors protect against
failures that resource limits cannot see: infinite loops (\code{goal\_delta}
stagnant across N cycles), stuck states (\code{reason\_output\_type} producing no
valid action for N cycles), and goal divergence (cosine distance between original
and current goal embedding above threshold). Each detector routes to its own
dedicated recovery node because recovery strategies are not interchangeable.
Chapter~3 asks the question that follows naturally: once the agent stops, whether
cleanly or by a guard, what state has it left behind, and how can that state be
resumed safely?
\end{takeawaybox}
