%% ============================================================
%%  Chapter 21 --- Human-in-the-Loop
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%%  Dependencies: langgraph >= 0.2.0,
%%                langchain-core >= 0.3.0,
%%                pydantic >= 2.7.0,
%%                fastapi >= 0.115.0,
%%                sse-starlette >= 2.1.0
%% ============================================================

\chapterquote{Automation earns trust action by action. The gate is not
a failure of confidence. It is the mechanism by which confidence is
built.}{anon}

The production failure this chapter prevents is an irreversible automated
action taken without human awareness. The parallel research graph of
Chapter~20 dispatches retrieval workers and synthesises an answer entirely
without pausing. That design is correct when the graph's output is a
document. It becomes a liability when the graph's output is an action: an
email sent to a client, a database record modified, a deployment triggered,
a purchase authorised. At the moment the graph reaches the action node it
has no mechanism to ask whether this particular action, with this particular
payload, is one a human would approve. The cost of the missing gate is not a
degraded result---it is an unrecoverable real-world event that no subsequent
checkpoint can undo. Deferring the gate until ``the agent misbehaves in
production'' inverts the correct engineering posture: the gate must exist
before the first real action runs, not after the first real mistake.

The solution has five layers. The first is an \emph{idempotency layer}:
every action node is designed to produce the same outcome whether it runs
once or ten times, because the Pregel scheduler may re-enter a node from
the top after a checkpoint restore, an interrupt, or a resume. The second
is a \emph{pause layer}: the \code{interrupt()} primitive, called inside a
node body after computing the action payload but before executing it, raises
a special exception that the runtime catches, serialises into the checkpoint,
and surfaces to the caller as an \code{Interrupt} event. The third is a
\emph{resume layer}: the caller injects a human decision back into the paused
thread by passing \code{Command(resume=value)} as the input to a second
\code{ainvoke} call on the same thread; the runtime re-enters the node,
reaches \code{interrupt()} again, and this time returns the stored decision
immediately without pausing. The fourth is a \emph{state-edit layer}: before
resuming, a human reviewer may call \code{graph.aupdate\_state()} to modify
the checkpoint's state at the breakpoint---correcting a payload, adjusting a
parameter, or overriding a field---without rerunning any nodes. The fifth is
a \emph{time-travel layer}: \code{get\_state\_history()} exposes the full
ordered checkpoint sequence so any past snapshot can be replayed, creating a
corrected execution branch from a prior point without mutating the original
run.

The sections are ordered by the dependency each layer has on the one before
it. Section~\ref{sec:hitl-idempotent} establishes idempotency first because
a node that is not idempotent cannot be safely interrupted or
resumed---re-entry will duplicate its side effects regardless of how
correctly the pause and resume mechanics work.
Section~\ref{sec:hitl-interrupt} introduces \code{interrupt()} after
idempotency is established, because the correct placement of the call---after
computing the payload, before executing the side effect---only makes sense
once the re-entry invariant is understood.
Section~\ref{sec:hitl-resume} shows \code{Command(resume=value)} after the
pause mechanics are established, because the resume contract---a new
\code{ainvoke} call, not a callback---requires understanding what
\code{interrupt()} serialised into the checkpoint.
Section~\ref{sec:hitl-edit} adds \code{aupdate\_state()} after resume because
state editing is a pre-resume operation whose result must be used as the
config for the subsequent \code{ainvoke} call.
Section~\ref{sec:hitl-timetravel} introduces time-travel after editing
because the \code{get\_state\_history()} iterator and the \code{as\_node}
metadata written by \code{aupdate\_state()} are prerequisites for finding the
correct replay snapshot.
Section~\ref{sec:hitl-policy} introduces \code{PolicyConfig} after all five
mechanics are established, because the policy is a configuration layer over
them---it does not change how they work, only whether a given node invokes
the pause path.
Section~\ref{sec:hitl-api} wires everything into a FastAPI surface.
Section~\ref{sec:hitl-full} provides the integration tests.

This chapter introduces eight terms defined in Table~\ref{tab:ch21-terms}.

\begin{table}[h]
\centering
\caption{Domain terms introduced in Chapter~21}
\label{tab:ch21-terms}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
\code{interrupt()} &
  \code{from langgraph.types import interrupt}; a synchronous function
  callable from both sync and async node bodies; raises
  \code{GraphInterrupt} on first call, causing Pregel to serialise the
  \code{value} argument into the current checkpoint and return an
  \code{Interrupt} event to the caller; on resume, returns the value
  stored by \code{Command(resume=...)} instead of raising. \\
\code{Command(resume=...)} &
  \code{from langgraph.types import Command}; when passed as the
  \emph{input} argument to \code{graph.ainvoke()} or
  \code{graph.astream()} on a paused thread, instructs the runtime to
  re-enter the interrupted node and return the \code{resume} value from
  the \code{interrupt()} call; routing is determined by
  \code{StateSnapshot.next}, not by \code{goto}. \\
\code{ActionRecord} &
  A \code{TypedDict} with fields \code{action\_id:~str},
  \code{node\_name:~str}, \code{payload:~dict}, and
  \code{status:~str}~(\code{"pending"} | \code{"approved"} |
  \code{"rejected"} | \code{"executed"}); one record per action node
  invocation, keyed by deterministic \code{action\_id} in
  \code{pending\_actions}. \\
\code{action\_id} &
  A 12-character hex prefix of
  \code{sha256(json.dumps(\{"node":~name,~"payload":~p\},
  sort\_keys=True))}; stable across re-runs of the same node with the
  same input; used as the merge key in the \code{\_merge\_actions}
  reducer to prevent duplicate records under re-entry. \\
\code{\_merge\_actions} reducer &
  Declared as \code{Annotated[list[ActionRecord],~\_merge\_actions]};
  performs a right-wins merge by \code{action\_id}: an incoming record
  with a known \code{action\_id} replaces the existing entry; a record
  with an unknown \code{action\_id} is appended; prevents duplicate
  \code{ActionRecord} entries on re-run. \\
\code{aupdate\_state()} &
  \code{await graph.aupdate\_state(config,~values,~as\_node=...)}
  merges \code{values} into the checkpoint via each field's declared
  reducer, as if a node had returned those values; returns a new
  \code{RunnableConfig} with an updated \code{checkpoint\_id} that
  must be used for the subsequent resume call. \\
\code{get\_state\_history()} &
  \code{graph.get\_state\_history(config)} returns an
  \code{Iterator[StateSnapshot]} in reverse chronological order
  (newest first); each \code{StateSnapshot} carries \code{.config}
  (with \code{checkpoint\_id}), \code{.values}, \code{.next}, and
  \code{.metadata} (containing \code{"writes"}, \code{"source"},
  \code{"step"}). \\
\code{PolicyConfig} &
  A Pydantic \code{BaseModel} that extends Chapter~14~\S14.3's runtime
  configuration model with \code{approval\_gates:~list[ApprovalGate]}
  and \code{auto\_approve\_all:~bool}; passed via
  \code{RunnableConfig["configurable"]["policy"]}; consulted by action
  nodes to decide whether to call \code{interrupt()}. \\
\bottomrule
\end{tabularx}
\end{table}

\medskip
\noindent\textbf{Source layout for this chapter:}
\begin{verbatim}
chapter21/
  state.py          # §21.1 — ApprovalState, ActionRecord,
                    #          _merge_actions, make_action_id
  action_node.py    # §21.2 — action_node with interrupt()
                    #          and idempotency guard
  runner.py         # §21.3, §21.4 — start_run, resume_run,
                    #                 edit_and_resume
  time_travel.py    # §21.5 — find_last_pre_edit_snapshot,
                    #          replay_from_snapshot
  policy.py         # §21.6 — ApprovalGate, PolicyConfig
  graph.py          # §21.8 — build_approval_graph,
                    #          synthesise_node
  api.py            # §21.7 — FastAPI app, SSE endpoint,
                    #          resume endpoint
  test_hitl.py      # §21.8 — three scenario tests
\end{verbatim}


%% ------------------------------------------------------------
\section{A Node That Can Be Interrupted Must Tolerate Re-entry}
\label{sec:hitl-idempotent}
%% ------------------------------------------------------------

The design problem is side-effect duplication under re-entry. The Pregel
scheduler checkpoints graph state \emph{before} calling a node function,
not after. When a node is interrupted---or when the process restarts after
a crash, or when a human's resume call re-enters the graph---the runtime
calls the node function again from the top with the pre-interrupt state as
its argument. Any external write that happens before the \code{interrupt()}
call---a database insert, an HTTP POST, a file write---executes again on
re-entry. A node that inserts a database record unconditionally on every
call will insert two records: one before the interrupt, one on resume. The
governing principle is that every action node must check whether its
intended side effect has already been committed before attempting to commit
it again.

The check-before-write pattern requires a stable key to look up. A key
generated with \code{uuid4()} inside the node body changes on every call:
the first invocation creates key~A, the second invocation creates
key~B, and the lookup of key~A finds nothing, so the second invocation
executes again unconditionally. The \emph{action\_id} must instead be
derived deterministically from the node's input---the name of the node
and the content of its payload---so that the same inputs always produce
the same key. A \code{sha256} hash of the JSON-serialised payload with
\code{sort\_keys=True} produces a 12-character hex prefix that is stable
across runs, resistant to field-order variation, and short enough to store
in every \code{ActionRecord}.

The \emph{pending\_actions} field that accumulates \code{ActionRecord}
entries across the graph's lifetime requires a reducer that merges by
\code{action\_id} rather than appending unconditionally. Without such a
reducer, a re-entered node that writes an updated \code{ActionRecord}
(changing \code{status} from \code{"pending"} to \code{"executed"}) would
append a second record rather than replacing the first, leaving two
records for the same logical action in the field---one stale, one
current---with no defined ordering between them.

\begin{lstlisting}[language=Python,
  caption={State schema, by-ID reducer, and deterministic action
           identifier},
  label={lst:hitl-state}]
# chapter21/state.py
import hashlib, json
from typing import Annotated, TypedDict
from langchain_core.messages import BaseMessage
from langgraph.graph.message import add_messages

class ActionRecord(TypedDict):
    action_id:  str   # sha256(node + json(payload))[:12]
    node_name:  str
    payload:    dict
    status:     str   # "pending"|"approved"|"rejected"|"executed"

def _merge_actions(
    left:  list[ActionRecord],
    right: list[ActionRecord],
) -> list[ActionRecord]:
    index = {r["action_id"]: r for r in left}
    for r in right:
        index[r["action_id"]] = r   # replace or insert
    return list(index.values())

class ApprovalState(TypedDict):
    task:            str
    messages:        Annotated[list[BaseMessage], add_messages]
    pending_actions: Annotated[list[ActionRecord], _merge_actions]
    approved_count:  int
    final_output:    str

def make_action_id(node_name: str, payload: dict) -> str:
    raw = json.dumps(
        {"node": node_name, "payload": payload}, sort_keys=True
    )
    return hashlib.sha256(raw.encode()).hexdigest()[:12]
\end{lstlisting}

Three decisions in Listing~\ref{lst:hitl-state} deserve explanation.
\code{ActionRecord} is a \code{TypedDict} rather than a Pydantic
\code{BaseModel} because it is a coordination record whose instances
are plain dicts at runtime; the \code{\_merge\_actions} reducer
operates directly on dict keys without needing \code{.model\_dump()},
and constructing instances with dict-literal syntax throughout the module
keeps the code uniform. The \code{\_merge\_actions} function is defined
before \code{ApprovalState} in the module because Python evaluates the
class body immediately: the \code{Annotated} expression references
\code{\_merge\_actions} by name, so the function must exist at class-
definition time or the module raises a \code{NameError} on import.
\code{make\_action\_id} uses \code{sort\_keys=True} on \code{json.dumps}
because Python dicts preserve insertion order, meaning the same logical
payload constructed from different code paths may produce different key
orderings and therefore different hashes; sorting eliminates that entire
class of non-determinism silently.

The idempotency contract is established. The remaining problem is the
pause mechanism itself: how to halt a node at a precise point, serialise
a structured payload for the human reviewer, and guarantee that the
external action does not execute until the reviewer responds.


%% ------------------------------------------------------------
\section{Execution Pauses at the Interrupt, Not at the Edge}
\label{sec:hitl-interrupt}
%% ------------------------------------------------------------

The design problem is the gap between when a routing decision is made and
when the action it implies is executed. The naive approach---pausing at a
graph edge using \code{interrupt\_before=["action\_node"]} at
\code{compile()} time---stops execution before the node runs and exposes
raw \code{ApprovalState} to the caller. That state contains the task
description but not the computed action payload, which the node would have
constructed from the task. The reviewer is asked to approve an abstract
intent rather than a concrete proposed action, forcing the approval
interface to reconstruct the payload independently---a duplication of
business logic between the graph and the interface. The governing principle
is that the node itself must compute the payload, present it to the
reviewer, and then pause; the pause lives inside the node body, not at
an edge.

\code{interrupt(value)} is a synchronous function callable from both sync
and async nodes. When the runtime encounters it for the first time, it
raises \code{GraphInterrupt}---a special exception caught by the Pregel
scheduler, never by user code. The scheduler serialises \code{value} into
the current checkpoint alongside the pre-node state, then returns control
to the caller: \code{graph.ainvoke()} returns the last committed state
dict, and \code{graph.astream()} with \code{stream\_mode="updates"} yields
a chunk whose \code{"\_\_interrupt\_\_"} key holds a tuple of \code{Interrupt}
objects. The node's return value is never written to state on this
path---the interrupt fires before the \code{return} statement executes, and
the Pregel runtime discards the partially-executed frame.

On the resume path, the runtime re-calls the node function from the top
with the same pre-interrupt state. The node re-executes all code before
\code{interrupt()}: the idempotency check passes (nothing is committed
yet), the payload is rebuilt from the same deterministic inputs, and the
call reaches \code{interrupt(value)} again. This time the scheduler
detects a stored resume value in the checkpoint and returns it immediately
without pausing. The return value of \code{interrupt()} on resume is
whatever was passed as \code{Command(resume=...)} by the caller---in the
following listing, the string \code{"approved"} or \code{"rejected"}.

\begin{warningbox}
Everything written to external systems \emph{before} \code{interrupt()}
in the node body may have already executed on the first pass; the
idempotency check from Section~\ref{sec:hitl-idempotent} must guard
every such write. Everything written to local variables \emph{after}
\code{interrupt()} will execute again on resume because the node
re-runs from the top. The correct structure is: build payload, check
idempotency, call \code{interrupt()}, then perform the external action
only after the resume value is returned.
\end{warningbox}

\begin{lstlisting}[language=Python,
  caption={Action node: idempotency guard, computed payload, inline
           interrupt, and post-resume branches},
  label={lst:hitl-action}]
# chapter21/action_node.py
from langgraph.types import interrupt
from langchain_core.runnables import RunnableConfig
from chapter21.state import ApprovalState, ActionRecord, make_action_id
from chapter21.policy import PolicyConfig

async def action_node(state: ApprovalState,
                      config: RunnableConfig) -> dict:
    payload = {"tool": "send_email",
               "to":   "team@example.com",
               "body": f"Report for: {state['task']}"}
    action_id = make_action_id("action_node", payload)

    for rec in state["pending_actions"]:
        if rec["action_id"] == action_id \
                and rec["status"] == "executed":
            return {}   # idempotency: already committed

    record: ActionRecord = {
        "action_id": action_id, "node_name": "action_node",
        "payload": payload,     "status":    "pending",
    }
    configurable = config.get("configurable") or {}
    policy: PolicyConfig = configurable.get("policy", PolicyConfig())

    if policy.requires_approval("action_node"):
        human_decision: str = interrupt(record)
    else:
        human_decision = "approved"

    if human_decision == "approved":
        done: ActionRecord = {**record, "status": "executed"}  # type: ignore[assignment]
        return {"pending_actions": [done],
                "approved_count":  state["approved_count"] + 1}
    rejected: ActionRecord = {**record, "status": "rejected"}  # type: ignore[assignment]
    return {"pending_actions": [rejected]}
\end{lstlisting}

Three decisions in Listing~\ref{lst:hitl-action} deserve explanation.
\code{record} is constructed as a plain dict literal annotated with
\code{ActionRecord} rather than a positional constructor call: \code{TypedDict}
types are plain dicts at runtime and do not support keyword-argument
construction like Pydantic models do; the literal form makes field names
visible at the construction site without requiring a post-hoc variable
assignment. The \code{type:~ignore[assignment]} comment on the status-
override lines is required because mypy cannot prove that a dict-spread
with a string override still satisfies the \code{TypedDict} contract at the
type-checking level, even though it is correct at runtime; a Pydantic model
would not require this comment, but the coordination-record character of
\code{ActionRecord} makes the \code{TypedDict} choice correct here
as discussed in Section~\ref{sec:hitl-idempotent}. The idempotency return
is an empty \code{dict} rather than an exception: an empty dict is a valid
node return that writes nothing to state, which is the correct outcome when
the action has already been committed; raising would trigger any registered
\code{RetryPolicy}, compounding the problem rather than resolving it.

The node is now pausable and idempotent. The remaining problem is the resume
contract: how the caller injects the human's decision back into the paused
thread and what distinguishes a correct resume call from one that silently
starts a new run from \code{START}.


%% ------------------------------------------------------------
\section{Resume Is a New Call, Not a Callback}
\label{sec:hitl-resume}
%% ------------------------------------------------------------

The design problem is the mismatch between the mental model of resuming a
paused coroutine and the actual mechanics of resuming a paused LangGraph
thread. A coroutine resumes at the \code{await} point that suspended it;
a LangGraph thread resumes by calling \code{graph.ainvoke()} a second time.
These look similar but differ in a critical way: passing a state dict as the
input to the second \code{ainvoke()} call silently discards the checkpoint
and starts a brand-new run from \code{START} on the same \code{thread\_id},
overwriting all prior state. There is no error and no warning. The governing
principle is that a resumed run's input must be \code{Command(resume=value)},
not a state dict; the config must carry the same \code{thread\_id} as the
original call; and the thread's pause state must be verified before
attempting to resume.

\code{Command(resume=value)} carries no \code{goto} field on a resume call.
The Pregel scheduler reads \code{StateSnapshot.next} from the checkpoint to
determine which node to re-enter; specifying \code{goto} on a resume
\code{Command} does not override this and its presence only misleads readers
into thinking routing is configurable at resume time when it is not. The
\code{value} passed to \code{resume} becomes the return value of
\code{interrupt()} inside the node; its type must match whatever the node
expects, because there is no schema validation between the caller and the
node at resume time---a type mismatch is a silent runtime bug discovered only
when the node's branch logic produces an unexpected result.

\begin{notebox}
\code{graph.get\_state(config).next} returns a tuple of node names that
will run on the next step. An empty tuple means the thread has either
completed or never been started. Attempting to resume a thread whose
\code{.next} is empty raises \code{GraphBubbleUp} at runtime; verifying
\code{.next} before calling \code{ainvoke(Command(resume=...))} converts a
confusing internal error into a caller-controlled precondition check with a
meaningful error message.
\end{notebox}

\begin{lstlisting}[language=Python,
  caption={Thread helpers: construct config, start a run, verify pause
           state, resume with decision},
  label={lst:hitl-runner}]
# chapter21/runner.py
from langgraph.types import Command
from langgraph.graph.state import CompiledStateGraph
from langchain_core.runnables import RunnableConfig
from chapter21.state import ApprovalState
from chapter21.policy import PolicyConfig

def _thread_config(thread_id: str,
                   policy: PolicyConfig) -> RunnableConfig:
    return {"configurable": {"thread_id": thread_id,
                             "policy":    policy}}

async def start_run(task: str, thread_id: str,
                    graph: CompiledStateGraph,
                    policy: PolicyConfig) -> dict:
    return await graph.ainvoke(
        ApprovalState(task=task, messages=[], pending_actions=[],
                      approved_count=0, final_output=""),
        config=_thread_config(thread_id, policy),
    )

async def resume_run(decision: str, thread_id: str,
                     graph: CompiledStateGraph,
                     policy: PolicyConfig) -> dict:
    config = _thread_config(thread_id, policy)
    if not graph.get_state(config).next:
        raise RuntimeError(
            f"Thread '{thread_id}' is not paused; cannot resume."
        )
    return await graph.ainvoke(
        Command(resume=decision), config=config,
    )
\end{lstlisting}

Two decisions in Listing~\ref{lst:hitl-runner} deserve explanation.
\code{\_thread\_config()} is a private helper rather than an inline dict
at each call site because the \code{policy} object must appear in
\code{configurable} on \emph{every} call---including the resume
call---and omitting it on the resume call means \code{action\_node} falls
back to a gate-free \code{PolicyConfig()}, silently executing without the
intended approval check on re-entry. \code{start\_run} returns the result
of \code{ainvoke} directly without testing whether the graph paused or
completed: the caller uses \code{graph.get\_state(config).next} to
distinguish the two outcomes, and encoding that distinction in
\code{start\_run}'s return type would require a sentinel value whose
semantics are less clear than the \code{StateSnapshot.next} field already
provided by the LangGraph API.

A decision that approves or rejects the action with the payload exactly as
computed is the common case. A human reviewer will often want to change the
payload before approving---correcting a recipient, adjusting a body,
overriding a parameter---without rerunning the node that computed it. The
next problem is how to modify graph state at the breakpoint without
re-executing any node.


%% ------------------------------------------------------------
\section{State at the Breakpoint Can Be Edited Before Resuming}
\label{sec:hitl-edit}
%% ------------------------------------------------------------

The design problem is the gap between what the node computed and what the
human reviewer actually wants to execute. The node constructed an
\code{ActionRecord} with a specific payload and stored it in
\code{pending\_actions} through the \code{\_merge\_actions} reducer. If
the reviewer wants to change the payload, there is no mechanism to do so
through the resume path alone: \code{Command(resume=value)} injects a
decision string, not a revised action record. The naive workaround---adding
an \code{edited\_payload} field to \code{ApprovalState} and having the
node read it on re-entry---pollutes the state schema with a coordination
artefact and requires the reviewer to write to a different field from the
one the execution step reads. The governing principle is that the reviewer
edits the state field that the execution step reads, using the same reducer
path that a node return would use.

\code{graph.aupdate\_state(config,~values,~as\_node=None)} writes \code{values}
into the checkpoint by applying each field's declared reducer, exactly as if
the named node had returned those values. For \code{pending\_actions},
passing \code{\{"pending\_actions":~[edited\_record]\}} where
\code{edited\_record} carries the same \code{action\_id} as the original
triggers the \code{\_merge\_actions} reducer's replacement path: the
existing record is replaced by the edited one. The call returns a new
\code{RunnableConfig} whose \code{configurable["checkpoint\_id"]} points to
the post-edit checkpoint. The subsequent resume call must use this returned
config---not the original---or it resumes from the pre-edit state silently
and the edit is lost without any error.

\begin{warningbox}
\code{aupdate\_state} returns a \code{RunnableConfig}. Discarding this
return value and calling \code{ainvoke(Command(resume=...),
config=original\_config)} silently resumes from the checkpoint that existed
before the edit. There is no API to retrieve the post-edit config after the
fact without scanning \code{get\_state\_history()}. Always assign the
return value and pass it to every subsequent call for this thread.
\end{warningbox}

The \code{as\_node} parameter names the node credited for the write in the
checkpoint's \code{metadata["writes"]} dict. Passing
\code{as\_node="\_\_human\_edit\_\_"} makes human modifications
distinguishable from node writes when iterating \code{get\_state\_history()}:
a snapshot whose \code{metadata["writes"]} contains the key
\code{"\_\_human\_edit\_\_"} was produced by a reviewer, not by automated
execution. This metadata distinction is the foundation of the time-travel
filtering in Section~\ref{sec:hitl-timetravel}.

\begin{lstlisting}[language=Python,
  caption={Edit state at the breakpoint then resume from the post-edit
           checkpoint},
  label={lst:hitl-edit}]
# chapter21/runner.py  (continued)
from chapter21.state import ActionRecord

async def edit_and_resume(thread_id: str,
                          edited_action: ActionRecord,
                          decision: str,
                          graph: CompiledStateGraph,
                          policy: PolicyConfig) -> dict:
    config = _thread_config(thread_id, policy)
    if not graph.get_state(config).next:
        raise RuntimeError(
            f"Thread '{thread_id}' is not paused; cannot edit."
        )
    # aupdate_state applies _merge_actions: same action_id = replace.
    new_config = await graph.aupdate_state(
        config,
        {"pending_actions": [edited_action]},
        as_node="__human_edit__",
    )
    # new_config carries the post-edit checkpoint_id.
    # Resuming with original config silently ignores the edit.
    return await graph.ainvoke(
        Command(resume=decision), config=new_config,
    )
\end{lstlisting}

Two decisions in Listing~\ref{lst:hitl-edit} deserve explanation.
\code{edit\_and\_resume} requires \code{edited\_action} to carry the same
\code{action\_id} as the record currently in \code{pending\_actions};
passing a record with a different \code{action\_id} appends a second record
via the \code{\_merge\_actions} reducer, leaving both the original and the
edited record in the field, after which the action node's idempotency check
finds the first matching record with \code{status~=~"executed"} on the next
run and silently ignores the edit entirely. The pause guard is repeated in
\code{edit\_and\_resume} rather than extracted to a shared helper because
the error message here uses the word ``edit'', not ``resume'': the
distinction matters when reading logs from a concurrent system where both
functions may be called on overlapping threads and the error context must
identify the call site without requiring a stack trace.

The reviewer can now pause, examine, and modify state at a breakpoint. The
remaining temporal problem is larger: what if the reviewer wants not to
modify the current breakpoint but to go back to a prior execution
point---before a previous action was taken, before a previous edit was
applied---and re-run forward from there?


%% ------------------------------------------------------------
\section{Every Checkpoint Is a Potential Starting Point}
\label{sec:hitl-timetravel}
%% ------------------------------------------------------------

The design problem is the irreversibility of a committed execution branch.
Once a node has run and the checkpoint has been written, the graph's
execution history is a linear sequence of state snapshots. If a reviewer
decides---after a completed run, or midway through a long run---that the
graph should have taken a different path at an earlier point, the only
mechanism available under a linear history model is to restart from scratch
with modified inputs. That is not always acceptable: a run that required
several expensive LLM calls and three human approvals should not need to be
replayed from the beginning to correct a decision made at step two. The
governing principle is that any checkpoint in the execution history is a
valid starting point for a new branch; replaying from a past checkpoint
produces a new execution lineage without modifying the original.

\code{graph.get\_state\_history(config)} returns an
\code{Iterator[StateSnapshot]} in reverse chronological order: the most
recent snapshot is yielded first. Each \code{StateSnapshot} carries a
\code{.config} field whose \code{configurable["checkpoint\_id"]}
identifies the exact checkpoint, a \code{.values} field containing the full
state at that point, a \code{.next} tuple naming the nodes that would run
next, and a \code{.metadata} dict. The \code{metadata["writes"]} key is a
dict mapping node-or-sentinel names to the values they returned; checking
whether \code{"\_\_human\_edit\_\_"} is in \code{metadata["writes"]}
locates checkpoints written by \code{aupdate\_state(...,
as\_node="\_\_human\_edit\_\_")}, making the human-edit history queryable
without additional instrumentation.

Replaying from a past snapshot is a single
\code{graph.ainvoke(None,~config=snapshot.config)} call. Passing \code{None}
as input means ``use the state already stored at this checkpoint''; the
graph advances forward from \code{snapshot.next}, creating a new checkpoint
sequence branching from \code{snapshot.config["configurable"]["checkpoint\_id"]}.
The original checkpoint chain is not modified. Both the original run and the
replay branch are visible in subsequent \code{get\_state\_history()} calls,
making the full decision history auditable.

\begin{lstlisting}[language=Python,
  caption={History iterator, pre-edit snapshot filter, and branch replay},
  label={lst:hitl-timetravel}]
# chapter21/time_travel.py
from langgraph.graph.state import CompiledStateGraph
from langgraph.checkpoint.base import StateSnapshot
from langchain_core.runnables import RunnableConfig

def find_last_pre_edit_snapshot(
    graph: CompiledStateGraph, thread_id: str,
) -> StateSnapshot | None:
    config: RunnableConfig = {
        "configurable": {"thread_id": thread_id}
    }
    # Yields newest-first; return first snapshot with no human-edit write.
    for snapshot in graph.get_state_history(config):
        writes = snapshot.metadata.get("writes") or {}
        if "__human_edit__" not in writes:
            return snapshot
    return None

async def replay_from_snapshot(
    graph:    CompiledStateGraph,
    snapshot: StateSnapshot,
) -> dict:
    # None input: use state already stored at this checkpoint.
    # snapshot.config already carries the correct thread_id.
    return await graph.ainvoke(None, config=snapshot.config)
\end{lstlisting}

Two decisions in Listing~\ref{lst:hitl-timetravel} deserve explanation.
\code{find\_last\_pre\_edit\_snapshot} iterates newest-first and returns
the \emph{first} snapshot without a human-edit write, which is the most
recent checkpoint that precedes any reviewer intervention; returning the
oldest pre-edit snapshot would force a replay from the very beginning of
the run, discarding all approved actions, whereas the most recent pre-edit
snapshot replays only the minimum necessary portion of the run.
\code{replay\_from\_snapshot} does not accept a \code{thread\_id}
argument because \code{snapshot.config} already embeds the correct
\code{thread\_id}: constructing a new config with a different
\code{thread\_id} would branch the replay onto a different thread, producing
two divergent histories under different identifiers instead of two branches
on the same thread where both are visible to a single
\code{get\_state\_history(config)} call.

The time-travel mechanics are now established. The remaining problem is
that the decision of whether a given node invokes the pause path is
currently hardcoded in \code{action\_node} as a \code{PolicyConfig} lookup.
That lookup will be identical in every action node across the system, and
the \code{PolicyConfig} object itself must be defined before it can be
passed into the graph.


%% ------------------------------------------------------------
\section{Approval Rules Belong in Configuration, Not in Code}
\label{sec:hitl-policy}
%% ------------------------------------------------------------

The design problem is hardcoded gate logic scattered across node
implementations. If every action node implements its own approval
condition---checking a feature flag, reading an environment variable,
testing a field in state---then adding a new action requires writing new
gate logic, changing the approval threshold for an existing action requires
finding and editing its node, and disabling all gates for automated tests
requires modifying code rather than changing a parameter. The governing
principle is that the set of nodes requiring human approval, and the
conditions under which approval is required, should be expressed in a
single declarative configuration object injected through
\code{RunnableConfig["configurable"]["policy"]}.

Chapter~14~\S14.3 introduced \code{PolicyConfig} as the book's runtime
configuration model for graph execution parameters. The \code{approval\_gates}
field added here follows the same injection pattern: the object is
constructed by the caller, placed in
\code{RunnableConfig["configurable"]["policy"]}, and read by each node
that may need to pause. \code{requires\_approval(node\_name)} encapsulates
the gate lookup so nodes never iterate \code{approval\_gates} directly; the
method returns \code{False} by default when no gate is registered for the
named node, making auto-approval the correct default for nodes whose
callers have not opted in to human review.

\begin{notebox}
\code{PolicyConfig} declares
\code{model\_config~=~ConfigDict(extra="allow")} so that Chapter~14's
existing fields coexist with the new \code{approval\_gates} field without
requiring callers to redeclare every prior field. \code{extra="allow"} means
Pydantic accepts any additional keyword arguments at construction time,
preserving backward compatibility as new fields are added across chapters.
\end{notebox}

The \code{ApprovalGate.timeout\_seconds} field specifies how many seconds
a thread may remain paused at an interrupt before the system should act
automatically. LangGraph's \code{interrupt()} primitive has no native
timeout; this field is a schema fact that a production deployment enforces
externally---a scheduled task scans for threads where
\code{snapshot.next~!=~()} and where the checkpoint's creation timestamp
exceeds the gate's \code{timeout\_seconds}, then calls
\code{resume\_run(decision="auto\_rejected",~...)} on those threads.
Chapter~22~\S22.4 implements that enforcement pattern.

\begin{lstlisting}[language=Python,
  caption={Approval gate model and policy lookup method},
  label={lst:hitl-policy}]
# chapter21/policy.py
from pydantic import BaseModel, Field, ConfigDict

class ApprovalGate(BaseModel):
    node_name:        str
    require_approval: bool       = True
    timeout_seconds:  int | None = 300

class PolicyConfig(BaseModel):
    """Extends Chapter 14 S14.3 PolicyConfig with approval gates."""
    model_config = ConfigDict(extra="allow")

    approval_gates:   list[ApprovalGate] = Field(default_factory=list)
    auto_approve_all: bool               = False

    def requires_approval(self, node_name: str) -> bool:
        if self.auto_approve_all:
            return False
        for gate in self.approval_gates:
            if gate.node_name == node_name:
                return gate.require_approval
        return False   # no gate registered: auto-approved
\end{lstlisting}

Two decisions in Listing~\ref{lst:hitl-policy} deserve explanation.
\code{auto\_approve\_all} defaults to \code{False} rather than \code{True}:
a permissive default would mean a system deployed without an explicit policy
configuration silently bypasses all human review, which is the production
failure mode the chapter opens by naming; an operator who wants no gates must
set \code{auto\_approve\_all=True} explicitly, making the choice auditable in
configuration rather than implicit in code. \code{requires\_approval} returns
\code{False} when no gate is registered for a node name, not \code{True}: the
correct posture is opt-in gates, not opt-out; requiring every new node to
declare an \code{ApprovalGate} would produce a deployment failure the first
time a new node is added without a corresponding gate, whereas an
auto-approved-by-default node is safe until explicitly gated and any
oversight is a deliberate product decision rather than a configuration
accident.

The approval policy is declarative and injectable. The remaining problem is
the interface: the graph runs asynchronously on a server, the reviewer is
in a browser, and the interrupt event must travel from the server to the
reviewer in real time---and the reviewer's decision must travel
back---without polling.


%% ------------------------------------------------------------
\section{The Interface Streams Interrupt Events and Receives Decisions}
\label{sec:hitl-api}
%% ------------------------------------------------------------

The design problem is the latency between when the graph pauses and when
the human reviewer learns that a decision is needed. A polling architecture
requires the frontend to query the server repeatedly; between polls the
reviewer is unaware the graph is waiting. The correct primitive for server-
to-client event push over HTTP is \emph{Server-Sent Events} (SSE): a long-
lived HTTP response whose body the server writes incrementally as events
occur. The client holds a single connection open, and the server writes an
interrupt event the moment \code{graph.astream()} yields the
\code{"\_\_interrupt\_\_"} chunk. The governing principle is that the
interrupt event must reach the reviewer at the moment the graph pauses,
not at the next poll interval.

\code{graph.astream(input,~config,~stream\_mode="updates")} yields \code{dict}
chunks as each node completes. An interrupt produces a chunk with key
\code{"\_\_interrupt\_\_"} whose value is a \emph{tuple} of \code{Interrupt}
named-tuples: \code{\{"\_\_interrupt\_\_":~(Interrupt(value=...,~resumable=True,~...),)\}}.
The payload is in \code{chunk["\_\_interrupt\_\_"][0].value}---the first element
of the tuple, which is the \code{ActionRecord} passed to \code{interrupt()}
in \code{action\_node}. Indexing with \code{[0]} is correct because
\code{interrupt()} in a single node produces exactly one \code{Interrupt}
per pause event; the tuple form exists to accommodate future multi-interrupt
nodes, not because a single-call node produces multiple elements.

\begin{warningbox}
\code{EventSourceResponse} from \code{sse-starlette} requires the
\code{"data"} field in each yielded dict to be a \emph{string}, not a
dict. Passing a raw dict raises a serialisation error at the
\code{EventSourceResponse} layer before any bytes reach the client. Every
\code{"data"} value must pass through \code{json.dumps()} before being
placed in the yield dict---including the \code{"done"} sentinel, which
must be \code{"\{\}"} (a string), not \code{\{\}} (a dict).
\end{warningbox}

The resume path is a separate \code{POST} endpoint that accepts a
\code{ResumeRequest} and calls either \code{resume\_run} or
\code{edit\_and\_resume} from \code{runner.py}. Before calling either,
the endpoint queries \code{graph.get\_state(config).next} to verify the
thread is actually paused; if \code{.next} is empty a resume attempt would
raise \code{GraphBubbleUp}---an internal runtime error that must never
reach the client as an unhandled 500. A 409 response is the correct HTTP
status for the precondition failure: the request is well-formed but the
server's current state conflicts with it.

\begin{lstlisting}[language=Python,
  caption={FastAPI app and SSE endpoint: interrupt event stream},
  label={lst:hitl-api-sse}]
# chapter21/api.py
import json
from fastapi import FastAPI, Depends, HTTPException
from sse_starlette.sse import EventSourceResponse
from pydantic import BaseModel
from langgraph.types import Command
# graph, state, policy, runner: see prior listings
from chapter21.graph import build_approval_graph
from chapter21.state import ApprovalState, ActionRecord
from chapter21.policy import PolicyConfig, ApprovalGate
from chapter21.runner import resume_run, edit_and_resume

app    = FastAPI()
_graph = build_approval_graph()   # process-scoped singleton

class StartRequest(BaseModel):
    task:      str
    thread_id: str
    policy:    PolicyConfig = PolicyConfig()

@app.post("/runs")
async def start_run_sse(
    req: StartRequest,
    graph = Depends(lambda: _graph),
) -> EventSourceResponse:
    async def _gen():
        cfg = {"configurable": {"thread_id": req.thread_id,
                                "policy":    req.policy}}
        async for chunk in graph.astream(
            ApprovalState(task=req.task, messages=[],
                          pending_actions=[], approved_count=0,
                          final_output=""),
            config=cfg, stream_mode="updates",
        ):
            if "__interrupt__" in chunk:
                val = chunk["__interrupt__"][0].value
                data = (val.model_dump()
                        if hasattr(val, "model_dump") else val)
                yield {"event": "interrupt",
                       "data": json.dumps(data)}
            else:
                yield {"event": "update",
                       "data": json.dumps(chunk, default=str)}
        yield {"event": "done", "data": "{}"}
    return EventSourceResponse(_gen())
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={Resume endpoint: pause-state guard and decision injection},
  label={lst:hitl-api-resume}]
# chapter21/api.py  (continued)
from langchain_core.runnables import RunnableConfig

class ResumeRequest(BaseModel):
    decision:       str                     # "approved" | "rejected"
    edited_action:  ActionRecord | None = None
    policy:         PolicyConfig        = PolicyConfig()

@app.post("/runs/{thread_id}/resume")
async def resume_endpoint(
    thread_id: str,
    req:       ResumeRequest,
    graph      = Depends(lambda: _graph),
) -> dict:
    cfg: RunnableConfig = {
        "configurable": {"thread_id": thread_id}
    }
    if not graph.get_state(cfg).next:
        raise HTTPException(
            status_code=409,
            detail=f"Thread '{thread_id}' is not paused.",
        )
    if req.edited_action is not None:
        return await edit_and_resume(
            thread_id=thread_id,
            edited_action=req.edited_action,
            decision=req.decision,
            graph=graph,
            policy=req.policy,
        )
    return await resume_run(
        decision=req.decision,
        thread_id=thread_id,
        graph=graph,
        policy=req.policy,
    )
\end{lstlisting}

Three decisions across Listings~\ref{lst:hitl-api-sse}
and~\ref{lst:hitl-api-resume} deserve explanation.
\code{\_graph} is a process-scoped singleton: the compiled graph object is
stateless between threads---all per-thread state lives in the
\code{MemorySaver} checkpointer---so a single instance safely handles all
concurrent requests without any locking. The \code{Depends(lambda:~\_graph)}
pattern rather than \code{Depends(get\_graph)} with a named function is
equivalent but avoids declaring a one-line free function solely to satisfy
FastAPI's \code{Depends} requirement; in a test, the lambda can be overridden
by overriding \code{app.dependency\_overrides}. The
\code{ResumeRequest.policy} field mirrors \code{StartRequest.policy}: the
resume path calls \code{\_thread\_config(thread\_id,~policy)} inside
\code{resume\_run} and \code{edit\_and\_resume}, which requires a
\code{PolicyConfig} in \code{configurable}; a frontend that omits
\code{policy} on resume would silently execute without the intended approval
gates on re-entry, reproducing the production failure mode named in the
chapter opener.

All five layers of the solution are independently implemented and testable.
The remaining problem is verification: confirming that the three HITL
workflows---approval, state edit, and time-travel replay---each produce the
correct state transitions under a fresh in-process checkpointer, before
any persistent storage or network transport is introduced.


%% ------------------------------------------------------------
\section{Full System: Approval, Edit, and Replay as Runnable Tests}
\label{sec:hitl-full}
%% ------------------------------------------------------------

The design problem is the absence of an executable specification for the
three HITL workflows. The five sections above establish the mechanics
individually, but no single runnable artefact confirms that they compose
correctly---that the idempotency check survives resume, that
\code{\_merge\_actions} replaces rather than appends after an edit, and that
\code{replay\_from\_snapshot} produces a new branch on the same thread. The
governing principle is that each workflow is a separate \code{pytest-asyncio}
test with its own fresh \code{MemorySaver} instance, asserting the highest-
level contract first and the per-field details second.

The graph assembled for this chapter is minimal: \code{action\_node} followed
by \code{synthesise\_node}, a single static edge between them, and a
\code{MemorySaver} checkpointer. There are no conditional edges and no retry
policies---those are Chapter~19 and Chapter~20 concerns. The \code{MemorySaver}
limitation is intentional: it is in-process and lost on restart, which makes
the retry-across-restart invariant provably absent. Verifying durability
across process restarts requires the \code{AsyncSqliteSaver} migration
introduced in Chapter~22~\S22.1, recorded here as a skipped test rather than
a silent gap.

\begin{lstlisting}[language=Python,
  caption={Graph assembly: MemorySaver checkpointer, two nodes, no
           conditional edges},
  label={lst:hitl-graph}]
# chapter21/graph.py
from langgraph.graph import StateGraph, START, END
from langgraph.graph.state import CompiledStateGraph
from langgraph.checkpoint.memory import MemorySaver
from langchain_core.runnables import RunnableConfig
from chapter21.state import ApprovalState
from chapter21.action_node import action_node

async def synthesise_node(state: ApprovalState,
                          config: RunnableConfig) -> dict:
    return {"final_output": f"Completed: {state['task']}"}

def build_approval_graph() -> CompiledStateGraph:
    b = StateGraph(ApprovalState)
    b.add_node("action_node", action_node)
    b.add_node("synthesise",  synthesise_node)
    b.add_edge(START,         "action_node")
    b.add_edge("action_node", "synthesise")
    b.add_edge("synthesise",  END)
    return b.compile(checkpointer=MemorySaver())
\end{lstlisting}

Two decisions in Listing~\ref{lst:hitl-graph} deserve explanation.
\code{build\_approval\_graph()} constructs a new \code{MemorySaver()} on
every call so that each test receives a completely isolated checkpointer;
sharing a single \code{MemorySaver} instance across tests would cause the
second test's thread to find prior run state in the checkpointer and silently
branch from an unexpected snapshot, producing false-positive passes when
the thread happens to be in the correct state by coincidence. \code{synthesise\_node}
returns a deterministic string rather than making an LLM call: the tests
assert that \code{final\_output} is non-empty after a completed run, and a
predictable string is a verifiable assertion target without requiring a live
model or a mock; replacing this stub with a real LLM call is a one-line
change that does not affect any other contract in the test suite.

\begin{lstlisting}[language=Python,
  caption={Scenario 1: approval gate --- pause, verify, approve, assert
           execution invariants},
  label={lst:hitl-test-approve}]
# chapter21/test_hitl.py
import pytest
from langgraph.types import Command
from chapter21.graph import build_approval_graph
from chapter21.state import ApprovalState, ActionRecord
from chapter21.policy import PolicyConfig, ApprovalGate
from chapter21.runner import resume_run
from langchain_core.runnables import RunnableConfig

_POLICY = PolicyConfig(approval_gates=[
    ApprovalGate(node_name="action_node", require_approval=True)
])

def _cfg(thread_id: str) -> RunnableConfig:
    return {"configurable": {"thread_id": thread_id,
                             "policy":    _POLICY}}

@pytest.mark.asyncio
async def test_approval_gate() -> None:
    graph = build_approval_graph()
    cfg   = _cfg("t-approve-001")
    await graph.ainvoke(
        ApprovalState(task="Send weekly report", messages=[],
                      pending_actions=[], approved_count=0,
                      final_output=""),
        config=cfg,
    )
    snapshot = graph.get_state(cfg)
    assert snapshot.next, "Graph must pause at interrupt"
    result = await graph.ainvoke(Command(resume="approved"), config=cfg)
    assert result["final_output"] != "", \
        "synthesise_node must write final_output after approval"
    executed = [r for r in result["pending_actions"]
                if r["status"] == "executed"]
    assert len(executed) == 1, \
        "Exactly one ActionRecord must reach executed status"
    assert result["approved_count"] == 1
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={Scenario 2: edit at breakpoint; Scenario 3: time-travel
           replay},
  label={lst:hitl-test-edit-replay}]
# chapter21/test_hitl.py  (continued)
from chapter21.runner import edit_and_resume
from chapter21.time_travel import replay_from_snapshot

@pytest.mark.asyncio
async def test_state_edit_at_breakpoint() -> None:
    graph = build_approval_graph()
    cfg   = _cfg("t-edit-001")
    await graph.ainvoke(
        ApprovalState(task="Send quarterly report", messages=[],
                      pending_actions=[], approved_count=0,
                      final_output=""),
        config=cfg,
    )
    assert graph.get_state(cfg).next, "Must be paused before edit"
    orig: ActionRecord = graph.get_state(cfg).values["pending_actions"][0]
    edited: ActionRecord = {**orig,  # type: ignore[assignment]
                            "payload": {**orig["payload"],
                                        "body": "EDITED BODY"}}
    result = await edit_and_resume(
        thread_id="t-edit-001", edited_action=edited,
        decision="approved", graph=graph, policy=_POLICY,
    )
    done = [r for r in result["pending_actions"]
            if r["status"] == "executed"]
    assert done[0]["payload"]["body"] == "EDITED BODY"

@pytest.mark.asyncio
async def test_time_travel_replay() -> None:
    graph   = build_approval_graph()
    cfg     = _cfg("t-replay-001")
    await graph.ainvoke(
        ApprovalState(task="Archive logs", messages=[],
                      pending_actions=[], approved_count=0,
                      final_output=""),
        config=cfg,
    )
    history = list(graph.get_state_history(cfg))
    assert len(history) >= 2  # current + at least one prior
    pre_interrupt = history[1]   # second-newest: before interrupt
    await replay_from_snapshot(graph, pre_interrupt)
    new_history = list(graph.get_state_history(cfg))
    assert len(new_history) > len(history), \
        "Replay must produce at least one new checkpoint on the thread"

@pytest.mark.asyncio
async def test_persistent_checkpointer_required() -> None:
    pytest.skip(
        "Requires AsyncSqliteSaver: Chapter 22 \\S22.1 migration"
    )
\end{lstlisting}

Three decisions across Listings~\ref{lst:hitl-test-approve}
and~\ref{lst:hitl-test-edit-replay} deserve explanation.
The module-level \code{\_POLICY} and \code{\_cfg} fixtures are shared
across all three tests because a divergence in gate configuration between
tests would make a failing test ambiguous: one cannot tell whether the
assertion failed because the HITL mechanic broke or because the gate was
not active. \code{test\_state\_edit\_at\_breakpoint} asserts
\code{done[0]["payload"]["body"]~==~"EDITED~BODY"} rather than asserting
\code{action\_id}
