%% ============================================================
%%  Chapter 4 — Structured Output, Validation & Self-Healing
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%% ============================================================

\chapterquote{If the contract between components is vague, the
system will fail in vague ways.}{Mary Poppendieck, adapted}

Chapter~1 built a four-node loop whose Reason node calls an LLM and
wraps the result in a discriminated union. Chapter~2 added termination
guards that fire when the loop runs too long or costs too much.
Chapter~3 showed how to persist and reload that loop across process
restarts. All three chapters assumed that the model returns parseable,
valid output---that the typed union the Reason node writes into state
is always a genuine \code{ToolRequest}, \code{FinalAnswer}, or
\code{ReasoningTrace}. That assumption is fragile.

Language models emit text, not typed objects. Even with function calling
or JSON mode enabled, a model can return extra commentary, omit a
required field, produce a number outside a declared range, or---in
failure cases---emit a completely unparseable string. None of these
failures are loud. They do not raise exceptions from the model API.
They surface as downstream breakage when the Action or Reflect node
tries to interpret a value that was never validated.

This chapter builds the last load-bearing piece of the runtime:
a contract layer between raw model output and the rest of the graph.
You will understand why that contract belongs in a dedicated graph node
rather than inside the Reason node itself, how to choose the right
Instructor strategy per model family, and how the repair loop uses the
graph's own belief mechanism to self-correct. Where repair fails, a
structured escalation path surfaces the failure to Langfuse with enough
context for offline diagnosis.

%% ------------------------------------------------------------
\section{Why Unstructured Output Is a Production Problem}
\label{sec:output-problem}
%% ------------------------------------------------------------

Consider what happens when the Reason node returns a string that cannot
be parsed into a \code{ReasonOutput}. In Chapter~1, Instructor handles
this inline: it retries the model call up to \code{max\_retries} times
and raises if all attempts fail. That raised exception propagates out of
the node function, crashes the graph execution, and leaves no record in
state---only an uncaught Python exception.

That is the best case. The worst case is more common: the model returns
something that partially matches the expected shape, coercion fills in
the gaps, and a semantically invalid object flows downstream. A
\code{confidence} field set to \code{2.5} passes without complaint unless
the field validator rejects it. A \code{tool\_name} set to an empty string
reaches the Action node, which raises an \code{UnregisteredToolError}
deep in the cycle---far from the model call that caused it.

The root cause is that parsing and business-rule validation are mixed
inside the Reason node without a recovery path. The fix requires
separating three concerns that are currently entangled: (1)~calling the
model and capturing raw output, (2)~validating that raw output against a
strict schema, and (3)~deciding what to do when validation fails.
Separating them produces a system where validation failures are
observable, recoverable, and bounded.

%% ------------------------------------------------------------
\section{Pydantic Output Models as Contracts}
\label{sec:output-models}
%% ------------------------------------------------------------

A contract, in this context, is an explicit, machine-checkable statement
of what a node may produce. The discriminated union from Chapter~1
already expresses three legal shapes. This chapter makes that union
strict: it rejects any value that satisfies the structural shape but
violates a business rule.

Pydantic~v2 strict mode is the first change. Without it, Pydantic
coerces compatible types: the string \code{"1"} becomes the integer
\code{1}, an integer \code{1} becomes the boolean \code{True}. Silent
coercion is useful in form-handling code. In the context of model output
it is dangerous---it hides the fact that the model produced a type it
was not supposed to produce, making failure patterns invisible in
Langfuse data.

The second change is field-level validators on values with natural
business bounds. \code{confidence} is bounded between \code{0.0} and
\code{1.0} because it is a probability. Without the validator, a model
that has learned to output \code{confidence: 1.5} when uncertain
produces silently miscalibrated escalation behaviour---the Reflect node
sees a high confidence and does not escalate, even though the value is
out of range.

\begin{lstlisting}[language=Python,
  caption={Strict Pydantic output contracts: one model per Reason output variant},
  label={lst:output-models}]
# src/agent/output_models.py

class ToolRequest(BaseModel):
    model_config = {"strict": True}
    type:      Literal["tool_request"] = "tool_request"
    tool_name: str
    arguments: dict[str, Any]

class FinalAnswer(BaseModel):
    model_config = {"strict": True}
    type:       Literal["final_answer"] = "final_answer"
    content:    str
    confidence: float

    @field_validator("confidence")
    @classmethod
    def confidence_in_range(cls, v: float) -> float:
        if not 0.0 <= v <= 1.0:
            raise ValueError(
                f"confidence must be 0.0--1.0, got {v}")
        return v

class ReasoningTrace(BaseModel):
    model_config = {"strict": True}
    type:  Literal["reasoning_trace"] = "reasoning_trace"
    trace: str

ReasonOutput = Annotated[
    ToolRequest | FinalAnswer | ReasoningTrace,
    Field(discriminator="type"),
]
\end{lstlisting}

One model per output variant---not one model with optional fields for
all variants. Optional fields create mixed states that are syntactically
valid and semantically incoherent: a record that has both a
\code{tool\_name} and a \code{content} field, or neither. A discriminated
union keyed by a \code{type} field rules those states out at the schema
level, before any field validator runs.

\begin{warningbox}
Do not add a catch-all \code{"other"} variant to handle unexpected
shapes. If the model emits a shape you did not anticipate, that is a
validation failure and it should trigger the repair loop. Routing it
downstream as an untyped blob hides a model quality problem and
produces bugs that are very hard to trace in production.
\end{warningbox}

%% ------------------------------------------------------------
\section{Choosing an Instructor Mode}
\label{sec:instructor-modes}
%% ------------------------------------------------------------

Instructor is a library that wraps any OpenAI-compatible client so that
\code{.create()} accepts a \code{response\_model=} argument and returns a
validated Pydantic object. The key question is: how does it make the
model produce structured output? The answer depends on what the model
provider supports, and choosing the wrong mode produces either silent
parse failures or unnecessary prompt overhead.

Instructor supports three strategies. Understanding what each one does
at the API level is necessary for making the right choice.

\textbf{TOOLS mode} uses the provider's native function-calling API. The
Instructor client serialises the Pydantic schema into a JSON Schema tool
definition and passes it in the \code{tools} parameter. The provider's
model inference layer uses the schema to constrain token generation at
the grammar level---the model literally cannot generate tokens that would
violate the JSON structure. The result comes back in a
\code{tool\_calls} message, and Instructor deserialises it into your
Pydantic model. This is the most reliable mode because the constraint is
applied by the inference engine, not by post-processing.

\textbf{JSON mode} uses the provider's native JSON output mode. The
Instructor client sets \code{response\_format=\{"type": "json\_object"\}}
and injects the schema into the system prompt. The model is instructed to
produce valid JSON, but the provider's grammar constraint only enforces
``valid JSON''---it does not enforce your specific schema. The result is
a JSON string that Instructor parses and validates. This mode is useful
for providers that support JSON mode but do not support function calling.

\textbf{MD\_JSON mode} is a prompt-only fallback. Instructor injects
the schema into the system prompt and asks the model to wrap its JSON
output in a markdown code fence (\code{```json ... ```}). No API-level
constraint is applied. The model produces whatever it wants, and
Instructor strips the code fence and attempts to parse the content.
This mode fails most often and should be treated as temporary until
you can switch to a model that supports TOOLS or JSON mode.

\begin{lstlisting}[language=Python,
  caption={Instructor mode selection: enum, mapping, and client factory},
  label={lst:instructor-config}]
# src/agent/instructor_config.py

class InstructorMode(str, Enum):
    TOOLS   = "tools"
    JSON    = "json"
    MD_JSON = "md_json"

_MODE_MAP: dict[InstructorMode, instructor.Mode] = {
    InstructorMode.TOOLS:   instructor.Mode.TOOLS,
    InstructorMode.JSON:    instructor.Mode.JSON,
    InstructorMode.MD_JSON: instructor.Mode.MD_JSON,
}

MODEL_TO_INSTRUCTOR_MODE: dict[str, InstructorMode] = {
    "gpt-4o":                InstructorMode.TOOLS,
    "gpt-4o-mini":           InstructorMode.TOOLS,
    "claude-3-5-sonnet":     InstructorMode.TOOLS,
    "mistral/mistral-large": InstructorMode.JSON,
    "ollama/llama3":         InstructorMode.MD_JSON,
}

def make_instructor_client(
    model: str,
    mode: InstructorMode | None = None,
) -> instructor.Instructor:
    base_client = litellm.Client()
    resolved = mode or MODEL_TO_INSTRUCTOR_MODE.get(
        model, InstructorMode.JSON
    )
    return instructor.from_litellm(
        client=base_client, mode=_MODE_MAP[resolved]
    )
\end{lstlisting}

The decision rule in prose: use TOOLS if the provider supports function
calling (OpenAI, Anthropic, most major providers). Use JSON if it supports
JSON mode but not function calling. Use MD\_JSON only as a last resort and
log every parse failure as a model quality signal.

Chapter~9 returns to this mapping when it wires LiteLLM's cost callbacks.
The mode affects token consumption: TOOLS mode adds schema tokens to the
request; MD\_JSON adds schema tokens to the system prompt. Both are
accounted for, but through different fields in the usage response.

%% ------------------------------------------------------------
\section{Validation as a Graph Node}
\label{sec:validation-node}
%% ------------------------------------------------------------

The central design decision in this chapter is to place validation in its
own node rather than inside the Reason node. This is not a refactoring
preference. It follows from three constraints that the graph runtime
imposes.

The first constraint is routing. LangGraph's conditional edges connect
nodes to nodes. If validation lives inside the Reason node, a validation
failure can only be handled by the Reason node itself---by raising an
exception or returning a special sentinel value. Neither option is clean.
Raising propagates a Python exception through the graph executor.
Returning a sentinel requires the Action node to understand and handle
validation state, which is not its responsibility. A dedicated
Validation node can be connected to a repair branch, an escalation
branch, and a pass-through branch using the same conditional edge
mechanism you already use everywhere else.

The second constraint is observability. Chapter~1 established that
\code{@observe} decorates node functions, not the graph runner,
specifically so that each node produces its own Langfuse span with
its own duration, metadata, and token usage. If validation is inline
inside the Reason node, every validation failure is buried inside the
Reason span. You cannot distinguish ``Reason was slow'' from ``Reason
was fast but Validation failed three times'' without parsing span
metadata. A dedicated Validation node produces a distinct span for
every validation attempt, with attempt number and error type as span
attributes.

The third constraint is reuse. Once the Validation node exists as a
standalone pure function, it can be inserted after any node that
produces model output---Reason, Reflect, or a summarization node added
in a later chapter. If validation is inlined, each node has its own
copy of the validation logic and its own copy of the retry bookkeeping.

Three private state fields carry the validation context across node
boundaries: \code{\_raw\_model\_output} holds the last raw JSON string the
model produced; \code{\_validation\_error} holds the last error string if
validation failed; and \code{\_validation\_attempt} counts consecutive
failures for the current cycle. These fields are prefixed with an
underscore by convention---they are internal orchestration signals, not
business state.

\begin{lstlisting}[language=Python,
  caption={Validation node: parse, record failure, or clear on success},
  label={lst:validation-node}]
# src/agent/nodes/validation.py

@observe(name="validation_node")
def validation_node(state: AgentState) -> dict:
    raw = state.get("_raw_model_output")
    if raw is None:
        return {}   # wiring error upstream; nothing to validate

    attempt = int(state.get("_validation_attempt", 0))
    try:
        parsed = ReasonOutput.model_validate_json(raw)
    except ValidationError as exc:
        error_str = str(exc)
        _emit_validation_failure_event(
            error=error_str,
            node_name="validation_node",
            attempt_number=attempt,
            model=state.get("_config_model", "unknown"),
            output_type="unknown",
            raw_output=raw,
        )
        return {
            "_validation_error":   error_str,
            "_validation_attempt": attempt + 1,
        }

    return {          # success: write parsed object, clear error fields
        "last_action":         parsed.model_dump(),
        "_validation_error":   None,
        "_validation_attempt": 0,
    }
\end{lstlisting}

On success the node writes the fully parsed union into \code{last\_action}
and resets the attempt counter to zero. On failure it increments the
counter and records the error string, but does not touch
\code{last\_action}. The routing function reads those two fields to decide
what comes next.

%% ------------------------------------------------------------
\section{A Bounded Repair Loop}
\label{sec:repair-loop}
%% ------------------------------------------------------------

A single validation failure is not a reason to give up. Models
occasionally hallucinate fields, emit a number slightly out of range,
or add commentary before the JSON object. A single retry, with the error
in context, resolves most of these cases.

The mechanism is the belief system introduced in Chapter~1. \code{beliefs}
is an append-only list in \code{AgentState}. Every node can append
observations to it; the Reason node reads the recent belief window when
assembling its prompt. This means injecting a validation error as a
belief is enough to inform the next Reason call of exactly what went
wrong---without any special-case prompt engineering, without a new
prompt template, and without touching the Reason node.

When the Validation node fails and the routing function decides to
retry, a small helper node appends a repair belief describing the error:

\begin{lstlisting}[language=Python,
  caption={Belief injection: the repair loop uses the existing belief mechanism},
  label={lst:repair-belief-injection}]
# src/agent/nodes/validation.py

def _build_repair_belief(error: str, attempt: int) -> str:
    return (
        f"[validation-retry attempt={attempt}] "
        f"ValidationError: {error}"
    )

def _inject_validation_belief(state: AgentState) -> dict:
    error   = state.get("_validation_error")
    attempt = int(state.get("_validation_attempt", 0))
    if not error:
        return {}
    return {
        "beliefs": [_build_repair_belief(error, attempt)]
    }
\end{lstlisting}

On the next Reason call, the model receives its standard prompt plus
the injected belief. It now knows that its previous output failed
validation and why. Most models correct the specific error they are
shown. The mechanism requires no changes to the Reason node---the
belief window already includes recent observations, and a validation
error is just another observation.

The number of retries is bounded by \code{validation\_max\_retries} in
the \code{TerminationPolicy} introduced in Chapter~2. Adding this field
to the existing policy object---rather than defining a new constant---
ensures that the retry ceiling survives checkpoint serialisation and
reload, is visible in Langfuse as a policy attribute, and can be
configured differently per task type without touching any node code.

\begin{lstlisting}[language=Python,
  caption={Adding a validation retry ceiling to TerminationPolicy},
  label={lst:termination-policy-validation}]
# src/agent/termination.py

class TerminationPolicy(BaseModel):
    max_cycles:             int
    token_ceiling:          int
    wall_clock_timeout_s:   int
    cost_cap_usd:           float
    validation_max_retries: int = 2

    @field_validator("validation_max_retries")
    @classmethod
    def retries_must_be_positive(cls, v: int) -> int:
        if v < 1:
            raise ValueError(
                f"validation_max_retries must be >= 1, got {v}")
        return v
\end{lstlisting}

The routing function after the Validation node reads the policy from
state and maps the current attempt count to one of three outcomes:
pass, retry, or escalate.

\begin{lstlisting}[language=Python,
  caption={Routing after validation: pass, retry, or escalate},
  label={lst:repair-loop}]
# src/agent/nodes/validation.py

def route_after_validation(state: AgentState) -> str:
    policy = TerminationPolicy.model_validate(
        state.get("_policy", {}))
    error   = state.get("_validation_error")
    attempt = int(state.get("_validation_attempt", 0))

    if error is None:
        return "pass"
    if attempt < policy.validation_max_retries:
        return "retry"
    return "escalate"
\end{lstlisting}

Each retry costs one additional Reason call. Chapter~2's token budget
tracks every call, so retries are automatically deducted from the
remaining budget. No special-case accounting is needed.

%% ------------------------------------------------------------
\section{Escalation on Ceiling Breach}
\label{sec:validation-escalation}
%% ------------------------------------------------------------

When \code{route\_after\_validation} returns \code{"escalate"}, the
repair loop has exhausted every attempt it was given. At this point a
fundamental assumption of the current run has been violated: the model
is not producing output that the runtime can interpret. Continuing is
not just ineffective---it risks consuming the full token budget on
retries that will also fail.

The escalation path reuses Chapter~1's existing \code{ESCALATE} decision.
This is not an accident of convenience. Chapter~1 established that
\code{reflect\_decision} is the single field the routing function reads
to decide where the graph goes next, and \code{ESCALATE} means ``route to
END and surface this run for human review''. Writing
\code{reflect\_decision = "ESCALATE"} from the Validation node means the
escalation path is consistent with every other termination path in the
graph---the same routing function, the same Langfuse trace structure,
and the same \code{termination\_status = "FAILURE"} in the
\code{CycleTrace}.

\begin{lstlisting}[language=Python,
  caption={Building the escalation state update on ceiling breach},
  label={lst:validation-escalation}]
# src/agent/nodes/validation.py

@dataclass
class ValidationEscalationRecord:
    node_name:        str
    attempt_count:    int
    last_error:       str
    raw_model_output: str | None
    model:            str
    timestamp_utc:    str = field(
        default_factory=lambda:
            datetime.now(timezone.utc).isoformat()
    )

def _build_escalation_update(
    state: AgentState,
    error: str,
    attempt: int,
    model: str,
) -> dict:
    record = ValidationEscalationRecord(
        node_name="validation_node",
        attempt_count=attempt,
        last_error=error,
        raw_model_output=state.get("_raw_model_output"),
        model=model,
    )
    summary = (
        f"Validation failed after {attempt} retries. "
        f"Last error: {error[:200]}"
    )
    return {
        "reflect_decision":    "ESCALATE",
        "final_answer":        summary,
        "_validation_error":   error,
        "_validation_attempt": attempt,
        "beliefs":    [f"[validation-escalated] {summary}"],
        "cycle_traces": [{"validation_escalation": asdict(record)}],
    }
\end{lstlisting}

After this update is written to state, Chapter~1's
\code{route\_from\_reflect} sees \code{reflect\_decision == "ESCALATE"} and
routes to \code{END}. The \code{ValidationEscalationRecord} in
\code{cycle\_traces} distinguishes this escalation from a voluntary
\code{ESCALATE} produced by the Reflect node---the
\code{termination\_status} field in the outer \code{CycleTrace} records
\code{"FAILURE"}, not \code{"ESCALATED"}.

%% ------------------------------------------------------------
\section{Langfuse Events for Validation Failures}
\label{sec:validation-langfuse}
%% ------------------------------------------------------------

Langfuse spans capture timing and sequence: you can see that the
Validation node ran, how long it took, and how many spans it produced
in a given trace. What spans do not capture well is the content of
individual failures---the specific error message, the model that
produced the bad output, and the raw output length. To aggregate those
dimensions across a fleet of agents you need a structured event with a
stable schema.

Spans and events serve different purposes. A span is a duration with
optional metadata, designed for latency analysis and trace trees. An
event is a point-in-time record with a fixed set of fields, designed for
aggregation and alerting. Every validation failure should emit one event.

\begin{lstlisting}[language=Python,
  caption={Emitting a structured Langfuse event on validation failure},
  label={lst:langfuse-event}]
# src/agent/nodes/validation.py

def _emit_validation_failure_event(
    error: str,
    node_name: str,
    attempt_number: int,
    model: str,
    output_type: str,
    raw_output: str | None,
) -> None:
    langfuse_context.update_current_observation(
        metadata={
            "event":            "validation_failure",
            "validation_error": error,
            "node_name":        node_name,
            "attempt_number":   attempt_number,
            "model":            model,
            "output_type":      output_type,
            "raw_output_len":   len(raw_output or ""),
        }
    )
\end{lstlisting}

The six fields in this event schema are the minimum needed to answer the
questions that matter in production: \emph{which model} caused the
failure, \emph{which output type} it was trying to produce, \emph{how
many retries} had already been attempted, and \emph{how long} the raw
output was (a proxy for the model having generated something completely
off-schema). From these fields you can build a dashboard that shows
validation failure rates broken down by model and output type, and alert
when the rate for a specific model exceeds a threshold.

\begin{notebox}
Set \code{LANGFUSE\_ENABLED=false} in your test environment so that the
\code{langfuse\_context} call becomes a no-op. The helper function has no
other side effects and tests for the Validation node do not need to mock
the Langfuse SDK.
\end{notebox}

%% ------------------------------------------------------------
\section{Graph Wiring and Tests}
\label{sec:validation-graph}
%% ------------------------------------------------------------

Inserting the Validation node requires a small change to the graph
wiring from Chapter~1. The Reason node no longer writes
\code{last\_action} directly; it writes \code{\_raw\_model\_output}
instead. The Validation node consumes that field, and the conditional
edge from Validation---not from Reason---decides whether execution
proceeds to Action, loops back to Reason, or exits to END.

\begin{lstlisting}[language=Python,
  caption={Updated StateGraph wiring with the Validation node},
  label={lst:graph-ch4}]
# src/agent/graph.py (excerpt)

b.add_node("validation_node",          validation_node)
b.add_node("inject_validation_belief", _inject_validation_belief)

b.add_edge("perceive_node",  "reason_node")
b.add_edge("reason_node",    "validation_node")

b.add_conditional_edges(
    "validation_node",
    route_after_validation,
    {
        "pass":    "action_node",
        "retry":   "inject_validation_belief",
        "escalate": END,
    },
)
b.add_edge("inject_validation_belief", "reason_node")
\end{lstlisting}

The \code{inject\_validation\_belief} node is the bridge between the
``retry'' branch and the next Reason call. It exists because appending
to \code{beliefs} is a state update, and state updates happen through
node return values. The node itself is the small helper shown in
Listing~\ref{lst:repair-belief-injection}; it reads the current error
and attempt count from state and returns a one-item belief list.

The unit tests verify three invariants: valid output clears all error
fields; invalid output increments the attempt counter; and exceeding the
retry ceiling routes to escalation rather than another retry.

\begin{lstlisting}[language=Python,
  caption={Node-level tests: valid pass, retry increment, ceiling breach},
  label={lst:validation-tests}]
# tests/test_validation.py

os.environ.setdefault("LANGFUSE_ENABLED", "false")

def test_valid_output_passes() -> None:
    state = make_test_state(
        _raw_model_output=FinalAnswer(
            content="done", confidence=0.9
        ).model_dump_json()
    )
    result = validation_node(state)
    assert result["last_action"]["type"] == "final_answer"
    assert result["_validation_error"] is None
    assert result["_validation_attempt"] == 0

def test_invalid_output_increments_attempt() -> None:
    raw = (
        '{"type":"final_answer","content":"done",'
        '"confidence": 2.5}'
    )
    state = make_test_state(
        _raw_model_output=raw, _validation_attempt=0)
    result = validation_node(state)
    assert result["_validation_attempt"] == 1
    assert "confidence" in result["_validation_error"]

def test_ceiling_breach_routes_to_escalate() -> None:
    raw = (
        '{"type":"final_answer","content":"done",'
        '"confidence": 2.5}'
    )
    policy = {"max_cycles": 10, "token_ceiling": 10000,
              "wall_clock_timeout_s": 60, "cost_cap_usd": 1.0,
              "validation_max_retries": 3}
    state = make_test_state(
        _raw_model_output=raw,
        _validation_attempt=3,
        _policy=policy,
    )
    node_result = validation_node(state)
    merged = {**state, **node_result}
    assert route_after_validation(merged) == "escalate"
\end{lstlisting}

Chapter~28 extends this suite to graph-level integration tests that
compile the full five-node graph---Perceive, Reason, Validation, Action,
Reflect---and assert on multi-cycle state evolution when the model
returns invalid output on the first attempt and valid output on the
second.

\begin{notebox}
The Validation node and the belief-injection helper are pure
\code{(state) -> dict} functions. They have no external dependencies
other than the Langfuse SDK (disabled in tests). Every assertion in the
suite above tests a state transformation, not a side effect. This
matches the testing pattern established in Chapter~1 and maintained
throughout the book: test the contract, not the implementation.
\end{notebox}
