%% ============================================================
%%  Chapter 19 --- Supervisor Pattern
%%  Production-Grade Agentic AI: LangGraph + Python Frameworks
%%  Dependencies: langgraph >= 0.2.0,
%%                langfuse >= 3.0.0,
%%                langchain-litellm >= 0.3.0,
%%                pydantic >= 2.7.0,
%%                langchain-core >= 0.3.0
%% ============================================================

\chapterquote{Routing is not an implementation detail. It is the architecture.}{anon}

The production failure this chapter prevents is coordination collapse at
scale. A flat LangGraph graph that distributes work across specialised
subgraphs via \code{add\_conditional\_edges} fails in a predictable way
as the system grows: every new worker requires edge declarations from every
other node that might route to it, retry paths need their own conditional
functions, and there is no single place in the codebase that can read
current work status and decide which worker should run next. The graph's
edge table becomes the coordination logic, and it cannot be updated without
touching every node that participates in routing. At three workers this is
inconvenient; at six it is a structural defect whose cost is paid in every
subsequent feature addition.

The solution has five layers. The first is a \emph{routing layer}: a
single supervisor node calls an LLM with a summary of the current work
state and returns a \code{Command} that names the next worker and
optionally updates state, atomically, before routing occurs. The second is
a \emph{composition layer}: each worker is a compiled subgraph registered
in the parent graph through an adapter node that translates state across
the boundary; the supervisor sees a flat node list regardless of how
complex each worker is internally. The third is an \emph{isolation layer}:
per-worker output, errors, and retry budget live in a namespaced
\code{subagent\_state} field with an \code{operator.or\_} reducer that
merges per-worker keys independently; no worker can overwrite another's
result. The fourth is a \emph{recovery layer}: per-worker retry budget is
stored in graph state rather than a decorator closure, making it
checkpointable, inspectable, and policy-configurable per worker without
code change. The fifth is an \emph{observability layer}: each worker
adapter wraps its subgraph call in a Langfuse SDK~v3 context-manager span,
producing nested child spans automatically through OpenTelemetry context
propagation without any coupling between the worker and the parent's
tracing client.

The sections are ordered by dependency. Section~\ref{sec:sup-command}
establishes \code{Command} first because every subsequent section returns
one. Section~\ref{sec:sup-subgraph} builds the subgraph composition
pattern next because the adapter's return type cannot be specified before
\code{Command} is understood. Section~\ref{sec:sup-state} defines the
shared state schema after composition because the isolation invariant only
becomes necessary once multiple subgraphs write to the same parent object.
Section~\ref{sec:sup-fallback} builds the retry and escalation logic after
state because per-worker retry budget is a field in \code{SubagentRecord}.
Section~\ref{sec:sup-langfuse} introduces observability after the full
node structure is established because the span lifecycle is entirely
contained within the adapter node. Section~\ref{sec:sup-full} assembles
all five layers without introducing new concepts.

This chapter introduces six terms defined in Table~\ref{tab:ch19-terms}.

\begin{table}[h]
\centering
\caption{Domain terms introduced in Chapter~19}
\label{tab:ch19-terms}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
\code{Command} &
  \code{from langgraph.types import Command}; a return value from a node
  that carries \code{goto} (the name of the next node, or \code{END}) and
  an optional \code{update} dict that is merged into graph state via
  reducers before routing occurs; the two operations are atomic. \\
Supervisor node &
  The single LangGraph node whose only job is to read current state,
  call an LLM to decide which worker should run next, and return a
  \code{Command}; it holds no explicit edges to workers in the graph
  definition---routing is entirely implicit through \code{Command.goto}. \\
Subgraph &
  A \code{StateGraph} compiled independently with
  \code{subgraph.compile()} at module load time and registered in the
  parent graph through an adapter node that translates
  \code{AgentState} to a worker-local \code{TypedDict} and back. \\
\code{subagent\_state} &
  An
  \code{Annotated[dict[str,~SubagentRecord],~operator.or\_]}
  field in \code{AgentState}; keyed by worker name; the
  \code{operator.or\_} reducer merges per-worker keys independently
  so concurrent updates cannot overwrite each other. \\
\code{SubagentRecord} &
  A \code{TypedDict} with fields \code{output:~str},
  \code{error:~str~|~None}, \code{retries:~int}, and
  \code{max\_retries:~int}; one instance per registered worker in
  \code{subagent\_state}. \\
\code{RunnableConfig} &
  \code{from langchain\_core.runnables import RunnableConfig}; the
  typed config dict passed as the second argument to every LangGraph
  node; its \code{metadata} key carries per-call observability context
  (e.g.\ a Langfuse \code{trace\_context} dict); its \code{configurable}
  key carries per-thread runtime parameters such as the checkpointer
  thread identifier. \\
\bottomrule
\end{tabularx}
\end{table}

\medskip
\noindent\textbf{Source layout for this chapter:}
\begin{verbatim}
chapter19/
  state.py                # §19.3 — AgentState, SubagentRecord,
                          #          operator.or_, initial_agent_state()
  supervisor_core.py      # §19.1, §19.4 — supervisor_node,
                          #                 RoutingDecision,
                          #                 route_or_escalate,
                          #                 escalate_node
  workers.py              # §19.2 — compiled subgraphs + adapter nodes
  observability.py        # §19.5 — Langfuse SDK v3 span helpers,
                          #          inject_trace_config
  supervisor_graph.py     # §19.6 — graph assembly, run_supervisor
\end{verbatim}


%% ------------------------------------------------------------
\section{Routing and State Updates Must Be Atomic}
\label{sec:sup-command}
%% ------------------------------------------------------------

The design problem is the gap between a routing decision and the state
that caused it. In a graph that uses \code{add\_conditional\_edges}, the
conditional function reads state and returns a destination name, but it
cannot update state: any change the routing logic wants to record---which
worker was chosen, why, or what the current retry count is---must be
written by the node before the conditional function runs. This two-step
pattern creates a window in which the graph holds a partially updated
state: the node has written its output, but the routing decision has not
yet been recorded. Under concurrent execution or after a checkpoint
restore, that window becomes a consistency hazard. The governing principle
is that a routing decision and the state update it implies must be a single
indivisible operation.

\code{Command} is the LangGraph primitive that closes this gap. A node
that returns \code{Command(goto="next\_node", update=\{"key": "value"\})}
instructs the graph runtime to merge \code{update} into state via the
field's reducer and then route to \code{goto}, in that order, as one
atomic step. No conditional edge function is needed: the graph has no
static edges between the supervisor and its workers. The supervisor node
is registered with \code{add\_node}; the workers are registered with
\code{add\_node}; the only static edge is \code{START} to
\code{"supervisor"}. Everything else is determined at runtime by
\code{Command.goto}.

The type annotation \code{Command[Literal[...]]} on the supervisor node is
not cosmetic. LangGraph uses it to validate the graph's routing structure
at compile time and to render the graph topology in visualisation tools.
A supervisor node that omits this annotation produces a graph that works at
runtime but cannot be statically validated or visualised, which removes two
diagnostic tools that matter most when the supervisor routes to an
unexpected node under an edge case.

The supervisor's routing decision comes from a structured LLM call. The
LLM receives a compact summary of current work state---not the raw
\code{AgentState} dict---and returns a \code{RoutingDecision} Pydantic
model whose \code{next\_worker} field names the worker to run, or the
sentinel string \code{"FINISH"} when all work is complete.
\code{"FINISH"} is the LLM's output vocabulary; it maps to \code{END},
the LangGraph module-level constant, inside \code{supervisor\_node}.
Passing the string \code{"FINISH"} directly to \code{Command.goto} would
cause a \code{NodeNotFoundError} at the first invocation where the
supervisor concludes work is done.

\begin{warningbox}
\code{with\_structured\_output} raises \code{ValidationError} if the LLM
returns a response that does not conform to \code{RoutingDecision}. This
happens under load when the model truncates its response, and when a
provider rate-limit returns an error payload rather than a JSON object.
Wrap the structured call in a \code{try/except} and route to
\code{"escalate"} on any failure---do not let a malformed routing response
propagate as an unhandled exception into the graph runtime.
\end{warningbox}

\begin{lstlisting}[language=Python,
  caption={Supervisor node with atomic \texttt{Command} routing},
  label={lst:sup-command}]
# chapter19/supervisor_core.py
from typing import Literal
from langgraph.types import Command
from langgraph.graph import END
from langchain_core.runnables import RunnableConfig
from langchain_litellm import ChatLiteLLM
from pydantic import BaseModel, ValidationError
from chapter19.state import AgentState

class RoutingDecision(BaseModel):
    next_worker: str  # worker node name, or "FINISH"
    reason:      str  # one sentence, stored in state for observability

_llm = ChatLiteLLM(model="anthropic/claude-sonnet-4-8")
_structured_llm = _llm.with_structured_output(RoutingDecision)

def _build_status_summary(state: AgentState) -> str:
    lines = [f"Task: {state['task']}"]
    for name, rec in state["subagent_state"].items():
        if rec["error"] is None and rec["output"] != "":
            status = "done"
        elif rec["error"]:
            status = f"error: {rec['error']}"
        else:
            status = "pending"
        lines.append(f"  {name}: {status} (retries {rec['retries']})")
    return "\n".join(lines)

async def supervisor_node(
    state: AgentState,
    config: RunnableConfig,
) -> Command[Literal["researcher", "writer",
                     "critic", "escalate"]]:
    try:
        decision: RoutingDecision = await _structured_llm.ainvoke(
            _build_status_summary(state)
        )
    except (ValidationError, Exception):
        return Command(goto="escalate")
    next_node = END if decision.next_worker == "FINISH" \
                else decision.next_worker
    return Command(
        goto=next_node,
        update={"routing_reason": decision.reason},
    )
\end{lstlisting}

Three decisions in Listing~\ref{lst:sup-command} deserve explanation.
\code{\_build\_status\_summary()} serialises \code{subagent\_state} into a
bounded string rather than passing the raw state dict to the LLM; this
decouples the prompt format from Python field names and prevents the
supervisor's routing behaviour from becoming sensitive to schema changes in
\code{AgentState}. The \code{\_structured\_llm} module-level binding calls
\code{with\_structured\_output} once at import time rather than on every
node invocation; \code{with\_structured\_output} constructs a tool-calling
chain at call time, and rebuilding it per invocation adds measurable
overhead at scale. The \code{except (ValidationError, Exception)} clause
is deliberately broad: a provider HTTP error, a JSON parse failure, and a
Pydantic schema mismatch all have the same correct response from the
supervisor---route to \code{"escalate"}---and distinguishing them here
would produce dead code.

A supervisor that can express any routing decision in a single \code{Command}
return is only useful if the workers it routes to are themselves independent
units. The next problem is composition: how to register a multi-step worker
as a single node without collapsing the state boundary between the worker
and the supervisor.


%% ------------------------------------------------------------
\section{Each Worker Is an Independent Graph}
\label{sec:sup-subgraph}
%% ------------------------------------------------------------

The design problem is the boundary between a worker's internal steps and
the supervisor's coordination logic. The naive approach is to define worker
logic as a Python function called directly from a node in the parent graph.
This makes the worker's internal steps invisible to LangGraph's
checkpointer, its errors impossible to classify at the subgraph level, and
its state schema entangled with the parent's. The governing principle is
that a worker is a graph: it has its own \code{TypedDict} state, its own
internal nodes, and its own compiled execution context, registered in the
parent as a single addressable node through an adapter that translates
state across the boundary.

A \emph{subgraph} is a \code{StateGraph} compiled once with
\code{subgraph.compile()} and then made available to the parent through
a Python function---the adapter node---that transforms
\code{AgentState} into the worker's local \code{TypedDict}, invokes the
compiled subgraph, and returns a \code{Command} encoding both the state
update and the routing decision. Calling \code{compile()} at module scope
rather than inside \code{add\_node} matters: \code{compile()} validates
the subgraph's internal edge structure and locks its state schema at import
time. A misconfigured worker fails immediately on startup rather than at
the first production call.

\begin{notebox}
\code{CompiledStateGraph.ainvoke()} returns a plain \code{dict}, not the
typed \code{TypedDict} used as the subgraph's state parameter. The
annotation \code{result: ResearchState = await research\_compiled.ainvoke(...)}
is a programmer's cast that passes type-checking but is not enforced at
runtime. Always access \code{result} with \code{.get()} or verify keys
before use.
\end{notebox}

The adapter node owns three responsibilities: translating \code{AgentState}
fields into the worker-local input, passing the \code{RunnableConfig}
unchanged to \code{ainvoke} so that Langfuse trace context propagates into
the subgraph, and calling \code{route\_or\_escalate()}---introduced in
Section~\ref{sec:sup-fallback}---to produce the \code{Command} that
carries the updated \code{SubagentRecord} back. Because the adapter must
choose between returning to the supervisor on success and routing to
escalation on failure, its return type is
\code{Command[Literal["supervisor",~"escalate"]]}, not \code{dict}. A node
that returns a \code{dict} cannot express routing; \code{Command} is the
only correct return type once routing is a responsibility of the node.

\begin{lstlisting}[language=Python,
  caption={Compiled subgraph and its parent-facing adapter node},
  label={lst:sup-subgraph}]
# chapter19/workers.py
from typing import Literal, TypedDict
from langgraph.graph import StateGraph, START, END
from langgraph.types import Command
from langchain_core.runnables import RunnableConfig
from langchain_litellm import ChatLiteLLM
from chapter19.state import AgentState, SubagentRecord

_llm = ChatLiteLLM(model="anthropic/claude-sonnet-4-8")

class ResearchState(TypedDict):
    query:    str
    findings: str
    error:    str | None

async def _research_step(
    state: ResearchState,
    config: RunnableConfig,
) -> ResearchState:
    try:
        findings = (await _llm.ainvoke(state["query"])).content
        return ResearchState(
            query=state["query"], findings=findings, error=None,
        )
    except Exception as exc:
        return ResearchState(
            query=state["query"], findings="", error=str(exc),
        )

_research_graph = StateGraph(ResearchState)
_research_graph.add_node("research_step", _research_step)
_research_graph.add_edge(START, "research_step")
_research_graph.add_edge("research_step", END)
research_compiled = _research_graph.compile()   # module scope

async def research_node(
    state: AgentState,
    config: RunnableConfig,
) -> Command[Literal["supervisor", "escalate"]]:
    result: dict = await research_compiled.ainvoke(
        ResearchState(query=state["task"], findings="", error=None),
        config=config,
    )
    record = SubagentRecord(
        output=result.get("findings", ""),
        error=result.get("error"),
        retries=state["subagent_state"]["researcher"]["retries"],
        max_retries=state["subagent_state"]["researcher"]["max_retries"],
    )
    from chapter19.supervisor_core import route_or_escalate
    return route_or_escalate("researcher", record)
\end{lstlisting}

Three decisions in Listing~\ref{lst:sup-subgraph} deserve explanation.
The \code{error} field in \code{ResearchState} is set by the worker itself
inside \code{\_research\_step} rather than caught by the adapter; this
keeps failure classification inside the subgraph where the exception type
is known, rather than forcing the adapter to infer meaning from a bare
\code{Exception} re-raised across a boundary. The \code{config=config}
passthrough to \code{ainvoke} is not optional: constructing a new
\code{RunnableConfig} inside the adapter would drop the \code{metadata}
dict carrying the Langfuse trace context, orphaning all spans the subgraph
creates from the parent trace. The \code{from chapter19.supervisor\_core
import route\_or\_escalate} import is deferred to inside the function body
to break the circular dependency between \code{workers.py} and
\code{supervisor\_core.py}; both modules reference each other's symbols,
and a module-level import in either file would raise \code{ImportError} at
startup before any graph is constructed.

With the subgraph boundary established, the next problem is what happens to
state on the parent side when multiple worker adapters write to
\code{subagent\_state}---potentially concurrently in a fan-out scenario.


%% ------------------------------------------------------------
\section{Worker State Must Not Reach Supervisor Fields}
\label{sec:sup-state}
%% ------------------------------------------------------------

The design problem is shared mutable state under concurrent writes. When
two worker adapters return simultaneously in a LangGraph fan-out, both
write to \code{subagent\_state}. Without a reducer, the graph applies
state updates as last-write-wins dict replacement: whichever adapter
finishes last overwrites the other's result entirely. The naive approach
is a flat set of top-level fields---\code{researcher\_output},
\code{writer\_output}---that avoids the concurrent-write problem by
giving each worker its own key, but at the cost of forcing the supervisor
to enumerate every worker's field by name. Adding a new worker then
requires modifying \code{AgentState}, every node that reads output by field
name, and the supervisor's routing prompt. The governing principle is that
worker-local data belongs in a single namespaced field whose reducer
guarantees per-key independence regardless of concurrency.

The \code{Annotated} type annotation is LangGraph's mechanism for attaching
a reducer to a state field. \code{Annotated[dict[str,~SubagentRecord],
operator.or\_]} binds \code{operator.or\_} as the merge function for
\code{subagent\_state}. \code{operator.or\_} on two dicts computes
\code{left | right}, equivalent to \code{\{**left,~**right\}} with
right-key precedence. When \code{research\_node} returns a \code{Command}
whose \code{update} contains \code{\{"subagent\_state": \{"researcher":
record\}\}}, the runtime calls \code{operator.or\_(existing, \{"researcher":
record\})}, merging only the \code{"researcher"} key while leaving all
other worker keys untouched.

\begin{warningbox}
A worker adapter must pass only the delta \code{\{worker\_name: record\}}
to \code{Command.update}, not the full merged \code{subagent\_state} dict.
Assembling the full dict inside the node reads state before the reducer
runs; under concurrent fan-out, this read happens before the other
worker's write, so merging the full dict will silently overwrite that
worker's result regardless of the reducer.
\end{warningbox}

\code{SubagentRecord} carries four fields. \code{output} and \code{error}
are the worker's result and failure message. \code{retries} and
\code{max\_retries} encode the per-worker retry policy: \code{max\_retries}
is set at initialisation time and never changed; \code{retries} is
incremented by \code{route\_or\_escalate()} when a failed worker is
re-routed rather than escalated. Per-worker retry budget differs
intentionally: a researcher calls an external retrieval pipeline and
network failures are common; a writer failure is almost always a prompt or
model problem that a retry will not fix. These are schema facts visible in
\code{initial\_agent\_state()}, not runtime configuration buried in a
decorator.

\begin{lstlisting}[language=Python,
  caption={State schema with \texttt{operator.or\_} reducer and per-worker record},
  label={lst:sup-state}]
# chapter19/state.py
import operator
from typing import Annotated, TypedDict

class SubagentRecord(TypedDict):
    output:      str
    error:       str | None
    retries:     int
    max_retries: int

class AgentState(TypedDict):
    task:           str
    final_answer:   str
    routing_reason: str
    subagent_state: Annotated[
        dict[str, SubagentRecord],
        operator.or_,
    ]

def _blank(max_retries: int) -> SubagentRecord:
    return SubagentRecord(
        output="", error=None, retries=0, max_retries=max_retries,
    )

def initial_agent_state(task: str) -> AgentState:
    return AgentState(
        task=task,
        final_answer="",
        routing_reason="",
        subagent_state={
            "researcher": _blank(max_retries=3),
            "writer":     _blank(max_retries=1),
            "critic":     _blank(max_retries=2),
        },
    )
\end{lstlisting}

Two decisions in Listing~\ref{lst:sup-state} deserve explanation.
\code{final\_answer} lives at the top level of \code{AgentState} rather
than inside the last worker's \code{SubagentRecord} because the supervisor
writes it when it concludes the task is done---at that moment the
supervisor does not know and should not care which worker produced the
final output; the field is a coordination result, not a worker output.
\code{initial\_agent\_state()} is a plain function rather than a default
factory on the \code{TypedDict} fields because \code{TypedDict} does not
support field defaults: attempting to declare \code{subagent\_state:
Annotated[dict[str,~SubagentRecord], operator.or\_] = \{\}} raises a
\code{TypeError} at class definition time.

The state schema is now stable. The remaining problem is what to do when a
worker's \code{SubagentRecord} arrives with a non-null \code{error} field:
the routing decision is not simply ``retry the same worker'' but a policy
that depends on how many times that worker has already failed.


%% ------------------------------------------------------------
\section{Retry Budget Lives in State, Not in a Decorator}
\label{sec:sup-fallback}
%% ------------------------------------------------------------

The design problem is invisible retry accounting. The conventional
approach---a \code{tenacity} decorator or a \code{while} loop inside the
node function---records retry attempts in a closure or a local variable.
That counter is not part of graph state: it cannot be checkpointed, it
resets to zero when the process restarts mid-task, it is invisible to the
supervisor's routing LLM, and it cannot be read by a monitoring system.
When the retry budget is exhausted and escalation is required, the
decorator raises an exception that the graph runtime catches as an
unhandled node error rather than routing to a named escalation node. The
governing principle is that retry accounting is coordination state and must
live where all coordination state lives---in \code{AgentState}---where it
is durable, observable, and policy-configurable per worker without code
change.

The design separates two concerns that decorators collapse together:
\emph{execution} and \emph{retry policy}. The worker subgraph executes
once and reports its result into \code{SubagentRecord.error}; it has no
knowledge of retry budget. The function \code{route\_or\_escalate()} reads
the updated record, checks whether \code{retries~<~max\_retries}, and
returns a \code{Command} that either routes back to the same worker with
an incremented \code{retries} count, returns to the supervisor, or routes
to \code{"escalate"}. The \code{Command.update} in all cases passes only
the delta \code{\{worker\_name:~record\}} so the \code{operator.or\_}
reducer merges exactly the one changed key.

Escalation routes to a named node rather than directly to \code{END}
because escalation is a discrete business event, not a graph termination
condition. A named \code{escalate\_node} appears as a span in Langfuse,
can be replaced with \code{interrupt()} for human-in-the-loop
approval---\code{interrupt()} is a LangGraph primitive introduced in
Chapter~21~\S21.2 that pauses graph execution and surfaces current state
to an external caller---and can be individually unit-tested. A direct
\code{Command(goto=END)} from the adapter would make all three of those
capabilities impossible without restructuring the graph.

\begin{lstlisting}[language=Python,
  caption={Per-worker retry logic as a pure function},
  label={lst:sup-fallback}]
# chapter19/supervisor_core.py  (continued)
from langgraph.types import Command
from langgraph.graph import END
from langchain_core.runnables import RunnableConfig
from chapter19.state import AgentState, SubagentRecord

def route_or_escalate(
    worker_name:    str,
    updated_record: SubagentRecord,
) -> Command:
    delta = {"subagent_state": {worker_name: updated_record}}
    if updated_record["error"] is None:
        return Command(goto="supervisor", update=delta)
    if updated_record["retries"] < updated_record["max_retries"]:
        retried = SubagentRecord(
            **{**updated_record,
               "retries": updated_record["retries"] + 1},
        )
        return Command(
            goto=worker_name,
            update={"subagent_state": {worker_name: retried}},
        )
    return Command(goto="escalate", update=delta)

async def escalate_node(
    state:  AgentState,
    config: RunnableConfig,
) -> Command:
    # Log all workers that exhausted their retry budget.
    failed = {
        name: rec for name, rec in state["subagent_state"].items()
        if rec["error"] is not None
    }
    # Production: emit to alerting here. Failed dict is serialisable.
    _ = failed
    return Command(goto=END)
\end{lstlisting}

Three decisions in Listing~\ref{lst:sup-fallback} deserve explanation.
\code{route\_or\_escalate()} is a plain synchronous function, not a
coroutine, because it makes no I/O calls; making it \code{async} would
require an \code{await} at every call site without gaining any concurrency,
and would make it impossible to call from a synchronous context such as a
unit test. The \code{SubagentRecord(**\{**updated\_record, "retries": ...\})}
construction is intentional: \code{TypedDict} does not support
keyword-override spreading (\code{SubagentRecord(**rec, retries=n)} is a
\code{TypeError} because \code{TypedDict} constructors do not accept
keyword argument override after a spread); the explicit dict merge
\code{\{**updated\_record, "retries": n\}} is the correct pattern.
\code{escalate\_node} collects failed workers into a local dict rather
than calling any I/O function inline; the alerting call belongs in the
production deployment layer (a LangGraph interrupt handler or a
post-processing hook), not embedded in a node that must remain testable
in isolation.

The retry and escalation logic is now correct and checkpointable. The
remaining problem is observability: the supervisor routing cycle and each
worker's subgraph invocation produce spans in isolation, and without an
enclosing parent trace those spans cannot be correlated into a single
end-to-end view of a supervision run.


%% ------------------------------------------------------------
\section{Observability Must Cross the Subgraph Boundary}
\label{sec:sup-langfuse}
%% ------------------------------------------------------------

The design problem is span orphaning under nested async execution. When the
supervisor routes to a worker adapter, and the adapter invokes a compiled
subgraph via \code{ainvoke}, the subgraph's internal LLM calls emit
OpenTelemetry spans in the process's active OTEL context. If no enclosing
Langfuse span exists in that context at the moment \code{ainvoke} is
called, those spans are either emitted as anonymous root traces or silently
discarded, making it impossible to correlate a subgraph call with the
supervision run that triggered it. The naive approach---a module-level
\code{Langfuse()} singleton---ties all concurrent supervision runs to the
same tracing context, producing incorrect span nesting under parallelism.
The governing principle is that observability context is per-invocation:
the adapter node opens a Langfuse span as a context manager around its
\code{ainvoke} call, and the Langfuse SDK~v3 OTEL backend nests every span
created inside that block as a child automatically.

Langfuse Python SDK~v3 (generally available June 2025) is built on
OpenTelemetry. When a span is opened with
\code{langfuse.start\_as\_current\_span()} as a context manager, that span
becomes the active OTEL span for the duration of the \code{with} block.
Any LLM call made inside the block that is instrumented via
LangChain's callback system---which LangGraph uses---emits its generation
span as a child of the active Langfuse span automatically, without any
additional configuration. This means the adapter's \code{async with}
block is the only integration point needed: no callbacks are registered,
no span IDs are threaded through state, and no worker internals are
modified. The worker subgraph is completely unaware of the tracing context.

The singleton accessor \code{get\_client()} returns the same
\code{Langfuse} instance throughout the process. Calling
\code{Langfuse()} a second time with the same public key reuses the
existing instance rather than creating a new one, but \code{get\_client()}
makes the singleton contract explicit and requires no constructor
arguments after the first initialisation.

\begin{notebox}
\code{RunnableConfig} has two separate namespaces that must not be mixed.
\code{config["configurable"]} is the LangGraph-managed namespace for
per-thread runtime keys such as the checkpointer's \code{thread\_id}.
\code{config["metadata"]} is the user-managed namespace for call-scoped
context. Placing a Langfuse \code{trace\_context} dict under
\code{"configurable"} causes the LangGraph checkpointer to attempt to use
it as a thread parameter; placing \code{thread\_id} under
\code{"metadata"} causes the checkpointer to miss it entirely.
\end{notebox}

The \code{inject\_trace\_config()} function in \code{observability.py}
constructs the \code{RunnableConfig} that every invocation of
\code{run\_supervisor()} passes to the graph. It opens a root Langfuse span
for the entire supervision run, reads the W3C-compliant trace identifier
from that span, and embeds it as a \code{trace\_context} dict under
\code{config["metadata"]}. Each worker adapter reads this dict and passes
it to \code{start\_as\_current\_span()} via the \code{trace\_context}
parameter, which instructs the SDK to attach the new span as a child of
the root span regardless of which async task is currently running.

\begin{lstlisting}[language=Python,
  caption={Langfuse SDK v3 root span, worker span, and config injection},
  label={lst:sup-langfuse}]
# chapter19/observability.py
from langfuse import get_client, Langfuse
from langchain_core.runnables import RunnableConfig

_lf = get_client()

def inject_trace_config(thread_id: str) -> RunnableConfig:
    root = _lf.start_span(name="supervisor_run")
    trace_ctx = {"trace_id": root.trace_id,
                 "parent_span_id": root.id}
    return {
        "configurable": {"thread_id": thread_id},
        "metadata":     {"langfuse_trace_context": trace_ctx,
                         "langfuse_root_span":     root},
    }

def close_root_span(config: RunnableConfig) -> None:
    root = (config.get("metadata") or {}).get("langfuse_root_span")
    if root is not None:
        root.end()
    _lf.flush()
\end{lstlisting}

\begin{lstlisting}[language=Python,
  caption={Adapter node with worker span scoped to \texttt{ainvoke} call},
  label={lst:sup-langfuse-adapter}]
# chapter19/workers.py  (research_node with observability)
async def research_node(
    state: AgentState,
    config: RunnableConfig,
) -> Command[Literal["supervisor", "escalate"]]:
    meta = config.get("metadata") or {}
    trace_ctx = meta.get("langfuse_trace_context")
    span_kwargs = ({"trace_context": trace_ctx}
                   if trace_ctx else {})
    async with _lf.start_as_current_span(
        name="researcher",
        input={"task": state["task"]},
        **span_kwargs,
    ) as worker_span:
        result: dict = await research_compiled.ainvoke(
            ResearchState(
                query=state["task"], findings="", error=None,
            ),
            config=config,
        )
        worker_span.update(
            output={"findings": result.get("findings", ""),
                    "error":    result.get("error")},
        )
    record = SubagentRecord(
        output=result.get("findings", ""),
        error=result.get("error"),
        retries=state["subagent_state"]["researcher"]["retries"],
        max_retries=state["subagent_state"]["researcher"]["max_retries"],
    )
    from chapter19.supervisor_core import route_or_escalate
    return route_or_escalate("researcher", record)
\end{lstlisting}

Three decisions across Listings~\ref{lst:sup-langfuse}
and~\ref{lst:sup-langfuse-adapter} deserve explanation.
\code{inject\_trace\_config()} uses \code{\_lf.start\_span()} (not
\code{start\_as\_current\_span()}) for the root span because the root span
must outlive the function call that creates it; a context manager would
close it at the end of \code{inject\_trace\_config()}, before any worker
has run. The root span is stored under
\code{config["metadata"]["langfuse\_root\_span"]} so \code{close\_root\_span()}
can call \code{.end()} on it after \code{graph.ainvoke()} returns; without
this explicit close, the root span's end time is never recorded and the
Langfuse UI shows an open span indefinitely. The adapter uses
\code{start\_as\_current\_span()} (not \code{start\_span()}) so that the
span becomes the active OTEL context during \code{ainvoke}; this is the
mechanism by which LangChain callback spans emitted inside
\code{research\_compiled} are automatically nested as children without any
explicit parent reference.

All five layers are individually correct and independently testable. The
final problem is assembly: how to register all nodes, verify the absence of
\code{add\_conditional\_edges}, and confirm that the complete graph
produces the expected span hierarchy end-to-end.


%% ------------------------------------------------------------
\section{Full System: Three Subagents Under One Supervisor}
\label{sec:sup-full}
%% ------------------------------------------------------------

The design problem is silent structural divergence between the intended
graph topology and the compiled one. With \code{Command}-based routing,
the graph has no explicit edges between the supervisor and its workers; a
programmer who adds a new worker and forgets to add it to the
\code{RoutingDecision} prompt compiles a valid graph where the supervisor's
LLM never routes to the new worker. There is no compiler error and no
runtime exception---the new worker is silently unreachable. The governing
principle is that the graph assembly function must be explicit about every
registered node, and the integration test must invoke the graph with a task
that exercises all three workers, asserting not just graph termination but
that all three workers produced output and left no open errors.

Before the full graph listing, the \code{writer\_node} and
\code{critic\_node} adapters in \code{workers.py} follow the exact same
structure as \code{research\_node}: each compiles its own
\code{StateGraph} at module scope, each adapter wraps its \code{ainvoke}
call in a \code{start\_as\_current\_span()} context manager keyed to the
\code{langfuse\_trace\_context} from \code{config["metadata"]}, and each
returns \code{route\_or\_escalate(worker\_name, record)}. The writer's
internal step consumes \code{state["subagent\_state"]["researcher"]["output"]}
to draft a summary; the critic's internal step evaluates that draft and
writes its verdict directly to \code{final\_answer} via
\code{Command(goto="supervisor", update=\{"final\_answer": verdict\})}.
No new structural concepts appear in either adapter; only the internal step
function and the state fields it reads differ.

\begin{lstlisting}[language=Python,
  caption={Graph assembly: all nodes registered, one static edge},
  label={lst:sup-full-graph}]
# chapter19/supervisor_graph.py
from langgraph.graph import StateGraph, START
from langgraph.graph.state import CompiledStateGraph
from langgraph.checkpoint.memory import MemorySaver
from chapter19.state import AgentState, initial_agent_state
from chapter19.supervisor_core import supervisor_node, escalate_node
from chapter19.workers import research_node, writer_node, critic_node
from chapter19.observability import inject_trace_config, close_root_span

def build_supervisor_graph() -> CompiledStateGraph:
    builder = StateGraph(AgentState)
    builder.add_node("supervisor",  supervisor_node)
    builder.add_node("researcher",  research_node)
    builder.add_node("writer",      writer_node)
    builder.add_node("critic",      critic_node)
    builder.add_node("escalate",    escalate_node)
    builder.add_edge(START, "supervisor")
    # No add_conditional_edges: all routing via Command.goto.
    return builder.compile(checkpointer=MemorySaver())

async def run_supervisor(task: str, thread_id: str) -> AgentState:
    graph  = build_supervisor_graph()
    config = inject_trace_config(thread_id=thread_id)
    try:
        result: AgentState = await graph.ainvoke(
            initial_agent_state(task),
            config=config,
        )
    finally:
        close_root_span(config)
    return result
\end{lstlisting}

Two decisions in Listing~\ref{lst:sup-full-graph} deserve explanation.
\code{MemorySaver} is the development checkpointer: it stores all graph
state in process memory and loses it on restart. It is the correct choice
here precisely because it makes the retry-in-state limitation concrete---a
restarted process loses \code{SubagentRecord.retries}, which is exactly
the failure mode that motivates replacing \code{MemorySaver} with
\code{AsyncSqliteSaver} from \code{langgraph.checkpoint.sqlite.aio} in
production, as Chapter~22~\S22.1 introduces. The \code{try/finally} block
in \code{run\_supervisor()} ensures \code{close\_root\_span()} is called
even if \code{graph.ainvoke()} raises; without it, the root Langfuse span
remains open in the Langfuse UI for every failed invocation, producing
misleading latency histograms.

\begin{lstlisting}[language=Python,
  caption={Integration test: all three workers produce output},
  label={lst:sup-full-run}]
# chapter19/test_supervisor.py
import pytest
from chapter19.supervisor_graph import run_supervisor

@pytest.mark.asyncio
async def test_all_workers_produce_output() -> None:
    result = await run_supervisor(
        task="Research and summarise quantum error correction.",
        thread_id="test-001",
    )
    assert result["final_answer"] != "", \
        "critic must write final_answer before FINISH"
    for worker in ("researcher", "writer", "critic"):
        rec = result["subagent_state"][worker]
        assert rec["error"] is None, \
            f"{worker} finished with error: {rec['error']}"
        assert rec["output"] != "", \
            f"{worker} produced no output"

@pytest.mark.asyncio
async def test_retry_state_persists_across_nodes() -> None:
    pytest.skip(
        "Requires AsyncSqliteSaver: Chapter 22 §22.1 migration"
    )
\end{lstlisting}

Two decisions in Listing~\ref{lst:sup-full-run} deserve explanation.
The first test asserts \code{final\_answer~!=~""} before checking
individual worker records because that is the highest-level contract: if
the supervisor never concluded and the critic never wrote a verdict, the
per-worker assertions are noise that obscures the root cause. The second
test is not empty---it calls \code{pytest.skip()} with an explicit message
naming the capability it requires; this makes the test file's position in
the \code{pytest} output a visible indicator of which Chapter~22 migration
has not yet been applied, rather than a silent absence that only a diligent
reader would notice.

\begin{notebox}
The supervisor graph as built here routes workers sequentially: the
supervisor activates one worker per cycle. This is correct for tasks where
worker B's input depends on worker A's output---research feeds writing,
writing feeds critique. When a task can be decomposed into independent
subtasks with no data dependency, sequential activation wastes wall-clock
time. \code{Send}---a LangGraph primitive that spawns a named node with a
custom input dict, introduced in Chapter~20 as the parallel fan-out
mechanism---replaces the single-worker-per-supervisor-cycle pattern with a
batch dispatch, allowing all independent workers to run concurrently. The
\code{operator.or\_} reducer on \code{subagent\_state} makes this
transition safe: parallel workers writing to different keys under the same
reducer produce a correct merged result without any change to the state
schema or the retry logic.
\end{notebox}

Chapter~20 takes the next step from this architecture: where this chapter
asks which single worker should run next, the next chapter asks which set
of workers can run simultaneously---and what the correct decomposition
model is for dispatching a task to multiple independent workers in a single
supervisor cycle rather than one at a time.
