%% ============================================================
%% Chapter 27 — Resilience & Error Handling
%% Production-Grade Agentic AI: LangGraph + Python Frameworks
%%
%% Deep-review fixes applied (June 2026):
%%   [F1]  wait_exponential_jitter signature corrected:
%%         (initial, max, exp_base, jitter) — "max" kwarg, not "min"
%%   [F2]  InMemorySaver is the current class name (langgraph.checkpoint.memory)
%%         confirmed via official LangGraph reference docs example
%%   [F3]  config_schema deprecated in LangGraph v0.6; context_schema is current
%%   [F4]  Langfuse trace URL: from langfuse_handler.get_trace_id(), NOT config["run_id"]
%%   [F5]  asynccontextmanager required (not contextmanager) for async node wrapping
%%   [F6]  route_cascade success path returns END sentinel, not bare string "end"
%%   [F7]  error_class="none" sentinel removed; field absent on happy path (NotRequired)
%%   [F8]  reraise + retry_error_callback are mutually exclusive (Tenacity docs confirmed)
%%   [F9]  NodeInterrupt is handled by LangGraph runtime, never reaches node except blocks
%%   [F10] RetryPolicy is a native LangGraph add_node param — mentioned in notebox
%%   [F11] add_conditional_edges path_map "__end__" -> END confirmed via reference docs
%%   [F12] DeadLetterRecord.model_validate() is Pydantic v2 API; parse_obj() removed
%%   [F13] failed_at timezone-aware assertion: tzinfo is not None (portable across impls)
%% ============================================================




A production LLM agent fails in ways that a conventional API service does
not. A conventional service either returns a result or raises a
deterministic error; the retry-or-die decision is straightforward. An agent
can fail at a vendor rate limit, at a content-policy boundary, at a network
timeout, or at the recursion ceiling of its own graph---and each of these
failure modes demands a qualitatively different response. Treating them all
as HTTP~500 is the single most common cause of agent reliability incidents:
a \code{RateLimitError} that deserves exponential back-off instead crashes
the run; a \code{ContentPolicyViolationError} that should dead-letter
immediately instead wastes quota retrying a prompt that will always be
rejected. The cost is not only a bad user experience: a crashed run loses
its checkpointed state, and the engineering team has no structured record of
what went wrong or at which tier of the system.

The solution in this chapter is organised into five layers. The
\emph{retry layer} wraps individual LLM calls in Tenacity decorators whose
backoff profiles are matched to the exception class they handle. The
\emph{routing layer} expresses the distinction between retriable (soft)
and terminal (hard) failures as LangGraph conditional edges, keeping routing
logic at the graph level rather than hidden in node try/except blocks. The
\emph{audit layer} snapshots every terminal failure into a typed
\code{DeadLetterRecord} Pydantic model that lives in agent state and is
linked to the Langfuse trace for that run. The \emph{observability layer}
instruments each node with Prometheus counters and defines alerting rules
that fire when per-node error rates breach a five-percent SLO budget. The
\emph{degradation layer} assembles those nodes into a four-tier cascade ---
full model, cheap model, static fallback, and escalation --- each tier
activating only when the tier above it has been exhausted.

The sections are ordered by dependency. The retry layer
(\S\ref{sec:retries-depend-on-failure-type}) must come before the routing
layer (\S\ref{sec:hard-vs-soft-branches}) because the routing layer reads
the \emph{result} of exhausted retries: a node writes \code{error\_class}
only after its Tenacity decorator has given up. The routing layer must come
before the dead-letter section (\S\ref{sec:dead-letter-state}) because the
dead-letter node is a routing destination, and its contract is only
meaningful once the edges that lead to it are defined. Prometheus
instrumentation (\S\ref{sec:error-budget-tracking}) follows the dead-letter
section because the counters are incremented inside the dead-letter node
itself. The degradation cascade (\S\ref{sec:degradation-cascade}) follows
observability because its nodes are instrumented. The capstone
(\S\ref{sec:capstone-resilient-agent}) assembles all prior sections and
verifies each failure path with failure injection tests.

\begin{table}[h]
\centering
\caption{Domain terms introduced in this chapter}
\label{tab:ch27-terms}
\begin{tabular}{lp{9.5cm}}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
\emph{soft failure} &
  A transient, retriable error condition --- such as a vendor rate limit or
  a network timeout --- where re-sending the same request after a delay has
  a non-zero probability of succeeding. \\
\emph{hard failure} &
  A terminal, non-retriable error condition --- such as a content policy
  violation --- where re-sending the same request will always produce the
  same rejection. \\
\emph{dead-letter state} &
  A structured record of a terminal failure stored inside the agent's
  checkpointed state, analogous to a dead-letter queue in message-broker
  architectures, but queryable via graph state history. \\
\emph{error budget} &
  The fraction of requests that a system is allowed to fail while still
  meeting its SLO; commonly expressed as $1 - \text{availability target}$
  (e.g.\ a 99.95\% SLO gives a 0.05\% error budget). \\
\emph{RetryError} &
  The exception raised by Tenacity when all retry attempts are exhausted
  and \code{reraise=False}; with \code{reraise=True} the original exception
  class is re-raised instead, which is the preferred pattern for typed
  exception flow. \\
\emph{degradation cascade} &
  A sequence of progressively cheaper fallback tiers, each activated only
  when the tier above it has been exhausted, providing partial service
  under conditions of full LLM unavailability. \\
\bottomrule
\end{tabular}
\end{table}

\medskip
\noindent\textbf{Source layout for this chapter:}

\begin{lstlisting}[language=bash, caption={Chapter 27 source tree},
  label={lst:ch27-tree}]
src/
  resilience/
    decorators.py     # Tenacity per-exception retry decorators (S27.1)
    dead_letter.py    # DeadLetterRecord model + dead_letter_node (S27.3)
  graph/
    state.py          # AgentState TypedDict with error fields (S27.2)
    routing.py        # route_after_llm, route_cascade (S27.2, S27.5)
    degradation.py    # four-tier cascade nodes (S27.5)
    agent.py          # build_resilient_graph factory (S27.6)
  observability/
    metrics.py        # Prometheus counters + instrument_node (S27.4)
k8s/
  alerts/
    agent-slo.yaml    # PrometheusRule CRD (S27.4)
tests/
  fakes.py            # FailureInjector (S27.6)
  test_resilient_agent.py  # failure injection tests (S27.6)
\end{lstlisting}

%%------------------------------------------------------------
\section{Retries Depend on Failure Type}
\label{sec:retries-depend-on-failure-type}
%%------------------------------------------------------------

The design problem here is that a single \code{@retry} decorator applied
to all LLM call failures treats a vendor rate limit and a network timeout
as if they were the same kind of problem. They are not: a
\code{RateLimitError} is a contention problem shared across every pod in
your deployment, and the canonical solution is \emph{full-jitter
exponential backoff} --- waiting a random interval drawn from an
exponentially growing window so that many pods do not retry in lockstep.
A \code{Timeout} is a transient connection blip, and the appropriate
response is a fast retry with modest additive jitter, not a 60-second wait.
Applying the slower strategy to the faster error inflates p99 latency;
applying the faster strategy to the rate-limit error amplifies the
thundering herd that caused the limit in the first place. The principle is
that each exception class requires its own retry decorator whose wait and
stop parameters match the semantics of that specific failure mode.

\emph{Tenacity} is a Python retry library whose composable primitives ---
\code{wait}, \code{stop}, \code{retry}, and \code{before\_sleep} --- are
passed as arguments to the \code{@retry} decorator factory. The \code{wait}
argument controls the inter-attempt delay; the \code{stop} argument controls
when to give up; the \code{retry} argument is a predicate that determines
which exception types trigger a retry attempt. Two wait strategies are
relevant here. \code{wait\_random\_exponential} implements the Full Jitter
algorithm: each wait is drawn uniformly from
\code{[min, multiplier * 2**n]}, where \code{n} is the attempt count;
Tenacity's own documentation labels this strategy ``suitable for resolving
contention between multiple processes for a shared resource'' and explicitly
contrasts it with \code{wait\_exponential}, which uses fixed intervals and
is ``not suitable'' for that case. \code{wait\_exponential\_jitter} uses
the formula \code{min(initial * 2**n + random.uniform(0, jitter), max)} ---
additive jitter on a deterministic exponential base --- and is appropriate
for per-instance timeouts that are not shared across pods.

Critically, \code{reraise} and \code{retry\_error\_callback} are mutually
exclusive in Tenacity: the \code{Retrying} class documentation shows both
as optional constructor arguments, but when \code{retry\_error\_callback}
is set it takes precedence and \code{reraise} is ignored. The preferred
pattern for typed exception flow is \code{reraise=True}, which re-raises
the original exception class rather than Tenacity's \code{RetryError}
wrapper --- the calling code can then inspect the exception's type to decide
routing, rather than unwrapping a \code{RetryError} to find what it
contains.

\begin{lstlisting}[language=Python,
  caption={Three decorators, three backoff profiles ---
    \texttt{src/resilience/decorators.py}},
  label={lst:tenacity-decorators}]
import logging
from tenacity import (
    retry, stop_after_attempt,
    wait_random_exponential,
    wait_exponential_jitter,
    wait_fixed,
    retry_if_exception_type,
    before_sleep_log, after_log,
)
from litellm.exceptions import (
    RateLimitError, Timeout, APIConnectionError,
)

log = logging.getLogger(__name__)

def rate_limit_retry(fn):
    """Full-jitter backoff for RateLimitError (multi-pod safe)."""
    return retry(
        wait=wait_random_exponential(multiplier=1, min=2, max=60),
        stop=stop_after_attempt(4),
        retry=retry_if_exception_type(RateLimitError),
        before_sleep=before_sleep_log(log, logging.INFO),
        after=after_log(log, logging.WARNING),
        reraise=True,
    )(fn)

def timeout_retry(fn):
    """Additive-jitter exponential backoff for Timeout."""
    return retry(
        wait=wait_exponential_jitter(initial=1, max=15),
        stop=stop_after_attempt(3),
        retry=retry_if_exception_type(Timeout),
        before_sleep=before_sleep_log(log, logging.WARNING),
        reraise=True,
    )(fn)

def connection_retry(fn):
    """Fixed short wait for APIConnectionError (transient TCP blips)."""
    return retry(
        wait=wait_fixed(1),
        stop=stop_after_attempt(2),
        retry=retry_if_exception_type(APIConnectionError),
        reraise=True,
    )(fn)
\end{lstlisting}

Three decisions in Listing~\ref{lst:tenacity-decorators} deserve
explanation. First, \code{wait\_random\_exponential} is used for
\code{RateLimitError} and \code{wait\_exponential\_jitter} for
\code{Timeout}: the Full Jitter algorithm distributes retries uniformly
across a wide window, preventing multiple pods from retrying simultaneously
after the same rate-limit response. \code{wait\_exponential\_jitter} adds a
smaller random term (\code{random.uniform(0, jitter=1.0)} by default) to a
deterministic base, which is appropriate for per-instance timeouts where
the pods are not competing for the same resource. The \code{wait\_exponential\_jitter}
signature is \code{(initial, max, exp\_base, jitter)} --- both \code{max}
and \code{jitter} are keyword arguments with defaults, so
\code{wait\_exponential\_jitter(initial=1, max=15)} is a valid call that
uses the default \code{jitter=1.0} and \code{exp\_base=2}. Second,
\code{Timeout} and \code{APIConnectionError} are handled by separate
decorators with different stop budgets:
\code{stop\_after\_attempt(3)} for timeouts versus
\code{stop\_after\_attempt(2)} for connection errors, because a connection
that fails twice is more likely to indicate a dead host than a timeout that
fails three times. Third, \code{before\_sleep\_log} fires before each sleep
interval and logs the retry countdown; \code{after\_log} fires after each
attempt and logs the outcome on exception --- both hooks are necessary for
full retry observability without writing custom callbacks.

\begin{notebox}
  \textbf{LangGraph's native \code{RetryPolicy} vs.\ Tenacity decorators.}
  Since LangGraph~0.2, \code{add\_node} accepts a \code{retry\_policy}
  parameter of type \code{RetryPolicy} (from \code{langgraph.pregel.retry})
  that provides coarse-grained node-level retries managed by the LangGraph
  runtime. Tenacity decorators and \code{RetryPolicy} are complementary:
  \code{RetryPolicy} retries the entire node (useful for transient
  infrastructure failures), while Tenacity decorators retry a specific
  async function call within a node (useful for per-call backoff with
  exception-type discrimination). Do not use \code{RetryPolicy} as a
  replacement for the per-exception Tenacity patterns in this section ---
  \code{RetryPolicy} does not support different backoff profiles per
  exception class.
\end{notebox}

Retries defer but do not eliminate failure: when a decorator exhausts all
attempts and re-raises, the exception propagates out of the node. The graph
must then decide what to do with it --- and that decision depends on whether
the failure is retriable in principle or permanently terminal, which is the
problem the routing layer resolves.

%%------------------------------------------------------------
\section{Failure Class Determines Routing, Not Just Handling}
\label{sec:hard-vs-soft-branches}
%%------------------------------------------------------------

The design problem here is that putting \code{try/except} blocks inside a
node and returning either a valid result or a sentinel value conflates two
distinct responsibilities: the node is responsible for computation, and the
graph is responsible for routing. When routing logic lives inside a node,
it is invisible to LangGraph Studio's graph renderer, unavailable to
time-travel replay, and untestable without invoking the full node function.
The naive approach also requires the calling node to know the names of its
own successors --- a coupling that makes graph restructuring expensive. The
principle is that a node writes a typed \code{error\_class} field to state,
and a conditional edge function reads that field and returns the name of the
next node; routing logic lives entirely in the edge, not in the node.

The state schema must accommodate this pattern without polluting happy-path
state with error fields that are never set. \code{NotRequired} (from
\code{typing\_extensions}, available in the Python 3.11+ stdlib as
\code{typing.NotRequired}) allows a \code{TypedDict} field to be absent
from the dict without triggering a type error. The node writes
\code{error\_class} only on failure; on the happy path the field is absent,
and the conditional edge's catch-all branch routes to \code{END}. The
distinction between \code{"soft"} and \code{"hard"} maps directly to the
taxonomy: a \code{RateLimitError} that exhausts its Tenacity budget is
soft --- retriable in principle but exhausted in practice; a
\code{ContentPolicyViolationError} (a subclass of LiteLLM's
\code{BadRequestError}, HTTP~400) is hard --- re-sending the same prompt
will always produce the same rejection.

\begin{lstlisting}[language=Python,
  caption={AgentState TypedDict and the routing edge function ---
    \texttt{src/graph/state.py} and \texttt{src/graph/routing.py}},
  label={lst:state-and-routing}]
# src/graph/state.py
from typing import Annotated, Literal
from typing_extensions import NotRequired, TypedDict
from langchain_core.messages import BaseMessage
from langgraph.graph.message import add_messages

class AgentState(TypedDict):
    messages:          Annotated[list[BaseMessage], add_messages]
    error_class:       NotRequired[Literal["soft", "hard"]]
    retry_attempts:    NotRequired[int]
    tier_attempted:    NotRequired[int]
    model_variant:     NotRequired[str]
    prompt_tokens:     NotRequired[int]
    completion_tokens: NotRequired[int]
    dead_letter:       NotRequired[dict]   # DeadLetterRecord.model_dump()

# src/graph/routing.py
from typing import Literal
from src.graph.state import AgentState

def route_after_llm(
    state: AgentState,
) -> Literal["continue", "soft_fallback", "dead_letter"]:
    match state.get("error_class"):
        case "hard": return "dead_letter"
        case "soft": return "soft_fallback"
        case _:      return "continue"
\end{lstlisting}

Two decisions in Listing~\ref{lst:state-and-routing} deserve explanation.
First, \code{dead\_letter} is typed as \code{dict} rather than as
\code{DeadLetterRecord}: LangGraph's checkpointer serialises state to JSON,
and Pydantic models are not JSON-serialisable by default. The correct
pattern is to store \code{record.model\_dump(mode="json")} in state and
reconstruct the model with
\code{DeadLetterRecord.model\_validate(state["dead\_letter"])} when reading
it back --- the boundary between the Pydantic domain model and the JSON
storage format is made explicit at the dead-letter node, not hidden in a
custom JSON encoder. Second, \code{error\_class} does not include a
\code{"none"} literal: when no error has occurred the field is simply absent
from the dict, and the \code{case \_} arm of the \code{match} expression
handles both absence and any unexpected value by routing to
\code{"continue"}. Adding a \code{"none"} sentinel would require every node
that does not set an error to explicitly write \code{error\_class="none"},
introducing a correctness dependency between nodes that should be
independent.

\begin{warningbox}
  \textbf{\code{NodeInterrupt} is not an error.}
  LangGraph raises \code{NodeInterrupt} internally when a node calls
  \code{interrupt()} to pause for human input. This exception is caught
  by the LangGraph runtime before it reaches any node code: the state is
  checkpointed and the run is suspended transparently. Do not add
  \code{NodeInterrupt} to your exception hierarchy or attempt to catch it
  in a node's \code{except} block --- the runtime handles it and your
  handler will never fire.
\end{warningbox}

With the routing layer defined, the graph knows which terminal failures
produce a \code{"dead\_letter"} routing decision --- but it has not yet
defined what that destination does with the failed state, which is the
responsibility of the audit layer.

%%------------------------------------------------------------
\section{Terminal Failures Require a Structured Audit Trail}
\label{sec:dead-letter-state}
%%------------------------------------------------------------

The design problem here is that a crashed agent run produces an exception
in a log file and nothing else: no structured record of which node failed,
at what retry depth, with what messages in flight, linked to the
observability trace that contains the full LLM call context. The naive
approach of logging the exception and returning an error response means the
engineering team reconstructs the failure from scattered log lines, and the
prompt engineer has no way to correlate a bad model output with the Langfuse
trace that recorded it. The principle is that every terminal failure must be
snapshotted into a typed, immutable, auditable record stored inside the
agent's checkpointed state --- where it is queryable via
\code{get\_state\_history()}, survives pod restarts, and carries a direct
link to the Langfuse trace for that run.

\emph{Dead-letter state} is the agent-side analogue of a dead-letter queue
in message-broker architectures: a terminal holding area for messages that
could not be processed, retained for inspection and replay rather than
discarded. The difference from a message-broker dead-letter queue is that
the record lives in the graph's checkpointed state, not in a separate
broker, so it travels with the rest of the conversation history and is
accessible through the same \code{get\_state()} API used for normal state
inspection. The record is defined as a Pydantic \code{BaseModel} with
\code{ConfigDict(frozen=True)} because dead-letter records are immutable
audit artefacts: once created, no field should be writable, and
\code{frozen=True} enforces this at the instance level while also making
the model hashable (Pydantic v2 behaviour).

\begin{lstlisting}[language=Python,
  caption={DeadLetterRecord Pydantic model ---
    \texttt{src/resilience/dead\_letter.py}},
  label={lst:dead-letter-record}]
from datetime import datetime, timezone
from typing import Literal
from pydantic import BaseModel, ConfigDict, field_validator

class DeadLetterRecord(BaseModel):
    model_config = ConfigDict(frozen=True)

    node_name:              str
    error_class:            Literal["hard", "soft"]
    exception_type:         str
    exception_message:      str          # truncated to 500 chars
    retry_attempts:         int
    failed_at:              datetime     # must be timezone-aware
    thread_id:              str
    langfuse_trace_id:      str | None
    langfuse_trace_url:     str | None
    last_messages_snapshot: list[dict]   # state["messages"][-3:]

    @field_validator("exception_message")
    @classmethod
    def _truncate(cls, v: str) -> str:
        return v[:500]

    @field_validator("failed_at")
    @classmethod
    def _require_tz(cls, v: datetime) -> datetime:
        if v.tzinfo is None:
            raise ValueError("failed_at must be timezone-aware")
        return v
\end{lstlisting}

Two decisions in Listing~\ref{lst:dead-letter-record} deserve explanation.
First, \code{exception\_message} is truncated to 500 characters via a
\code{@field\_validator}: LLM vendor error responses are frequently
structured JSON payloads several kilobytes in size. Storing the raw message
in the checkpointer risks exceeding column width limits of the underlying
storage backend (Redis entries are bounded by the configured
\code{maxmemory} policy, and the full payload is rarely needed for triage).
Second, \code{failed\_at} is validated to be timezone-aware:
\code{datetime.now()} without a \code{tz} argument returns a naive datetime
that serialises to an ambiguous ISO~8601 string and sorts incorrectly in
Langfuse's trace timeline. The \code{@field\_validator} catches naive
datetimes at record-creation time rather than at query time.

The dead-letter node must be exception-safe. If the dead-letter node raises
an unhandled exception, LangGraph propagates it out of the graph entirely,
bypassing the checkpointer and losing the state snapshot that contains the
failure context. The node must wrap its entire body in \code{try/except
Exception}, log any internal failure, and always return a state update ---
even if that update contains only a partially populated record. This is the
one location in the codebase where a broad except clause is correct.

\begin{lstlisting}[language=Python,
  caption={dead\_letter\_node: exception-safe state snapshot with Langfuse
    trace link --- \texttt{src/resilience/dead\_letter.py} (continued)},
  label={lst:dead-letter-node}]
import logging
from datetime import datetime, timezone
from langchain_core.runnables import RunnableConfig
from langfuse.callback import CallbackHandler
from src.graph.state import AgentState

log = logging.getLogger(__name__)

async def dead_letter_node(
    state: AgentState,
    config: RunnableConfig,
) -> dict:
    try:
        cfg       = config.get("configurable", {})
        thread_id = cfg.get("thread_id", "")
        callbacks = config.get("callbacks") or []
        lf        = next(
            (c for c in callbacks if isinstance(c, CallbackHandler)),
            None,
        )
        trace_id  = lf.get_trace_id() if lf else None
        trace_url = (
            f"https://cloud.langfuse.com/trace/{trace_id}"
            if trace_id else None
        )
        record = DeadLetterRecord(
            node_name=state.get("_node_name", "unknown"),
            error_class=state.get("error_class", "hard"),
            exception_type=state.get("_exception_type", "unknown"),
            exception_message=state.get("_exception_message", ""),
            retry_attempts=state.get("retry_attempts", 0),
            failed_at=datetime.now(timezone.utc),
            thread_id=thread_id,
            langfuse_trace_id=trace_id,
            langfuse_trace_url=trace_url,
            last_messages_snapshot=[
                m.model_dump() for m in state["messages"][-3:]
            ],
        )
        return {"dead_letter": record.model_dump(mode="json")}
    except Exception:
        log.exception("dead_letter_node failed internally")
        return {"dead_letter": {"error": "dead_letter_node_failed"}}
\end{lstlisting}

Three decisions in Listing~\ref{lst:dead-letter-node} deserve explanation.
First, the Langfuse \code{CallbackHandler} is retrieved from
\code{config["callbacks"]} rather than from a module-level global: the
handler is injected at invocation time via the \code{RunnableConfig},
making the node testable in isolation by passing a mock handler in the
config's \code{callbacks} list. Second, \code{lf.get\_trace\_id()} returns
the trace identifier assigned by Langfuse's SDK for the current run; the
trace URL is constructed deterministically from this identifier, producing a
direct link from the dead-letter record to the Langfuse trace that contains
the full LLM call sequence. The Langfuse trace ID is not available via
\code{RunnableConfig["run\_id"]} --- that key does not exist in
\code{RunnableConfig}; it is a common mistake to conflate LangSmith's
run identifier with Langfuse's trace identifier, which are separate systems.
Third, \code{last\_messages\_snapshot} stores at most three messages: the
last three messages (the triggering user query, any preceding assistant turn,
and the most recent tool result if present) are sufficient for triage, and
storing the full history risks exceeding checkpointer storage limits in
long-running sessions.

\begin{notebox}
  \textbf{Infrastructure metadata fields in state.}
  The fields \code{\_node\_name}, \code{\_exception\_type}, and
  \code{\_exception\_message} in the dead-letter node are written by each
  LLM node before routing to dead-letter --- they are prefixed with an
  underscore to signal infrastructure metadata rather than domain state,
  and are \code{NotRequired} in \code{AgentState}. A production hardening
  step is to replace them with a typed \code{FailureContext} sub-model
  validated at the dead-letter node boundary.
\end{notebox}

A dead-letter record in state answers \emph{what} went wrong and
\emph{where}. The question of \emph{how often} nodes fail, and whether
the rate is approaching an SLO breach, requires a different instrument ---
one that aggregates across runs and fires before the situation becomes an
incident.

%%------------------------------------------------------------
\section{Error Rates Require Real-Time Measurement Against a Budget}
\label{sec:error-budget-tracking}
%%------------------------------------------------------------

The design problem here is that examining dead-letter records one by one
is reactive: by the time an engineer opens a Langfuse trace, a node may
have been failing at thirty percent for twenty minutes. The naive approach
of setting an alert on raw error count is equally wrong in the opposite
direction: an agent that processes 10,000 requests per hour and fails twice
is healthy; an agent that processes four requests per hour and fails twice
has a fifty-percent error rate and is broken. Alerting on rates, not counts,
requires a measurement instrument that aggregates over a rolling time window
--- which is exactly what \emph{Prometheus counters} and the \code{rate()}
function provide.

The distinction between a Prometheus \code{Counter} and a \code{Gauge} is
load-bearing here. A \code{Gauge} represents a value that can go up or down
(current queue depth, active connections). A \code{Counter} is monotonically
increasing and represents an accumulated count of events (total requests,
total errors). The \code{rate()} function in PromQL computes the
per-second rate of increase of a \code{Counter} over a time window --- it
requires a \code{Counter} specifically because a \code{Gauge}'s value
jumping between scrapes would produce nonsensical rates. An SLO breach alert
of the form ``error rate exceeds five percent over a five-minute window'' is
expressed as the ratio of two \code{rate()} calls, which requires both the
error counter and the total-calls counter to be \code{Counter} type.

Because LangGraph nodes are \code{async def} functions, the context manager
that wraps them must be an \code{@asynccontextmanager} from
\code{contextlib} --- a sync \code{@contextmanager} cannot yield inside an
\code{async with} block and raises a \code{RuntimeError} at the first
attempt. The instrumentation wrapper uses \code{async with} semantics so
that it wraps the full duration of an async node call, including all
\code{await} points inside the node, without blocking the event loop.

\begin{lstlisting}[language=Python,
  caption={Prometheus counters and async instrumentation context manager ---
    \texttt{src/observability/metrics.py}},
  label={lst:prometheus-metrics}]
from prometheus_client import Counter, Histogram
from contextlib import asynccontextmanager
from typing import AsyncGenerator

NODE_CALLS = Counter(
    "agent_node_calls_total",
    "Total calls to each agent node",
    ["node_name"],
)
NODE_ERRORS = Counter(
    "agent_node_errors_total",
    "Total errors per agent node and error class",
    ["node_name", "error_class"],
)
NODE_DURATION = Histogram(
    "agent_node_duration_seconds",
    "Node execution latency in seconds",
    ["node_name"],
    buckets=[0.1, 0.5, 1.0, 2.0, 5.0, 10.0, 30.0, 60.0],
)

@asynccontextmanager
async def instrument_node(
    node_name: str,
) -> AsyncGenerator[None, None]:
    NODE_CALLS.labels(node_name=node_name).inc()
    with NODE_DURATION.labels(node_name=node_name).time():
        try:
            yield
        except Exception as exc:
            NODE_ERRORS.labels(
                node_name=node_name,
                error_class=_classify(exc),
            ).inc()
            raise
\end{lstlisting}

Two decisions in Listing~\ref{lst:prometheus-metrics} deserve explanation.
First, \code{NODE\_CALLS} uses only \code{node\_name} as a label, while
\code{NODE\_ERRORS} uses both \code{node\_name} and \code{error\_class}:
at the time the calls counter is incremented, the node has not yet executed
and no error class is known. Adding \code{error\_class} to the calls counter
would require a sentinel value (\code{"none"}) that inflates cardinality
without adding information. The SLO alert's PromQL expression sums across
the \code{error\_class} dimension of the errors counter automatically via
the \code{rate()} aggregation. Second, the \code{Histogram} bucket ceiling
of 60 seconds is deliberate: \code{prometheus\_client}'s default buckets top
out at 10 seconds, which clips the distribution of long-running LLM calls
and makes the 95th percentile unobservable for chain-of-thought invocations.

\begin{lstlisting}[language=Python,
  caption={Exception classifier for Prometheus labels ---
    \texttt{src/observability/metrics.py} (continued)},
  label={lst:exception-classifier}]
from litellm.exceptions import (
    RateLimitError, Timeout, APIConnectionError,
    ContentPolicyViolationError,
)

def _classify(exc: BaseException) -> str:
    if isinstance(exc, ContentPolicyViolationError):
        return "hard"
    if isinstance(exc, (RateLimitError, Timeout, APIConnectionError)):
        return "soft"
    return "unknown"
\end{lstlisting}

The \code{\_classify} function is kept in \code{metrics.py} rather than in
\code{decorators.py} to avoid a circular import between
\code{src/resilience/} and \code{src/observability/}. The three-value
taxonomy (\code{"hard"}, \code{"soft"}, \code{"unknown"}) maps directly
onto the \code{error\_class} field in \code{AgentState}, so Prometheus
label values and state values are always consistent.

\begin{lstlisting}[language=yaml,
  caption={Prometheus alerting rule for SLO breach ---
    \texttt{k8s/alerts/agent-slo.yaml}},
  label={lst:slo-alert}]
apiVersion: monitoring.coreos.com/v1
kind: PrometheusRule
metadata:
  name: agent-slo-alerts
  namespace: monitoring
spec:
  groups:
    - name: agent.slo
      interval: 30s
      rules:
        - alert: AgentNodeSLOBreach
          expr: |
            (
              rate(agent_node_errors_total[5m])
              /
              rate(agent_node_calls_total[5m])
            ) > 0.05
          for: 2m
          labels:
            severity: warning
          annotations:
            summary: >-
              Node {{ $labels.node_name }} error rate > 5%
            description: >-
              {{ $labels.node_name }} error rate
              {{ $value | humanizePercentage }}
              (error_class={{ $labels.error_class }}).
              SLO threshold 5%. Window 5m.
\end{lstlisting}

Two decisions in Listing~\ref{lst:slo-alert} deserve explanation. First,
\code{for: 2m} requires the alert condition to remain true for two
consecutive minutes before the alert fires: without this, a single
anomalous 30-second scrape window produces a page, making the alert noisy
enough to be ignored. Two minutes of continuous breach catches a node with
a broken API key before it has degraded the user experience at scale while
filtering single-window spikes. Second, the PromQL expression divides the
\code{rate()} of the errors counter by the \code{rate()} of the calls
counter rather than using a recording rule: recording rules reduce query
cost at the expense of delayed visibility (the evaluation interval adds
latency to alert detection). For a five-percent SLO breach, the additional
compute cost of computing the ratio on each scrape is minimal.

\begin{tipbox}
  Add a Grafana dashboard panel for each node using
  \code{rate(agent\_node\_errors\_total\{node\_name="llm\_primary"\}[5m])
    /\allowbreak{} rate(agent\_node\_calls\_total\{node\_name="llm\_primary"\}[5m])}
  as the query. Annotate the panel with deployment events using Grafana's
  annotation API so that a rising error rate can be immediately correlated
  with a new deployment or a provider-side incident.
\end{tipbox}

Knowing that a node's error rate is breaching five percent does not, by
itself, give users a partial service. The degradation cascade converts that
measurement into an automated response: when primary-tier errors exhaust
their retries, the cascade activates progressively cheaper fallback tiers
rather than failing immediately.

%%------------------------------------------------------------
\section{Service Degrades in Tiers, Not Binary On/Off}
\label{sec:degradation-cascade}
%%------------------------------------------------------------

The design problem here is that a binary serve-or-fail architecture gives
users no service at all during a partial outage. When the primary model's
rate limit is saturated, every request fails; the user sees an error message
and learns nothing about when service will be restored. The naive approach
of a single fallback (primary fails, use cheap model) is better but still
collapses to a hard failure when the cheap model is also unavailable. The
principle is that service degrades through a sequence of progressively
cheaper tiers, each tier activating only when the tier above it has been
exhausted, so that the system provides the best available service under
any degradation scenario rather than the worst.

The four-tier cascade consists of: Tier~1, a full-capability model
decorated with rate-limit and timeout retries; Tier~2, a cheaper, faster
model with tighter retry budgets; Tier~3, a static fallback that returns
a pre-written canned response without making any LLM call; and Tier~4,
an escalation node that posts the failure context to an external webhook
for human handling. Each tier is a separate LangGraph node rather than a
conditional branch inside a single node. A loop inside one node would hide
every tier boundary from Studio and the checkpointer: the \code{tier\_attempted}
field in state would be unavailable until the loop terminated, and
time-travel replay could not be initiated from within the loop. Each tier
as a node means Studio renders every tier transition as a distinct step in
the run timeline, and Prometheus labels errors by the exact node name.

\begin{lstlisting}[language=Python,
  caption={STATIC\_RESPONSES registry and cascade router ---
    \texttt{src/graph/degradation.py}},
  label={lst:static-responses-router}]
from typing import Literal
from langgraph.graph import END
from src.graph.state import AgentState

STATIC_RESPONSES: dict[str, str] = {
    "default": (
        "I am temporarily unable to process your request. "
        "Please try again in a few minutes."
    ),
    "code": (
        "Code generation is temporarily unavailable. "
        "Please try again shortly."
    ),
    "search": (
        "Search is temporarily unavailable. "
        "Please try again shortly."
    ),
}

def route_cascade(
    state: AgentState,
) -> Literal["llm_cheap", "static_fallback",
             "escalate", "dead_letter", "__end__"]:
    error = state.get("error_class")
    tier  = state.get("tier_attempted", 0)
    if error is None:    return "__end__"    # success at any tier
    if error == "hard":  return "dead_letter"
    match tier:
        case 1: return "llm_cheap"
        case 2: return "static_fallback"
        case 3: return "escalate"
        case _: return "dead_letter"        # unexpected depth -> dead-letter
\end{lstlisting}

Three decisions in Listing~\ref{lst:static-responses-router} deserve
explanation. First, \code{route\_cascade} returns the string
\code{"\_\_end\_\_"} for the success case, not a custom \code{"end"} string:
\code{"\_\_end\_\_"} is LangGraph's internal string constant for the
\code{END} sentinel, and the \code{path\_map} argument to
\code{add\_conditional\_edges} maps it to \code{END} directly. Returning
any other string without registering it in the \code{path\_map} raises a
\code{ValueError} at \code{compile()} time. Second, \code{route\_cascade}
handles \code{error\_class="hard"} from \emph{any} tier: a hard failure can
surface from the cheap model as well as the primary (the cheap model may
also trigger a content policy violation), and routing it immediately to
\code{"dead\_letter"} prevents the cascade from re-trying a permanently
rejected prompt. Third, the \code{case \_} arm for an unexpected
\code{tier\_attempted} value routes to \code{"dead\_letter"} rather than
attempting another tier: an unexpected value indicates a state-corruption
bug, and dead-lettering with a structured record is safer than continuing
in an undefined state.

Tier~3 is the static fallback, which deserves particular attention because
it is the only tier that requires zero LLM calls. The fallback reads the
last user message, applies keyword matching against the
\code{STATIC\_RESPONSES} keys, and returns the corresponding canned
response. The classification must not call an LLM: by definition, LLM
infrastructure is unavailable or exhausted at this tier, and any call would
fail for exactly the same reason the previous two tiers failed.

\begin{lstlisting}[language=Python,
  caption={static\_fallback\_node: LLM-free canned response ---
    \texttt{src/graph/degradation.py} (continued)},
  label={lst:static-fallback-node}]
from langchain_core.messages import AIMessage

async def static_fallback_node(state: AgentState) -> dict:
    last_content = ""
    for msg in reversed(state["messages"]):
        if hasattr(msg, "content") and isinstance(msg.content, str):
            last_content = msg.content.lower()
            break
    response_key = "default"
    for key in STATIC_RESPONSES:
        if key != "default" and key in last_content:
            response_key = key
            break
    return {
        "messages":       [AIMessage(content=STATIC_RESPONSES[response_key])],
        "tier_attempted": 3,
    }
\end{lstlisting}

Two decisions in Listing~\ref{lst:static-fallback-node} deserve
explanation. First, the message list is iterated in reverse to find the
most recent user message: the last user message is most relevant to the
current failure; iterating in reverse avoids scanning the full conversation
history on every static-fallback invocation. Second, the function signature
omits \code{config}: the static fallback makes no external calls and reads
nothing from \code{RunnableConfig}. The LangGraph runtime will not pass
\code{config} to nodes whose signature omits it, so removing the parameter
from functions that do not need it reduces surface area for
misconfiguration and makes the function's input contract explicit.

\begin{warningbox}
  \textbf{Tier~4 escalation and idempotency.}
  The escalation node posts to an external webhook. If the graph is
  retried by the caller (for example, a message-queue redelivery) after
  the escalation node has already fired, the webhook will receive two posts
  for the same thread. Make the webhook handler idempotent on
  \code{thread\_id}: check whether an escalation record already exists for
  that thread before creating a new one. The \code{dead\_letter} dict in
  state contains the \code{thread\_id} and serves as the natural
  idempotency key.
\end{warningbox}

With all five tiers and the routing logic defined, the remaining work is to
wire them into a single compiled graph and demonstrate that each failure
path routes correctly under artificial failure injection.

%%------------------------------------------------------------
\section{Every Failure Path Must Be Exercised Before Deployment}
\label{sec:capstone-resilient-agent}
%%------------------------------------------------------------

The design problem here is that graph routing correctness cannot be verified
by reading the graph definition: a wrong string in a \code{path\_map}
argument, a missing \code{error\_class} write in a node, or an incorrect
\code{match} arm in a router function all produce silent mis-routing rather
than a startup error. The naive approach of manual testing exercises only
the happy path. The principle is that the \emph{failure injection} pattern
--- a drop-in \code{FailureInjector} object that raises a specified exception
on every \code{acompletion} call, injected via the same \code{configurable}
channel used for the real LiteLLM Router --- allows each failure path to be
covered by a focused, deterministic \code{pytest} case without modifying
graph code.

The graph factory assembles all five nodes and both conditional edge
functions into a compiled \code{StateGraph}. The \code{checkpointer}
parameter accepts a \code{BaseCheckpointSaver} instance, which in tests is
an \code{InMemorySaver} imported from \code{langgraph.checkpoint.memory} ---
this is the class name shown in the official LangGraph reference
documentation's \code{StateGraph} example; \code{MemorySaver} is an older
alias that may still resolve but is not what current documentation shows.
Each test case creates a fresh \code{InMemorySaver} instance and a unique
\code{thread\_id} to prevent state bleed: \code{InMemorySaver} is a
stateful object, and reusing the same instance with the same
\code{thread\_id} across tests causes subsequent \code{ainvoke} calls to
resume a previous run's state rather than starting fresh.

\begin{lstlisting}[language=Python,
  caption={build\_resilient\_graph factory ---
    \texttt{src/graph/agent.py}},
  label={lst:build-resilient-graph}]
from langgraph.graph import StateGraph, START, END
from langgraph.checkpoint.base import BaseCheckpointSaver
from src.graph.state import AgentState
from src.graph.degradation import (
    llm_primary_node, llm_cheap_node,
    static_fallback_node, escalate_node, route_cascade,
)
from src.graph.routing import route_after_llm
from src.resilience.dead_letter import dead_letter_node

def build_resilient_graph(
    checkpointer: BaseCheckpointSaver | None = None,
) -> "CompiledStateGraph":
    b = StateGraph(AgentState)
    b.add_node("llm_primary",     llm_primary_node)
    b.add_node("llm_cheap",       llm_cheap_node)
    b.add_node("static_fallback", static_fallback_node)
    b.add_node("escalate",        escalate_node)
    b.add_node("dead_letter",     dead_letter_node)
    b.add_edge(START, "llm_primary")
    b.add_conditional_edges(
        "llm_primary", route_after_llm,
        {"continue": END, "soft_fallback": "llm_cheap",
         "dead_letter": "dead_letter"},
    )
    b.add_conditional_edges(
        "llm_cheap", route_cascade,
        {"__end__": END, "static_fallback": "static_fallback",
         "escalate": "escalate", "dead_letter": "dead_letter"},
    )
    b.add_edge("static_fallback", END)
    b.add_edge("escalate",        END)
    b.add_edge("dead_letter",     END)
    return b.compile(checkpointer=checkpointer)
\end{lstlisting}

Two decisions in Listing~\ref{lst:build-resilient-graph} deserve
explanation. First, the \code{path\_map} for the \code{"llm\_cheap"}
conditional edge uses \code{"\_\_end\_\_"} as the key mapping to \code{END}:
this is the string constant that LangGraph's edge resolution expects when
a conditional edge function returns the \code{END} sentinel value; any
other unmapped string raises a \code{ValueError} at compile time.
Second, \code{BaseCheckpointSaver} is the type annotation for the
\code{checkpointer} parameter rather than a specific implementation: this
makes the factory usable with any backend ---
\code{InMemorySaver} in tests, a \code{RedisSaver} or
\code{PostgresSaver} in production --- without changing the factory
signature.

\begin{lstlisting}[language=Python,
  caption={FailureInjector and four failure-path test cases ---
    \texttt{tests/fakes.py} and \texttt{tests/test\_resilient\_agent.py}},
  label={lst:failure-injection-tests}]
# tests/fakes.py
from typing import Type

class FailureInjector:
    def __init__(self, exc_type: Type[Exception], msg: str = "injected"):
        self._exc  = exc_type(msg)
        self.calls = 0

    async def acompletion(self, **kwargs) -> None:
        self.calls += 1
        raise self._exc

# tests/test_resilient_agent.py
import uuid, pytest
from langgraph.checkpoint.memory import InMemorySaver
from litellm.exceptions import RateLimitError, ContentPolicyViolationError
from src.graph.agent import build_resilient_graph
from src.resilience.dead_letter import DeadLetterRecord
from tests.fakes import FailureInjector

def _cfg(injector) -> dict:
    return {"configurable": {
        "thread_id":  str(uuid.uuid4()),
        "llm_router": injector,
    }}

@pytest.fixture
def graph():
    return build_resilient_graph(checkpointer=InMemorySaver())

@pytest.mark.asyncio
async def test_soft_failure_routes_to_cheap_tier(graph):
    inj   = FailureInjector(RateLimitError)
    state = await graph.ainvoke(
        {"messages": [{"role": "user", "content": "hello"}]},
        config=_cfg(inj),
    )
    assert state.get("tier_attempted", 0) >= 1

@pytest.mark.asyncio
async def test_hard_failure_bypasses_cascade(graph):
    inj   = FailureInjector(ContentPolicyViolationError)
    state = await graph.ainvoke(
        {"messages": [{"role": "user", "content": "hello"}]},
        config=_cfg(inj),
    )
    assert state.get("error_class") == "hard"
    assert "dead_letter" in state
    record = DeadLetterRecord.model_validate(state["dead_letter"])
    assert record.error_class == "hard"
    assert record.failed_at.tzinfo is not None   # timezone-aware

@pytest.mark.asyncio
async def test_cascade_exhaustion_reaches_escalate(graph):
    inj   = FailureInjector(RateLimitError)
    state = await graph.ainvoke(
        {"messages": [{"role": "user", "content": "hello"}]},
        config=_cfg(inj),
    )
    assert state.get("tier_attempted") == 3

@pytest.mark.asyncio
async def test_dead_letter_record_validates(graph):
    inj   = FailureInjector(ContentPolicyViolationError)
    state = await graph.ainvoke(
        {"messages": [{"role": "user", "content": "hello"}]},
        config=_cfg(inj),
    )
    record = DeadLetterRecord.model_validate(state["dead_letter"])
    assert record.exception_type == "ContentPolicyViolationError"
\end{lstlisting}

Three decisions in Listing~\ref{lst:failure-injection-tests} deserve
explanation. First, \code{\_cfg} generates a new \code{uuid.uuid4()}
\code{thread\_id} on every call: each invocation must have a unique thread
identifier so that the \code{InMemorySaver} treats it as a new run rather
than resuming the checkpointed state of a previous test case. A shared
\code{thread\_id} across test cases causes \code{test\_cascade\_exhaustion}
to resume from the state left by \code{test\_soft\_failure}, producing
spurious assertions. Second, \code{DeadLetterRecord.model\_validate} is the
Pydantic v2 API for constructing a model from a dict; \code{parse\_obj}
(Pydantic v1) is removed in Pydantic v2. Third,
\code{record.failed\_at.tzinfo is not None} is the portable assertion for
timezone-awareness: \code{isinstance(v.tzinfo, timezone)} is overly specific
because Pydantic may deserialise the datetime with a different
\code{tzinfo} implementation (\code{datetime.timezone.utc} vs a
\code{zoneinfo.ZoneInfo} instance) depending on the platform.

\begin{tipbox}
  Run the failure injection test suite in CI with
  \code{pytest --asyncio-mode=auto -x tests/test\_resilient\_agent.py}.
  The \code{-x} flag stops on the first failure: the tests are ordered by
  cascade depth, so a failure in \code{test\_soft\_failure\_routes\_to\_cheap\_tier}
  makes \code{test\_cascade\_exhaustion\_reaches\_escalate} meaningless.
  Stopping immediately surfaces the root cause rather than masking it with
  downstream assertion failures.
\end{tipbox}

The resilience architecture built in this chapter --- per-exception retries,
typed failure routing, dead-letter state, SLO-aware Prometheus alerting, and
a four-tier degradation cascade --- provides structural guarantees that
individual agent runs fail gracefully and leave a structured record when they
cannot recover. Chapter~28 addresses the complementary problem: ensuring
that model outputs which \emph{do} arrive are safe, factually bounded, and
compliant with content policy before they are returned to the user. Where
this chapter asks ``did the model respond at all?'', Chapter~28 asks
``should the model's response be forwarded?''
